### 2020-09-02 lgpc_JBES.Rmd
Initial commit, revising for JBES.

---
title: "The locally Gaussian partial correlation"
author: Håkon Otneim^[Corresponding author. Department of Business and Management Science, NHH Norwegian School of Economics. Helleveien 30, 5045 Bergen, Norway. \texttt{hakon.otneim@nhh.no}] and Dag Tjøstheim^[University of Bergen]
output: 
  bookdown::pdf_document2:
      dev: tikz
      keep_tex: yes
      number_sections: yes
      toc: no
      includes:
          in_header: latex_preamble.tex
classoption: letterpaper
geometry: margin = 1in
bibliography: "bibliography.bib"
biblio-style: "apalike"
link-citations: yes
indent: yes
fontsize: 12pt
abstract: > 
   It is well known in econometrics and other fields that the dependence structure for jointly Gaussian variables can be fully captured using correlations, and that the \textit{conditional} dependence structure in the same way can be described using partial correlations. The partial correlation does not, however, characterize conditional dependence in many non-Gaussian populations. This paper introduces the local Gaussian partial correlation (LGPC), a new measure of conditional dependence. It is a local version of the partial correlation coefficient that characterizes conditional dependence in a large class of populations. It has some useful and novel properties besides: The LGPC reduces to the ordinary partial correlation for jointly normal variables, and it distinguishes between positive and negative conditional dependence. Furthermore, the LGPC can be used to study departures from conditional independence in specific parts of the distribution. We provide several examples of this, both simulated and real, and derive estimation theory under a local likelihood estimation framework. Finally,  we indicate how the LGPC can be used to construct a powerful test for conditional independence, which, for example, can be used to detect nonlinear Granger causality in time series. \newline \textbf{Keywords:} \emph{Conditional dependence, conditional independence test, Granger causality}
---

# INTRODUCTION {#introduction}

Estimation of conditional dependence and testing for conditional independence are important topics in classical as well as modern statistics. In the last two decades, for instance, there have been intense developments in using conditional dependence for probabilistic networks, conditional multivariate time series analysis as well as copula analysis. For jointly Gaussian variables, conditional dependence is measured by the partial correlation coefficient. Given three jointly Gaussian stochastic variables $X_1$, $X_2$ and $X_3$, $X_1$ and $X_2$ are conditionally independent given $X_3$ if and only if the partial correlation between $X_1$ and $X_2$ given $X_3$ is equal to zero. There is a rich literature on applications of the partial correlation coefficient to Gaussian networks, path analysis and causality. The Gaussian assumption is strict, however, and it is easy to find non-Gaussian examples where the partial correlation function completely fails in describing conditional dependence, as will be seen later in this article.

The purpose of this paper is to introduce a new concept for measuring conditional dependence. This concept retains all the properties of the ordinary partial correlation in the Gaussian case, but seeks to avoid the weaknesses of this measure in the non-Gaussian case. This is done by fitting a *family* of Gaussian distributions to a given continuous multivariate distribution and exploiting the simple conditioning rules of the Gaussian distribution *locally*. This approach produces a new measure of conditional dependence, *the local Gaussian partial correlation (LGPC)* which, being directly related to the ordinary partial correlation, is easy to interpret, and reduces to the very same ordinary partial correlation in the Gaussian case. Moreover, it distinguishes between positive and negative conditional dependence whereas competing non-linear measures report only the *strength* of the conditional dependence on some non-negative scale.

The local view gives much more flexibility. It allows the conditional dependence to be stronger or weaker in certain regions of a multivariate distribution than in others. This is of particular interest in finance, where description of tail behavior is important. The local aspect also makes it possible to focus tests of conditional independence to selected parts of the distribution, which may potentially increase the power of the test.

The local approach has been shown to be advantageous in other areas of statistics, such as the measurement of nonlinear dependence, density estimation and spectral analysis, see for instance @tjostheim2013local, @otneim2017locally and @jordanger2017nonlinear. The present paper, then, represents the first attempt to model conditional dependence locally. There are several other approaches to modelling conditional dependence and testing for conditional independence in the non-Gaussian case in the literature. For the most part, they revolve around test statistics in conditional independence tests. Such statistics are usually based on measures of distance between some property of the sample and the corresponding property of the population under the null hypothesis of conditional independence. As a consequence, the  conditional dependence measures are not always easy to interpret, and can, as mentioned above, not distinguish between negative and positive conditional dependence. We present a comprehensive set of references to this literature in Section \@ref(chap:testing).

Definitions and properties are presented in Section \@ref(chap:conditional), whereas estimation is covered in Section \@ref(chap:estimation). Two examples illustrating the strengths of our approach are given in Section \@ref(chap:example). Section \@ref(chap:testing) gives a brief glimpse of the potential in testing.

An integral part of our article is the online supplement. It deals with a number of issues complementing the treatment in the main paper. Among them are more examples, extensions of the main concept to more general situations, precise formulations of some theoretical results with proofs, as well as discussions of Pitman efficiency for testing and validity of the use of the bootstrap.

# THE LOCAL GAUSSIAN PARTIAL CORRELATION {#chap:conditional}

Let $\X = (X_1, \ldots, X_p)$ be a random vector. Denote by $(\X_1,\X_2,\X_3)$ a partition of $\X$ into vectors of dimension $d_1$, $d_2$ and $d_3$ respectively, such that $\Xone = (\X_1, \X_2) = (X_1, \ldots,X_{d_1 + d_2})$ consists of the first $d_1+d_2$ components in $\X$ and $\Xtwo = \X_3 = (X_{d_1+d_2+1}, \ldots, X_{d_1+d_2+d_3})$ contains the remaining $d_3$ variables, where $p = d_1+d_2+d_3$. We assume that the mean vector $\fmu$ and covariance matrix $\fSigma$ of $\X$ exist, and partition them correspondingly, writing 
\begin{equation}
\fmu = \begin{pmatrix} \fmu_1 \\ \fmu_2 \\ \fmu_3 \end{pmatrix} \,\,\,\, \textrm{ and } \,\,\,\, \fSigma = \begin{pmatrix} \fSigma_{11} & \fSigma_{12} & \fSigma_{13} \\ \fSigma_{21} & \fSigma_{22} & \fSigma_{23} \\ \fSigma_{31} & \fSigma_{32} & \fSigma_{33}\end{pmatrix},
(\#eq:matrixpartition)
\end{equation}
where $\fSigma_{ij}$ is the covariance matrices of $(\X_i, \X_j)$, $i,j = 1,2,3$. There are two main concepts of correlation when $\Xtwo = \X_3$ is given, the *partial* and the *conditional* correlation, and they coincide in several joint distributions, among them the Gaussian. See e.g. @baba2004partial for details. We will use the partial correlation as a starting point when defining the LGPC. The partial variance-covariance matrix of $\Xone = (\X_1, \X_2)$ given $\Xtwo = \X_3$ is 
\begin{equation}
\fSigma_{12|3} = \fSigma^{11} - \fSigma^{12}\left(\fSigma^{-1}\right)^{22}\fSigma^{21},
(\#eq:matrixpartial)
\end{equation} 
where
$$\fSigma^{11} = \begin{pmatrix} \fSigma_{11} & \fSigma_{12} \\ \fSigma_{21} & \fSigma_{22} \end{pmatrix}, \,\, \fSigma^{12} = \begin{pmatrix} \fSigma_{13} \\ \fSigma_{23} \end{pmatrix}, \,\,\fSigma^{21} = \begin{pmatrix} \fSigma_{31} & \fSigma_{32} \end{pmatrix} \, \textrm{ and } \, \fSigma^{22} = \fSigma_{33},$$
and $\fSigma_{12|3}$ is the covariance matrix in the conditional (Gaussian) distribution of $\Xone$ given $\X_3$ if $\X$ is jointly normal. The partial correlation matrix between $\X_1$ and $\X_2$ given $\X_3$ is naturally defined as

\begin{equation}
\R_{12|3} = \D^{-1/2}\fSigma_{12|3}\D^{-1/2},
(\#eq:partialcorrelation)
\end{equation}
where $\D = \textrm{diag}(\fSigma_{12|3})$. We identify in the same way the partial correlation matrix \@ref(eq:partialcorrelation) with the correlation matrix in the conditional (Gaussian) distribution of $\Xone$ given $\Xtwo$ if $\X$ is jointly normal. Equations \@ref(eq:matrixpartial) and \@ref(eq:partialcorrelation) will serve as the starting point for our definition of the *local* partial correlation.

## Definition {#chap:definition}

In all of the following we assume that the components of $\X$ are continuous with joint density function $f_{\X}$. Given a point $\x$, we approximate $f_{\X}$ in a neighborhood of $\x$ by a multivariate Gaussian density

\begin{equation} 
\psi(\x, \fv) = \frac{1}{(2\pi)^{p/2}|\fSigma(\x)|^{1/2}} \exp \left\{-\frac{1}{2}(\fv - \fmu(\x))^T\fSigma^{-1}(\x)(\fv - \fmu(\x))\right\},
(\#eq:localgaussian)
\end{equation}
where $\x = (x_1\ldots, x_p)$, $\fmu(\x) = \{\mu_j(\x)\}$ and $\fSigma(\x) = \{\sigma_{jk}(\x)\}$ for $j,k=1,\ldots, p$. Moving to another point $\y$, there is another (generally different) Gaussian approximation $\psi(\y, \fv)$. In this way we approximate $f_{\X}$ by a family of multivariate Gaussian densities defined by a set of smooth parameter functions $\{\fmu(\x), \fSigma(\x)\}$, and if $f_{\X}$ is itself a Gaussian density, then the parameter functions collapse to constants corresponding to the true parameter values, and $\psi(\x) \equiv f_{\X}(\x)$. @hjort1996locally provide the general framework for estimating such parameter functions non-parametrically from a given data set using a local likelihood procedure, and the basic idea in the following treatment is to replace the components in the partial covariance matrix \@ref(eq:matrixpartial) by their locally estimated counterparts in order to obtain a local measure of conditional dependence.

Before we get that far, however, we will introduce a useful transformation technique that improves and simplifies the estimation of the LGPC. Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional] show that the estimation of the local parameter functions $\{\fmu(\x), \fSigma(\x)\}$ becomes easier by transforming each $X_j$ to a standard normal variable $Z_j = \Phi^{-1}(U_j)$, where $U_j$ is a uniform variable $U_j = F_{j}(X_j)$ with $F_{j}$ being the cumulative distribution function of $X_j$. Define the random vector $\Z$ by this transformation of $\X = (\Xone, \Xtwo) = (\X_1, \X_2, \X_3) = (X_1, X_2, \ldots, X_p)$ to marginal standard normality:

\begin{equation}
\Z = \Big(\Phi^{-1}\left(F_{1}(X_1)\right), \Phi^{-1}\left(F_{2}(X_2)\right), \ldots, \Phi^{-1}\left(F_{p}(X_p)\right)\Big).
(\#eq:trans)
\end{equation}
The transformation enables us to simplify the local Gaussian representation \@ref(eq:localgaussian) by writing the density $f_{\Z}$ of $\Z$ as 
\begin{equation}
f_{\Z}(\z) = \psi(\z, R(\z)) = \frac{1}{|2\pi\R(\z)|^{1/2}}\exp\left\{-\frac{1}{2}\z^T\R^{-1}(\z)\z\right\},
(\#eq:lgdeapprox)
\end{equation}  
where, as in Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional] in a simplified representation, we have fixed local means and standard deviations $\mu_j(\z) = 0$ and $\sigma_j^2(\z) = 1$, $j = 1,\ldots,p$, and where $\R(\z) = \{\rho_{jk}(\z)\}$ is the *local correlation matrix*.

In practice, we do not know $F_{j}$, but we can instead use the empirical distribution function 
$$F_{j,n}(x) = \frac{1}{n-1}\sum_{i=1}^n 1\left(X_{ji} \leq x\right),$$
where $1(\cdot)$ is the indicator function, $n$ is the number of observations, and $X_{ji}$ is the $i$th observation of $X_j$. This results in pseudo standard normal variables $Z_{j,n} = \Phi^{-1}(F_{j,n}(X_j))$.

In the following, we will not always distinguish between $Z_j$ and $Z_{j,n}$. In fact, we conclude our asymptotic analysis in Section \@ref(chap:asymptotic) by showing that under certain technical conditions, the error made by estimating $\R(\z)$ using the empirically transformed variables $Z_{j,n}$, instead of $Z_j$, is smaller in the limit than the estimation error made when estimating the local correlations themselves.

We will in this paper refer to $\X$ and its probability density function $f_{\X}$ as being on the \(x\)-scale, and to $\Z$ and its probability density function $f_{\Z}$ as being on the \(z\)-scale. For further discussion of the simplified $z$-representation we refer to @otneim2017locally.

Denote by $(\Zone, \Ztwo) = (\Z_1, \Z_2, \Z_3)$ the partitioning of $\Z$ corresponding to the partitioning $(\Xone, \Xtwo) = (\X_1, \X_2, \X_3)$ of $\X$. A natural definition of the *local* partial covariance matrix of $\Zone|\Ztwo$ is the local version of eq. \@ref(eq:matrixpartial).

\begin{equation}
\fSigma_{12|3}(\z) = \R^{11}(\zone) - \R^{12}(\z)\left(\R^{22}(\ztwo)\right)^{-1}\R^{21}(\z).
(\#eq:matrixlocalpartial)
\end{equation}

If $d_1=d_2=1$ then $\fSigma_{12|3}(\z)$ is a $2\times 2$ matrix, and we define the local Gaussian partial correlation $\alpha(\z)$ between the two variables in $\Zone = (Z_1,Z_2)$ given $\Ztwo = \Z_3$ in accordance with the ordinary (global) partial correlation provided by eq. \@ref(eq:partialcorrelation):

\begin{equation}
\alpha(\z) = R_{12|3}(\z) = \frac{\Big\{\fSigma_{12|3}(\z)\Big\}_{12}}{\Big\{\fSigma_{12|3}(\z)\Big\}^{1/2}_{11}\Big\{\fSigma_{12|3}(\z)\Big\}^{1/2}_{22}},
(\#eq:definition)
\end{equation}
which, if $\Ztwo = Z_3$ is scalar, reduces to
\begin{equation}
\alpha(\z) = \rho_{12|3}(z_1, z_2|z_3) = \frac{\rho_{12}(z_1, z_2) - \rho_{13}(z_1,z_3)\rho_{23}(z_2,z_3)}{\sqrt{1 - \rho^2_{13}(z_1,z_3)}\sqrt{1 - \rho^2_{23}(z_2, z_3)}}.
(\#eq:scalardefinition)
\end{equation}

It is of course possible to introduce an LGPC $\alpha(\x)$ directly on the $x$-scale, but that representation is in many ways harder to handle both computationally and asymptotically. For a multivariate Gaussian distribution, we have that $\alpha_{\X}(\x) = \alpha_{\Z}(\z) = \alpha$. In the remainder of this paper, we write mainly in terms of the $z$-representation using the LGPC $\alpha(\z) = \alpha_{\Z}(\z)$, but when we write the local partial correlation between $X_1$ and $X_2$ given $\X_3 = \x_3$ at the point $(x_1,x_2,\x_3)$, this is simply $\alpha(\z)$ inserted with $z_j = \Phi^{-1}(F_{j}(x_j))$, $j=1,\ldots,p$.

In the more general case where $d_1 > 1$ and/or $d_2 > 1$ then $\fSigma_{12|3}(\z)$ is a $(d_1+d_2)\times (d_1+d_2)$ matrix that describes the non-linear conditional dependence among the variables in $\Zone = (\Z_1, \Z_2)$ given $\Ztwo = \Z_3$ (or alternatively $\Xone = (\X_1, \X_2)$ given $\Xtwo = \X_3$), which in particular can be used to analyze the conditional dependence between two *sets* of random variables having dimensions $d_1$ and $d_2$ respectively, given a third set of variables having dimension $d_3$. This case poses no conceptual challenges to our approach, but it requires a rather large investment in new notation, and does not lead to simple formulas like eq. \@ref(eq:scalardefinition). We will therefore, for the most part, focus on the local Gaussian partial correlation between two stochastic scalar variables in the remainder of this paper. We provide the complete description of the general multivariate case in Sections A.1 and A.2 of the online supplement.

## Properties {#properties}

The local Gaussian partial correlation is closely related to the partial correlation between jointly normally distributed variables. We see this also from the following properties that we will establish next:

1. The LGPC $\alpha(\z)$ satisfies $-1 \leq \alpha(\z) \leq 1$.
2. If $\X$ is jointly normally distributed, then the LGPC coincides with the ordinary (global) partial correlation.
3. For stochastic vectors having joint density function on $z$-scale of the form \@ref(eq:lgdeapprox), the LGPC $\alpha(\z)$ is identically equal to zero if and only if $X_1$ and $X_2$ are conditionally independent given $\X_3$. Note that $\alpha(\z) \equiv 0$ if and only if $\alpha(\x) \equiv 0$. \label{prop:char}
4. The LGPC $\alpha(\z)$ is invariant with regard to a set of monotone transformations $\Y = \bm{h}(\X) = (h_1(X_1), \ldots, h_p(X_p))$.

Property 1 is trivially true if $\R(\z)$ is a valid correlation matrix. By removing the $\z$-dependence in the local correlations, it follows immediately from eq. \@ref(eq:lgdeapprox) and from the results referred to in Section \@ref(chap:definition) that property 2 holds.

To see that conditional independence between $X_1$ and $X_2$ given $\X_3$ is equivalent to $\alpha(\z) \equiv 0$, or equivalently $\alpha(\x) \equiv 0$, we need to follow a few simple steps that would also work in the global Gaussian case. Working on the standard normal $z$-scale and assuming that $(z_1, z_2, \z_3)$ has density function $f_{\Z}(\z)$ as given by \@ref(eq:lgdeapprox), it follows from @otneim2017conditional that we can calculate conditional distributions in the same way as in the global Gaussian case. For example, the conditional density of $Z_1|\Z_3$ at the point $z_1$ is given by

$$f_{z_1|\z_3}(z_1) = \frac{1}{|2\pi\sigma_{1|3}(z_1)|^{1/2}}\times\exp \left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1)}{\sigma_{1|3}(z_1)}\right)^2\right\},$$
where

\begin{equation}
\mu_{1|3}(z_1) = \R_{12}(z_1, \z_3)\R_{22}(\z_3)^{-1}\z_3 \,\,\, \textrm{ and } \,\,\, \sigma_{1|3}(z_1) = \fSigma_{Z_1|\Z_3},
(\#eq:localconditionalpar)
\end{equation}
and where the latter expression is defined analogously to \@ref(eq:matrixlocalpartial). The conditional density $f_{Z_2|\Z_3}(z_2|\z_3)$ is defined in the same way. If $Z_1$ and $Z_2$ are conditionally independent given $\Z_3$, then

\begin{align*}
f_{Z_1,Z_2|\Z_3}(z_1, z_2|\z_3) &= f_{Z_1|\Z_3}(z_1|\z_3)f_{Z_2|\Z_3}(z_2|\z_3) \\
&= \frac{1}{2\pi\sqrt{1-\sigma_{1|3}(z_1)}\sqrt{1-\sigma_{2|3}(z_2)}} \\ 
&\qquad \qquad \times\exp\left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1)}{\sigma_{1|3}(z_1)}\right)^2\right\}\exp\left\{-\frac{1}{2}\left(\frac{z_2 - \mu_{2|3}(z_2)}{\sigma_{2|3}(z_2)}\right)^2\right\},
\end{align*}
and we at once identify the conditional density of $(Z_1,Z_2)|\Z_3$ as another Gaussian distribution, but without any cross-term involving $z_1z_2$, which then implies that the off-diagonal element in the partial covariance matrix \@ref(eq:matrixlocalpartial) is identically equal to zero. We see immediately from our definition \@ref(eq:definition) that the LGPC is also identically equal to zero.

The converse statement follows straightforwardly by considering the conditional density of $(Z_1, Z_2|\Z_3)$. <!--Let $f_{\Z}(\z)$ be on the locally Gaussian form \@ref(eq:lgdeapprox). Again, we have from @otneim2017conditional that the conditional density of $(Z_1, Z_2|\Z_3)$ is locally associated with a Gaussian distribution, with local conditional mean vector and local conditional covariance matrix given by expressions corresponding to \@ref(eq:localconditionalpar). The off-diagonal in the local conditional covariance matrix is zero if $\alpha(\z) \equiv 0$, or equivalently $\alpha(\x) \equiv 0$, in which case one may factorize the joint conditional density of $(Z_1, Z_2|\Z_3)$ into two factors, one depending only on $z_1$, and one depending only on $z_2$. Hence, $Z_1$ and $Z_2$ are conditionally independent given $\Z_3$.-->

Let the transformation $\Y = \bm{h}(\X)$ define the \(y\)-scale in the same way as the transformation \@ref(eq:trans) defines the \(z\)-scale. We define the LGPC in terms of the marginally standard normal \(\Z\)-variables, which for the \(\Y\)-variables can be calculated as

$$\Z = \Big(\Phi^{-1}\left(F_{Y_1}(Y_1)\right), \ldots, \Phi^{-1}\left(F_{Y_p}(Y_p)\right)\Big),$$
but if $h_j(\cdot)$ is monotone for all $j = 1,\ldots,p$,

$$F_{Y_j}(y_j) = \mathbb{P}(h_j(X_j) \leq y_j) = \mathbb{P}(X_j \leq h_j^{-1}(y_j)) = \mathbb{P}(X_j \leq x_j) = F_{X_j}(x_j),$$
where $x_j$ is the point on the \(x\)-scale corresponding to the point $y_j$ on the \(y\)-scale. The \(\Z\)-variables are the same for the stochastic variables $\X$ and $\Y = \bm{h}(\X)$. Their LGPC-function $\alpha(\z)$ must therefore be the same as well according to our definition in the preceding section. Entirely analogous arguments for the properties 1.-4. in the general multivariate case can be constructed from the supplementary material, Sections A.1 and A.2.

# ESTIMATING CONDITIONAL DEPENDENCE USING THE LGPC {#chap:estimation}

## Estimation by local likelihood {#ll}

From the defining equations \@ref(eq:matrixlocalpartial) and \@ref(eq:definition) it is seen that the basic building blocks for the  *local Gaussian partial correlation* are the *local Gaussian correlation functions*, that populate the local correlation matrix $\R(\z)$ in the density function $f_{\Z}$ in \@ref(eq:lgdeapprox). Consider the random vector $\X = (X_1, X_2, \ldots, X_p)$ having joint probability density function (pdf) $f(\x)$, and its transformed counterpart $\Z = \left(\Phi^{-1}(F_1(X_1)), \ldots, \Phi^{-1}(F_p(X_p))\right)$, on the marginally standard normal \(z\)-scale. The relation between the pdf of $\X$ and the pdf of $\Z$ is given by

\begin{equation}
f_{\X}(\x) = f_{\Z}\left(\Phi^{-1}\left(F_1(x_1)\right), \ldots, \Phi^{-1}\left(F_p(x_p)\right)\right)\prod_{j=1}^p 
\frac{f_j(x_j)}{\phi\left(\Phi^{-1}\left(F_j(x_j)\right)\right)},
(\#eq:fxfz)
\end{equation}
where $f_j(\cdot)$, $j = 1,\ldots, p$ are the marginal density functions of $\X$, and $\phi(\cdot)$ is the standard normal density function. Let $\X_1, \ldots, \X_n$ be a random sample identically distributed as $\X$ and construct the pseudo standard normal observations $\hZ_1, \ldots, \hZ_n$ as

\begin{equation}
\hZ_t = \left(\Phi^{-1}\left(\hF_1(X_{1,t})\right), \ldots, \Phi^{-1}\left(\hF_p(X_{p,t})\right)\right),
(\#eq:pseudoobservations)
\end{equation}
where $\X_t = (X_{1,t}, X_{2,t}, \ldots, X_{p,t})$, and $\hF_j(\cdot)$ is an estimate of the marginal distribution function of $X_j$, for example the empirical distribution function. Next, we must produce an estimate $\hR(\z)$ of the local correlation matrix, and we will consider two variations of this task in  subsections \@ref(chap:trivariate-full) and \@ref(chap:bivariate) with a more general treatment in Section A.3 of the supplement. It is then, following \@ref(eq:matrixlocalpartial), natural to estimate the local Gaussian partial covariance matrix as

\begin{equation}
\hfSigma_{12|3}(\z) = \hR^{11}(\z) -  \hR^{12}(\z)\left(\hR^{22}(\z)\right)^{-1}\hR^{21}(\z),
(\#eq:localpartialcov)
\end{equation}
and if we assume that $X_1$ and $X_2$ constitute the two first elements in $\X$, then, following \@ref(eq:definition), the local Gaussian partial correlation between $X_1$ and $X_2$ given $\Xtwo = \X_3$ on the $z$-scale is estimated by
\begin{equation}
\widehat\alpha(\z) = \frac{\Big\{\hfSigma_{12|3}(\z)\Big\}_{12}}{\Big\{\hfSigma_{12|3}(\z)\Big\}_{11}^{1/2}\Big\{\hfSigma_{12|3}(\z)\Big\}_{22}^{1/2}}.
(\#eq:estimate)
\end{equation}

A corresponding value of $\halpha(\x)$ at the point $\x = \widehat F^{-1}(\Phi(z))$ is obtained by inserting $\z = \Phi^{-1}(\widehat F(\x))$ in \@ref(eq:estimate). <!-- The *curse of dimensionality* dictates that nonparametric estimates of multivariate regression and density functions quickly deteriorates as the number of variables increases. This is also true for local correlation surfaces, but it can be alleviated somewhat by introducing heavier smoothing towards the global Gaussian model.

@otneim2017locally propose a different strategy. By estimating each local correlation $\rho_{ij}(\z)$, $1\leq i<j\leq p$, sequentially in the bivariate problem involving only the variables $Z_i$ and $Z_j$, they effectively impose a pairwise dependence structure in order to circumvent the curse of dimensionality. This is indeed possible when calculating the LGPC as well, and it can be done for an arbitrary dimension $p$. It is described in more detail in Section \@ref(chap:bivariate).

In the case that $p = 3$, and all variables involved are scalars, we do provide in this paper the empirical option and corresponding estimation theory to fit a full trivariate normal distribution locally to $(\Zone, \Ztwo) = (Z_1,Z_2,Z_3)$ without any structural simplifications. This constitutes a novelty in the local Gaussian literature, and we present the procedure in the following sub-section. -->

### Estimation of $\R(\z)$ when $p = 3$ and $\Xtwo$ is a scalar {#chap:trivariate-full}

If dim$(\X) = 3$, then dim$(\Xtwo)=1$, and $\R(\z)$ is a $3\times3$ symmetric matrix of local correlation functions having arguments $\z = (z_1,z_2,z_3)$:
\begin{equation}
\R(\z) = \begin{pmatrix} 1 & \rho_{12}(\z) & \rho_{13}(\z) \\ \rho_{12}(\z) & 1 & \rho_{23}(\z) \\ \rho_{13}(\z) & \rho_{23}(\z) & 1\end{pmatrix},
\label{eq:trivariatecorr}
\end{equation}
containing three parameter functions that must be estimated from data. Note that here we deviate from the pairwise scheme found in Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional] as well as most of this article where $\rho_{ij}(\z) = \rho_{ij}(z_i, z_j)$. In equation \@ref(eq:trivariatecorr), $\z = (z_1,z_2,z_3)$. Based on the sample \@ref(eq:pseudoobservations), $\R(\z)$ is estimated by fitting the parametric family $\psi(z, \R)$, defined in \@ref(eq:lgdeapprox), locally to the density $f_{\Z}(\z)$ by maximizing the *local likelihood function* (cf. @hjort1996locally),

\begin{equation}
\hR(\z) = \arg\max_{\R} n^{-1} \sum_{t=1}^nK_{\hh}(\hZ_t - \z)\log\psi(\hZ_t, \R) - \int K_{\hh}(\y - \z)\psi(\y, \R)\,\textrm{d}\y
(\#eq:locallik3)
\end{equation}
in each point $\z$, where $K_{\hh}(\x) = |\hh|^{-1/2}K(\hh^{-1/2}\x)$, $K$ is a non-negative and radially symmetric kernel function that satisfies $\int K(\x)\,\textrm{d}\x=1$, and $\hh$ is a diagonal $3\times3$ matrix of bandwidths that serve as smoothing parameters. The estimate $\hf_{\Z}(\z) = \psi(\z, \hR(\z))$ aims at minimizing a locally weighted Kullback-Leibler distance to $f_{\Z}(\z)$, and we refer to @hjort1996locally and @tjostheim2013local for much more details about this construction, and to Section \@ref(chap:asymptotic) for the relevant asymptotic estimation theory.

Finally, we obtain the estimated LGPC $\halpha(\z)$ by plugging the estimated local correlations $\hR(\z)$ into equations \@ref(eq:localpartialcov) and \@ref(eq:estimate) and a corresponding value of $\halpha(\x)$ at the point $\x = \widehat F^{-1}(\Phi(\z))$ by inserting $\z = \Phi^{-1}({\widehat F(\x)})$.

### Estimation of $\R(\z)$ when $\Xtwo$ is a vector {#chap:bivariate}

The complexity of the estimation problem \@ref(eq:locallik3) increases sharply with the number of variables involved. The basic idea of @otneim2017locally to avoid the curse of dimensionality is to model local correlations in $\R(\z)$ as functions of their corresponding *pair* of variables only:

\begin{equation}
\R(\z) = \left\{\rho_{jk}(z_j, z_k)\right\}, \qquad j,k = 1, 2, \ldots, p,
(\#eq:loccormat)
\end{equation}
which reduces the estimation of the $p$-variate correlation functions $\rho_{jk}(\z)$ to a series of bivariate estimation problems. We suggest to use this approach when modelling local dependence if the number of variables in $\X$ is larger than three. This allows nonlinear dependence between variables to be approximated by the pairwise structure \@ref(eq:loccormat), while still being computationally tractable. An analogue to the pairwise approximation is the additive approximation in nonparametric regression.

We transform the observations to the $z$-scale as in \@ref(eq:pseudoobservations), but now estimate the individual components in $\R(\z)$ one by one, taking only into consideration the *pair* of variables in question: 
\begin{equation}
\hrho_{jk}(z_j, z_k) = \arg\max_{\rho_{jk}} n^{-1} \sum_{t=1}^nK_{\hh}(\hZ_t - \z)\log(\hZ_t, \rho_{jk}) - \int K_{\hh}(\y - \z)\psi(\y, \rho_{jk})\,\textrm{d}\y, 
(\#eq:locallik2)
\end{equation}
where all running variables and samples are bivariate subsets corresponding to the indices $(j,k)$, $\psi(\cdot, \rho)$ is the bivariate version of \@ref(eq:lgdeapprox), and $\hh$ now is a $2\times2$ diagonal matrix of bandwidths. After estimating all local correlations in this way, we proceed to calculate the LGPC using equations \@ref(eq:localpartialcov) and \@ref(eq:estimate) as above.

## Asymptotic theory {#chap:asymptotic}

Equations \@ref(eq:matrixlocalpartial), \@ref(eq:definition) and their empirical counterparts \@ref(eq:localpartialcov) and \@ref(eq:estimate) demonstrate clearly that the LGPC is nothing more than a deterministic function of the local correlation matrix, in the same way as the ordinary partial correlation is a function of the ordinary correlation matrix. The asymptotic behavior of $\halpha(\z)$ can thus be derived directly from the asymptotic behavior of $\hR(\z)$, which, for the most part, has been established in earlier works, see e.g. @tjostheim2013local and @otneim2017locally. Some key results follow below, and we refer to the supplementary material, Sections B--D for proofs of the asymptotic distributional results in this subsection, and to Section A.3 for a discussion of more general situations and for a result on consistency.

We start by assuming that the marginal distribution functions $F_1, \ldots, F_p$ are known, and then state two results on the joint asymptotic normality of the estimated local correlation matrix $\hR(\z)$ under a mixing condition. We will then apply the delta method in order to derive the limiting distribution of the estimated LGPC, and finally show that we may replace $F_1,\ldots,F_p$ with their empirical counterparts $\hF_1,\ldots,\hF_p$ without changing these asymptotic results.

Consider first the full trivariate fit $\hR(\z)$ described in Section \@ref(chap:trivariate-full) with $\z = (z_1,z_2,z_3)$. We estimate the local correlation matrix $\R(\z)$ by maximizing the local likelihood function in \@ref(eq:locallik3), which we denote by $L_n(\R(\z), \z)$, with the only exception that we, for now, assume the true transformation between the \(x\)- and the \(z\)-scale to be known. For a fixed matrix of bandwidths $\hh$, denote by $\R_{\hh}(z)$ the local correlation that satisfies, as $n \rightarrow \infty$,

\begin{equation}
\nabla L_n(\R_{\hh}, \z) \rightarrow \int K_{\hh}(\y - \z)\fu(\y, \R_{\hh}(\z))\{f_{\Z}(\y) - \psi(\y, \R_{\hh})\}\,\textrm{d}\y = 0,
(\#eq:popfixedh3)
\end{equation}
where $f_{\Z}(\z) = f_{Z_1,Z_2,Z_3}(z_1,z_2,z_3)$ is the density function of $\Z$, and $\fu(\cdot, \rho_{\hh})$ is the column vector of local score functions $\nabla \log\psi(\cdot, \rho_{\hh})$, and where the gradient is taken with respect to the parameters $\rho_{12}(\z)$, $\rho_{13}(\z)$ and $\rho_{23}(\z)$. Denote by $\frho = \frho(\z) = \{\rho_{12}(\z), \rho_{13}(\z), \rho_{23}(\z)\}$ the vector of local correlations. The joint limiting distribution of $\hfrho$ (for convenience only stated on the \(z\)-scale below) is given by the following result as $n \rightarrow \infty$ and $\hh \rightarrow \bm{0}$:

```{theorem, loccor3}
Let $\{\Z_t\}_{t=1}^n$ be observations on the marginally standard normally distributed random vector $\Z$ having joint density $f_{\Z}(\z)$. Assume that the following conditions hold:

1. For any sequence of bandwidth matrices $\hh_n$ tending to zero element-wise, there exists for the trivariate marginally standard Gaussian vector $\Z$ a unique set of local correlations $\frho_{\hh}(\z)$ that satisfies \@ref(eq:popfixedh3), and there exists a $\frho_0(\z)$ such that $\frho_{\hh}(\z) \rightarrow \frho_0(\z)$.
2. The process generating $\{\Z_t\}_{t=1}^n$ is $\alpha$-mixing with mixing coefficients $\alpha(m)$ satisfying $\sum_{m\geq1}m^{\lambda}\alpha(m)^{1-2/\delta} < \infty$ for some $\lambda > 1-2/\delta$ and $\delta>2$.
3. $n\rightarrow\infty$, and each bandwidth $b = b_n$ tends to zero such that $nb^{\frac{\lambda+2-2/\delta}{\lambda + 2/\delta}} = O(n^{\epsilon_0})$ for some constant $\epsilon_0 > 0$. 
4. In a given point $\z$, the parameter space $\Theta$ for each local correlation $\rho(\z)$ is a compact subset of $(-1,1)$.
5. The non-negative kernel function $K(\cdot)$ sasisfies $\sup_{\z}K(\z) < \infty$, $\int K(\y)\,\textrm{d}\y < \infty$, $\partial/\partial z_iK(\z) <\infty$, and $\lim_{z_i \rightarrow \infty}|z_iK(\z)| = 0$ for $i = 1,2,3$.

Then, by writing $b_{n1} = b_{n2} = b_{n3} = b \rightarrow 0$ assuming they all converge to zero at the same rate, 
\begin{equation}
\sqrt{nb^5}\Jb\Mb^{-1/2}\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
(\#eq:an3)
\end{equation}
where $\bm{I}$ is the $3\times3$ identity matrix,
\begin{align*}
\Jb &= \int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\fu^T(\y, \frho_{\hh}(\z))\psi(\y,\frho_{\hh}(\z))\,\textrm{d}\y \\
& \qquad\qquad\qquad\qquad - \int K_{\hh}(\y - \z)\nabla\fu(\y,\frho_{\hh}(\z))\Big\{f_{\Z}(y) - \psi(\y,\frho_{\hh}(\z))\Big\}\textrm{d}\y,
\end{align*}
and
\begin{align*}
\Mb &= b^3\int K^2_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\fu^T(\y,\frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y \\
& \qquad\qquad - b^3\int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y \int K_{\hh}(\y - \z)\fu^T(\y, \frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y.
\end{align*}
```

Theorem \@ref(thm:loccor3) and its assumptions above are variations of the assumptions and the results presented by @hjort1996locally, and in particular Theorem 3 by @tjostheim2013local. The analysis of the asymptotic behavior of $\Jb$ and $\Mb$ in the latter reference reveals that the local parameter estimates converge more slowly than, for example, the trivariate kernel density estimator (that converges as $1/(nb^3)^{1/2}$), and it is a straightforward exercise to modify their proof to suit our particular situation, which demonstrates that the three local correlations converge jointly in distribution at the rate $1/(nb^5)^{1/2}$, see Section B in the supplement for details.

Furthermore, @tjostheim2013local show that an analytic expression of the leading term in the covariance matrix $\Jb^{-1}\Mb(\Jb^{-1})^T$ is not practically available. In applications we may circumvent this problem by using the bootstrap. The validity of the bootstrap can be proved using a direct adaptations of the techniques of @lacal2018estimating. It is spelled out in some detail for the pairwise case in Theorems E.1, E.2 and corollaries E.3 and E.4 in Section E of the online supplement. The same arguments can be used in the present 3-parameter case (in fact, @lacal2018estimating treat the more general five-parameter case).

The pairwise estimation structure that we describe in Section \@ref(chap:bivariate) simplifies the asymptotic analysis of $\hR(\z)$ considerably. The result is essentially available in the literature as Theorem 2 in @otneim2017conditional, who state under similar assumptions as in Theorem \@ref(thm:loccor3) that <!--so we modify the notation slightly, and reproduce Theorem 3 in @otneim2017conditional. For a pair of variables $(X_j,X_k)$ and its corresponding pair of transformed variables $(Z_j, Z_k)$ (the relation between the $x$- and $z$-scale is again assumed to be known at this stage). We then estimate the single local correlation $\rho_{jk}(x_j,x_k) = \rho_{ij}(z_j,z_k)$ by maximizing the local likelihood function in \@ref(eq:locallik2), denoted by $L_n(\rho_{ij}, \z)$, For a fixed matrix of bandwidths $\hh$, now a diagonal $2\times2$-matrix, define $\rho_{\hh}$ in a similar way as above as the solution to

\begin{equation}
\frac{\partial L_n(\rho_{\hh}, \z)}{\partial\rho} \rightarrow \int K_{b}(\y - \z)u(\y, \rho_{\hh})\{f_{jk}(\y) - \psi(\y, \rho_{\hh})\}\,\textrm{d}\y = 0
(\#eq:popfixedh)
\end{equation}
as $n \rightarrow \infty$, but now $f_{jk}(\z)$ is the joint density of $(Z_j, Z_k)$, $u(\cdot, \rho_{\hh})$ is the local score function $\partial \log\psi(\cdot, \rho_{\hh})/\partial\rho$, and $K(\cdot)$ and $\psi(\cdot)$ are bivariate versions of the kernel function and the simplified Gaussian density associated with \@ref(eq:lgdeapprox) respectively. Denote by $\frho = \{\rho_{jk}\}_{j<k}$ the vector of all local correlations defined between all pairs of variables in $\X$. The joint limiting distribution of $\hfrho(\z)$ is then given by the following result [@otneim2017locally]:

\vspace{.5cm}

```{theorem, loccor}
Let $\{\Z\}_n$ be observations on the marginally standard normally distributed random vector $\Z$, and for each pair of variables $(Z_j, Z_k)$, $j<k\leq p$, assume that the following conditions hold:

1. For any sequence of bandwidth matrices $\hh = \hh_n$ tending to zero element-wise, there exists for the bivariate marginally standard Gaussian vector $(Z_j, Z_k)$ a unique $\rho_{\hh}(\z)$ that satisfies \@ref(eq:popfixedh), and there exists a $\rho_0(\z)$ such that $\rho_{\hh}(\z) \rightarrow \rho_0(\z)$.
2. For each pair $(j,k)$ of variables, with $1\leq j\leq p$, $1\leq k\leq p$, $j\neq k$, $\{(Z_j,Z_k)\}_n$ is $\alpha$-mixing with mixing coefficients satisfying $\sum_{m\geq1}m^{\lambda}\alpha(m)^{1-2/\delta} < \infty$ for some $\lambda > 1-2/\delta$ and $\delta>2$.
3. $n\rightarrow\infty$, and each bandwidth $b$ tends to zero such that $nb^{\frac{\lambda+2-2/\delta}{\lambda + 2/\delta}} = O(n^{\epsilon_0})$ for some constant $\epsilon_0 > 0$. 
4. For a given location $\z$, the parameter space $\Theta$ for each local correlation $\rho(\z)$ is a compact subset of $(-1,1)$.
5. The kernel function $K(\cdot)$ sasisfies $\sup_{\z}|K(\z)| < \infty$, $\int |K(\y)|\,\textrm{d}\y < \infty$, $\partial/\partial z_iK(\z) <\infty$, and $\lim_{z_i \rightarrow \infty}|z_iK(\z)| = 0$ for $i = 1,2$.

Then, by letting $b_n^2$ (assuming identical convergence rates) mean the product $b_{nj}b_{nk}$ corresponding to the pairs of variables defining the components in $\frho$, -->
\begin{equation}
\sqrt{nb_n^2}\left(\hfrho(\z) - \frho_0(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \fOmega),
\label{eq:ot2018}
\end{equation}
<!--
where $\fOmega$ is a diagonal matrix with components
\begin{equation}
\fOmega^{(\ell,\ell)} = 
\frac{f_{\ell}(\z_{\ell})\int K^2(\y_{\ell})\,\textrm{d}\y_{\ell}}
{u^2(\z_{\ell},\rho_{0,\ell}(\z_{\ell}))}
\psi^2(\z_{\ell}, \rho_{0,\ell}(\z_{\ell})),
(\#eq:asymcov)
\end{equation}
where $\ell = 1,\ldots,p(p-1/2)$ runs over all pairs of variables, and $f_{\ell}(\cdot)$ is the bivariate marginal density function of the pair $\Z_{\ell}$.
```

\vspace{.5cm}

For a proof, we refer to @otneim2017conditional.--> where the asymptotic covariance matrix $\fOmega$ is defined in an analogous way as in Theorem \@ref(thm:loccor3) There are two important differences between these two results, however. First of all, we have that the convergence rate in the simplified pairwise case is equal to the usual nonparametric rate of $1/(nb^2)^{1/2}$. This is because we estimate the local correlations one by one based on pairs of variables. The second difference is that the leading term of the asymptotic covariance matrix in Theorem 2 by @otneim2017conditional has a simple analytic expression that can be used to construct, for instance, confidence intervals. We refer to Section A.3 in the online supplement for further details regarding the limiting distribution of the LGPC as well conditions for consistency of the estimates.

We can now calculate the limiting distribution of the LGPC by means of the delta method. Denote by $g: \mathbb{R}^{p(p-1)/2} \rightarrow \mathbb{R}$ the translation \@ref(eq:definition) from the vector of local correlations between the components in $\Z$ (or, equivalently, $\X$) to the local partial correlation $\alpha(\z)$ between $Z_1$ and $Z_2$ given $Z_3, \ldots, Z_p$. From the expression of our estimate \@ref(eq:estimate) we have that $\halpha(\z) = g(\hfrho(\z))$, so we use the delta method to see that

\begin{equation}
\sqrt{nb_n^m}\left(\halpha(\z) - \alpha(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}\left(0, \nabla g(\frho)^T\fLambda\nabla g(\frho)\right),
(\#eq:limitlgpc)
\end{equation}
where $m$ is equal to 2 or 5, and $\fLambda$ is either equal to the leading term of $\Jb^{-1}\Mb(\Jb^{-1})^T$, or $\fOmega$ in \@ref(eq:ot2018), depending on whether we use the full trivariate locally Gaussian fit described in Section \@ref(chap:trivariate-full) or the sequentially pairwise simplification described in Section \@ref(chap:bivariate). In any case, it is only a matter of basic differentiation to work out an expression for $\nabla g(\cdot)$, and this task has been deferred to Section C of the online supplement in the case where $d_1 = d_2 = 1$.

Finally, we present the limiting distribution of the LGPC between $Z_1$ and $Z_2$ given $Z_3, \ldots, Z_p$ when the components in $\Z = (Z_1, \ldots, Z_p)$ are replaced by marginally standard normal *pseudo-observations* as defined in equation \@ref(eq:pseudoobservations). The following result ensures that Theorems \@ref(thm:loccor3) and Theorem 3 by @otneim2017conditional, and thus eq. \@ref(eq:limitlgpc), still holds.

\vspace{.5cm}

```{theorem, pseudo}
Under the conditions in Theorems \@ref(thm:loccor3) and Theorem 2 by @otneim2017conditional, assume further that the kernel function $K(\cdot)$ has bounded support. Replacing the marginally standard normal vector $\Z$ with the approximately marginally standard normal vector $\hZ$, does not change the conclusions in these theorems.
```

```{proof}
See Section D of the online supplement.
```

# EXAMPLES {#chap:example}

We will in this section provide some demonstrations on how the estimated LGPC can be used to reveal nonlinear departures from conditional independence and to measure that conditional dependence. All plots are presented on the $x$-scale.

First, we provide some reference pictures from the simplest situation imaginable. Let $X_1$, $X_2$ and  $X_3$ be independent standard (and thus also jointly) normal variables. Since $X_1$ is conditionally independent from $X_2$ given $X_3$ (written $X_1 \perp X_2 | X_3$) in this particular case, we also know from property 3 in Section 2.2 that the LGPC between $X_1$ and $X_2$ is zero everywhere.

\begin{figure}
    \subfloat[$n = $ 100\label{fig:gaussian-example-plots1}]{\includegraphics[width=0.31\linewidth]{figures/gaussian-example-plots-1} }
    \subfloat[$n = $ 500\label{fig:gaussian-example-plots2}]{\includegraphics[width=0.31\linewidth]{figures/gaussian-example-plots-2} }
    \subfloat[$n = $ 2000\label{fig:gaussian-example-plots3}]{\includegraphics[width=0.31\linewidth]{figures/gaussian-example-plots-3} }
    \caption{Estimated local Gaussian partial correlation between $X_1$ and $X_2$ given $X_3 = 0$, where $(X_1, X_2, X_3)$ is a jointly standard normally distributed vector. The fully trivariate model of Section 3.1.1 is used.}\label{fig:gaussian-example-plots}
\end{figure}

In Figure \@ref(fig:gaussian-example-plots) we see the estimated LGPC between $X_1$ and $X_2$ given $X_3=0$ for three samples, having sample size 100, 500 and 2000 respectively, mapped out on a grid. In this and all other examples we select bandwidths based on a simple plug-in formula that follows naturally from classical asymptotic arguments, $b = cn^{-1/9}$ for the full trivariate fit, and $b = cn^{-1/6}$ for the bivariate simplification, where the constant $c$ controls the amount of smoothing and must be chosen appropriately based on the task at hand. @otneim2016multivariate argues that $c=1.75$ is a good choice for multivariate density estimation, and we will in the next section see that we obtain good power in our test for conditional independence if $c$ is somewhat smaller than that. For the visual display of conditional dependence maps, however, we tend to prefer a fair amount of smoothing, and in Figure \@ref(fig:gaussian-example-plots) we have used $c=4$. @otneim2017locally also suggest a algorithm for bandwidth selection in the density estimation context, maximizing the leave-one-out cross validated log-likelihood of the sample;
$$\bb_{CV} = \textrm{arg max}_{\bb} \sum_{t = 1}^n \widehat{f}_{\bb}^{(-t)}(\Z_t),$$
where $\widehat f_{\bb}^{(-t)}(\Z_t)$ is the estimated local Gaussian density estimate calculated without, and evaluated in, the sample point $\Z_t$, using the bandwidth matrix $\bb$. A theory of efficient and accurate bandwidth selection for local likelihood estimation remains a topic for future research.

As expected, $\widehat\alpha(\x) = \widehat\alpha(\z)$ is close to zero in all three plots in Figure \@ref(fig:gaussian-example-plots) and a significance test carried out as in Section \@ref(chap:testing) does not reject conditional independence at any reasonable significance level.

\begin{figure}[p]
    \subfloat[Observed pairs $(X_1, X_2)$ from the conditionally \newline Gaussian model.\label{fig:simex2-plots1}]{\includegraphics[width=0.49\linewidth]{figures/simex2-plots-1} }
    \subfloat[The estimated LGPC between $X_1$ and $X_2$ given $\rho = -0.9$.\label{fig:simex2-plots2}]{\includegraphics[width=0.49\linewidth]{figures/simex2-plots-2} }\newline
    \subfloat[The estimated LGPC between $X_1$ and $X_2$ \newline  given $\rho = 0$\label{fig:simex2-plots3}]{\includegraphics[width=0.49\linewidth]{figures/simex2-plots-3} }
    \subfloat[The estimated LGPC between $X_1$ and $X_2$ \newline given $\rho = 0.9$\label{fig:simex2-plots4}]{\includegraphics[width=0.49\linewidth]{figures/simex2-plots-4} }
    \caption{The conditionally Gaussian model}\label{fig:simex2-plots}
\end{figure}

In the next simulated example we will demonstrate an important feature of the LGPC: It is able to *distinguish between positive and negative conditional relationships*, which, to our knowledge, has until now not been possible beyond the linear and jointly Gaussian setting using the ordinary partial correlation. Our approach also allows exploration of conditionally different dependence patterns across different levels of the conditioning variable.

Generate $X_3 = \rho \sim U(-1,1)$, and then generate $(X_1, X_2)$ from the bivariate Gaussian distribution having standard normal marginals, and correlation coefficient equal to $\rho$. We observe $\X = (X_1, X_2, \rho)$ and seek to visualize the dependence between $X_1$ and $X_2$ conditional on $\rho$.

We see the results in Figure \@ref(fig:simex2-plots). In panel (a), we generate $n=1000$ simulated pairs $(X_1, X_2)$ from this model. In panel (b)-(d) we see the estimated LGPC plotted over suitable grids, and where the conditioning variable $X_3 = \rho$ has been fixed at the respective values $-0.9$, $0$ and $0.9$, and we see clearly how the dependence between $X_1$ and $X_2$ dramatically changes. It is straightforward to see that $X_1$ and $X_2$ are uncorrelated and that the ordinary partial correlation between $X_1$ and $X_2$ given $X_3$ is equal to zero. Furthermore, we note that the pairwise simplification defined in Section 3.1.2 would also not be able to measure the *conditional* relationship between $X_1$ and $X_2$ given $X_3$ in this case, as we see clearly from eq. \@ref(eq:scalardefinition), because $X_1$ and $X_2$ are both marginally independent from $X_3$. This means that the two pairwise correlations $\rho_{13}(z_1,z_3)$ and $\rho_{23}(z_2,z_3)$ are equal to zero.

We conclude this section by using the LGPC to analyze real data in an example regarding Granger causality between financial time series. Granger causality has been a central concept in economics and econometrics ever since its inception by @granger1969investigating. In layman's terms, the time series $\{X_t\}$ Granger causes $\{Y_t\}$ if past values of $\{X_t\}$ are helpful when predicting future values of $\{Y_t\}$. Formally, $\{X_t\}$ Granger causes $\{Y_t\}$ if [@granger1980testing]

\begin{equation}
Y_t \not\perp \mathcal{I}^*(t-1) \,\,|\,\, \mathcal{I}_{-X}^*(t-1),
(\#eq:granger)
\end{equation}
where $\perp$ denotes independence, $\not\perp$ denotes dependence, and where $\mathcal{I}^*(t-1)$ is all information available at time $t-1$ and $\mathcal{I}_{-X}^*(t-1)$ is the same information, with the exception of the values of $\{X_t\}$ up to, but not including, time $t$. Of course, the hypothesis \@ref(eq:granger) can not be tested in practice in its full generality. By taking only effects up to the first lag into account, we may formulate a sufficient (but not necessary) condition for \@ref(eq:granger): $Y_t \not\perp X_{t-1} \,\,|\,\, Y_{t-1}$, the converse of which constitutes a testable null hypothesis of Granger non-causality:

\begin{equation}
\textrm{H}_0: Y_t \perp X_{t-1} \,\,|\,\, Y_{t-1}.
(\#eq:grangernull)
\end{equation}
There are many ways to carry out this test. The simplest, and original one, is based on the further restriction of linear relationships between $\{X_t\}$ and $\{Y_t\}$ and is thus a test for *conditional uncorrelatedness* rather than conditional independence. Nonparametric tests that have power against many nonlinear types of conditional dependence have also been developed, and we refer to Section \@ref(chap:testing) for further references to this literature and a new test for conditional independence based on the LGPC.

@fran:moug:vale:2010 investigate how stock returns of large firms may Granger cause the stock returns of small firms and vice versa from a nonlinear standpoint. They point to earlier evidence that such nonlinear lead-lag relationships have been found and investigate this further using the test for Granger non-causality by @baek:broc:1992, and using a long series of US data they find evidence that there is a bi-directional causal link between the stock prices of the firms in the most valuable quintile (20 percent) and the firms in the least valuable quintile. We will repeat this exercise for stocks traded on the Oslo Stock Exchange in Norway, but this time from a point of view using the estimated LGPC.

Denote by $R_{1,t}$ the log-returns on a value-weighted portfolio consisting of the most valuable quintile of firms listed on the Oslo Stock Exchange as of 27 April 2020 (the big firms), and denote by $R_{2,t}$ the corresponding log-returns for the least valuable companies (the small firms). We will investigate the possibility of Granger causality from the small to the big firms, the hypothesis being that smaller firms are more market sensitive and quickly react to changes in the economy, leading the larger firms that more slowly adapt to small variations. The null hypothesis of non-causality of the first order \eqref{eq:grangernull} in this particular case means that

\begin{equation}
R_{1,t} \perp R_{2, t-1} \,\,|\,\, R_{1,t-1},
\end{equation}
which we may test formally using the approaches described in the following section or the references therein. Let us rather concentrate on *describing* this conditional dependence relationship by estimating the local Gaussian partial correlation between $R_{1,t}$ and $R_{2,t-1}$ given a few different values of $R_{2,t-1}$, which gives considerable more information than a single $p$-value of a test. We collect daily observations on Oslo Stock Exchange^[Thanks to \emph{Børsprosjektet (The Stock Exchange Project)} at the Norwegian School of Economics for providing the data.] from 1 January 2018 until 31 December 2019. In Figure \ref{fig:order1} we see the estimated LGPC along the diagonal $R_{1,t} = R_{2,t-1}$ for three different values $R_{1,t-1} = -0.02, 0$ and $0.02$, along with their block-bootstrapped 95% confidence bands. The constant $\alpha(\z) = 0$ is indicated as a dotted line.

\begin{figure}[t]
\centering
\includegraphics{figures/granger-order1}
\caption{The estimated LGPC between $R_{1,t}$ and $R_{2,t-1}$ given $R_{1,t-1}$ along the diagonal of the sample space, for three values of the conditioning variable}
\label{fig:order1}
\end{figure}

One could perhaps make the argument that $R_{1,t}$ and $R_{2,t-1}$ seem to be somewhat positively dependent in the right half of the graph for *small* values of the conditioning variable $R_{1,t-1}$. Roughly translated to this particular situation this means that if the large-firm portfolio gives a negative return on day $t-1$, then its value on day $t$ depend on the performance small-firm portfolio on day $t-1$ in a nonlinear and non-Gaussian way that is characterized by positive dependence in the first quadrant. The evidence displayed in Figure \ref{fig:order1} is quite weak, however, because the confidence bands include zero in most of the sample space.

@fran:moug:vale:2010 go on to analyze if there is Granger causality from small to big firms in the US data if another lag is taken into account, which in general means finding evidence against the hypothesis

\begin{equation}
R_{1,t} \perp (R_{2,t-1}, R_{2, t-2}) \,|\, (R_{1, t-1}, R_{1, t-2}),
\label{eq:order2}
\end{equation}
departures from which can be quantified and tested for in a linear fashion using multiple regressions, or tested for directly by means of a nonlinear test as we discuss in the following Section. Let us here, on the other hand, use the LGPC to quantify the conditional dependence relationship \eqref{eq:order2} nonlinearly for the Norwegian returns data in order to obtain more detailed information. This means estimating a $3\times3$ partial correlation matrix $\hfalpha(\z)$, which we again do along the diagonal $R_{1,t} = R_{2,t-1} = R_{2, t-2}$ in this particular case.

\begin{figure}[t]
\centering
\includegraphics{figures/granger-order2}
\caption{The estimated local Gaussian partial correlations for the three variable pairs in $(R_{1,t}, R_{2,t-1}, R_{2, t-2})$ given $(R_{1, t-1}, R_{1, t-2})$ estimated along the diagonal in the sample space. The first two plots represent the pairs $(R_{1,t}, R_{2,t-1})$ and $(R_{1,t}, R_{2,t-2})$ which are of particular interest when analyzing the conditional dependence relationship \eqref{eq:order2}. The legend in Figure \ref{fig:order1} is still valid, except that the solid line now represents the LGPC conditional on \emph{both} $R_{1, t-1}$ and $R_{1, t-2}$ being equal to $-0.02$.}
\label{fig:order2}
\end{figure}

Including another lag does not change the estimated conditional relationship between $R_{1,t}$ and $R_{2,t-1}$ much compared to the first order estimate displayed in Figure \ref{fig:order1}. There does, however, seem to be a clearer picture when considering the LGPC between $R_{1,t}$ and $R_{2,t-2}$, which seems to be positive in the third quadrant for all values of the conditioning variables. In this particular problem this means that conditioning on the log return of the small-firm portfolio on days $t-1$ and $t-2$, the large-firm log-return on day $t$ depends on the small-firm  log-return on day $t-2$ in a nonlinear way, and more specifically; this conditional relationship can be described using the LGPC as stronger dependence in the lower left tail of the relevant conditional distribution. We present another example of Granger testing in addition to another simulated experiment in Section J of the online supplement.

# TESTING FOR CONDITIONAL INDEPENDENCE {#chap:testing}

We should point out here that the purpose of the LGPC is not primarily to compete with the nonparametric conditional independence *testing* literature, but rather to provide a new tool for the measurement of nonlinear conditional dependence. The following section could however be seen as an indication of the potential of using the LGPC to test for conditional independence. It remains to be investigated whether further improvements can be realized by using a 5-parameter specification and/or a focused version of the test; see e.g. Tables 3--5 of @lacal2017local for the substantial improvements in a similar focused independence test.

We restrict ourselves to the case where $X_1$ and $X_2$ are scalars, since this is the situation considered in the papers we compare with, but we reiterate (end of Section \@ref(chap:definition)) that there are no conceptual difficulties in extending to the vector case as is seen in the online supplement. The theory underlying LGPC testing is essentially based on @lacal2018estimating, and is outlined in Sections F--H of the supplement.

## The recent fauna of nonparametric tests {#fauna}

Property 3 in Section \@ref(properties) states that the LGPC characterizes conditional dependence within a quite large class of distributions: Two stochastic vectors $X_1$ and $X_2$ are independent given $\X_3$ if and only if the locally Gaussian partial correlation between them is identically equal to zero everywhere on the sample space of $(X_1, X_2, \X_3)$. The road is therefore short to a test for conditional independence that may have power against a great deal of nonlinear alternatives compared to a test based on the ordinary partial correlation coefficient.

The last decade or so has seen the publication of many new tests for conditional independence. Su and White have published a series of such tests: @su2007consistent is based on detecting differences between estimated characteristic functions (which is also the method used by @wang2017characteristic). @su2008nonparametric is based on estimating the Hellinger distance between conditional density estimates, @su2012conditional use local polynomial quantile regression to test for conditional independence, and @su2014testing present a test based on empirical likelihood. @huang2010testing introduces the maximal nonlinear conditional correlation which is used in a test for conditional independence, in turn extended to dependent data by @cheng2012conditional. @song2009testing constructs a test via Rosenblatt transformations, while @bergsma2011nonparametric and @bouezmarni2012nonparametric present new tests for conditional independence based on copula constructions. @boue:taam:2014 test for conditional independence by measuring the $L_2$ distance between estimated conditional distribution functions, and @patra2016on use empirical transformations to translate conditional independence to joint independence, tests for which exist in abundance. @wang2015conditional introduce the *conditional distance correlation* which they use to construct a test for conditional independence. Linton and Gozalo [-@linton1997conditional; -@linton2014testing] formulate conditional independence in terms of probability statements which then forms the basis of a test. Also, @delgado2001significance develop a test for conditional independence in a nonparametric regression framework. Finally, we mention testing by way of characterizing conditional independence via reproducing kernel Hilbert spaces (RKHS), see, for example, @zhang2012kernel and references therein.

A test based on the LGPC is quite different from the methods quoted above. It is semi-parametric and does not rely on traditional kernel estimates of density-, \mbox{distribution-,} or characteristic functions. Furthermore, due to our transformation of the data \@ref(eq:trans) to marginal standard normality, we do not necessarily have to specify weight functions in our test functional to lessen the impact of outliers - unless of course we wish to test for conditional independence in a specific portion of the sample space.

On the other hand, we have to pay a price when imposing structure on the dependence. Certain types of conditional dependence will remain invisible to our test statistic, but simulation experiments show that our test performs on par with, and sometimes better than existing, fully nonparametric tests.

## A test for conditional independence based on the LGPC {#thetest}

We construct a test statistic for testing

$$\textrm{H}_0: X_1 \perp X_2 \,\, | \,\, \X_3$$
or, equivalently, a test statistic in terms of the marginally Gaussian pseudo observations,

\begin{equation}
\textrm{H}_0: Z_1 \perp Z_2 \,\, | \,\, \Z_3
(\#eq:ci-test)
\end{equation}
by aggregating our local measure of dependence over the sample space of $\X$ (or $\Z$). A natural test statistic on the $z$-scale is

\begin{equation}
T_{n,b} = \int_S h\left( \widehat \alpha_b(\z)  \right) \di F_n(\z),
(\#eq:teststatistic)
\end{equation}
where $h(\cdot)$ is a real-valued function that is typically even and non-negative for most standard applications, and $S \subseteq \mathbb{R}^p$ is an integration area that can be altered in order to focus the test to specific portions of the sample space. Standard laws of large numbers ensure that, under regularity conditions, see Theorem F.1 of the supplement, $T_{n,b}$ converges in probability towards its population value $$T = \int_S h\left(\alpha(\z)\right) \di F(\z).$$

Departures from many types of conditional independence lead to large values of $T_{n,b}$ that, if larger than a critical value, leads to the rejection of \@ref(eq:ci-test). Asymptotically, one might expect that approximate $p$-values for our test can be extracted from the limiting distribution of the test statistic $T_{n,b}$, which we derive in the online supplement Section F. The limiting asymptotic distribution of $T_{n,b}-T$ is given in Theorem F.2 for the 1-parameter general pairwise case, whereas Theorem F.3 treats the 3-coordinate (3-parameter) case with $d_1=d_2=d_3=1$. It is interesting to note from the discussion immediately following the proofs of these theorems that the convergence rate in the first case is $n^{-1/2}$, and in the second case is $\left(nb^2\right)^{-1/2}$. This means that the tests are asymptotically able to pick up alternatives approaching the null hypothesis of conditional independence at these rates. This is measured by the Pitman criterion. A discussion of the Pitman criterion for other tests of conditional independence is given on pages 11--12 immediately following Theorem 2 of @wang2017characteristic. The rate $n^{-1/2}b^{-d_3/4}$ of Wang and Hong compares favorable with other tests. The rate $n^{-1/2}$ obtained by us must be interpreted with caution, however, because it is based on a pairwise model that is more restrictive. We go into more detail regarding asymptotic power and Pitman efficiency in Section G of the online supplement.

Several authors, for example @terasvirta2010modelling, have noted, however, that asymptotic analysis of nonparametric (or semiparametric) test statistics on the form \@ref(eq:teststatistic) tends to be too crude for finite sample applications. We have therefore approximated the distribution of $T_{n,b}$ using the bootstrap. The validity of the bootstrap in such contexts has been examined by @lacal2018estimating and we adapt their arguments to our situation in the online supplement, Section H.

In accordance with our treatment so far, let $\{Z_{1,t}, Z_{2, t}, \Z_{3, t}\}$, $t = 1,\ldots,n$ be observations on the $p$-variate stochastic vector $\Z$, $p\geq3$, with $Z_1$ and $Z_2$ being scalar and $\Z_3$ being $(p-2)$-variate. In order to calculate the statistic $T_{n,b}$ for testing the null hypothesis $Z_1 \perp Z_2 \,\,|\,\, \Z_3$, we must estimate the joint local correlation matrix $\fSigma(\z)$ of $\Z = (Z_1, Z_2, \Z_3)$. We exploit this in the following algorithm designed to produce approximate resampled versions of $T_{n,b}$ under the null hypothesis of conditional independence. This procedure is a variation of the so-called local bootstrap which has been used by several authors when testing for conditional independence. A brief survey of this procedure is given in Section I of the online supplement.

1. Use the local correlation estimates to estimate the conditional densities $f_{Z_1|\Z_3}(\cdot)$ and $f_{Z_2|\Z_3}(\cdot)$ by means of the method by @otneim2017conditional. In practice, in order to reduce the computational load, we evaluate $\widehat f_{Z_1|\Z_3}$ and $\widehat f_{Z_2|\Z_3}$ on a fine grid on their support, over which a continuous representation of the estimates are produced using cubic splines.
2. Using the accept-reject algorithm, generate $B$ samples, each of size $n$, from $\widehat f_{Z_1|\Z_3}$ and $\widehat f_{Z_2|\Z_3}$, leading to $B$ replicates 
$$Z_m^* = \Big\{\{Z_{1t}^*, Z_{2t}^*, \Z_{3t}\}_i\Big\}_m, \qquad t = 1,\ldots,n, \,\, m = 1,\ldots,B,$$
of $\Z$ under the null hypothesis.
3. Calculate $\{T_{n,b}\}_m^*$, $m = 1,\ldots,B$ for the replicated data sets and obtain the approximate $p$-value for the conditional independence test.

## Comparing with other tests {#chap:testexamples}

@su2008nonparametric formulate a simulation experiment for evaluating their nonparametric test for conditional independence by generating data from 10 different data generating processes (DGPs), and then check the power and level for their test. Many of the later works that were discussed in Section \@ref(fauna) contain very similar experiments, and some of them present simulation results for tests published prior to @su2008nonparametric. This allows us to present comprehensive comparisons between the new test presented above, and many alternatives.

Let $(\epsilon_{1,t}, \epsilon_{2,t}, \epsilon_{3,t})$ be IID observations from a $\mathcal{N}(0, I_3)$-distribution, where $I_3$ is the $3\times3$ identity matrix. We will test H$_0: X_{1,t} \perp X_{2,t} \,\,|\,\, X_{3,t}$ in 10 cases listed in Table \@ref(tab:dgps) taken from @su2008nonparametric, that cover various types of linear and nonlinear time series dependence.

\begin{table}[t]
\begin{tabular}{ll}
\hline
1.& $(X_{1,t}, X_{2,t}, X_{3,t}) = (\epsilon_{1,t}, \epsilon_{2,t}, \epsilon_{3,t})$.\\
2.& $X_{1,t} = 0.5X_{1, t-1} + \epsilon_{1,t}$, $X_{2,t} = 0.5X_{2,t-1} + \epsilon_{2,t}$, $X_{3,t} = X_{1,t-1}$.\\
3.& $X_{1,t} = \epsilon_{1,t}\sqrt{0.01 + 0.5X_ {t-1}^2}$, $X_{2,t} = 0.5X_{2, t-1} + \epsilon_{2,t}$, $X_{3,t} = X_{1,t-1}$.\\
4.& $X_{1,t} = \epsilon_{1,t}\sqrt{h_{1,t}}$, $X_{2,t} = \epsilon_{2,t}\sqrt{h_{2,t}}$, $X_{3,t} = X_{1,t-1}$, \\ & $h_{1,t}  = 0.01 + 0.9h_{1,t-1} + 0.05X_{1,t-1}^2$, $h_{2,t} = 0.01 + 0.9h_{2,t-1} + 0.05X_{2,t-1}^2$. \\
5.& $X_{1,t} = 0.5X_{1,t-1} + 0.5X_{2,t} + \epsilon_{1,t}$, $X_{2,t} = 0.5X_{2,t-1} + \epsilon_{2,t}$, $X_{3,t} = X_{1,t-1}$.\\
6.& $X_{1,t} = 0.5X_{1,t-1} + 0.5X_{2,t}^2 + \epsilon_{1,t}$, $X_{2,t} = 0.5X_{2,t-1} + \epsilon_{2,t}$, $X_{3,t} = X_{1,t-1}$.\\
7.& $X_{1,t} = 0.5X_{1,t-1}0.5X_{2,t} + \epsilon_{1,t}$, $X_{2,t} = 0.5X_{2,t-1} + \epsilon_{2,t}$, $X_{3,t} = X_{1,t-1}$.\\
8.& $X_{1,t} = 0.5X_{1,t-1} + 0.5X_{2,t}\epsilon_{1,t}$, $X_{2,t} = 0.5X_{2,t-1} + \epsilon_{2,t}$, $X_{3,t} = X_{1,t-1}$.\\
9.& $X_{1,t} = \epsilon_{1,t}\sqrt{0.01 + 0.5X_{1,t-1}^2 + 0.25X_{2,t}}$, $X_{2,t} = X_{2,t-1} + \epsilon_{2,t}$, $X_{3,t} = X_{1,t-1}$.\\
10. & $X_{1,t} = \epsilon_{1,t}\sqrt{h_{1,t}}$, $X_{2,t} = \epsilon_{2,t}\sqrt{h_{2,t}}$, $X_{3,t} = X_{1,t-1}$, $h_{1,t} = 0.01 + 0.1h_{1, t-1} + 0.4X_{1,t-1}^2 +$ \\ & $0.5X_{2,t}^2$, $h_{2,t} = 0.01 + 0.9h_{2, t-1} + 0.5Y_t^2$.\\
\hline
\end{tabular}
\caption{Data generating processes in simulation experiment}
\label{tab:dgps}
\end{table}

The null hypotheses of conditional independence between $X_{1,t}$ and $X_{2,t}$ given $X_{3,t}$ is true for DGPs 1--4, and these will be used to check the level of the test, while for DGPs 5--10 we measure the power. By evaluating our test using the sample sizes $n = 100$ and $n = 200$ at the $5\%$ level, we can harvest a great number of corresponding results from the literature, and they are presented below.

The first set of simulation results are collected from @su2008nonparametric: In Table \ref{tab:n100} LIN is a standard linear Granger causality test, CM and KS are tests by @linton1997conditional that use statistics of the two versions of the nonparametric test developed by @delgado2001significance. HEL refers to the Hellinger distance test that @su2008nonparametric present. Next, we move to @su2014testing, who provide simulations for their test for conditional independence based on the empirical likelihood (SEL), as well as the test by @su2007consistent that is based on properties of the conditional characteristic function (CHF). @cheng2012conditional provide simulations for their tests based on the maximal conditional correlation (MCC). In the online supplement to this article we provide a richer version of Table \@ref(tab:n100) below, namely Table 2*, that includes more results for the HEL- and MCC-tests at different levels of smoothing.

\input{tables/power_table_dim3_n100_short}

Finally, we include results from simulations using our new test based on the LGPC, using the trivariate specification defined in Section \@ref(chap:trivariate-full), for two different levels of smoothing, and include them in Table \ref{tab:n100}. We highlight the new results in grey to indicate that they, as opposed to all the other numbers, have not appeared in the literature before. Also, to the best of our knowledge, these results have not been compared simultaneously before.

We see in Table \ref{tab:n100} that our test has the correct level and is quite powerful against all alternative specifications of conditional dependence, in the additive models 5 and 6, as well as the remaining examples 7 to 10, that are more multiplicative in nature.

When $n=200$ we can include simulation results reported by @bouezmarni2012nonparametric for their conditional independence test based on measuring the Hellinger distance between copula density estimates (shortened BRT from the names of the authors) as well as the results reported by @boue:taam:2014 on their test based on $L_2$ distances between estimated conditional distribution functions, which is abbreviated by BT. We include in Table 3* in the online supplement all results from these papers. The general picture is the same as in Table \ref{tab:n100}. In particular for the case with $c=1.4$, our new test exhibits the best over-all performance among the examples listed, which, again to the best of our knowledge, include all such simulation results that have been published to date. It is seen that the standard linear Granger causality test (LIN) has a miserable performance in these examples.

@su2014testing define two extensions to a subset of the data generating processes defined in Table 1 in which the conditioning variable $\Xtwo$ is a vector with two and three components respectively. We refer to the online supplement for details regarding these extensions and numerical results. The new test for conditional independence based on the LGPC produces competitive results also in these cases. We point out again, however, that testing for conditional independence is merely one, albeit an important one, of several chief objectives for introducing the LGPC. Possible refinements of the LGPC testing is mentioned in the beginning of Section \@ref(fauna).

# CONCLUSION AND OUTLOOK

The purpose of this paper is two-fold. First we define and develop the LGPC, a signed local measure of conditional dependence that has useful properties and is easy to interpret. It ca measure both positive and negative conditional dependence. We then explore the possibility of using the LGPC in estimation of conditional dependence and indicate its potential in testing for conditional independence. The results are promising, and suggest that the newly developed tool provides powerful improvements to existing methods.

For 3 scalar variables $X_1, X_2, X_3$ a full trivariate analysis depending on all 3 coordinates can be undertaken as in Section \@ref(chap:trivariate-full). If any of the components $X_1, X_2$ or $X_3$ are vectors, a pairwise approximation and simplification has been described in Section \@ref(chap:bivariate). This can be likened to additive approximation in nonparametric regression.

Both the full trivariate approach and the pairwise simplification have been mainly analyzed at the $z$-scale using the transformed variables $\Z = \Phi^{-1}(F(\X))$. The corresponding LGPC $\alpha(\z)$ is invariant to monotone transformations of the marginals. The LGPC $\alpha(\x)$ on the $x$-scale can be obtained from $\alpha(\z)$ by taking $\x = F^{-1}(\Phi(\z))$, and it is in general different from $\alpha(\z)$ and is not invariant to monotone transformations of the marginals. An analogue is the different transformation properties of the Pearson correlation and the copula structure in describing joint dependence. For multivariate Gaussian variables, $\alpha(\x) \equiv \alpha(\z) \equiv \alpha$, which is the ordinary global partial correlation.

There is a potential for much further work. A next natural step may be structural equations as used in network analysis and Pearl-type causality [@pearl2000causality], and ultimately further exploring the relationship or lack of relationship between that type of causality and Granger causality.

All methods presented in this paper have been implemented in the R programming language [@r], and is available in the package **lg** [@otneim2019lg]. In addition, we provide code and data for easy replication of the results presented here in an online appendix.

**Acknowledgements.** The authors gratefully acknowledge the editor as well as two anonymus referees, whose comments have greatly improved this paper.

# REFERENCES

### 2020-09-02 lgpc_JBES.tex
Initial commit, revising for JBES.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
  letterpaper]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={The locally Gaussian partial correlation},
  pdfauthor={Håkon Otneim and Dag Tjøstheim},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin = 1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{setspace}
\usepackage{sectsty}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{bm}

\doublespacing
\sectionfont{\centering}

\newcommand{\X}{\bm{X}}
\newcommand{\Xone}{\bm{X}^{(1)}}
\newcommand{\Xtwo}{\bm{X}^{(2)}}
\newcommand{\xtwo}{\bm{x}^{(2)}}
\newcommand{\x}{\bm{x}}
\newcommand{\hX}{\widehat{X}}
\newcommand{\tX}{\tilde{X}}

\newcommand{\Z}{\bm{Z}}
\newcommand{\z}{\bm{z}}
\newcommand{\hz}{\widehat{z}}
\newcommand{\hfz}{\widehat{\bm{z}}}
\newcommand{\hZ}{\widehat{\bm{Z}}}
\newcommand{\hnZ}{\widehat{Z}}
\newcommand{\Zone}{\bm{Z}^{(1)}}
\newcommand{\Ztwo}{\bm{Z}^{(2)}}
\newcommand{\zone}{\bm{z}^{(1)}}
\newcommand{\ztwo}{\bm{z}^{(2)}}

\newcommand{\D}{\bm{D}}
\newcommand{\hD}{\widehat{\bm{D}}}

\newcommand{\y}{\bm{y}}
\newcommand{\Y}{\bm{Y}}
\newcommand{\hY}{\widehat{Y}}

\newcommand{\fv}{\bm{v}}
\newcommand{\fu}{\bm{u}}

\newcommand{\fC}{\bm{C}}
\newcommand{\fA}{\bm{A}}
\newcommand{\fV}{\bm{V}}
\newcommand{\M}{\bm{M}}
\newcommand{\J}{\bm{J}}

\newcommand{\R}{\bm{R}}
\newcommand{\hR}{\widehat{\bm{R}}}

\newcommand{\hF}{\widehat{F}}
\newcommand{\hf}{\widehat{f}}

\newcommand{\hrho}{\widehat{\rho}}
\newcommand{\frho}{\bm{\rho}}
\newcommand{\hfrho}{\widehat{\bm{\rho}}}

\newcommand{\hh}{\bm{b}}
\newcommand{\bb}{\bm{b}}

\newcommand{\fmu}{\bm{\mu}}
\newcommand{\hfmu}{\widehat{\bm{\mu}}}
\newcommand{\hsigma}{\widehat{\sigma}}
\newcommand{\fSigma}{\bm{\Sigma}}
\newcommand{\hfSigma}{\widehat{\bm{\Sigma}}}
\newcommand{\halpha}{\widehat{\alpha}}
\newcommand{\falpha}{\bm{\alpha}}
\newcommand{\hfalpha}{\widehat{\bm{\alpha}}}
\newcommand{\fOmega}{\bm{\Omega}}
\newcommand{\fLambda}{\bm{\Lambda}}
\newcommand{\fepsilon}{\bm{\epsilon}}

\newcommand{\Jb}{\bm{J}_{\hh}}
\newcommand{\Mb}{\bm{M}_{\hh}}

\newcommand{\Cov}{\textrm{Cov}}
\newcommand{\E}{\textrm{E}}
\newcommand{\Var}{\textrm{Var}} 
\newcommand{\di}{\,\textrm{d}}
\newcommand{\Corr}{\textrm{Corr}}

\usepackage{color}
\usepackage{colortbl}

\definecolor{Gray}{gray}{0.9}

\title{The locally Gaussian partial correlation}
\author{Håkon Otneim\footnote{Corresponding author. Department of Business and Management Science, NHH Norwegian School of Economics. Helleveien 30, 5045 Bergen, Norway. \texttt{hakon.otneim@nhh.no}} and Dag Tjøstheim\footnote{University of Bergen}}
\date{}

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\newtheorem{conjecture}{Conjecture}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}
\maketitle
\begin{abstract}
It is well known in econometrics and other fields that the dependence structure for jointly Gaussian variables can be fully captured using correlations, and that the \textit{conditional} dependence structure in the same way can be described using partial correlations. The partial correlation does not, however, characterize conditional dependence in many non-Gaussian populations. This paper introduces the local Gaussian partial correlation (LGPC), a new measure of conditional dependence. It is a local version of the partial correlation coefficient that characterizes conditional dependence in a large class of populations. It has some useful and novel properties besides: The LGPC reduces to the ordinary partial correlation for jointly normal variables, and it distinguishes between positive and negative conditional dependence. Furthermore, the LGPC can be used to study departures from conditional independence in specific parts of the distribution. We provide several examples of this, both simulated and real, and derive estimation theory under a local likelihood estimation framework. Finally, we indicate how the LGPC can be used to construct a powerful test for conditional independence, which, for example, can be used to detect nonlinear Granger causality in time series. \newline \textbf{Keywords:} \emph{Conditional dependence, conditional independence test, Granger causality}
\end{abstract}

\hypertarget{introduction}{%
\section{INTRODUCTION}\label{introduction}}

Estimation of conditional dependence and testing for conditional independence are important topics in classical as well as modern statistics. In the last two decades, for instance, there have been intense developments in using conditional dependence for probabilistic networks, conditional multivariate time series analysis as well as copula analysis. For jointly Gaussian variables, conditional dependence is measured by the partial correlation coefficient. Given three jointly Gaussian stochastic variables \(X_1\), \(X_2\) and \(X_3\), \(X_1\) and \(X_2\) are conditionally independent given \(X_3\) if and only if the partial correlation between \(X_1\) and \(X_2\) given \(X_3\) is equal to zero. There is a rich literature on applications of the partial correlation coefficient to Gaussian networks, path analysis and causality. The Gaussian assumption is strict, however, and it is easy to find non-Gaussian examples where the partial correlation function completely fails in describing conditional dependence, as will be seen later in this article.

The purpose of this paper is to introduce a new concept for measuring conditional dependence. This concept retains all the properties of the ordinary partial correlation in the Gaussian case, but seeks to avoid the weaknesses of this measure in the non-Gaussian case. This is done by fitting a \emph{family} of Gaussian distributions to a given continuous multivariate distribution and exploiting the simple conditioning rules of the Gaussian distribution \emph{locally}. This approach produces a new measure of conditional dependence, \emph{the local Gaussian partial correlation (LGPC)} which, being directly related to the ordinary partial correlation, is easy to interpret, and reduces to the very same ordinary partial correlation in the Gaussian case. Moreover, it distinguishes between positive and negative conditional dependence whereas competing non-linear measures report only the \emph{strength} of the conditional dependence on some non-negative scale.

The local view gives much more flexibility. It allows the conditional dependence to be stronger or weaker in certain regions of a multivariate distribution than in others. This is of particular interest in finance, where description of tail behavior is important. The local aspect also makes it possible to focus tests of conditional independence to selected parts of the distribution, which may potentially increase the power of the test.

The local approach has been shown to be advantageous in other areas of statistics, such as the measurement of nonlinear dependence, density estimation and spectral analysis, see for instance Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}), Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}) and Jordanger and Tjøstheim (\protect\hyperlink{ref-jordanger2017nonlinear}{2020}). The present paper, then, represents the first attempt to model conditional dependence locally. There are several other approaches to modelling conditional dependence and testing for conditional independence in the non-Gaussian case in the literature. For the most part, they revolve around test statistics in conditional independence tests. Such statistics are usually based on measures of distance between some property of the sample and the corresponding property of the population under the null hypothesis of conditional independence. As a consequence, the conditional dependence measures are not always easy to interpret, and can, as mentioned above, not distinguish between negative and positive conditional dependence. We present a comprehensive set of references to this literature in Section \ref{chap:testing}.

Definitions and properties are presented in Section \ref{chap:conditional}, whereas estimation is covered in Section \ref{chap:estimation}. Two examples illustrating the strengths of our approach are given in Section \ref{chap:example}. Section \ref{chap:testing} gives a brief glimpse of the potential in testing.

An integral part of our article is the online supplement. It deals with a number of issues complementing the treatment in the main paper. Among them are more examples, extensions of the main concept to more general situations, precise formulations of some theoretical results with proofs, as well as discussions of Pitman efficiency for testing and validity of the use of the bootstrap.

\hypertarget{chap:conditional}{%
\section{THE LOCAL GAUSSIAN PARTIAL CORRELATION}\label{chap:conditional}}

Let \(\X = (X_1, \ldots, X_p)\) be a random vector. Denote by \((\X_1,\X_2,\X_3)\) a partition of \(\X\) into vectors of dimension \(d_1\), \(d_2\) and \(d_3\) respectively, such that \(\Xone = (\X_1, \X_2) = (X_1, \ldots,X_{d_1 + d_2})\) consists of the first \(d_1+d_2\) components in \(\X\) and \(\Xtwo = \X_3 = (X_{d_1+d_2+1}, \ldots, X_{d_1+d_2+d_3})\) contains the remaining \(d_3\) variables, where \(p = d_1+d_2+d_3\). We assume that the mean vector \(\fmu\) and covariance matrix \(\fSigma\) of \(\X\) exist, and partition them correspondingly, writing
\begin{equation}
\fmu = \begin{pmatrix} \fmu_1 \\ \fmu_2 \\ \fmu_3 \end{pmatrix} \,\,\,\, \textrm{ and } \,\,\,\, \fSigma = \begin{pmatrix} \fSigma_{11} & \fSigma_{12} & \fSigma_{13} \\ \fSigma_{21} & \fSigma_{22} & \fSigma_{23} \\ \fSigma_{31} & \fSigma_{32} & \fSigma_{33}\end{pmatrix},
\label{eq:matrixpartition}
\end{equation}
where \(\fSigma_{ij}\) is the covariance matrices of \((\X_i, \X_j)\), \(i,j = 1,2,3\). There are two main concepts of correlation when \(\Xtwo = \X_3\) is given, the \emph{partial} and the \emph{conditional} correlation, and they coincide in several joint distributions, among them the Gaussian. See e.g.~Baba, Shibata, and Sibuya (\protect\hyperlink{ref-baba2004partial}{2004}) for details. We will use the partial correlation as a starting point when defining the LGPC. The partial variance-covariance matrix of \(\Xone = (\X_1, \X_2)\) given \(\Xtwo = \X_3\) is
\begin{equation}
\fSigma_{12|3} = \fSigma^{11} - \fSigma^{12}\left(\fSigma^{-1}\right)^{22}\fSigma^{21},
\label{eq:matrixpartial}
\end{equation}
where
\[\fSigma^{11} = \begin{pmatrix} \fSigma_{11} & \fSigma_{12} \\ \fSigma_{21} & \fSigma_{22} \end{pmatrix}, \,\, \fSigma^{12} = \begin{pmatrix} \fSigma_{13} \\ \fSigma_{23} \end{pmatrix}, \,\,\fSigma^{21} = \begin{pmatrix} \fSigma_{31} & \fSigma_{32} \end{pmatrix} \, \textrm{ and } \, \fSigma^{22} = \fSigma_{33},\]
and \(\fSigma_{12|3}\) is the covariance matrix in the conditional (Gaussian) distribution of \(\Xone\) given \(\X_3\) if \(\X\) is jointly normal. The partial correlation matrix between \(\X_1\) and \(\X_2\) given \(\X_3\) is naturally defined as

\begin{equation}
\R_{12|3} = \D^{-1/2}\fSigma_{12|3}\D^{-1/2},
\label{eq:partialcorrelation}
\end{equation}
where \(\D = \textrm{diag}(\fSigma_{12|3})\). We identify in the same way the partial correlation matrix \eqref{eq:partialcorrelation} with the correlation matrix in the conditional (Gaussian) distribution of \(\Xone\) given \(\Xtwo\) if \(\X\) is jointly normal. Equations \eqref{eq:matrixpartial} and \eqref{eq:partialcorrelation} will serve as the starting point for our definition of the \emph{local} partial correlation.

\hypertarget{chap:definition}{%
\subsection{Definition}\label{chap:definition}}

In all of the following we assume that the components of \(\X\) are continuous with joint density function \(f_{\X}\). Given a point \(\x\), we approximate \(f_{\X}\) in a neighborhood of \(\x\) by a multivariate Gaussian density

\begin{equation} 
\psi(\x, \fv) = \frac{1}{(2\pi)^{p/2}|\fSigma(\x)|^{1/2}} \exp \left\{-\frac{1}{2}(\fv - \fmu(\x))^T\fSigma^{-1}(\x)(\fv - \fmu(\x))\right\},
\label{eq:localgaussian}
\end{equation}
where \(\x = (x_1\ldots, x_p)\), \(\fmu(\x) = \{\mu_j(\x)\}\) and \(\fSigma(\x) = \{\sigma_{jk}(\x)\}\) for \(j,k=1,\ldots, p\). Moving to another point \(\y\), there is another (generally different) Gaussian approximation \(\psi(\y, \fv)\). In this way we approximate \(f_{\X}\) by a family of multivariate Gaussian densities defined by a set of smooth parameter functions \(\{\fmu(\x), \fSigma(\x)\}\), and if \(f_{\X}\) is itself a Gaussian density, then the parameter functions collapse to constants corresponding to the true parameter values, and \(\psi(\x) \equiv f_{\X}(\x)\). Hjort and Jones (\protect\hyperlink{ref-hjort1996locally}{1996}) provide the general framework for estimating such parameter functions non-parametrically from a given data set using a local likelihood procedure, and the basic idea in the following treatment is to replace the components in the partial covariance matrix \eqref{eq:matrixpartial} by their locally estimated counterparts in order to obtain a local measure of conditional dependence.

Before we get that far, however, we will introduce a useful transformation technique that improves and simplifies the estimation of the LGPC. Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, \protect\hyperlink{ref-otneim2017conditional}{2018}) show that the estimation of the local parameter functions \(\{\fmu(\x), \fSigma(\x)\}\) becomes easier by transforming each \(X_j\) to a standard normal variable \(Z_j = \Phi^{-1}(U_j)\), where \(U_j\) is a uniform variable \(U_j = F_{j}(X_j)\) with \(F_{j}\) being the cumulative distribution function of \(X_j\). Define the random vector \(\Z\) by this transformation of \(\X = (\Xone, \Xtwo) = (\X_1, \X_2, \X_3) = (X_1, X_2, \ldots, X_p)\) to marginal standard normality:

\begin{equation}
\Z = \Big(\Phi^{-1}\left(F_{1}(X_1)\right), \Phi^{-1}\left(F_{2}(X_2)\right), \ldots, \Phi^{-1}\left(F_{p}(X_p)\right)\Big).
\label{eq:trans}
\end{equation}
The transformation enables us to simplify the local Gaussian representation \eqref{eq:localgaussian} by writing the density \(f_{\Z}\) of \(\Z\) as
\begin{equation}
f_{\Z}(\z) = \psi(\z, R(\z)) = \frac{1}{|2\pi\R(\z)|^{1/2}}\exp\left\{-\frac{1}{2}\z^T\R^{-1}(\z)\z\right\},
\label{eq:lgdeapprox}
\end{equation}\\
where, as in Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, \protect\hyperlink{ref-otneim2017conditional}{2018}) in a simplified representation, we have fixed local means and standard deviations \(\mu_j(\z) = 0\) and \(\sigma_j^2(\z) = 1\), \(j = 1,\ldots,p\), and where \(\R(\z) = \{\rho_{jk}(\z)\}\) is the \emph{local correlation matrix}.

In practice, we do not know \(F_{j}\), but we can instead use the empirical distribution function
\[F_{j,n}(x) = \frac{1}{n-1}\sum_{i=1}^n 1\left(X_{ji} \leq x\right),\]
where \(1(\cdot)\) is the indicator function, \(n\) is the number of observations, and \(X_{ji}\) is the \(i\)th observation of \(X_j\). This results in pseudo standard normal variables \(Z_{j,n} = \Phi^{-1}(F_{j,n}(X_j))\).

In the following, we will not always distinguish between \(Z_j\) and \(Z_{j,n}\). In fact, we conclude our asymptotic analysis in Section \ref{chap:asymptotic} by showing that under certain technical conditions, the error made by estimating \(\R(\z)\) using the empirically transformed variables \(Z_{j,n}\), instead of \(Z_j\), is smaller in the limit than the estimation error made when estimating the local correlations themselves.

We will in this paper refer to \(\X\) and its probability density function \(f_{\X}\) as being on the \(x\)-scale, and to \(\Z\) and its probability density function \(f_{\Z}\) as being on the \(z\)-scale. For further discussion of the simplified \(z\)-representation we refer to Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}).

Denote by \((\Zone, \Ztwo) = (\Z_1, \Z_2, \Z_3)\) the partitioning of \(\Z\) corresponding to the partitioning \((\Xone, \Xtwo) = (\X_1, \X_2, \X_3)\) of \(\X\). A natural definition of the \emph{local} partial covariance matrix of \(\Zone|\Ztwo\) is the local version of eq. \eqref{eq:matrixpartial}.

\begin{equation}
\fSigma_{12|3}(\z) = \R^{11}(\zone) - \R^{12}(\z)\left(\R^{22}(\ztwo)\right)^{-1}\R^{21}(\z).
\label{eq:matrixlocalpartial}
\end{equation}

If \(d_1=d_2=1\) then \(\fSigma_{12|3}(\z)\) is a \(2\times 2\) matrix, and we define the local Gaussian partial correlation \(\alpha(\z)\) between the two variables in \(\Zone = (Z_1,Z_2)\) given \(\Ztwo = \Z_3\) in accordance with the ordinary (global) partial correlation provided by eq. \eqref{eq:partialcorrelation}:

\begin{equation}
\alpha(\z) = R_{12|3}(\z) = \frac{\Big\{\fSigma_{12|3}(\z)\Big\}_{12}}{\Big\{\fSigma_{12|3}(\z)\Big\}^{1/2}_{11}\Big\{\fSigma_{12|3}(\z)\Big\}^{1/2}_{22}},
\label{eq:definition}
\end{equation}
which, if \(\Ztwo = Z_3\) is scalar, reduces to
\begin{equation}
\alpha(\z) = \rho_{12|3}(z_1, z_2|z_3) = \frac{\rho_{12}(z_1, z_2) - \rho_{13}(z_1,z_3)\rho_{23}(z_2,z_3)}{\sqrt{1 - \rho^2_{13}(z_1,z_3)}\sqrt{1 - \rho^2_{23}(z_2, z_3)}}.
\label{eq:scalardefinition}
\end{equation}

It is of course possible to introduce an LGPC \(\alpha(\x)\) directly on the \(x\)-scale, but that representation is in many ways harder to handle both computationally and asymptotically. For a multivariate Gaussian distribution, we have that \(\alpha_{\X}(\x) = \alpha_{\Z}(\z) = \alpha\). In the remainder of this paper, we write mainly in terms of the \(z\)-representation using the LGPC \(\alpha(\z) = \alpha_{\Z}(\z)\), but when we write the local partial correlation between \(X_1\) and \(X_2\) given \(\X_3 = \x_3\) at the point \((x_1,x_2,\x_3)\), this is simply \(\alpha(\z)\) inserted with \(z_j = \Phi^{-1}(F_{j}(x_j))\), \(j=1,\ldots,p\).

In the more general case where \(d_1 > 1\) and/or \(d_2 > 1\) then \(\fSigma_{12|3}(\z)\) is a \((d_1+d_2)\times (d_1+d_2)\) matrix that describes the non-linear conditional dependence among the variables in \(\Zone = (\Z_1, \Z_2)\) given \(\Ztwo = \Z_3\) (or alternatively \(\Xone = (\X_1, \X_2)\) given \(\Xtwo = \X_3\)), which in particular can be used to analyze the conditional dependence between two \emph{sets} of random variables having dimensions \(d_1\) and \(d_2\) respectively, given a third set of variables having dimension \(d_3\). This case poses no conceptual challenges to our approach, but it requires a rather large investment in new notation, and does not lead to simple formulas like eq. \eqref{eq:scalardefinition}. We will therefore, for the most part, focus on the local Gaussian partial correlation between two stochastic scalar variables in the remainder of this paper. We provide the complete description of the general multivariate case in Sections A.1 and A.2 of the online supplement.

\hypertarget{properties}{%
\subsection{Properties}\label{properties}}

The local Gaussian partial correlation is closely related to the partial correlation between jointly normally distributed variables. We see this also from the following properties that we will establish next:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The LGPC \(\alpha(\z)\) satisfies \(-1 \leq \alpha(\z) \leq 1\).
\item
  If \(\X\) is jointly normally distributed, then the LGPC coincides with the ordinary (global) partial correlation.
\item
  For stochastic vectors having joint density function on \(z\)-scale of the form \eqref{eq:lgdeapprox}, the LGPC \(\alpha(\z)\) is identically equal to zero if and only if \(X_1\) and \(X_2\) are conditionally independent given \(\X_3\). Note that \(\alpha(\z) \equiv 0\) if and only if \(\alpha(\x) \equiv 0\). \label{prop:char}
\item
  The LGPC \(\alpha(\z)\) is invariant with regard to a set of monotone transformations \(\Y = \bm{h}(\X) = (h_1(X_1), \ldots, h_p(X_p))\).
\end{enumerate}

Property 1 is trivially true if \(\R(\z)\) is a valid correlation matrix. By removing the \(\z\)-dependence in the local correlations, it follows immediately from eq. \eqref{eq:lgdeapprox} and from the results referred to in Section \ref{chap:definition} that property 2 holds.

To see that conditional independence between \(X_1\) and \(X_2\) given \(\X_3\) is equivalent to \(\alpha(\z) \equiv 0\), or equivalently \(\alpha(\x) \equiv 0\), we need to follow a few simple steps that would also work in the global Gaussian case. Working on the standard normal \(z\)-scale and assuming that \((z_1, z_2, \z_3)\) has density function \(f_{\Z}(\z)\) as given by \eqref{eq:lgdeapprox}, it follows from Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}) that we can calculate conditional distributions in the same way as in the global Gaussian case. For example, the conditional density of \(Z_1|\Z_3\) at the point \(z_1\) is given by

\[f_{z_1|\z_3}(z_1) = \frac{1}{|2\pi\sigma_{1|3}(z_1)|^{1/2}}\times\exp \left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1)}{\sigma_{1|3}(z_1)}\right)^2\right\},\]
where

\begin{equation}
\mu_{1|3}(z_1) = \R_{12}(z_1, \z_3)\R_{22}(\z_3)^{-1}\z_3 \,\,\, \textrm{ and } \,\,\, \sigma_{1|3}(z_1) = \fSigma_{Z_1|\Z_3},
\label{eq:localconditionalpar}
\end{equation}
and where the latter expression is defined analogously to \eqref{eq:matrixlocalpartial}. The conditional density \(f_{Z_2|\Z_3}(z_2|\z_3)\) is defined in the same way. If \(Z_1\) and \(Z_2\) are conditionally independent given \(\Z_3\), then

\begin{align*}
f_{Z_1,Z_2|\Z_3}(z_1, z_2|\z_3) &= f_{Z_1|\Z_3}(z_1|\z_3)f_{Z_2|\Z_3}(z_2|\z_3) \\
&= \frac{1}{2\pi\sqrt{1-\sigma_{1|3}(z_1)}\sqrt{1-\sigma_{2|3}(z_2)}} \\ 
&\qquad \qquad \times\exp\left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1)}{\sigma_{1|3}(z_1)}\right)^2\right\}\exp\left\{-\frac{1}{2}\left(\frac{z_2 - \mu_{2|3}(z_2)}{\sigma_{2|3}(z_2)}\right)^2\right\},
\end{align*}
and we at once identify the conditional density of \((Z_1,Z_2)|\Z_3\) as another Gaussian distribution, but without any cross-term involving \(z_1z_2\), which then implies that the off-diagonal element in the partial covariance matrix \eqref{eq:matrixlocalpartial} is identically equal to zero. We see immediately from our definition \eqref{eq:definition} that the LGPC is also identically equal to zero.

The converse statement follows straightforwardly by considering the conditional density of \((Z_1, Z_2|\Z_3)\).

Let the transformation \(\Y = \bm{h}(\X)\) define the \(y\)-scale in the same way as the transformation \eqref{eq:trans} defines the \(z\)-scale. We define the LGPC in terms of the marginally standard normal \(\Z\)-variables, which for the \(\Y\)-variables can be calculated as

\[\Z = \Big(\Phi^{-1}\left(F_{Y_1}(Y_1)\right), \ldots, \Phi^{-1}\left(F_{Y_p}(Y_p)\right)\Big),\]
but if \(h_j(\cdot)\) is monotone for all \(j = 1,\ldots,p\),

\[F_{Y_j}(y_j) = \mathbb{P}(h_j(X_j) \leq y_j) = \mathbb{P}(X_j \leq h_j^{-1}(y_j)) = \mathbb{P}(X_j \leq x_j) = F_{X_j}(x_j),\]
where \(x_j\) is the point on the \(x\)-scale corresponding to the point \(y_j\) on the \(y\)-scale. The \(\Z\)-variables are the same for the stochastic variables \(\X\) and \(\Y = \bm{h}(\X)\). Their LGPC-function \(\alpha(\z)\) must therefore be the same as well according to our definition in the preceding section. Entirely analogous arguments for the properties 1.-4. in the general multivariate case can be constructed from the supplementary material, Sections A.1 and A.2.

\hypertarget{chap:estimation}{%
\section{ESTIMATING CONDITIONAL DEPENDENCE USING THE LGPC}\label{chap:estimation}}

\hypertarget{ll}{%
\subsection{Estimation by local likelihood}\label{ll}}

From the defining equations \eqref{eq:matrixlocalpartial} and \eqref{eq:definition} it is seen that the basic building blocks for the \emph{local Gaussian partial correlation} are the \emph{local Gaussian correlation functions}, that populate the local correlation matrix \(\R(\z)\) in the density function \(f_{\Z}\) in \eqref{eq:lgdeapprox}. Consider the random vector \(\X = (X_1, X_2, \ldots, X_p)\) having joint probability density function (pdf) \(f(\x)\), and its transformed counterpart \(\Z = \left(\Phi^{-1}(F_1(X_1)), \ldots, \Phi^{-1}(F_p(X_p))\right)\), on the marginally standard normal \(z\)-scale. The relation between the pdf of \(\X\) and the pdf of \(\Z\) is given by

\begin{equation}
f_{\X}(\x) = f_{\Z}\left(\Phi^{-1}\left(F_1(x_1)\right), \ldots, \Phi^{-1}\left(F_p(x_p)\right)\right)\prod_{j=1}^p 
\frac{f_j(x_j)}{\phi\left(\Phi^{-1}\left(F_j(x_j)\right)\right)},
\label{eq:fxfz}
\end{equation}
where \(f_j(\cdot)\), \(j = 1,\ldots, p\) are the marginal density functions of \(\X\), and \(\phi(\cdot)\) is the standard normal density function. Let \(\X_1, \ldots, \X_n\) be a random sample identically distributed as \(\X\) and construct the pseudo standard normal observations \(\hZ_1, \ldots, \hZ_n\) as

\begin{equation}
\hZ_t = \left(\Phi^{-1}\left(\hF_1(X_{1,t})\right), \ldots, \Phi^{-1}\left(\hF_p(X_{p,t})\right)\right),
\label{eq:pseudoobservations}
\end{equation}
where \(\X_t = (X_{1,t}, X_{2,t}, \ldots, X_{p,t})\), and \(\hF_j(\cdot)\) is an estimate of the marginal distribution function of \(X_j\), for example the empirical distribution function. Next, we must produce an estimate \(\hR(\z)\) of the local correlation matrix, and we will consider two variations of this task in subsections \ref{chap:trivariate-full} and \ref{chap:bivariate} with a more general treatment in Section A.3 of the supplement. It is then, following \eqref{eq:matrixlocalpartial}, natural to estimate the local Gaussian partial covariance matrix as

\begin{equation}
\hfSigma_{12|3}(\z) = \hR^{11}(\z) -  \hR^{12}(\z)\left(\hR^{22}(\z)\right)^{-1}\hR^{21}(\z),
\label{eq:localpartialcov}
\end{equation}
and if we assume that \(X_1\) and \(X_2\) constitute the two first elements in \(\X\), then, following \eqref{eq:definition}, the local Gaussian partial correlation between \(X_1\) and \(X_2\) given \(\Xtwo = \X_3\) on the \(z\)-scale is estimated by
\begin{equation}
\widehat\alpha(\z) = \frac{\Big\{\hfSigma_{12|3}(\z)\Big\}_{12}}{\Big\{\hfSigma_{12|3}(\z)\Big\}_{11}^{1/2}\Big\{\hfSigma_{12|3}(\z)\Big\}_{22}^{1/2}}.
\label{eq:estimate}
\end{equation}

A corresponding value of \(\halpha(\x)\) at the point \(\x = \widehat F^{-1}(\Phi(z))\) is obtained by inserting \(\z = \Phi^{-1}(\widehat F(\x))\) in \eqref{eq:estimate}.

\hypertarget{chap:trivariate-full}{%
\subsubsection{\texorpdfstring{Estimation of \(\R(\z)\) when \(p = 3\) and \(\Xtwo\) is a scalar}{Estimation of \textbackslash R(\textbackslash z) when p = 3 and \textbackslash Xtwo is a scalar}}\label{chap:trivariate-full}}

If dim\((\X) = 3\), then dim\((\Xtwo)=1\), and \(\R(\z)\) is a \(3\times3\) symmetric matrix of local correlation functions having arguments \(\z = (z_1,z_2,z_3)\):
\begin{equation}
\R(\z) = \begin{pmatrix} 1 & \rho_{12}(\z) & \rho_{13}(\z) \\ \rho_{12}(\z) & 1 & \rho_{23}(\z) \\ \rho_{13}(\z) & \rho_{23}(\z) & 1\end{pmatrix},
\label{eq:trivariatecorr}
\end{equation}
containing three parameter functions that must be estimated from data. Note that here we deviate from the pairwise scheme found in Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, \protect\hyperlink{ref-otneim2017conditional}{2018}) as well as most of this article where \(\rho_{ij}(\z) = \rho_{ij}(z_i, z_j)\). In equation \eqref{eq:trivariatecorr}, \(\z = (z_1,z_2,z_3)\). Based on the sample \eqref{eq:pseudoobservations}, \(\R(\z)\) is estimated by fitting the parametric family \(\psi(z, \R)\), defined in \eqref{eq:lgdeapprox}, locally to the density \(f_{\Z}(\z)\) by maximizing the \emph{local likelihood function} (cf.~Hjort and Jones (\protect\hyperlink{ref-hjort1996locally}{1996})),

\begin{equation}
\hR(\z) = \arg\max_{\R} n^{-1} \sum_{t=1}^nK_{\hh}(\hZ_t - \z)\log\psi(\hZ_t, \R) - \int K_{\hh}(\y - \z)\psi(\y, \R)\,\textrm{d}\y
\label{eq:locallik3}
\end{equation}
in each point \(\z\), where \(K_{\hh}(\x) = |\hh|^{-1/2}K(\hh^{-1/2}\x)\), \(K\) is a non-negative and radially symmetric kernel function that satisfies \(\int K(\x)\,\textrm{d}\x=1\), and \(\hh\) is a diagonal \(3\times3\) matrix of bandwidths that serve as smoothing parameters. The estimate \(\hf_{\Z}(\z) = \psi(\z, \hR(\z))\) aims at minimizing a locally weighted Kullback-Leibler distance to \(f_{\Z}(\z)\), and we refer to Hjort and Jones (\protect\hyperlink{ref-hjort1996locally}{1996}) and Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) for much more details about this construction, and to Section \ref{chap:asymptotic} for the relevant asymptotic estimation theory.

Finally, we obtain the estimated LGPC \(\halpha(\z)\) by plugging the estimated local correlations \(\hR(\z)\) into equations \eqref{eq:localpartialcov} and \eqref{eq:estimate} and a corresponding value of \(\halpha(\x)\) at the point \(\x = \widehat F^{-1}(\Phi(\z))\) by inserting \(\z = \Phi^{-1}({\widehat F(\x)})\).

\hypertarget{chap:bivariate}{%
\subsubsection{\texorpdfstring{Estimation of \(\R(\z)\) when \(\Xtwo\) is a vector}{Estimation of \textbackslash R(\textbackslash z) when \textbackslash Xtwo is a vector}}\label{chap:bivariate}}

The complexity of the estimation problem \eqref{eq:locallik3} increases sharply with the number of variables involved. The basic idea of Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}) to avoid the curse of dimensionality is to model local correlations in \(\R(\z)\) as functions of their corresponding \emph{pair} of variables only:

\begin{equation}
\R(\z) = \left\{\rho_{jk}(z_j, z_k)\right\}, \qquad j,k = 1, 2, \ldots, p,
\label{eq:loccormat}
\end{equation}
which reduces the estimation of the \(p\)-variate correlation functions \(\rho_{jk}(\z)\) to a series of bivariate estimation problems. We suggest to use this approach when modelling local dependence if the number of variables in \(\X\) is larger than three. This allows nonlinear dependence between variables to be approximated by the pairwise structure \eqref{eq:loccormat}, while still being computationally tractable. An analogue to the pairwise approximation is the additive approximation in nonparametric regression.

We transform the observations to the \(z\)-scale as in \eqref{eq:pseudoobservations}, but now estimate the individual components in \(\R(\z)\) one by one, taking only into consideration the \emph{pair} of variables in question:
\begin{equation}
\hrho_{jk}(z_j, z_k) = \arg\max_{\rho_{jk}} n^{-1} \sum_{t=1}^nK_{\hh}(\hZ_t - \z)\log(\hZ_t, \rho_{jk}) - \int K_{\hh}(\y - \z)\psi(\y, \rho_{jk})\,\textrm{d}\y, 
\label{eq:locallik2}
\end{equation}
where all running variables and samples are bivariate subsets corresponding to the indices \((j,k)\), \(\psi(\cdot, \rho)\) is the bivariate version of \eqref{eq:lgdeapprox}, and \(\hh\) now is a \(2\times2\) diagonal matrix of bandwidths. After estimating all local correlations in this way, we proceed to calculate the LGPC using equations \eqref{eq:localpartialcov} and \eqref{eq:estimate} as above.

\hypertarget{chap:asymptotic}{%
\subsection{Asymptotic theory}\label{chap:asymptotic}}

Equations \eqref{eq:matrixlocalpartial}, \eqref{eq:definition} and their empirical counterparts \eqref{eq:localpartialcov} and \eqref{eq:estimate} demonstrate clearly that the LGPC is nothing more than a deterministic function of the local correlation matrix, in the same way as the ordinary partial correlation is a function of the ordinary correlation matrix. The asymptotic behavior of \(\halpha(\z)\) can thus be derived directly from the asymptotic behavior of \(\hR(\z)\), which, for the most part, has been established in earlier works, see e.g.~Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) and Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}). Some key results follow below, and we refer to the supplementary material, Sections B--D for proofs of the asymptotic distributional results in this subsection, and to Section A.3 for a discussion of more general situations and for a result on consistency.

We start by assuming that the marginal distribution functions \(F_1, \ldots, F_p\) are known, and then state two results on the joint asymptotic normality of the estimated local correlation matrix \(\hR(\z)\) under a mixing condition. We will then apply the delta method in order to derive the limiting distribution of the estimated LGPC, and finally show that we may replace \(F_1,\ldots,F_p\) with their empirical counterparts \(\hF_1,\ldots,\hF_p\) without changing these asymptotic results.

Consider first the full trivariate fit \(\hR(\z)\) described in Section \ref{chap:trivariate-full} with \(\z = (z_1,z_2,z_3)\). We estimate the local correlation matrix \(\R(\z)\) by maximizing the local likelihood function in \eqref{eq:locallik3}, which we denote by \(L_n(\R(\z), \z)\), with the only exception that we, for now, assume the true transformation between the \(x\)- and the \(z\)-scale to be known. For a fixed matrix of bandwidths \(\hh\), denote by \(\R_{\hh}(z)\) the local correlation that satisfies, as \(n \rightarrow \infty\),

\begin{equation}
\nabla L_n(\R_{\hh}, \z) \rightarrow \int K_{\hh}(\y - \z)\fu(\y, \R_{\hh}(\z))\{f_{\Z}(\y) - \psi(\y, \R_{\hh})\}\,\textrm{d}\y = 0,
\label{eq:popfixedh3}
\end{equation}
where \(f_{\Z}(\z) = f_{Z_1,Z_2,Z_3}(z_1,z_2,z_3)\) is the density function of \(\Z\), and \(\fu(\cdot, \rho_{\hh})\) is the column vector of local score functions \(\nabla \log\psi(\cdot, \rho_{\hh})\), and where the gradient is taken with respect to the parameters \(\rho_{12}(\z)\), \(\rho_{13}(\z)\) and \(\rho_{23}(\z)\). Denote by \(\frho = \frho(\z) = \{\rho_{12}(\z), \rho_{13}(\z), \rho_{23}(\z)\}\) the vector of local correlations. The joint limiting distribution of \(\hfrho\) (for convenience only stated on the \(z\)-scale below) is given by the following result as \(n \rightarrow \infty\) and \(\hh \rightarrow \bm{0}\):

\begin{theorem}
\protect\hypertarget{thm:loccor3}{}{\label{thm:loccor3} }Let \(\{\Z_t\}_{t=1}^n\) be observations on the marginally standard normally distributed random vector \(\Z\) having joint density \(f_{\Z}(\z)\). Assume that the following conditions hold:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  For any sequence of bandwidth matrices \(\hh_n\) tending to zero element-wise, there exists for the trivariate marginally standard Gaussian vector \(\Z\) a unique set of local correlations \(\frho_{\hh}(\z)\) that satisfies \eqref{eq:popfixedh3}, and there exists a \(\frho_0(\z)\) such that \(\frho_{\hh}(\z) \rightarrow \frho_0(\z)\).
\item
  The process generating \(\{\Z_t\}_{t=1}^n\) is \(\alpha\)-mixing with mixing coefficients \(\alpha(m)\) satisfying \(\sum_{m\geq1}m^{\lambda}\alpha(m)^{1-2/\delta} < \infty\) for some \(\lambda > 1-2/\delta\) and \(\delta>2\).
\item
  \(n\rightarrow\infty\), and each bandwidth \(b = b_n\) tends to zero such that \(nb^{\frac{\lambda+2-2/\delta}{\lambda + 2/\delta}} = O(n^{\epsilon_0})\) for some constant \(\epsilon_0 > 0\).
\item
  In a given point \(\z\), the parameter space \(\Theta\) for each local correlation \(\rho(\z)\) is a compact subset of \((-1,1)\).
\item
  The non-negative kernel function \(K(\cdot)\) sasisfies \(\sup_{\z}K(\z) < \infty\), \(\int K(\y)\,\textrm{d}\y < \infty\), \(\partial/\partial z_iK(\z) <\infty\), and \(\lim_{z_i \rightarrow \infty}|z_iK(\z)| = 0\) for \(i = 1,2,3\).
\end{enumerate}

Then, by writing \(b_{n1} = b_{n2} = b_{n3} = b \rightarrow 0\) assuming they all converge to zero at the same rate,
\begin{equation}
\sqrt{nb^5}\Jb\Mb^{-1/2}\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
\label{eq:an3}
\end{equation}
where \(\bm{I}\) is the \(3\times3\) identity matrix,
\begin{align*}
\Jb &= \int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\fu^T(\y, \frho_{\hh}(\z))\psi(\y,\frho_{\hh}(\z))\,\textrm{d}\y \\
& \qquad\qquad\qquad\qquad - \int K_{\hh}(\y - \z)\nabla\fu(\y,\frho_{\hh}(\z))\Big\{f_{\Z}(y) - \psi(\y,\frho_{\hh}(\z))\Big\}\textrm{d}\y,
\end{align*}
and
\begin{align*}
\Mb &= b^3\int K^2_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\fu^T(\y,\frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y \\
& \qquad\qquad - b^3\int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y \int K_{\hh}(\y - \z)\fu^T(\y, \frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y.
\end{align*}
\end{theorem}

Theorem \ref{thm:loccor3} and its assumptions above are variations of the assumptions and the results presented by Hjort and Jones (\protect\hyperlink{ref-hjort1996locally}{1996}), and in particular Theorem 3 by Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}). The analysis of the asymptotic behavior of \(\Jb\) and \(\Mb\) in the latter reference reveals that the local parameter estimates converge more slowly than, for example, the trivariate kernel density estimator (that converges as \(1/(nb^3)^{1/2}\)), and it is a straightforward exercise to modify their proof to suit our particular situation, which demonstrates that the three local correlations converge jointly in distribution at the rate \(1/(nb^5)^{1/2}\), see Section B in the supplement for details.

Furthermore, Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) show that an analytic expression of the leading term in the covariance matrix \(\Jb^{-1}\Mb(\Jb^{-1})^T\) is not practically available. In applications we may circumvent this problem by using the bootstrap. The validity of the bootstrap can be proved using a direct adaptations of the techniques of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}). It is spelled out in some detail for the pairwise case in Theorems E.1, E.2 and corollaries E.3 and E.4 in Section E of the online supplement. The same arguments can be used in the present 3-parameter case (in fact, Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) treat the more general five-parameter case).

The pairwise estimation structure that we describe in Section \ref{chap:bivariate} simplifies the asymptotic analysis of \(\hR(\z)\) considerably. The result is essentially available in the literature as Theorem 2 in Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}), who state under similar assumptions as in Theorem \ref{thm:loccor3} that
\begin{equation}
\sqrt{nb_n^2}\left(\hfrho(\z) - \frho_0(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \fOmega),
\label{eq:ot2018}
\end{equation}
where the asymptotic covariance matrix \(\fOmega\) is defined in an analogous way as in Theorem \ref{thm:loccor3} There are two important differences between these two results, however. First of all, we have that the convergence rate in the simplified pairwise case is equal to the usual nonparametric rate of \(1/(nb^2)^{1/2}\). This is because we estimate the local correlations one by one based on pairs of variables. The second difference is that the leading term of the asymptotic covariance matrix in Theorem 2 by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}) has a simple analytic expression that can be used to construct, for instance, confidence intervals. We refer to Section A.3 in the online supplement for further details regarding the limiting distribution of the LGPC as well conditions for consistency of the estimates.

We can now calculate the limiting distribution of the LGPC by means of the delta method. Denote by \(g: \mathbb{R}^{p(p-1)/2} \rightarrow \mathbb{R}\) the translation \eqref{eq:definition} from the vector of local correlations between the components in \(\Z\) (or, equivalently, \(\X\)) to the local partial correlation \(\alpha(\z)\) between \(Z_1\) and \(Z_2\) given \(Z_3, \ldots, Z_p\). From the expression of our estimate \eqref{eq:estimate} we have that \(\halpha(\z) = g(\hfrho(\z))\), so we use the delta method to see that

\begin{equation}
\sqrt{nb_n^m}\left(\halpha(\z) - \alpha(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}\left(0, \nabla g(\frho)^T\fLambda\nabla g(\frho)\right),
\label{eq:limitlgpc}
\end{equation}
where \(m\) is equal to 2 or 5, and \(\fLambda\) is either equal to the leading term of \(\Jb^{-1}\Mb(\Jb^{-1})^T\), or \(\fOmega\) in \eqref{eq:ot2018}, depending on whether we use the full trivariate locally Gaussian fit described in Section \ref{chap:trivariate-full} or the sequentially pairwise simplification described in Section \ref{chap:bivariate}. In any case, it is only a matter of basic differentiation to work out an expression for \(\nabla g(\cdot)\), and this task has been deferred to Section C of the online supplement in the case where \(d_1 = d_2 = 1\).

Finally, we present the limiting distribution of the LGPC between \(Z_1\) and \(Z_2\) given \(Z_3, \ldots, Z_p\) when the components in \(\Z = (Z_1, \ldots, Z_p)\) are replaced by marginally standard normal \emph{pseudo-observations} as defined in equation \eqref{eq:pseudoobservations}. The following result ensures that Theorems \ref{thm:loccor3} and Theorem 3 by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}), and thus eq. \eqref{eq:limitlgpc}, still holds.

\vspace{.5cm}

\begin{theorem}
\protect\hypertarget{thm:pseudo}{}{\label{thm:pseudo} }Under the conditions in Theorems \ref{thm:loccor3} and Theorem 2 by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}), assume further that the kernel function \(K(\cdot)\) has bounded support. Replacing the marginally standard normal vector \(\Z\) with the approximately marginally standard normal vector \(\hZ\), does not change the conclusions in these theorems.
\end{theorem}

\begin{proof}
\iffalse{} {Proof. } \fi{}See Section D of the online supplement.
\end{proof}

\hypertarget{chap:example}{%
\section{EXAMPLES}\label{chap:example}}

We will in this section provide some demonstrations on how the estimated LGPC can be used to reveal nonlinear departures from conditional independence and to measure that conditional dependence. All plots are presented on the \(x\)-scale.

First, we provide some reference pictures from the simplest situation imaginable. Let \(X_1\), \(X_2\) and \(X_3\) be independent standard (and thus also jointly) normal variables. Since \(X_1\) is conditionally independent from \(X_2\) given \(X_3\) (written \(X_1 \perp X_2 | X_3\)) in this particular case, we also know from property 3 in Section 2.2 that the LGPC between \(X_1\) and \(X_2\) is zero everywhere.

\begin{figure}
    \subfloat[$n = $ 100\label{fig:gaussian-example-plots1}]{\includegraphics[width=0.31\linewidth]{figures/gaussian-example-plots-1} }
    \subfloat[$n = $ 500\label{fig:gaussian-example-plots2}]{\includegraphics[width=0.31\linewidth]{figures/gaussian-example-plots-2} }
    \subfloat[$n = $ 2000\label{fig:gaussian-example-plots3}]{\includegraphics[width=0.31\linewidth]{figures/gaussian-example-plots-3} }
    \caption{Estimated local Gaussian partial correlation between $X_1$ and $X_2$ given $X_3 = 0$, where $(X_1, X_2, X_3)$ is a jointly standard normally distributed vector. The fully trivariate model of Section 3.1.1 is used.}\label{fig:gaussian-example-plots}
\end{figure}

In Figure \ref{fig:gaussian-example-plots} we see the estimated LGPC between \(X_1\) and \(X_2\) given \(X_3=0\) for three samples, having sample size 100, 500 and 2000 respectively, mapped out on a grid. In this and all other examples we select bandwidths based on a simple plug-in formula that follows naturally from classical asymptotic arguments, \(b = cn^{-1/9}\) for the full trivariate fit, and \(b = cn^{-1/6}\) for the bivariate simplification, where the constant \(c\) controls the amount of smoothing and must be chosen appropriately based on the task at hand. Otneim (\protect\hyperlink{ref-otneim2016multivariate}{2016}) argues that \(c=1.75\) is a good choice for multivariate density estimation, and we will in the next section see that we obtain good power in our test for conditional independence if \(c\) is somewhat smaller than that. For the visual display of conditional dependence maps, however, we tend to prefer a fair amount of smoothing, and in Figure \ref{fig:gaussian-example-plots} we have used \(c=4\). Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}) also suggest a algorithm for bandwidth selection in the density estimation context, maximizing the leave-one-out cross validated log-likelihood of the sample;
\[\bb_{CV} = \textrm{arg max}_{\bb} \sum_{t = 1}^n \widehat{f}_{\bb}^{(-t)}(\Z_t),\]
where \(\widehat f_{\bb}^{(-t)}(\Z_t)\) is the estimated local Gaussian density estimate calculated without, and evaluated in, the sample point \(\Z_t\), using the bandwidth matrix \(\bb\). A theory of efficient and accurate bandwidth selection for local likelihood estimation remains a topic for future research.

As expected, \(\widehat\alpha(\x) = \widehat\alpha(\z)\) is close to zero in all three plots in Figure \ref{fig:gaussian-example-plots} and a significance test carried out as in Section \ref{chap:testing} does not reject conditional independence at any reasonable significance level.

\begin{figure}[p]
    \subfloat[Observed pairs $(X_1, X_2)$ from the conditionally \newline Gaussian model.\label{fig:simex2-plots1}]{\includegraphics[width=0.49\linewidth]{figures/simex2-plots-1} }
    \subfloat[The estimated LGPC between $X_1$ and $X_2$ given $\rho = -0.9$.\label{fig:simex2-plots2}]{\includegraphics[width=0.49\linewidth]{figures/simex2-plots-2} }\newline
    \subfloat[The estimated LGPC between $X_1$ and $X_2$ \newline  given $\rho = 0$\label{fig:simex2-plots3}]{\includegraphics[width=0.49\linewidth]{figures/simex2-plots-3} }
    \subfloat[The estimated LGPC between $X_1$ and $X_2$ \newline given $\rho = 0.9$\label{fig:simex2-plots4}]{\includegraphics[width=0.49\linewidth]{figures/simex2-plots-4} }
    \caption{The conditionally Gaussian model}\label{fig:simex2-plots}
\end{figure}

In the next simulated example we will demonstrate an important feature of the LGPC: It is able to \emph{distinguish between positive and negative conditional relationships}, which, to our knowledge, has until now not been possible beyond the linear and jointly Gaussian setting using the ordinary partial correlation. Our approach also allows exploration of conditionally different dependence patterns across different levels of the conditioning variable.

Generate \(X_3 = \rho \sim U(-1,1)\), and then generate \((X_1, X_2)\) from the bivariate Gaussian distribution having standard normal marginals, and correlation coefficient equal to \(\rho\). We observe \(\X = (X_1, X_2, \rho)\) and seek to visualize the dependence between \(X_1\) and \(X_2\) conditional on \(\rho\).

We see the results in Figure \ref{fig:simex2-plots}. In panel (a), we generate \(n=1000\) simulated pairs \((X_1, X_2)\) from this model. In panel (b)-(d) we see the estimated LGPC plotted over suitable grids, and where the conditioning variable \(X_3 = \rho\) has been fixed at the respective values \(-0.9\), \(0\) and \(0.9\), and we see clearly how the dependence between \(X_1\) and \(X_2\) dramatically changes. It is straightforward to see that \(X_1\) and \(X_2\) are uncorrelated and that the ordinary partial correlation between \(X_1\) and \(X_2\) given \(X_3\) is equal to zero. Furthermore, we note that the pairwise simplification defined in Section 3.1.2 would also not be able to measure the \emph{conditional} relationship between \(X_1\) and \(X_2\) given \(X_3\) in this case, as we see clearly from eq. \eqref{eq:scalardefinition}, because \(X_1\) and \(X_2\) are both marginally independent from \(X_3\). This means that the two pairwise correlations \(\rho_{13}(z_1,z_3)\) and \(\rho_{23}(z_2,z_3)\) are equal to zero.

We conclude this section by using the LGPC to analyze real data in an example regarding Granger causality between financial time series. Granger causality has been a central concept in economics and econometrics ever since its inception by Granger (\protect\hyperlink{ref-granger1969investigating}{1969}). In layman's terms, the time series \(\{X_t\}\) Granger causes \(\{Y_t\}\) if past values of \(\{X_t\}\) are helpful when predicting future values of \(\{Y_t\}\). Formally, \(\{X_t\}\) Granger causes \(\{Y_t\}\) if (Granger \protect\hyperlink{ref-granger1980testing}{1980})

\begin{equation}
Y_t \not\perp \mathcal{I}^*(t-1) \,\,|\,\, \mathcal{I}_{-X}^*(t-1),
\label{eq:granger}
\end{equation}
where \(\perp\) denotes independence, \(\not\perp\) denotes dependence, and where \(\mathcal{I}^*(t-1)\) is all information available at time \(t-1\) and \(\mathcal{I}_{-X}^*(t-1)\) is the same information, with the exception of the values of \(\{X_t\}\) up to, but not including, time \(t\). Of course, the hypothesis \eqref{eq:granger} can not be tested in practice in its full generality. By taking only effects up to the first lag into account, we may formulate a sufficient (but not necessary) condition for \eqref{eq:granger}: \(Y_t \not\perp X_{t-1} \,\,|\,\, Y_{t-1}\), the converse of which constitutes a testable null hypothesis of Granger non-causality:

\begin{equation}
\textrm{H}_0: Y_t \perp X_{t-1} \,\,|\,\, Y_{t-1}.
\label{eq:grangernull}
\end{equation}
There are many ways to carry out this test. The simplest, and original one, is based on the further restriction of linear relationships between \(\{X_t\}\) and \(\{Y_t\}\) and is thus a test for \emph{conditional uncorrelatedness} rather than conditional independence. Nonparametric tests that have power against many nonlinear types of conditional dependence have also been developed, and we refer to Section \ref{chap:testing} for further references to this literature and a new test for conditional independence based on the LGPC.

Francis, Mougoue, and Panchenko (\protect\hyperlink{ref-fran:moug:vale:2010}{2010}) investigate how stock returns of large firms may Granger cause the stock returns of small firms and vice versa from a nonlinear standpoint. They point to earlier evidence that such nonlinear lead-lag relationships have been found and investigate this further using the test for Granger non-causality by Baek and Brock (\protect\hyperlink{ref-baek:broc:1992}{1992}), and using a long series of US data they find evidence that there is a bi-directional causal link between the stock prices of the firms in the most valuable quintile (20 percent) and the firms in the least valuable quintile. We will repeat this exercise for stocks traded on the Oslo Stock Exchange in Norway, but this time from a point of view using the estimated LGPC.

Denote by \(R_{1,t}\) the log-returns on a value-weighted portfolio consisting of the most valuable quintile of firms listed on the Oslo Stock Exchange as of 27 April 2020 (the big firms), and denote by \(R_{2,t}\) the corresponding log-returns for the least valuable companies (the small firms). We will investigate the possibility of Granger causality from the small to the big firms, the hypothesis being that smaller firms are more market sensitive and quickly react to changes in the economy, leading the larger firms that more slowly adapt to small variations. The null hypothesis of non-causality of the first order \eqref{eq:grangernull} in this particular case means that

\begin{equation}
R_{1,t} \perp R_{2, t-1} \,\,|\,\, R_{1,t-1},
\end{equation}
which we may test formally using the approaches described in the following section or the references therein. Let us rather concentrate on \emph{describing} this conditional dependence relationship by estimating the local Gaussian partial correlation between \(R_{1,t}\) and \(R_{2,t-1}\) given a few different values of \(R_{2,t-1}\), which gives considerable more information than a single \(p\)-value of a test. We collect daily observations on Oslo Stock Exchange\footnote{Thanks to \emph{Børsprosjektet (The Stock Exchange Project)} at the Norwegian School of Economics for providing the data.} from 1 January 2018 until 31 December 2019. In Figure \ref{fig:order1} we see the estimated LGPC along the diagonal \(R_{1,t} = R_{2,t-1}\) for three different values \(R_{1,t-1} = -0.02, 0\) and \(0.02\), along with their block-bootstrapped 95\% confidence bands. The constant \(\alpha(\z) = 0\) is indicated as a dotted line.

\begin{figure}[t]
\centering
\includegraphics{figures/granger-order1}
\caption{The estimated LGPC between $R_{1,t}$ and $R_{2,t-1}$ given $R_{1,t-1}$ along the diagonal of the sample space, for three values of the conditioning variable}
\label{fig:order1}
\end{figure}

One could perhaps make the argument that \(R_{1,t}\) and \(R_{2,t-1}\) seem to be somewhat positively dependent in the right half of the graph for \emph{small} values of the conditioning variable \(R_{1,t-1}\). Roughly translated to this particular situation this means that if the large-firm portfolio gives a negative return on day \(t-1\), then its value on day \(t\) depend on the performance small-firm portfolio on day \(t-1\) in a nonlinear and non-Gaussian way that is characterized by positive dependence in the first quadrant. The evidence displayed in Figure \ref{fig:order1} is quite weak, however, because the confidence bands include zero in most of the sample space.

Francis, Mougoue, and Panchenko (\protect\hyperlink{ref-fran:moug:vale:2010}{2010}) go on to analyze if there is Granger causality from small to big firms in the US data if another lag is taken into account, which in general means finding evidence against the hypothesis

\begin{equation}
R_{1,t} \perp (R_{2,t-1}, R_{2, t-2}) \,|\, (R_{1, t-1}, R_{1, t-2}),
\label{eq:order2}
\end{equation}
departures from which can be quantified and tested for in a linear fashion using multiple regressions, or tested for directly by means of a nonlinear test as we discuss in the following Section. Let us here, on the other hand, use the LGPC to quantify the conditional dependence relationship \eqref{eq:order2} nonlinearly for the Norwegian returns data in order to obtain more detailed information. This means estimating a \(3\times3\) partial correlation matrix \(\hfalpha(\z)\), which we again do along the diagonal \(R_{1,t} = R_{2,t-1} = R_{2, t-2}\) in this particular case.

\begin{figure}[t]
\centering
\includegraphics{figures/granger-order2}
\caption{The estimated local Gaussian partial correlations for the three variable pairs in $(R_{1,t}, R_{2,t-1}, R_{2, t-2})$ given $(R_{1, t-1}, R_{1, t-2})$ estimated along the diagonal in the sample space. The first two plots represent the pairs $(R_{1,t}, R_{2,t-1})$ and $(R_{1,t}, R_{2,t-2})$ which are of particular interest when analyzing the conditional dependence relationship \eqref{eq:order2}. The legend in Figure \ref{fig:order1} is still valid, except that the solid line now represents the LGPC conditional on \emph{both} $R_{1, t-1}$ and $R_{1, t-2}$ being equal to $-0.02$.}
\label{fig:order2}
\end{figure}

Including another lag does not change the estimated conditional relationship between \(R_{1,t}\) and \(R_{2,t-1}\) much compared to the first order estimate displayed in Figure \ref{fig:order1}. There does, however, seem to be a clearer picture when considering the LGPC between \(R_{1,t}\) and \(R_{2,t-2}\), which seems to be positive in the third quadrant for all values of the conditioning variables. In this particular problem this means that conditioning on the log return of the small-firm portfolio on days \(t-1\) and \(t-2\), the large-firm log-return on day \(t\) depends on the small-firm log-return on day \(t-2\) in a nonlinear way, and more specifically; this conditional relationship can be described using the LGPC as stronger dependence in the lower left tail of the relevant conditional distribution. We present another example of Granger testing in addition to another simulated experiment in Section J of the online supplement.

\hypertarget{chap:testing}{%
\section{TESTING FOR CONDITIONAL INDEPENDENCE}\label{chap:testing}}

We should point out here that the purpose of the LGPC is not primarily to compete with the nonparametric conditional independence \emph{testing} literature, but rather to provide a new tool for the measurement of nonlinear conditional dependence. The following section could however be seen as an indication of the potential of using the LGPC to test for conditional independence. It remains to be investigated whether further improvements can be realized by using a 5-parameter specification and/or a focused version of the test; see e.g.~Tables 3--5 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}) for the substantial improvements in a similar focused independence test.

We restrict ourselves to the case where \(X_1\) and \(X_2\) are scalars, since this is the situation considered in the papers we compare with, but we reiterate (end of Section \ref{chap:definition}) that there are no conceptual difficulties in extending to the vector case as is seen in the online supplement. The theory underlying LGPC testing is essentially based on Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}), and is outlined in Sections F--H of the supplement.

\hypertarget{fauna}{%
\subsection{The recent fauna of nonparametric tests}\label{fauna}}

Property 3 in Section \ref{properties} states that the LGPC characterizes conditional dependence within a quite large class of distributions: Two stochastic vectors \(X_1\) and \(X_2\) are independent given \(\X_3\) if and only if the locally Gaussian partial correlation between them is identically equal to zero everywhere on the sample space of \((X_1, X_2, \X_3)\). The road is therefore short to a test for conditional independence that may have power against a great deal of nonlinear alternatives compared to a test based on the ordinary partial correlation coefficient.

The last decade or so has seen the publication of many new tests for conditional independence. Su and White have published a series of such tests: Su and White (\protect\hyperlink{ref-su2007consistent}{2007}) is based on detecting differences between estimated characteristic functions (which is also the method used by Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2017})). Su and White (\protect\hyperlink{ref-su2008nonparametric}{2008}) is based on estimating the Hellinger distance between conditional density estimates, Su and White (\protect\hyperlink{ref-su2012conditional}{2012}) use local polynomial quantile regression to test for conditional independence, and Su and White (\protect\hyperlink{ref-su2014testing}{2014}) present a test based on empirical likelihood. Huang (\protect\hyperlink{ref-huang2010testing}{2010}) introduces the maximal nonlinear conditional correlation which is used in a test for conditional independence, in turn extended to dependent data by Cheng and Huang (\protect\hyperlink{ref-cheng2012conditional}{2012}). Song (\protect\hyperlink{ref-song2009testing}{2009}) constructs a test via Rosenblatt transformations, while Bergsma (\protect\hyperlink{ref-bergsma2011nonparametric}{2010}) and Bouezmarni, Rombouts, and Taamouti (\protect\hyperlink{ref-bouezmarni2012nonparametric}{2012}) present new tests for conditional independence based on copula constructions. Bouezmarni and Taamouti (\protect\hyperlink{ref-boue:taam:2014}{2014}) test for conditional independence by measuring the \(L_2\) distance between estimated conditional distribution functions, and Patra, Sen, and Székely (\protect\hyperlink{ref-patra2016on}{2016}) use empirical transformations to translate conditional independence to joint independence, tests for which exist in abundance. Wang et al. (\protect\hyperlink{ref-wang2015conditional}{2015}) introduce the \emph{conditional distance correlation} which they use to construct a test for conditional independence. Linton and Gozalo (\protect\hyperlink{ref-linton1997conditional}{1997}, \protect\hyperlink{ref-linton2014testing}{2014}) formulate conditional independence in terms of probability statements which then forms the basis of a test. Also, Delgado and Manteiga (\protect\hyperlink{ref-delgado2001significance}{2001}) develop a test for conditional independence in a nonparametric regression framework. Finally, we mention testing by way of characterizing conditional independence via reproducing kernel Hilbert spaces (RKHS), see, for example, Zhang et al. (\protect\hyperlink{ref-zhang2012kernel}{2012}) and references therein.

A test based on the LGPC is quite different from the methods quoted above. It is semi-parametric and does not rely on traditional kernel estimates of density-, \mbox{distribution-,} or characteristic functions. Furthermore, due to our transformation of the data \eqref{eq:trans} to marginal standard normality, we do not necessarily have to specify weight functions in our test functional to lessen the impact of outliers - unless of course we wish to test for conditional independence in a specific portion of the sample space.

On the other hand, we have to pay a price when imposing structure on the dependence. Certain types of conditional dependence will remain invisible to our test statistic, but simulation experiments show that our test performs on par with, and sometimes better than existing, fully nonparametric tests.

\hypertarget{thetest}{%
\subsection{A test for conditional independence based on the LGPC}\label{thetest}}

We construct a test statistic for testing

\[\textrm{H}_0: X_1 \perp X_2 \,\, | \,\, \X_3\]
or, equivalently, a test statistic in terms of the marginally Gaussian pseudo observations,

\begin{equation}
\textrm{H}_0: Z_1 \perp Z_2 \,\, | \,\, \Z_3
\label{eq:ci-test}
\end{equation}
by aggregating our local measure of dependence over the sample space of \(\X\) (or \(\Z\)). A natural test statistic on the \(z\)-scale is

\begin{equation}
T_{n,b} = \int_S h\left( \widehat \alpha_b(\z)  \right) \di F_n(\z),
\label{eq:teststatistic}
\end{equation}
where \(h(\cdot)\) is a real-valued function that is typically even and non-negative for most standard applications, and \(S \subseteq \mathbb{R}^p\) is an integration area that can be altered in order to focus the test to specific portions of the sample space. Standard laws of large numbers ensure that, under regularity conditions, see Theorem F.1 of the supplement, \(T_{n,b}\) converges in probability towards its population value \[T = \int_S h\left(\alpha(\z)\right) \di F(\z).\]

Departures from many types of conditional independence lead to large values of \(T_{n,b}\) that, if larger than a critical value, leads to the rejection of \eqref{eq:ci-test}. Asymptotically, one might expect that approximate \(p\)-values for our test can be extracted from the limiting distribution of the test statistic \(T_{n,b}\), which we derive in the online supplement Section F. The limiting asymptotic distribution of \(T_{n,b}-T\) is given in Theorem F.2 for the 1-parameter general pairwise case, whereas Theorem F.3 treats the 3-coordinate (3-parameter) case with \(d_1=d_2=d_3=1\). It is interesting to note from the discussion immediately following the proofs of these theorems that the convergence rate in the first case is \(n^{-1/2}\), and in the second case is \(\left(nb^2\right)^{-1/2}\). This means that the tests are asymptotically able to pick up alternatives approaching the null hypothesis of conditional independence at these rates. This is measured by the Pitman criterion. A discussion of the Pitman criterion for other tests of conditional independence is given on pages 11--12 immediately following Theorem 2 of Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2017}). The rate \(n^{-1/2}b^{-d_3/4}\) of Wang and Hong compares favorable with other tests. The rate \(n^{-1/2}\) obtained by us must be interpreted with caution, however, because it is based on a pairwise model that is more restrictive. We go into more detail regarding asymptotic power and Pitman efficiency in Section G of the online supplement.

Several authors, for example Teräsvirta, Tjøstheim, and Granger (\protect\hyperlink{ref-terasvirta2010modelling}{2010}), have noted, however, that asymptotic analysis of nonparametric (or semiparametric) test statistics on the form \eqref{eq:teststatistic} tends to be too crude for finite sample applications. We have therefore approximated the distribution of \(T_{n,b}\) using the bootstrap. The validity of the bootstrap in such contexts has been examined by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) and we adapt their arguments to our situation in the online supplement, Section H.

In accordance with our treatment so far, let \(\{Z_{1,t}, Z_{2, t}, \Z_{3, t}\}\), \(t = 1,\ldots,n\) be observations on the \(p\)-variate stochastic vector \(\Z\), \(p\geq3\), with \(Z_1\) and \(Z_2\) being scalar and \(\Z_3\) being \((p-2)\)-variate. In order to calculate the statistic \(T_{n,b}\) for testing the null hypothesis \(Z_1 \perp Z_2 \,\,|\,\, \Z_3\), we must estimate the joint local correlation matrix \(\fSigma(\z)\) of \(\Z = (Z_1, Z_2, \Z_3)\). We exploit this in the following algorithm designed to produce approximate resampled versions of \(T_{n,b}\) under the null hypothesis of conditional independence. This procedure is a variation of the so-called local bootstrap which has been used by several authors when testing for conditional independence. A brief survey of this procedure is given in Section I of the online supplement.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Use the local correlation estimates to estimate the conditional densities \(f_{Z_1|\Z_3}(\cdot)\) and \(f_{Z_2|\Z_3}(\cdot)\) by means of the method by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}). In practice, in order to reduce the computational load, we evaluate \(\widehat f_{Z_1|\Z_3}\) and \(\widehat f_{Z_2|\Z_3}\) on a fine grid on their support, over which a continuous representation of the estimates are produced using cubic splines.
\item
  Using the accept-reject algorithm, generate \(B\) samples, each of size \(n\), from \(\widehat f_{Z_1|\Z_3}\) and \(\widehat f_{Z_2|\Z_3}\), leading to \(B\) replicates
  \[Z_m^* = \Big\{\{Z_{1t}^*, Z_{2t}^*, \Z_{3t}\}_i\Big\}_m, \qquad t = 1,\ldots,n, \,\, m = 1,\ldots,B,\]
  of \(\Z\) under the null hypothesis.
\item
  Calculate \(\{T_{n,b}\}_m^*\), \(m = 1,\ldots,B\) for the replicated data sets and obtain the approximate \(p\)-value for the conditional independence test.
\end{enumerate}

\hypertarget{chap:testexamples}{%
\subsection{Comparing with other tests}\label{chap:testexamples}}

Su and White (\protect\hyperlink{ref-su2008nonparametric}{2008}) formulate a simulation experiment for evaluating their nonparametric test for conditional independence by generating data from 10 different data generating processes (DGPs), and then check the power and level for their test. Many of the later works that were discussed in Section \ref{fauna} contain very similar experiments, and some of them present simulation results for tests published prior to Su and White (\protect\hyperlink{ref-su2008nonparametric}{2008}). This allows us to present comprehensive comparisons between the new test presented above, and many alternatives.

Let \((\epsilon_{1,t}, \epsilon_{2,t}, \epsilon_{3,t})\) be IID observations from a \(\mathcal{N}(0, I_3)\)-distribution, where \(I_3\) is the \(3\times3\) identity matrix. We will test H\(_0: X_{1,t} \perp X_{2,t} \,\,|\,\, X_{3,t}\) in 10 cases listed in Table \ref{tab:dgps} taken from Su and White (\protect\hyperlink{ref-su2008nonparametric}{2008}), that cover various types of linear and nonlinear time series dependence.

\begin{table}[t]
\begin{tabular}{ll}
\hline
1.& $(X_{1,t}, X_{2,t}, X_{3,t}) = (\epsilon_{1,t}, \epsilon_{2,t}, \epsilon_{3,t})$.\\
2.& $X_{1,t} = 0.5X_{1, t-1} + \epsilon_{1,t}$, $X_{2,t} = 0.5X_{2,t-1} + \epsilon_{2,t}$, $X_{3,t} = X_{1,t-1}$.\\
3.& $X_{1,t} = \epsilon_{1,t}\sqrt{0.01 + 0.5X_ {t-1}^2}$, $X_{2,t} = 0.5X_{2, t-1} + \epsilon_{2,t}$, $X_{3,t} = X_{1,t-1}$.\\
4.& $X_{1,t} = \epsilon_{1,t}\sqrt{h_{1,t}}$, $X_{2,t} = \epsilon_{2,t}\sqrt{h_{2,t}}$, $X_{3,t} = X_{1,t-1}$, \\ & $h_{1,t}  = 0.01 + 0.9h_{1,t-1} + 0.05X_{1,t-1}^2$, $h_{2,t} = 0.01 + 0.9h_{2,t-1} + 0.05X_{2,t-1}^2$. \\
5.& $X_{1,t} = 0.5X_{1,t-1} + 0.5X_{2,t} + \epsilon_{1,t}$, $X_{2,t} = 0.5X_{2,t-1} + \epsilon_{2,t}$, $X_{3,t} = X_{1,t-1}$.\\
6.& $X_{1,t} = 0.5X_{1,t-1} + 0.5X_{2,t}^2 + \epsilon_{1,t}$, $X_{2,t} = 0.5X_{2,t-1} + \epsilon_{2,t}$, $X_{3,t} = X_{1,t-1}$.\\
7.& $X_{1,t} = 0.5X_{1,t-1}0.5X_{2,t} + \epsilon_{1,t}$, $X_{2,t} = 0.5X_{2,t-1} + \epsilon_{2,t}$, $X_{3,t} = X_{1,t-1}$.\\
8.& $X_{1,t} = 0.5X_{1,t-1} + 0.5X_{2,t}\epsilon_{1,t}$, $X_{2,t} = 0.5X_{2,t-1} + \epsilon_{2,t}$, $X_{3,t} = X_{1,t-1}$.\\
9.& $X_{1,t} = \epsilon_{1,t}\sqrt{0.01 + 0.5X_{1,t-1}^2 + 0.25X_{2,t}}$, $X_{2,t} = X_{2,t-1} + \epsilon_{2,t}$, $X_{3,t} = X_{1,t-1}$.\\
10. & $X_{1,t} = \epsilon_{1,t}\sqrt{h_{1,t}}$, $X_{2,t} = \epsilon_{2,t}\sqrt{h_{2,t}}$, $X_{3,t} = X_{1,t-1}$, $h_{1,t} = 0.01 + 0.1h_{1, t-1} + 0.4X_{1,t-1}^2 +$ \\ & $0.5X_{2,t}^2$, $h_{2,t} = 0.01 + 0.9h_{2, t-1} + 0.5Y_t^2$.\\
\hline
\end{tabular}
\caption{Data generating processes in simulation experiment}
\label{tab:dgps}
\end{table}

The null hypotheses of conditional independence between \(X_{1,t}\) and \(X_{2,t}\) given \(X_{3,t}\) is true for DGPs 1--4, and these will be used to check the level of the test, while for DGPs 5--10 we measure the power. By evaluating our test using the sample sizes \(n = 100\) and \(n = 200\) at the \(5\%\) level, we can harvest a great number of corresponding results from the literature, and they are presented below.

The first set of simulation results are collected from Su and White (\protect\hyperlink{ref-su2008nonparametric}{2008}): In Table \ref{tab:n100} LIN is a standard linear Granger causality test, CM and KS are tests by Linton and Gozalo (\protect\hyperlink{ref-linton1997conditional}{1997}) that use statistics of the two versions of the nonparametric test developed by Delgado and Manteiga (\protect\hyperlink{ref-delgado2001significance}{2001}). HEL refers to the Hellinger distance test that Su and White (\protect\hyperlink{ref-su2008nonparametric}{2008}) present. Next, we move to Su and White (\protect\hyperlink{ref-su2014testing}{2014}), who provide simulations for their test for conditional independence based on the empirical likelihood (SEL), as well as the test by Su and White (\protect\hyperlink{ref-su2007consistent}{2007}) that is based on properties of the conditional characteristic function (CHF). Cheng and Huang (\protect\hyperlink{ref-cheng2012conditional}{2012}) provide simulations for their tests based on the maximal conditional correlation (MCC). In the online supplement to this article we provide a richer version of Table \ref{tab:n100} below, namely Table 2*, that includes more results for the HEL- and MCC-tests at different levels of smoothing.

\input{tables/power_table_dim3_n100_short}

Finally, we include results from simulations using our new test based on the LGPC, using the trivariate specification defined in Section \ref{chap:trivariate-full}, for two different levels of smoothing, and include them in Table \ref{tab:n100}. We highlight the new results in grey to indicate that they, as opposed to all the other numbers, have not appeared in the literature before. Also, to the best of our knowledge, these results have not been compared simultaneously before.

We see in Table \ref{tab:n100} that our test has the correct level and is quite powerful against all alternative specifications of conditional dependence, in the additive models 5 and 6, as well as the remaining examples 7 to 10, that are more multiplicative in nature.

When \(n=200\) we can include simulation results reported by Bouezmarni, Rombouts, and Taamouti (\protect\hyperlink{ref-bouezmarni2012nonparametric}{2012}) for their conditional independence test based on measuring the Hellinger distance between copula density estimates (shortened BRT from the names of the authors) as well as the results reported by Bouezmarni and Taamouti (\protect\hyperlink{ref-boue:taam:2014}{2014}) on their test based on \(L_2\) distances between estimated conditional distribution functions, which is abbreviated by BT. We include in Table 3* in the online supplement all results from these papers. The general picture is the same as in Table \ref{tab:n100}. In particular for the case with \(c=1.4\), our new test exhibits the best over-all performance among the examples listed, which, again to the best of our knowledge, include all such simulation results that have been published to date. It is seen that the standard linear Granger causality test (LIN) has a miserable performance in these examples.

Su and White (\protect\hyperlink{ref-su2014testing}{2014}) define two extensions to a subset of the data generating processes defined in Table 1 in which the conditioning variable \(\Xtwo\) is a vector with two and three components respectively. We refer to the online supplement for details regarding these extensions and numerical results. The new test for conditional independence based on the LGPC produces competitive results also in these cases. We point out again, however, that testing for conditional independence is merely one, albeit an important one, of several chief objectives for introducing the LGPC. Possible refinements of the LGPC testing is mentioned in the beginning of Section \ref{fauna}.

\hypertarget{conclusion-and-outlook}{%
\section{CONCLUSION AND OUTLOOK}\label{conclusion-and-outlook}}

The purpose of this paper is two-fold. First we define and develop the LGPC, a signed local measure of conditional dependence that has useful properties and is easy to interpret. It ca measure both positive and negative conditional dependence. We then explore the possibility of using the LGPC in estimation of conditional dependence and indicate its potential in testing for conditional independence. The results are promising, and suggest that the newly developed tool provides powerful improvements to existing methods.

For 3 scalar variables \(X_1, X_2, X_3\) a full trivariate analysis depending on all 3 coordinates can be undertaken as in Section \ref{chap:trivariate-full}. If any of the components \(X_1, X_2\) or \(X_3\) are vectors, a pairwise approximation and simplification has been described in Section \ref{chap:bivariate}. This can be likened to additive approximation in nonparametric regression.

Both the full trivariate approach and the pairwise simplification have been mainly analyzed at the \(z\)-scale using the transformed variables \(\Z = \Phi^{-1}(F(\X))\). The corresponding LGPC \(\alpha(\z)\) is invariant to monotone transformations of the marginals. The LGPC \(\alpha(\x)\) on the \(x\)-scale can be obtained from \(\alpha(\z)\) by taking \(\x = F^{-1}(\Phi(\z))\), and it is in general different from \(\alpha(\z)\) and is not invariant to monotone transformations of the marginals. An analogue is the different transformation properties of the Pearson correlation and the copula structure in describing joint dependence. For multivariate Gaussian variables, \(\alpha(\x) \equiv \alpha(\z) \equiv \alpha\), which is the ordinary global partial correlation.

There is a potential for much further work. A next natural step may be structural equations as used in network analysis and Pearl-type causality (Pearl \protect\hyperlink{ref-pearl2000causality}{2000}), and ultimately further exploring the relationship or lack of relationship between that type of causality and Granger causality.

All methods presented in this paper have been implemented in the R programming language (R Core Team \protect\hyperlink{ref-r}{2020}), and is available in the package \textbf{lg} (Otneim \protect\hyperlink{ref-otneim2019lg}{2019}). In addition, we provide code and data for easy replication of the results presented here in an online appendix.

\textbf{Acknowledgements.} The authors gratefully acknowledge the editor as well as two anonymus referees, whose comments have greatly improved this paper.

\hypertarget{references}{%
\section*{REFERENCES}\label{references}}
\addcontentsline{toc}{section}{REFERENCES}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-baba2004partial}{}%
Baba, Kunihiro, Ritei Shibata, and Masaaki Sibuya. 2004. ``Partial Correlation and Conditional Correlation as Measures of Conditional Independence.'' \emph{Australian \& New Zealand Journal of Statistics} 46 (4): 657--64.

\leavevmode\hypertarget{ref-baek:broc:1992}{}%
Baek, Ehung, and William Brock. 1992. ``A General Test for Nonlinear Granger Causality: Bivariate Model.'' \emph{Iowa State University and University of Wisconsin at Madison Working Paper}.

\leavevmode\hypertarget{ref-bergsma2011nonparametric}{}%
Bergsma, Wicher. 2010. ``Nonparametric Testing of Conditional Independence by Means of the Partial Copula.'' \emph{Unpublished Manuscript}. \url{http://dx.doi.org/10.2139/ssrn.1702981}.

\leavevmode\hypertarget{ref-bouezmarni2012nonparametric}{}%
Bouezmarni, Taoufik, Jeroen V K Rombouts, and Abderrahim Taamouti. 2012. ``Nonparametric Copula-Based Test for Conditional Independence with Applications to Granger Causality.'' \emph{Journal of Business and Economic Statistics} 30 (2): 275--87.

\leavevmode\hypertarget{ref-boue:taam:2014}{}%
Bouezmarni, Taoufik, and Abderrahim Taamouti. 2014. ``Nonparametric Tests for Conditional Independence Using Conditional Distributions.'' \emph{Journal of Nonparametric Statistics} 26 (4): 697--719.

\leavevmode\hypertarget{ref-cheng2012conditional}{}%
Cheng, Yu-Hsiang, and Tzee-Ming Huang. 2012. ``A Conditional Independence Test for Dependent Data Based on Maximal Conditional Correlation.'' \emph{Journal of Multivariate Analysis} 107: 210--26.

\leavevmode\hypertarget{ref-delgado2001significance}{}%
Delgado, Miguel A, and Wenceslao González Manteiga. 2001. ``Significance Testing in Nonparametric Regression Based on the Bootstrap.'' \emph{The Annals of Statistics} 29 (5): 1469--1507.

\leavevmode\hypertarget{ref-fran:moug:vale:2010}{}%
Francis, Bill B, Mbodja Mougoue, and Valentyn Panchenko. 2010. ``Is There a Symmetric Nonlinear Causal Relationship Between Large and Small Firms?'' \emph{Journal of Empirical Finance} 17 (1): 23--38.

\leavevmode\hypertarget{ref-granger1969investigating}{}%
Granger, Clive WJ. 1969. ``Investigating Causal Relations by Econometric Models and Cross-Spectral Methods.'' \emph{Econometrica}, 424--38.

\leavevmode\hypertarget{ref-granger1980testing}{}%
---------. 1980. ``Testing for Causality: A Personal Viewpoint.'' \emph{Journal of Economic Dynamics and Control} 2: 329--52.

\leavevmode\hypertarget{ref-hjort1996locally}{}%
Hjort, Nils Lid, and MC Jones. 1996. ``Locally Parametric Nonparametric Density Estimation.'' \emph{The Annals of Statistics}, 1619--47.

\leavevmode\hypertarget{ref-huang2010testing}{}%
Huang, Tzee-Ming. 2010. ``Testing Conditional Independence Using Maximal Nonlinear Conditional Correlation.'' \emph{The Annals of Statistics} 38 (4): 2047--91.

\leavevmode\hypertarget{ref-jordanger2017nonlinear}{}%
Jordanger, Lars Arne, and Dag Tjøstheim. 2020. ``Nonlinear Spectral Analysis: A Local Gaussian Approach.'' \emph{arXiv Preprint arXiv:1708.02166v4, Revised version submitted to the Journal of the American Statistical Association}.

\leavevmode\hypertarget{ref-lacal2017local}{}%
Lacal, Virginia, and Dag Tjøstheim. 2017. ``Local Gaussian Autocorrelation and Tests for Serial Independence.'' \emph{Journal of Time Series Analysis} 38 (1): 51--71.

\leavevmode\hypertarget{ref-lacal2018estimating}{}%
---------. 2018. ``Estimating and Testing Nonlinear Local Dependence Between Two Time Series.'' \emph{Journal of Business \& Economic Statistics}, 1--13.

\leavevmode\hypertarget{ref-linton1997conditional}{}%
Linton, Oliver, and Pedro Gozalo. 1997. ``Conditional Independence Restrictions: Testing and Estimation.'' Unpublished manuscript.

\leavevmode\hypertarget{ref-linton2014testing}{}%
---------. 2014. ``Testing Conditional Independence Restrictions.'' \emph{Econometric Reviews} 33 (5-6): 523--52.

\leavevmode\hypertarget{ref-otneim2016multivariate}{}%
Otneim, Håkon. 2016. ``Multivariate and Conditional Density Estimation Using Local Gaussian Approximations.'' PhD thesis, University of Bergen; The University of Bergen.

\leavevmode\hypertarget{ref-otneim2019lg}{}%
Otneim, Håkon. 2019. \emph{\texttt{lg}: Locally Gaussian Distributions: Estimation and Methods}. \url{https://CRAN.R-project.org/package=lg}.

\leavevmode\hypertarget{ref-otneim2017locally}{}%
Otneim, Håkon, and Dag Tjøstheim. 2017. ``The Locally Gaussian Density Estimator for Multivariate Data.'' \emph{Statistics and Computing} 27 (6): 1595--1616.

\leavevmode\hypertarget{ref-otneim2017conditional}{}%
---------. 2018. ``Conditional Density Estimation Using the Local Gaussian Correlation.'' \emph{Statistics and Computing} 28 (2): 303--21.

\leavevmode\hypertarget{ref-patra2016on}{}%
Patra, Rohit K, Bodhisattva Sen, and Gábor J Székely. 2016. ``On a Nonparametric Notion of Residual and Its Applications.'' \emph{Statistics and Probability Letters} 109: 208--13.

\leavevmode\hypertarget{ref-pearl2000causality}{}%
Pearl, Judea. 2000. \emph{Causality: Models, Reasoning and Inference}. Vol. 29. Springer.

\leavevmode\hypertarget{ref-r}{}%
R Core Team. 2020. \emph{R: A Language and Environment for Statistical Computing}. Vienna, Austria: R Foundation for Statistical Computing. \url{https://www.R-project.org/}.

\leavevmode\hypertarget{ref-song2009testing}{}%
Song, Kyungchul. 2009. ``Testing Conditional Independence via Rosenblatt Transforms.'' \emph{The Annals of Statistics} 37 (6B): 4011--45.

\leavevmode\hypertarget{ref-su2007consistent}{}%
Su, Liangjun, and Halbert White. 2007. ``A Consistent Characteristic Function-Based Test for Conditional Independence.'' \emph{Journal of Econometrics} 141 (2): 807--34.

\leavevmode\hypertarget{ref-su2008nonparametric}{}%
---------. 2008. ``A Nonparametric Hellinger Metric Test for Conditional Independence.'' \emph{Econometric Theory} 24 (4): 829--64.

\leavevmode\hypertarget{ref-su2014testing}{}%
---------. 2014. ``Testing Conditional Independence via Empirical Likelihood.'' \emph{Journal of Econometrics} 182 (1): 27--44.

\leavevmode\hypertarget{ref-su2012conditional}{}%
Su, Liangjun, and Halbert L White. 2012. ``Conditional Independence Specification Testing for Dependent Processes with Local Polynomial Quantile Regression.'' In \emph{Essays in Honor of Jerry Hausman}, 355--434. Emerald Group Publishing Limited.

\leavevmode\hypertarget{ref-terasvirta2010modelling}{}%
Teräsvirta, Timo, Dag Tjøstheim, and Clive W. J. Granger. 2010. \emph{Modelling Nonlinear Economic Time Series}. Oxford University Press.

\leavevmode\hypertarget{ref-tjostheim2013local}{}%
Tjøstheim, Dag, and Karl Ove Hufthammer. 2013. ``Local Gaussian Correlation: A New Measure of Dependence.'' \emph{Journal of Econometrics} 172 (1): 33--48.

\leavevmode\hypertarget{ref-wang2017characteristic}{}%
Wang, Xia, and Yongmiao Hong. 2017. ``Characteristic Function Based Testing for Conditional Independence: A Nonparametric Regression Approach.'' \emph{Econometric Theory}, 1--35.

\leavevmode\hypertarget{ref-wang2015conditional}{}%
Wang, Xueqin, Wenliang Pan, Wenhao Hu, Yuan Tian, and Heping Zhang. 2015. ``Conditional Distance Correlation.'' \emph{Journal of the American Statistical Association} 110 (512): 1726--34.

\leavevmode\hypertarget{ref-zhang2012kernel}{}%
Zhang, Kun, Jonas Peters, Dominik Janzing, and Bernhard Schölkopf. 2012. ``Kernel-Based Conditional Independence Test and Application in Causal Discovery.'' \emph{arXiv Preprint arXiv:1202.3775}.

\end{document}

### 2020-09-02 lgpc_JBES_supp.Rmd
Initial commit, revising for JBES.

---
title: "Online supplement accompanying 'The Locally Gaussian Partial Correlation'"
author: Håkon Otneim^[Corresponding author. Department of Business and Management Science, NHH Norwegian School of Economics. Helleveien 30, 5045 Bergen, Norway. \texttt{hakon.otneim@nhh.no}] and Dag Tjøstheim^[University of Bergen]
output: 
  bookdown::pdf_document2:
      dev: tikz
      keep_tex: yes
      number_sections: yes
      toc: no
      includes:
          in_header: latex_preamble_supp.tex
classoption: letterpaper
geometry: margin = 1in
bibliography: "bibliography.bib"
biblio-style: "apalike"
link-citations: yes
indent: yes
fontsize: 12pt
---

<!-- # MORE EXAMPLES {#chap:examples} -->

<!-- ## Simulated examples {#chap:simulated-examples} -->

# THE MULTIVARIATE LGPC

## Definition

The literature on nonlinear conditional dependence modeling is mostly concerned with the simple case where the vector of conditionally independent variables consists of two scalar random variables, $\Zone = (Z_1, Z_2)$, and this case is therefore the focus of our treatment of the LGPC in the main article. This means that we estimate a single local Gaussian partial correlation function $\widehat \alpha(\z)$. One of the most attractive properties of the LGPC, however, is the close parallel to the multivariate Gaussian distribution, and this is also the case when describing conditional dependence among more than two variables, which in our case leads to the estimation of a *matrix* of local Gaussian partial correlations. Besides needing a slightly more involved notation, this is a straightforward exercise in our framework as we will see below.

Let $\Z$ be a vector of continuous random variables, that we split in two components $\Z = (\Zone, \Ztwo)$, but this time $\Zone = (Z_1, \ldots, Z_{d_1 + d_2})$ is itself a vector with dimension $d_1 + d_2$ that is split in two groups; the $d_1$-dimensional vector $\Z_1$ and the $d_2$-dimensional vector $\Z_2$. The vector $\Ztwo = \Z_3$ is a vector of $d_3$ random variables. Denote by $\fmu$ and $\fSigma$ the expectation vector and the covariance matrix of $\Z$, now partitioned according to our partition of $\Z = (\Z_1, \Z_2, \Z_3)$ as

$$\fmu = \begin{pmatrix} \fmu_1 \\ \fmu_2 \\ \fmu_3 \end{pmatrix}, \qquad \fSigma = \begin{pmatrix} \fSigma_{11} & \fSigma_{12} & \fSigma_{13} \\ \fSigma_{21} & \fSigma_{22} & \fSigma_{23} \\ \fSigma_{31} & \fSigma_{32} & \fSigma_{33} \end{pmatrix},$$
where $\fSigma_{11}$ is the covariance matrix of $\Z_1$, $\fSigma_{23}$ contains the covariances between $\Z_2$ and $\Z_3$, and so forth, and where naturally $\fSigma^T = \fSigma$, $\fSigma_{12}^T = \fSigma_{21}$, $\fSigma_{13}^T = \fSigma_{31}$, and $\fSigma_{23}^T = \fSigma_{32}$.

All of these parameters can be described and estimated locally in the exact same way as laid out in the main article, resulting from the locally Gaussian joint density

\begin{equation}
\psi(\z) = \frac{1}{(2\pi)^{(d_1+d_2+d_3)/2}|\fSigma(\z)|^{1/2}} \exp \left\{-\frac{1}{2}(\z - \fmu(\z))^T\fSigma^{-1}(\z)(\z - \fmu(\z))\right\}
(\#eq:localgaussian0)
\end{equation}
and the locally Gaussian *conditional* density

\begin{align*}
\psi(\z_1, \z_2 | \Z_3 = \z_3) & = \frac{1}{(2\pi)^{(d_1+d_2)/2}|\fSigma_{12|3}(\z)|^{1/2}} \nonumber  \\ 
\qquad\qquad\qquad & \times \exp \left\{-\frac{1}{2}(\z - \fmu_{12|3}(\z))^T\fSigma_{12|3}^{-1}(\z)(\z - \fmu_{12|3}(\z))\right\}.
\end{align*}
In order to simplify notation, in the following we will drop the $z$-dependence in the parameters. In order to define $\fmu_{12|3}$ and $\fSigma_{12|3}$ let $\fSigma^{11}$, $\fSigma^{12}$, $\fSigma^{21}$ and $\fSigma^{22}$ be matrices of dimension $(d_1 + d_2) \times (d_1 + d_2)$, $(d_1 + d_2) \times d_3$, $d_3 \times (d_1 + d_2)$ and $d_3 \times d_3$ respectively given by

$$\fSigma^{11} = \begin{pmatrix} \fSigma_{11} & \fSigma_{12} \\ \fSigma_{21} & \fSigma_{22}\end{pmatrix}, \fSigma^{12} = \begin{pmatrix} \fSigma_{13} \\ \fSigma_{23}\end{pmatrix}, \fSigma^{21} = \begin{pmatrix} \fSigma_{31} & \fSigma_{32}\end{pmatrix}, \textrm{ and } \, \fSigma^{22} = \fSigma_{33}$$
such that 
$$\fSigma = \begin{pmatrix} \fSigma^{11} & \fSigma^{12} \\ \fSigma^{21} & \fSigma^{22}\end{pmatrix},$$
It follows from standard arguments (cf. @otneim2017conditional) that
\begin{equation}
\fmu_{12|3} = \begin{pmatrix} \fmu_1 \\ \fmu_2 \end{pmatrix} + \fSigma^{12}\left(\fSigma^{22}\right)^{-1}\left(\ztwo - \fmu_3\right) = \begin{pmatrix} \fmu_1 \\ \fmu_2 \end{pmatrix} + \begin{pmatrix} \fSigma_{13}\fSigma_{33}^{-1} \left(\ztwo - \fmu_3\right) \\  \fSigma_{23}\fSigma_{33}^{-1} \left(\ztwo - \fmu_3\right)\end{pmatrix}
\end{equation}
and

\begin{align}
\fSigma_{12|3} &= \fSigma^{11} - \fSigma^{12}\left(\fSigma^{22}\right)^{-1}\fSigma^{21} = \fSigma^{11} - \begin{pmatrix} \fSigma_{13} \\ \fSigma_{23} \end{pmatrix} \fSigma_{33}^{-1} \begin{pmatrix} \fSigma_{31} & \fSigma_{32} \end{pmatrix}\nonumber \\
&= \begin{pmatrix} \fSigma_{11} - \fSigma_{13}\fSigma_{33}^{-1} \fSigma_{31} & \fSigma_{12} - \fSigma_{13}\fSigma_{33}^{-1} \fSigma_{32} \\ \fSigma_{21} - \fSigma_{23}\fSigma_{33}^{-1} \fSigma_{31} & \fSigma_{22} - \fSigma_{23}\fSigma_{33}^{-1} \fSigma_{32}\end{pmatrix}. 
(\#eq:blocksigma)
\end{align}

We proceed now to establish the notion of conditional independence between $\Z_1$ and $\Z_2$ using the LGPC in the multivariate case. The conditional densities of $\Z_1|\Z_3$ and $\Z_2|\Z_3$ in terms of the local Gaussian parameters are given by

\begin{align*}
f(\z_1 |\z_3) &= \frac{1}{(2\pi)^{d_1/2}|\fSigma_{1|3}|^{1/2}} \exp \left\{-\frac{1}{2}(\z_1 - \fmu_{1|3})^T\fSigma_{1|3}^{-1}(\z_1 - \fmu_{1|3})\right\}, \\
f(\z_2 |\z_3) &= \frac{1}{(2\pi)^{d_2/2}|\fSigma_{2|3}|^{1/2}} \exp \left\{-\frac{1}{2}(\z_2 - \fmu_{2|3})^T\fSigma_{2|3}^{-1}(\z_2 - \fmu_{2|3})\right\} 
\end{align*}
where

\begin{equation}
\fmu_{1|3} = \fmu_1 + \fSigma^{13}\left(\fSigma^{33}\right)^{-1}\left(\z_3 - \fmu_3\right), \,\, \fmu_{2|3} = \fmu_2 + \fSigma^{23}\left(\fSigma^{33}\right)^{-1}\left(\z_3 - \fmu_3\right)  
\end{equation}
and

\begin{equation}
\fSigma_{1|3} = \fSigma_{11}-\fSigma_{13}\fSigma_{33}^{-1}\fSigma_{31}, \,\, \fSigma_{2|3} = \fSigma_{22}-\fSigma_{23}\fSigma_{33}^{-1}\fSigma_{32}.
(\#eq:localsigma)
\end{equation}
Under conditional independence between $\Z_1$ and $\Z_2$ given $\Z_3$, the joint conditional density of $(\Z_1, \Z_2|\Z_3)$ is given by

\begin{align*}
f(\z_1, \z_2|\z_3) &= f(\z_1|\z_3)f(\z_2|\z_3) \\
&= \frac{1}{(2\pi)^{(d_1+d_2)/2}|\fSigma_{1|3}|^{1/2}|\fSigma_{2|3}|^{1/2}} \\
& \qquad \times \exp \Big\{-\frac{1}{2}\big[(\z_1 - \fmu_{1|3})^T\fSigma_{1|3}^{-1}(\z_1 - \fmu_{1|3}) \,\, + \\
& \qquad\qquad\qquad\qquad\qquad\qquad(\z_2 - \fmu_{2|3})^T\fSigma_{2|3}^{-1}(\z_2 - \fmu_{2|3})\big]\Big\}.
\end{align*}
Comparing expressions \@ref(eq:localgaussian0) -- \@ref(eq:localsigma) we see that we have conditional independence if and only if the matrix $\fSigma_{12|3}$ in \@ref(eq:blocksigma) is block diagonal, i.e. if and only if 
$$\fSigma_{12} - \fSigma_{13}\fSigma_{33}^{-1} \fSigma_{32} = \fSigma_{21} - \fSigma_{23}\fSigma_{33}^{-1} \fSigma_{31} = \bm{0}.$$

## Some special cases {#special-cases}

Let $\D = \textrm{diag}\left\{\fSigma_{12|3}\right\}$. The local Gaussian partial correlation matrix can be expressed as the lower left hand block matrix of $\D^{-1/2}\fSigma_{12|3}\D^{-1/2}$, and so the LGPC matrix is thus given as the upper right hand block of the matrix
\begin{equation}
\D^{-1/2}\fSigma_{12|3}\D^{-1/2} = \D^{-1/2}\begin{pmatrix} \fSigma_{11} - \fSigma_{13}\fSigma_{33}^{-1} \fSigma_{31} & \fSigma_{12} - \fSigma_{13}\fSigma_{33}^{-1} \fSigma_{32} \\ \fSigma_{21} - \fSigma_{23}\fSigma_{33}^{-1} \fSigma_{31} & \fSigma_{22} - \fSigma_{23}\fSigma_{33}^{-1} \fSigma_{32}\end{pmatrix}\D^{-1/2}.
(\#eq:matrixdef)
\end{equation}
Let us consider this expression for a few special cases.

\begin{itemize}
\item[\textbf{a)}] \textbf{Let} $\bm{d_1 = d_2 = d_3 = 1}$. Consider the partial covariance matrix, where we again drop the $z$-dependence in all parameters: 
$$\fSigma_{12|3} = \fSigma^{11} - \fSigma^{12}\left(\fSigma^{22}\right)^{-1}\fSigma^{21}.$$
Here, since all variables are normalized with variance one, we have that
$$\fSigma = \begin{pmatrix} 1 & \rho_{12} & \rho_{13} \\ \rho_{12} & 1 & \rho_{23} \\ \rho_{31} & \rho_{32} & 1 \end{pmatrix},$$
and 
$$\fSigma^{11} = \begin{pmatrix} 1 & \rho_{12} \\ \rho_{21} & 1 \end{pmatrix}, \,\, \fSigma^{12} = \begin{pmatrix} \rho_{13} \\ \rho_{23} \end{pmatrix},\,\, \left(\fSigma^{22}\right)^{-1} = 1, \,\, \fSigma^{21} = \begin{pmatrix} \rho_{31} & \rho_{32} \end{pmatrix}.$$
Hence,
\begin{align*}
\fSigma^{11} - \fSigma^{12}\left(\fSigma^{22}\right)^{-1}\fSigma^{21} &=  \begin{pmatrix} 1 & \rho_{12} \\ \rho_{21} & 1 \end{pmatrix} - \begin{pmatrix} \rho_{13} \\ \rho_{23} \end{pmatrix} \begin{pmatrix} \rho_{31} & \rho_{32} \end{pmatrix} \\
&= \begin{pmatrix} 1 - \rho_{13}^2 & \rho_{12} - \rho_{13}\rho_{23} \\ \rho_{12} - \rho_{13}\rho_{23} & 1 - \rho_{23}^2 \end{pmatrix}.
\end{align*}
This is a covariance matrix, and not a correlation matrix, so normalizing with standard deviations results in the partial correlation matrix
\begin{equation}
\begin{pmatrix}
1 & \frac{\rho_{12} - \rho_{13}\rho_{23}}{\sqrt{1-\rho_{13}^2}\sqrt{1-\rho_{23}^2}} \\  \frac{\rho_{12} - \rho_{13}\rho_{23}}{\sqrt{1-\rho_{13}^2}\sqrt{1-\rho_{23}^2}} & 1 
\end{pmatrix}.
\label{eq:partialmatrix}
\end{equation}
This result can also be obtained by direct application of \eqref{eq:matrixdef}:
$$\begin{pmatrix}1-\rho_{13}^2 & 0 \\ 0 & 1-\rho_ {23}^2\end{pmatrix}^{-1/2} \begin{pmatrix}  1 - \rho_{13}^2 & \rho_{12} - \rho_{13}\rho_{23} \\ \rho_{12} - \rho_{13}\rho_{23} & 1 - \rho_{23}^2  \end{pmatrix} \begin{pmatrix}1-\rho_{13}^2 & 0 \\ 0 & 1-\rho_ {23}^2\end{pmatrix}^{-1/2}.$$

\item[\textbf{b)}] \textbf{Let} $\bm{d_3=1}$ \textbf{and let} $\bm{d_1}$ \textbf{and} $\bm{d_2}$ \textbf{be arbitrary}. In this case 
$$\fSigma^{11} = \begin{pmatrix} \fSigma_{11} & \fSigma_{12} \\ \fSigma_{21} & \fSigma_{22}\end{pmatrix},$$
where
\begin{align*}
&\fSigma_{11} = \left\{ \rho_{ij}^{1,1} \right\}, i = 1, \ldots, d_1; j = 1, \ldots, d_1,   &\fSigma_{12} = \left\{ \rho_{ij}^{1,2} \right\}, i = 1, \ldots, d_1; j = 1, \ldots, d_2 \\
&\fSigma_{21} = \left\{ \rho_{ij}^{2,1} \right\}, i = 1, \ldots, d_2; j = 1, \ldots, d_1,   &\fSigma_{22} = \left\{ \rho_{ij}^{2,2} \right\}, i = 1, \ldots, d_2; j = 1, \ldots, d_2.
\end{align*}
Since $\fSigma^{22} = \fSigma_{33} = 1,$
$$\fSigma^{12}\left(\fSigma^{22}\right)^{-1}\fSigma^{21} = \fSigma^{12}\fSigma^{21} = \begin{pmatrix} \fSigma_{13} \\ \fSigma_{23} \end{pmatrix} \begin{pmatrix}\fSigma_{31} \fSigma_{32} \end{pmatrix},$$
where 
\begin{align*}
&\fSigma_{13} = \left\{ \rho_{i3}^{1,3} \right\}, i = 1, \ldots, d_1;   &\fSigma_{23} = \left\{ \rho_{j3}^{2,3} \right\}, j = 1, \ldots, d_2 \\
&\fSigma_{31} = \left\{ \rho_{3i}^{3,1} \right\}, i = 1, \ldots, d_1;   &\fSigma_{32} = \left\{ \rho_{3j}^{3,2} \right\}, j = 1, \ldots, d_2.
\end{align*}
Furthermore, 
$$\fSigma_{12|3} = \begin{pmatrix}\fSigma_{11} & \fSigma_{12} \\ \fSigma_{21} & \fSigma_{22}\end{pmatrix} - \begin{pmatrix} \fSigma_{13} \\ \fSigma_{23}\end{pmatrix}\begin{pmatrix} \fSigma_{31} & \fSigma_{32}\end{pmatrix}.$$
The diagonal of $\fSigma_{11}$ and $\fSigma_{22}$ are populated with ones, whereas, using symmetry of correlations, $\begin{pmatrix} \fSigma_{13} & \fSigma_{23}\end{pmatrix}^T\begin{pmatrix} \fSigma_{31} & \fSigma_{32}\end{pmatrix}$ has $\left(\rho_{i3}^{1,2}\right)^2$, $i=1,\ldots,d_1$ and $\left(\rho_{3j}^{3,2}\right)^2$, $j=1,\ldots,d_2$ along the diagonal, such that $\D$ is the $(d_1+d_2)\times(d_1+d_2)$ diagonal matrix with the concatenated sequence $1-\left(\rho_{i3}^{1,2}\right)^2, i=1,\ldots,d_1; 1- \left(\rho_{3j}^{3,2}\right)^2, j=1,\ldots,d_2$ along the diagonal.

Pre- and post multiplying $\fSigma_{12|3}$ by $\D^{-1/2}$ converts $\fSigma_{12|3}$ into a correlation matrix. All of the off-diagonal elements in this matrix has the same form as the off-diagonal elements in \eqref{eq:partialmatrix} so that
\begin{align}
\rho(Z_{1,i},Z_{1,j}|Z_3) &= \frac{\rho(Z_{1,i},Z_{1,j}) - \rho(Z_{1,i},Z_3)\rho(Z_{1,j},Z_3)}{\sqrt{1-\rho^2(Z_{1,i},Z_3)} \sqrt{1-\rho^2(Z_{1,j},Z_3)}} 
, \; i,j = 1,\ldots,d_1, \label{eq:offdiag1}\\
\rho(Z_{2,i},Z_{2,j}|Z_3) &= \frac{\rho(Z_{2,i},Z_{2,j}) - \rho(Z_{2,i},Z_3)\rho(Z_{2,j},Z_3)}{\sqrt{1-\rho^2(Z_{2,i},Z_3)} \sqrt{1-\rho^2(Z_{2,j},Z_3)}}
, \; i,j = 1,\ldots,d_2; \label{eq:offdiag2} \\
\rho(Z_{1,i},Z_{2,j}|Z_3) &= \frac{\rho(Z_{1,i},Z_{2,j}) - \rho(Z_{1,i},Z_3)\rho(Z_{2,j},Z_3)}{\sqrt{1-\rho^2(Z_{1,i},Z_3)} \sqrt{1-\rho^2(Z_{2,j},Z_3)}}
, \; i = 1,\ldots,d_1; j=1,\ldots,d_2. \label{eq:offdiag3}
\end{align}

Testing for conditional independence between $\Z_1$ and $\Z_2$ given $\Z_3$ is in our context the same as testing for zero local partial correlation between all pairs of variables, a test for which is presented in the main article and discussed in technical detail later in this supplement. One can also, based on \eqref{eq:offdiag1} and \eqref{eq:offdiag2}, test for conditional independence between combinations of components given $\Z_3$. Equations \eqref{eq:offdiag1}--\eqref{eq:offdiag3} above give explicit expressions for the conditional correlation matrix of $\Z_1$ and $\Z_2$ given $\Z_3$.

\item[\textbf{c)}] \textbf{Let} $\bm{d_1 = d_2 = 1}$ \textbf{and let} $\bm{d_3}$ \textbf{be arbitrary}. This is the case usually treated in the literature, also the focus of the discussions in the main article. In this case
$$\fSigma^{11} = \begin{pmatrix} 1 & \rho_{12} \\ \rho_{21} & 1 \end{pmatrix}, \fSigma_{13} = \left\{\rho_{1i}^{1,3}\right\}, i=1,\ldots,d_3; \fSigma_{23} = \left\{\rho_{2i}^{2,3}\right\}; i=1,\ldots,d_3,$$
hence $\begin{pmatrix} \fSigma_{13} \fSigma_{23}\end{pmatrix}^T$ is of dimension $2\times d_3$, whereas $\begin{pmatrix} \fSigma_{31} \fSigma_{32}\end{pmatrix}$ is of dimension $d_3\times 2$ such that
$$\begin{pmatrix} \fSigma_{13} \\ \fSigma_{23} \end{pmatrix} \fSigma_{33}^{-1} \begin{pmatrix} \fSigma_{31} & \fSigma_32 \end{pmatrix}$$
is of dimension $2\times2$. However, as is immediately seen, no formulas like \eqref{eq:offdiag1}--\eqref{eq:offdiag3} will appear; they will be much more complicated and include sums and products of various correlations.

\item[\textbf{d)}] \textbf{Let} $\bm{d_1, d_2}$ \textbf{and} $\bm{d_3}$ \textbf{be arbitrary}. This is the general case, and it seems impossible to obtain explicit formulas for the local Gaussian partial correlations, but the general fact still holds of course: The local partial correlation matrix of $\Z_1$and $\Z_2$ given $\Z_3$ can be identified by the lower left hand block matrix of $\D^{-1/2}\fSigma_{12|3}\D^{-1/2}$, and can be estimated using pairwise local correlations in \eqref{eq:matrixdef}.
\end{itemize}

## Estimation and estimation theory in more general situations {#chap:estimation}

Sections \@ref(app-loccor) -- \@ref(proofpseudo) of this supplement deal with the results stated in Section 3 of the main article where $X_1$ and $X_2$ are scalars. The present subsection looks briefly at more general situations.

Recall that the matrix $D$ defined in the beginning of Section \@ref(special-cases) is given by $D = \textrm{diag}(\fSigma_{12|3})$. Let us block divide the $(d_1+d_2)\times (d_1+d_2)$ diagonal matrix $\D$ into
$$\D = \begin{pmatrix} \D_{11} & \bm{0} \\ \bm{0} & \D_{22}\end{pmatrix}.$$
The estimated partial correlation matrix between $\X_1$ and $\X_2$ given $\X_3$ is then given as 
\begin{equation}
\hD_{11}^{-1/2}\left(\hfSigma_{21} - \hfSigma_{23}\hfSigma_{33}^{-1}\hfSigma_{31}\right)\hD_{22}^{-1/2} \stackrel{\textrm{def}}{=} g(\hfSigma).
(\#eq:lgpc-matrix)
\end{equation}

We present the asymptotic theory for the local likelihood estimator of the local partial correlations in two situations in the main article with details of proof in Sections \@ref(app-loccor) -- \@ref(proofpseudo):

\begin{itemize}
\item[\textbf{(i)}] If $\bm{d_1 = d_2 = d_3=1}$, then we can estimate the full trivariate local likelihood estimate, meaning that all the local correlations depend on all three coordinates $z=(z_1,z_2,z_3)$. The asymptotic theory for this estimator is covered by Theorem 3.1 in the main article.

\item[\textbf{(ii)}] If $\bm{d_1 = d_2 = 1}$ \textbf{and} $\bm{d_3 > 1}$, we estimate all correlations in a pairwise manner as laid out in the main article. The limiting distribution of the (scalar) local Gaussian partial correlation is in general given by 
\begin{equation}
\sqrt{nb_n^2}\left(\widehat{\alpha}(z))-\alpha(z)\right)  \stackrel {{\cal L}}{\to} {\cal N}(0,\nabla g(\frho)^{T} {\fSigma} \nabla g (\frho)),
\end{equation}
see eq. (22) in the main article, and see Theorem 2 in Otneim and Tjøstheim (2018) for a definition of $\fSigma$. The computation of $\nabla g$ is outlined in Section \ref{app-asvar}.
\end{itemize}

These results can be extended to the multivariate cases that we treat in this supplement in Section \@ref(special-cases) as well. The assumptions remain identical to the assumptions in Theorem 3.1 in the main article and Theorem 3 in @otneim2017conditional.

\begin{itemize}
\item[\textbf{(iii)}] If $\bm{d_1}$ \textbf{and} $\bm{d_2}$ \textbf{are arbitrary, and} $\bm{d_3 = 1}$, the local partial correlations take the form \eqref{eq:offdiag3}. The limiting distribution of $\widehat\rho(Z_{1i}, Z_{2j}|Z_3)$ follows from the delta rule applied to the joint distribution of $\{\widehat\rho(Z_{1i}, Z_{2j}), \widehat\rho(Z_{1i}, Z_3), \widehat\rho(Z_{2j}, Z_3)\}$, the joint distribution of which again follows directly from Theorem 2 in Otneim and Tjøstheim (2018), who also note that the asymptotic correlation between two local correlation estimates is of smaller order if the two index pairs are distinct. The computation of $\nabla g$ is fairly straightforward in this case, by taking the appropriate derivatives in \eqref{eq:offdiag3}.

\item[(\textbf{iv})] If $\bm{d_1, d_2}$ \textbf{and} $\bm{d_3}$ \textbf{are all arbitrary}, we have the general case, where, in principle, for each $i=1,\ldots,d_1$ and $j=1,\ldots,d_2$, the local partial correlation $\rho(Z_{1i}, Z_{2j}|\Z_3)$ can be expressed as a deterministic function $g_{ij}$ of a local correlation vector $\frho$ using equation \eqref{eq:lgpc-matrix}, the asymptotic normality for which is proved as before using the delta rule and Theorem 2 of Otneim and Tjøstheim (2018). It is, however, difficult to obtain explicit expressions of the asymptotic covariance matrix in this case.
\end{itemize}
The joint asymptotic normality of the pairwise partial correlation estimates can be proved using the Cramer-Wold device. But, again, obtaining an expression for the asymptotic covariance matrix is hard in the general case.

Observe, finally, that a more general pairwise scheme can be devised, as treated in detail by @lacal2017local, @lacal2018estimating and @jordanger2017nonlinear:

\begin{itemize}
\item[\textbf{(v)}] In this scheme, we do not use a pairwise set-up for the local correlation only (the one-parameter case), but for all parameters in a bivariate normal distribution; the correlation $\rho(x_1, x_2)$, the expectations $\mu_1(x_1,x_2), \mu_2(x_1,x_2)$ and the variances $\sigma_1(x_1,x_2),\sigma_2(x_1,x_2)$ (the five-parameter case). This setup can also be simplified and fitted to a $p$-variate distribution by estimating $\mu_i = \mu_i(x_1)$ and $\sigma_i = \sigma_i(x_i)$, $i=1,\ldots,p$, marginally. Tjøstheim and Hufthammer (2013) give the asymptotic theory in this case. 
\end{itemize}

Below is a brief summary of the relevant local likelihood theory for the general pairwise case of (ii)--(iv). We also state a general consistency result in Theorem \@ref(thm:A1) at the end of this section.

The pairwise log-likelihood in cases (ii) -- (iv) is given by

\begin{align}
L_{n,b}(\Z_{in}, \Z_{jn}; \rho_b(z_i, z_j)) &= \frac{1}{n}\sum_{t=1}^n K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
& \qquad\qquad\qquad - \int K_b(v-z_i, w- z_j)\psi(v,w;\rho_b(z_i, z_j)) \, \di v \di w,
(\#eq:loglikelihood)
\end{align}
where $\Z_{in} = (Z_{i,1}, Z_{i,2}, \ldots, Z_{i,n})$ and $\Z_{jn} = (Z_{j,1}, Z_{j,2}, \ldots, Z_{j,n})$ are the observation vectors for components $i$ and $j$ respectively, $n$ is the number of observations, $z_i, z_j, v, w \in \mathbb{R}$, $\rho_b = \rho_{ij,b}(z_i,z_j)$ is the local Gaussian correlation parameter, $K_b(v-z_i, w-z_j) = (b_1b_2)^{-1}K(b_1^{-1}(v-z_i))K(b_2^{-1}(w-z_j))$ is a product kernel with bandwidth $b=b_n=(b_1,b_2) = (b_{1n}, b_{2n})$, and $K$ is an non-negative kernel function with compact support such that $\int K(v,w) = 1$. For a fixed bandwidth $b$, $\rho_{0,b}$ is the true local parameter, that is, the minimizer of the local penalty function

\begin{equation}
\int K_b(v-z_i, w-z_j)\big[\psi(v,w; \rho_b(z_i, z_j)) - \log(\psi(v,w; \rho_b(z_i, z_j)))f(v,w)\big] \di v \di w,
(\#eq:penalty)
\end{equation}
where $f$ is the density function of $(Z_i, Z_j)$. Note in particular that $\rho_{0,b} \rightarrow \rho_0$ as $b\rightarrow0$, which is the true local correlation parameter. The reader may consult @tjostheim2013local for more details and regularity conditions in this regard. The corresponding estimate $\rho_{n, b}(z_i,z_j)$ is obtained by maximizing the log-likelihood in \@ref(eq:loglikelihood). Note that, writing $\rho$ for $\rho_b$,
\begin{equation}
L_{n,b}(\rho) = \frac{1}{n}\sum_{t=1}^n L^{(b)}(Z_{i,t}, Z_{j,t}; \rho),
\end{equation}
where
\begin{align}
L^{(b)}(Z_{i,t}, Z_{j,t}; \rho) &= K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
& \qquad\qquad\qquad \int K_b(v-z_i, w- z_j)\rho_b(v,w;\psi_b(z_i, z_j)) \, \di v \di w.
\label{eq:A16}
\end{align}

We must next proceed by introducing some new concepts before stating the equivalents of assumptions A1 -- A2 by @lacal2018estimating. These assumptions are similar but distinct from those used in Theorem 3.1 in the main article and Theorem 2 in @otneim2017conditional. They will also be used to prove the consistency result in Theorem \@ref(thm:A1) at the end of this section, and they will be part of the assumptions used to prove the validity of the bootstrap in Section \@ref(validity-bootstrap).

For a given function $g$, the process $\{g(Z_{i,t}, Z_{j,t}; \rho)\}$ is said to be $r$-dominated on $\Theta$ uniformly in $t$, where $t = 1,\ldots,n$, if there exists a function $D_t: \mathbb{R} \rightarrow \mathbb{R}$ such that $|g(Z_{i,t}, Z_{j,t}; \rho)| \leq D_t$ for all $\rho \in \Theta$, $t = 1,\ldots, n$ and $D_t$ is $\mathcal{B}$-measurable such that for some $\Delta > 0$, $||D_t||_r \stackrel{\textrm{def}}{=} \left(\int |D_t(u)|^r\di u\right)^{1/r} \leq \Delta < \infty$ for all $t = 1,\ldots,n$, where $\mathcal{B}$ is the Borel $\sigma$-algebra. In this particular case the location $(z_i, z_j)$ is fixed, and $\Theta = [-1,1]$. Moreover, the process $\{g(Z_{i,t}, Z_{j,t}; \rho)\}$ is said to be Lipshitz continuous on $\Theta$ if

$$|\{g(Z_{i,t}, Z_{j,t}; \rho)\} - \{g(Z_{i,t}, Z_{j,t}; \rho')\}| \leq C_t|\rho - \rho'|$$
almost surely in $\mathbb{P}$ for all $\rho,\rho'\in\Theta$ and with a sufficiently large constant $M$ such that $C_t\leq M$. Here $\mathbb{P}$ is the probability measure on the probability space in which $\{Z_{i,t}, Z_{j,t}\}$ is defined. We can now state conditions A1 -- A2 which are basic assumptions for proving consistency and asymptotic normality of $\rho_{n,b}$:

\begin{itemize}
\item[\textbf{A1}] $(\Omega, \mathcal{F}, \mathbb{P})$ is a complete probability space on which is defined a sequence of random functions $\{L_{n,b}: \Omega\rightarrow\overline{\mathbb{R}}\}$, where $\overline{\mathbb{R}}$ is the extended real line. The parameter set $\Theta$ is compact (trivially fulfilled in our case because $\Theta = [-1,1]$). The random process $\{Z_{i,t}, Z_{j,t}\}$ is $\mathcal{F}_t$-measurable, stationary and $\alpha$-mixing with mixing coefficients $\alpha_k = O(\alpha^k)$, $\alpha \in (0,1)$ for all $t$, and with a bivariate density having support in all of $\mathbb{R}^2$. Here $\mathcal{F}_t$ is the $\sigma$-algebra generated by $\{Z_{i,s}, Z_{j,s}, s \leq t\}$.
\item[\textbf{A2}] The parameter value $\rho_{0,b}$ is the unique maximizer of $\E(L_{n,b}(\rho)): \Theta \rightarrow \overline{\mathbb{R}}$, with $\E(L_{n,b}(\rho))$ is defined in \eqref{eq:penalty}.
\end{itemize}

The conditions A1-- A2 are quite mild and discussed by @lacal2018estimating. Note that they have been formulated for a scalar parameter $\rho = \rho_{ij}$. To have them fulfilled for all possible combinations of $i,j$, the assumption A1, for example can be easily formulated for the $\{\Z_t\}$ process. Consistency for the local parameter estimates is required for consistency of the bootstrap:

\begin{thm} Let assumptions A1 and A2 hold, and let $\{L^{(b)}(Z_{i,t}, Z_{j,t};\rho)\}, \rho \in \Theta$ be Lipshitz continuous on $\Theta$ and 4-dominated on $\Theta$ uniformly in $t$ and $n$. Moreover, assume that as $b=b_n\rightarrow 0$, $\sigma^2_b \stackrel{\textrm{def}}{=} \Var\left(L^{(b)}(Z_{i,t}, Z_{j,t};\rho)\right) \rightarrow \sigma^2 < \infty$ and $\mu_b \stackrel{\textrm{def}}{=} \E\left(L^{(b)}(Z_{i,t}, Z_{j,t};\rho)\right) \rightarrow \mu < \infty$. Then $\rho_{n,b} - \rho_{0,b} \stackrel{\mathbb{P}}{\rightarrow} 0$ for a fixed $b$ and $\rho_{n,b} - \rho_{0} \stackrel{\mathbb{P}}{\rightarrow} 0$ as $b \rightarrow 0$.
\label{thm:A1}
\end{thm}

The proof of Theorem \@ref(thm:A1) is mostly identical to the proof of Theorem 2 of @lacal2018estimating with trivial modifications to take into account the scalar nature of the parameter $\rho_{ij}$ in the present case. A distributional limit theorem analogous to Theorem 2.2 in @lacal2018estimating can also be rephrased and proved as in that paper, but we refrain from doing this given the distributional results already stated in Section 3 of the main part of the paper.

# DETAILS REGARDING THE PROOF OF THEOREM 3.1 {#app-loccor}

For fixed bandwidths we have from standard arguments, such as those provided by @hjort1996locally, that 
\begin{equation}
\sqrt{nb^3}\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(0, \Jb^{-1}\Mb(\Jb^{-1})^T),
(\#eq:fixed-h)
\end{equation}
which corresponds to the usual rate for nonparametric density estimation. As $\hh\rightarrow0$, however, we must take extra care when considering the asymptotic covariance matrix. @tjostheim2013local, Section 4,  write the Taylor expansions of $\Mb$ and $\Jb$, being functionals of three variables $(x_1, x_2, x_3)$ or $(z_1,z_2,z_3)$, as
$$\Mb = I_{\bm{M}} + II_{\bm{M}} + o(b^2) \,\,\, \textrm{ and } \Jb = I_{\bm{J}} + II_{\bm{J}} + o(b^2),$$
and consider each term. For example, we write their first term in the expansion of $\Mb$ in the three-variate case as follows:
$$I_{\bm{M}} \sim \int K^2(w_1, w_2, w_3)Ab_wb_w^TA^Tf(\z + b_1w_1 + b_2w_2 + b_3w_3)\,\textrm{d}w_1\,\textrm{d}w_2\,\textrm{d}w_3,$$
where $K$ is a product kernel, and, in our case, $b_w$ is the vector defined by $b_w^T = \begin{pmatrix} 1 & b_1w_1 & b_2w_2 & b_3w_3 \end{pmatrix}$, and $A$ is the $3\times 4$ matrix $A = \begin{pmatrix} v & v_{z_1} & v_{z_2} & v_{z_3}\end{pmatrix}$ with $\fv^T =  \begin{pmatrix} v_1 & v_2 & v_3 \end{pmatrix}$, $v_i = \partial L_n/\partial \rho_i$, where the index $i=1,2,3$ represents the three local correlations, and $v_{z_i} = \partial \fv/\partial z_i$. In the next step, they compute the matrix

\noindent $\int K^2(w_1, w_2, w_3)b_wb_w^T\,\textrm{d}w_1\,\textrm{d}w_2\,\textrm{d}w_3$, which, by omitting all constant factors and exploiting that $\int w^k K^2(w)\,\textrm{d}w = 0$ for $k = 1,3$, in our case becomes the diagonal matrix 
$$H = \begin{pmatrix} 1 & 0 & 0 & 0 \\ 0 & b_1^2w_1^2 & 0 & 0 \\ 0 & 0 & b_2^2w_2^2 & 0 \\ 0 & 0 & 0 & b_3^2w_3^2 \end{pmatrix}.$$

The second term defining $\Mb$ has smaller order than $I_{\bm{M}}$ as $b \rightarrow 0$, while the first term of $\Jb$ can be treated exactly as $I_{\bm{M}}$ above resulting in a matrix of order $b^2$.

The second term in $\Jb$ can similarly be written as $II_J = B(\psi(\z, \theta_h(\z)) - f(\z)) + o(b^2)$, where $B = \nabla \fv(\z, \theta_b(\z))$ are second derivatives of the local likelihood function with respect to the local correlations, and $\psi(\z, \theta_b(\z)) - f(\z) = O(b^2)$ according to @hjort1996locally. Hence, it follows that the leading term of $\Jb^{-1}\Mb(\Jb^{-1})^T$ is $O(b^{-2})$, which again means that the convergence rate in eq. \@ref(eq:fixed-h) above must be modified accordingly in order to balance the convergence in eq. (20) in the main article. Extracting an analytic expression for the leading term of the asymptotic covariance matrix is, as mentioned by @tjostheim2013local,  possible by means of symbolic manipulation software, but the expression is very complicated and of little practical use. See also the corresponding proof of Theorem 3 in @tjostheim2013local.

# THE ASYMPTOTIC VARIANCE OF THE LGPC {#app-asvar}

The limiting distribution of the LGPC is given in eq. (22) in the main article. In this section, we will present the basic steps needed to work out the value of $\nabla g(\frho)$ in the simplest case where $d_1 = d_2=1$. In the more general situation it is not feasible to obtain analytic expressions for the asymptotic variance, and the bootstrap should be used. For the sake of this particular argument, we may simplify notation quite a bit, leaving out what we do not need: We drop the $\z$-dependence, and simply write $\fSigma = \fSigma_{12|3}(\z)$ for the $2\times2$ local partial covariance matrix between $Z_1$ and $Z_2$ given $(Z_3,\ldots,Z_p)$. The vector of local correlations $\frho$ is indexed by $k$, which in turn means that $k$ corresponds to a specific *pair* $(i,j)$: $\rho_k = \rho_{ij}$, with $\rho_{ij} = \rho_{ji}$. Let $\fSigma^{(k)}$ be the element-wise matrix of partial derivatives
$$\fSigma^{(k)} = \frac{\partial \fSigma}{\partial \rho_k} = \frac{\partial \fSigma}{\partial \rho_{ij}},$$
which means that element $k$ in the gradient $\nabla g(\frho)$ is given by
\begin{equation}
\left\{\nabla g(\frho) \right\}_k = \frac{\fSigma_{11}\fSigma_{22}\fSigma^{(k)}_{12} - \fSigma_{12}\left(\fSigma_{11}^{(k)}\fSigma_{22} + \fSigma_{11}\fSigma_{22}^{(k)}\right)}{\left(\fSigma_{11}\fSigma_{22}\right)^2},
(\#eq:gradientelement)
\end{equation}  
where the double subscript to $\fSigma$ and $\fSigma^{(k)}$ here means matrix elements. In order to avoid confusion with the double subscripts defining matrix *partitions* in the preceding sections, we re-label the matrix blocks $\R^{11}$, $\R^{12}$ and $\R^{22}$ to $\R_1$, $\R_2$ and $\R_3$ respectively, with $\R^{21} = \R_2^T$. Following basic differentiation rules for matrices (see e.g. @van2007parameter), we have that, element-wise,
\begin{equation}
\frac{\partial\left(\R_2\R_3^{-1}\R_2^T\right)}{\partial\rho_k} = \frac{\partial\R_2}{\partial\rho_k}\R_3^{-1}\R_3^T - \R_2\R_3^{-1}\frac{\partial \R_3}{\partial \rho_k}\R_3^{-1}\R_2^T + \R_2\R_3^{-1}\frac{\partial\R_2^T}{\partial\rho_k}.
(\#eq:derivative)
\end{equation}
The value of the gradient \@ref(eq:gradientelement) depends on $k$, and when differentiating (7) with respect to $\rho_k$ there are three different cases we must consider:

1. If $\rho_k = \rho_{12}$ the second term in (7) vanishes, because $\rho_{12}$ is only present in $\R_1$. Indeed, 
$$\R^{11} = \begin{pmatrix}1 & \rho_{12} \\ \rho_{12} & 1 \end{pmatrix},$$ 
which means that 
$$\fSigma^{(k)} = \begin{pmatrix}0&1\\1&0\end{pmatrix} - \begin{pmatrix}0&0\\0&0\end{pmatrix},$$ 
and we easily see from \@ref(eq:gradientelement) that in this particular case, $\left\{\nabla g(\frho) \right\}_k = 1$.
2. If $\rho_k = \rho_{ij}$ where $i \in \{1,2\}$ and $j \in \{3, \ldots,p\}$, then $\rho_{ij}$ is an element in $\R_2$, but not $\R_1$ or $\R_3$. For that reason, we are left with the second term in (7), and furthermore, the middle term in the derivative \@ref(eq:derivative) above will also vanish. The $2\times(p-2)$ matrix $\partial\R_{12}/\partial\rho_{ij}$ will contain only zeros except for a $1$ in the position of $\rho_{ij}$; and $\partial\R^{21}/\partial\rho_{ij}$ is its transposed. Write $\fC = \left(\R^{22}\right)^{-1}\R^{21}$, and we see that $\fSigma^{(k)}$ in this case is a simple linear combination of the elements in $\fC$: 
$$\fSigma^{(k)} = \frac{\partial\R^{12}}{\partial\rho_{ij}}\fC + \fC^T\frac{\partial\R^{21}}{\partial\rho_{ij}},$$ 
from which we select the elements needed to calculate $\left\{\nabla g(\frho) \right\}_k$ through \@ref(eq:gradientelement).  
3. In this case, we look at the derivative of $g(\frho)$ with respect to the local correlations located in $\R_3$, that is, $\rho_{ij}$ for $i,j \in \{3,\ldots,p\}$. We only need to keep the middle term in \@ref(eq:derivative). We have that
$$\fSigma^{(k)} = -\fC^T\frac{\partial\R_3}{\partial\rho_j}\fC,$$ 
where $\partial\R_3/\partial\rho_k$ is a symmetric $(p-2)\times(p-2)$ matrix of zeros everywhere, except for 1s in the positions corresponding to $\rho_k$ with respect to which we perform the differentiation. Finally, equation \@ref(eq:derivative) provides the final expression for $\left\{\nabla g(\frho) \right\}_k$.

# PROOF OF THEOREM 3.2 {#proofpseudo}

Theorem 3.2 states that using marginally standard normal pseudo-observations instead of exactly standard normally distributed observations for estimating the LGPC does not change the conclusions in Theorem 3.1 in the main article and Theorem 2 in @otneim2017conditional. The result may be expected, as the empirical distribution functions $\widehat{F}_j$, $j = 1,\ldots,p$ converge at a rate faster than the local correlations ($n^{-\frac{1}{2}}$ vs. $(nb^5)^{-\frac{1}{2}}$ or $(nb^2)^{-\frac{1}{2}}$) depending on which theorem we refer to), but we must attend to some details nevertheless.

Proving joint asymptotic normality of the local Gaussian correlations with marginally normally distributed variables $(Z_1, \ldots, Z_p) = \left(\Phi^{-1}(F_1(X_1)), \ldots, \Phi^{-1}(F_p(X_p))\right)$, where $F_j$ is the cdf of $X_j$ for $i = 1,\ldots,p$, relies on proving asymptotic normality for the variables
$$Y_n(\z) = \frac{1}{n}\sum_{t=1}^nK_b(\Z_t - \z)\fv(\Z_t, \frho_0),$$
where $\fv(\cdot) = \partial L(\cdot)/\partial\frho$ in the same way as in Section \@ref(app-loccor) in this supplement. This has been done by @otneim2017locally in the iid case using arguments from @schervish1995theory, and by @otneim2017conditional in the $\alpha$-mixing case, using arguments from @fan2008nonlinear. We need here to prove asymptotic normality of the variables

$$\hY_n(\z) = \frac{1}{n}\sum_{t=1}^nK_b(\hZ_t - \z)\fv(\hZ_t,\frho_0),$$
with $\hnZ_t = \Phi^{-1}(F_n(\X_t))$, and $F_n(\X_t)$ is the application of the marginal empirical distribution functions to the corresponding components of the vector $\X_t$.

In order to ease notation we state the proof in the simplified case only, where the local correlations are estimated using *pairs* $(i,j)$ of variables. It can be extended to the trivariate case by adding the appropriate terms in the derivations below, without changing the final conclusion. Write $\hY_n(\z) = Y_n(\z) - (Y_n(\z) - \hY_n(\z))$ and do a Taylor expansion of $Y_n(\z)$ around $F_n(\X_t)$. Since we use product kernels, we then need derivatives of (assuming without loss of generality that $K(x) = K(-x)$)

$$H_{\z,\hh} = K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right)K\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right)v(\Phi^{-1}(\y)),$$
where  $\y = F(\x)$, $y_i = F(x_i)$, $i=1,2$, and we write $K(x_1,x_2) = K(x_1)K(x_2)$. Writing $k(z) = K'(z)$ and $v_{y_i} = \partial v_i/\partial y_i,$
\begin{align*}
\frac{\partial H_{\bm{z},b}}{\partial y_1} 
&= k\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right) \frac{v(\Phi^{-1}(\y))}{b\phi(\Phi^{-1}(y_1))} \\
& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad +  K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right) \frac{v_1(\Phi^{-1}(\y))}{\phi(\Phi^{-1}(y_1))}, \\
\frac{\partial H_{\bm{z},b}}{\partial y_2} 
&= K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right) k\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right) \frac{v(\Phi^{-1}(\y))}{b\phi(\Phi^{-1}(y_2))} \\
& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad+  K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right) \frac{v_2(\Phi^{-1}(\y))}{\phi(\Phi^{-1}(y_1))}.
\end{align*}
A typical term in the Taylor expansion of $Y_n(\z)$ takes the following form:

\begin{align*}
& K_b\left(\Phi^{-1}(F(\X_t)) - \z\right)v\left(\Phi^{-1}(F(\X_t)), \rho_0\right) = \\
& \qquad K_b\left(\Phi^{-1}(F_n(\X_t)) - \z\right)v\left(\Phi^{-1}(F_n(\X_t)), \rho_0\right) \\
& \qquad\qquad + k\left(\frac{z_1 - \Phi^{-1}(F_n^*(X_{1,t}))}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(F_n^*(X_{2,t}))}{b}\right) \frac{v(\Phi^{-1}(F_n^*(\X_t)))}{b\phi(\Phi^{-1}(F_n^*(X_{1,t})))}\left(F(X_{1,t}) - F_n(X_{1,t})\right) \\
& \qquad\qquad + K\left(\frac{z_1 - \Phi^{-1}(F_n^*(X_{1,t}))}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(F_n^*(X_{2,t}))}{b}\right) \frac{v_1(\Phi^{-1}(F_n^*(\bm{X}_t)))}{\phi(\Phi^{-1}(F_n^*(X_{1,t})))}\left(F(X_{1,t}) - F_n(X_{1,t})\right) \\
&\qquad + \textrm{ analogous terms for the second (and possibly third) index involving } \left(F(X_{2,t}) - F_n(X_{2,t})\right), 
\end{align*}
where $F_n^*$ comes from the mean value theorem. The challenge in proving the desired result is to control the behavior of quantities on the form $1/\phi(\Phi^{-1}(F(x)))$, because this fraction tends to infinity as $|x| \rightarrow \infty$. The key to this problem is the assumption that the kernel function $K(\cdot)$, and thus its derivative $k$, have compact support. This means, that in the expressions above, $K = k \equiv 0$ if for some $M > 0$ 
$$\frac{|z_1 - \Phi^{-1}(F_n^*(X_{1,t}))|}{b} \geq M,$$
or
$$|z_1 - \Phi^{-1}(F_n^*(X_{1,t}))| \geq Mb,$$
or, by removing the absolute value signs,
$$ \Phi^{-1}(F_n^*(X_{1,t})) \leq z_1 - Mb\,\, \textrm{ or }\,\, \Phi^{-1}(F_n^*(X_{1,t})) \geq z_1 + Mb,$$
where $M$ may be large. The same reasoning applies of course to the second index as well. Letting $n\rightarrow\infty$ and using consistency of the empirical distribution function, it follows that the kernel function term is zero if

$$X_{1,t} \leq F^{-1}(\Phi(z_1 - Mb)) \,\,\, \textrm{ or } \,\,\, X_{1,t} \geq F^{-1}(\Phi(z_1 + Mb)).$$
The same reasoning applies to the derivative of the kernel function, where we consider the function $k/b$ instead of $K$. All of this implies that the challenge of controlling the magnitude of the Taylor terms above as $x\rightarrow\infty$ disappears, since, taking boundedness of the other terms into account,

$$ k\left(\frac{z_1 - \Phi^{-1}(F_n^*(X_{t,1}))}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(F_n^*(X_{t, 2}))}{b}\right) \frac{v(\Phi^{-1}(F_n^*(\X_t)))}{b\phi(\Phi^{-1}(F_n^*(X_{t,1})))}$$
and
$$K\left(\frac{z_1 - \Phi^{-1}(F_n^*(X_{t,1}))}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(F_n^*(X_{t, 2}))}{b}\right) \frac{v_1(\Phi^{-1}(F_n^*(\X_t)))}{\phi(\Phi^{-1}(F_n^*(X_{t,1})))}$$
are bounded almost surely as $n\rightarrow\infty$ and $b\rightarrow 0$.

All this means that if we can prove that for $i=1,2$,
$$\frac{1}{n}\sum_{t=1}^n(F_n(X_{i,t}) - F(X_{i,t})) \stackrel{\mathbb{P}}{\rightarrow} 0$$
at an appropriate rate, then we are done by Slutsky's Theorem.

In the following we omit the index $i=1,2$. We need a bound on the supremum of $|F_n(X_t) - F(X_t)|$ and $|F_n(x) - F(x)|$ as $n \rightarrow \infty$. According to @arcones1995law, the best law of the iterated logarithm has been obtained by @rio1995functional. One statement of this is as follows:

Let $f(x) \in L_p$, $2<p\leq\infty$, $\sum\alpha_k k^{2/(p-2)} < \infty$, then the sequence $f(X_t)$ satisfies the compact LIL (*law of the iterated logarithms*), i.e.
$$\left\{\left(n2\log\log n\right)^{-\frac{1}{2}} \sum_{t=1}^n(f(X_t) - E[f(X_t)]) \right\}_{n=1}^{\infty}$$
is relatively compact and its limit set is $[-\sigma, \sigma]$ where
$$\sigma^2 = \textrm{Var}\Big(f(X_t)\Big) + 2\sum_{s=1}^{\infty}\textrm{Cov}\Big(f(X_t), f(X_s)\Big).$$
For a fixed $x$, take $f(X_t) = I(X_t\leq x)$. Then E$(f(X_t)) = F(x)$ and $f(x)\in L_p$ for all $p$, and
$$\left\{\left(n2\log\log n\right)^{-\frac{1}{2}} \sum_{s=1}^n \left(I(X_s \leq x) - F(x) \right) \right\}$$
has its limit set in $[-\sigma, \sigma]$. Here 
$$\textrm{Var}(I(X_t \leq x)) = F(x)(1-F(x)) \leq \frac{1}{4},$$
and according to Davydov's Lemma (see e.g. @fan2008nonlinear p. 278),
$$\big| \textrm{Cov}(I(X_t \leq x), I(X_s \leq x))\big| \leq 8 \|I(X_t \leq x)\|_p \|I(X_s \leq x)\|_q \alpha(|s-t)^{1-p^{-1}-q^{-1}}, \,\,\, p,q>1, \,\,\, \frac{1}{p}+\frac{1}{q} < 1,$$
with $\| I(X_t \leq x)\|_p = \left(\textrm{E}|I(X_t \leq x)|^p\right)^{1/p}$. We may for example take $p=q=4$, leading to $\| I(X_t \leq x) \|_4 = F(x)^{1/4} \leq 1$.

Since we have assumed exponential mixing (much weaker conditions suffice), then Rio's condition $\sum \alpha_k k^{2/(p-2)} < \infty$ is satisfied, and the bounds of $-\sigma$ and $\sigma$ can be made independent of $x$.

Having 
$$F_n(x) = \frac{1}{n}\sum_{t=1}^n I(X_t \leq x)$$
means that
$$n^{\frac{1}{2}}[2\log \log n]^{-\frac{1}{2}} (F_n(x) - F(x))$$
has its limits in $[-\sigma, \sigma]$ or that the supremum of $|F_n(x) - F(x)|$ is almost surely bounded by $n^{-1/2}(2\log\log n)^{1/2}$.

This immediately implies that
$$\frac{1}{n}\sum_{t=1}^n\left(F_n(X_t) - F(X_t)\right)$$
is almost surely (and consequentially in probability) bounded by $n^{-\frac{1}{2}}(2 \log \log n)^{\frac{1}{2}}$ which for all conceivable bandwidths  is faster than the convergence rates of the CLT for $Y_n(\z)$, which is $(nb^2)^{-\frac{1}{2}}$.

# THE VALIDITY OF THE BOOTSTRAP FOR ESTIMATION {#validity-bootstrap}

The asymptotic theory for the LGPC involves complicated expressions that may be difficult to implement in practice. It is then natural to use the bootstrap instead to construct confidence intervals. In order to accommodate the time series case, we must use the block bootstrap, and we are thus faced with the problem of proving the validity of the block bootstrap.

We have seen that the building blocks of the asymptotic estimation theory of the LGPC between the vectors $\Z_1$ and $\Z_2$ given $\Z_3$ are the pairwise local Gaussian correlations contained in the local correlation matrix $\fSigma$. Let us simply use the notation $\rho_{ij} = \rho_{ij}(z_i, z_j)$ for one of these local correlation and $\hrho_{ij}$ for its estimate as outlined in the main article. If we can prove that the block bootstrap is valid for such a pairwise local likelihood estimate, then it follows fairly straightforwardly that it is also valid for the appropriate vector of such pairwise estimates used in the asymptotic theory laid out above. Note that the 3-coordinate case (i) in Section \@ref(chap:estimation) requires essentially the same arguments.

We can use the bootstrap theory by @lacal2018estimating more or less directly, which in turn is based on Gon\c{c}alves and White [-@gonccalves2002bootstrap; -@gonccalves2004maximum], but we spell out some details  below. First, we need to specify the analogue of the conditions A1--A5 by @lacal2018estimating, where we adapt their notation to our situation by replacing the parameter $\theta$ with $\rho_{ij}$, and the time series $\{X_t, Y_t\}$ by our multivariate observations $\{\Z_t\}$, possibly a multivariate time series. Since $\rho = \rho_{ij}$ is estimated using the two corresponding data vectors only, we can, for the time being and without loss of generality, assume that the observation series is bivariate: ${\Z_t} = \{Z_{i,t}, Z_{j,t}\}$. Note further that it suffices to consider the true pseudo-observations $\{\Z_t\}$ instead of the empirical pseudo observations $\{\hZ_t\}$ in the following because the difference between them is asymptotically negligible according to Theorem 3.2 in the main article.

To state the bootstrap results we need a few more concepts and three additional conditions. We block bootstrap the entire vector series $\{\Z_t\}$. This automatically leads to the same block-bootstrap for any pair of components which we have denoted by $\{Z_{i,t}, Z_{j,t}\}$ above. @lacal2018estimating use both the block bootstrap as well as the stationary bootstrap, and the results below hold irrespective of this particular choice. We will therefore not distinguish between them here, but rather refer the reader to @lacal2018estimating for further details on this distinction. Bootstrapped quantities will be denoted with a star ($*$) in the following.

Before we introduce the specific bootstrap conditions **A4** and **A5** below, we need to state the following condition. The notation of equations \@ref(eq:loglikelihood)--\@ref(eq:A16) is used.

\begin{itemize}
\item[\textbf{A3}] For a fixed bandwidth $b$, and using stationarity,
$$J_{n,b} = \E\left(-\nabla^2L_{n,b}(\rho_{0,b})\right) = J_b = \E\left(-\nabla^2 L^{(b)}(Z_{i,t}, Z_{j,t};\rho_{0,b})\right)$$
is of order $O(1)$, and continuous on $\Theta$ uniformly in $n$. Further,
$$M_{n,b} = \Var\left(\sqrt{nb_1b_2}\nabla L_{n,b}(\rho_0,b)\right)$$
is $O(1)$ and uniformly positive. Here $\nabla$ is the gradient with respect to $\rho$, which in the present scalar parameter case is simply the differential operator $\di/\di\rho$.
\end{itemize}

The bootstrap lives on the probability space $(\Lambda, \mathcal{G}, \mathbb{P}^*)$. The outcomes $\lambda \in \Lambda$ depend on the outcomes $\omega \in \Omega$ in the original probability space, and we can now state two more assumptions:

\begin{itemize}
\item[\textbf{A4}] The bootstrap probability space $(\Lambda, \mathcal{G}, \mathbb{P}^*)$ is a complete probability space for all $\omega \in \Omega$. Let $L_{n,b}$ be defined as in \eqref{eq:loglikelihood} with $\rho \in \Theta$. Let $\{L^*_{n,b}: \Lambda\times\Omega\times\Theta \rightarrow \overline{\mathbb{R}}\}$ be a sequence of random functions such that $L_{n,b}^*(\rho) = L_{n,b}((\Z_{in}^{*}(\lambda,\omega),\Z_{jn}^*(\lambda,\omega)),\rho)$ where $Z_{i,t}^*(\lambda,\omega) = Z_{i\tau_t^{Z}(\lambda)}(\omega)$, $Z_{j,t}^*(\lambda,\omega)= Z_{j\tau_t^{Z}(\lambda)}(\omega)$, where $\tau_t^{Z}: \Lambda \to \mathbb{N},\; \omega \in \Omega, \; \lambda \in \Lambda$ are vectors of random indices representing the block bootstrap operation on $\{\Z_t\}$. Moreover, the block length $l_n$ is such that $l_n=o(\sqrt{n})$ as $n \to \infty$. 
\item[\textbf{A5}] For $\rho \in \Theta$ and for every $t=1,\ldots,n$, $\{L^{(b)}(Z_{i,t},Z_{j,t}),\rho\}$ is Lipschitz continuous on $\Theta$, $\{\nabla L^{(b)}((Z_{i,t},Z_{j,t}),\rho)\}$ is 6-dominated  on $\Theta$ uniformly in $t,n$, and $\{\nabla^2 L^{(b)}((Z_{i,t},Z_{j,t}),\rho)\}$ is Lipschitz continuous on $\Theta$ and 2-dominated on $\Theta$ uniformly in $t,n$ ($r$-dominated is defined after \eqref{eq:A16}).
\end{itemize}

Assumption A4 sets the stage for the block bootstrapping, whereas A5 is needed in proving the validity of the bootstrap central limit theorem. The assumption A4 is not restrictive, and again we refer to @lacal2018estimating for more discussion.

In order to state the bootstrap validity theorems, we need to specify convergence in the space $(\Lambda, {\mathcal G}, \mathbb{P}_{\omega}^*)$ . This is done as by Gon\c{c}alves and White [-@gonccalves2002bootstrap; -@gonccalves2004maximum]. First, for a random variable $Y_n^*$, we write $Y_n^* \stackrel {\mathbb{P}_{\omega}^*,\mathbb{P}}{\rightarrow} 0$ if for any $\varepsilon, \delta > 0$,
$$
\lim_{n \to \infty} \mathbb{P}(\omega:\mathbb{P}_{\omega}^*(\lambda:|Y_n^*(\lambda,\omega)| > \varepsilon) > \delta) = 0.
$$ 
Further, we write $Y_n^* \stackrel {d_{\mathbb{P}_{\omega}^*}}{\longrightarrow} {\mathcal N}(0,1)\;\mbox{prob}-\mathbb{P}$, if for every sub-sequence $\{n^{\prime}\}$, there exists a further sub-sequence $\{n^{\prime \prime}\}$ such that $Y_{n^{\prime \prime}} \stackrel {d_{\mathbb{P}_{\omega}^{*}}} {\longrightarrow} {\cal N}(0,1)$ almost surely , see Gon\c{c}alves and White (2004, page 210). This definition is based on the fact that convergence in probability implies almost sure convergence for such kinds of sub-sequences (see Theorem 20.5 of @bill:2008).

We proceed by stating the validity of the bootstrap in terms of consistency and a central limit theorem:

\begin{thm}
Let assumptions A1, A2 and A4 hold, and  let $\{L^{(b)}((Z_{i,t},Z_{j,t}),\rho)\}$, $\rho \in \Theta$ be Lipschitz continuous and 4-dominated on $\Theta$ uniformly in $t,n$. Moreover, assume that, as $b=b_n \to 0, \sigma_b^2 \stackrel{\textrm{def}}{=} \Var(L^{(b)}((Z_{i,t},Z_{j,t}),\rho) \to \sigma^2 < \infty$ and $\mu_b \stackrel{\textrm{def}}{=} \E(L^{(b)}(Z_{i,t},Z_{j,t}),\rho)) \to \mu < \infty$, for all $t=1,\ldots,n$ and $\rho \in \Theta$. 
Then,
\begin{equation}
\rho_{n,b}^* - \rho_{b,n} \stackrel {\mathbb{P}_{\omega}^{*},\mathbb{P}}{\to} 0.
\label{eq:thmB}
\end{equation}
\label{thm:E1}
\end{thm}

\begin{thm}
Under the conditions 1--5 of Theorems 3.1 and 3.2 in the  main article and assumptions A1--A5 stated above,

\begin{equation}
\sqrt{nb_1b_2} M_{n,b}^{-1/2}J_{n,b}(\rho_{n,b}^{*} - \rho_{n,b}) \stackrel {d_{\mathbb{P}_{\omega}^{*}}}{\to} {\cal N}(0,1)\mbox{prob}-\mathbb{P}.
\end{equation}
\label{thm:E2}
\end{thm}

Theorems \@ref(thm:E1) and \@ref(thm:E2) are formulated in their simplest case here, which concerns only one local parameter. A more complicated version is given by @lacal2018estimating concerning the five parameter case. The proofs are essentially the same and are spelled out in some detail in the supplementary material of that paper as Theorems D1 and D2.

Theorems \@ref(thm:E1) and \@ref(thm:E2) concern only one local correlation. The local *partial* correlation, however, is a function of a vector of local correlations. The form of this function and the vector of local correlations depend on the generality and the type of model considered as seen from the alternatives (i) -- (v) in Section \@ref(chap:estimation) of this supplement. Corollaries \@ref(cor:E3) and \@ref(cor:E4) cover the general situation with pairwise model and pairwise coordinates, including alternative (i)', which is alternative (i) but restricted to pairwise coordinates. Corollary \@ref(cor:E3) treats bootstrap consistency and Corollary \@ref(cor:E4) gives a bootstrap CLT:

\begin{cor}
Assume the conditions of Theorem \ref{thm:E1}. Let $\alpha_{n,b}(i,j|\Z_3) \stackrel{\textrm{def}}{=} \widehat \rho(Z_{1i}, Z_{2j}|\Z_3)$ be the $i,j$th element of the estimated partial correlation matrix of $(\Z_1, \Z_2)$ given $\Z_3$. This depends on scalar coordinates $(z_1, z_2)$ and a vector coordinate $\z_3$. Note that $\alpha_{n,b}(i,j|\Z_3)$ is a continuous function of a vector of local correlations according to one of the alternatives (i)', (ii)--(v) in Section \ref{chap:estimation}. Then $\alpha_{n,b}(i,j|\Z_3) - \alpha^*_{n,b}(i,j|\Z_3) \buildrel {\mathbb{P}_{\omega}^{*},\mathbb{P}} \over {\longrightarrow} 0$. 
\label{cor:E3}
\end{cor}

*Proof.* Denote by $\frho = \frho(\Z)$ the vector of local correlations according to one of the alternatives (i)', (ii)--(v). This vector has a finite number of components and equation \@ref(eq:thmB) guarantees the bootstrap consistency of each component. The bootstrap consistency for the vector follows immediately and the proof is concluded by the continuous mapping theorem. $\blacksquare$

The next results gives the asymptotic normality of the bootstrapped local partial correlation. We let $\alpha(i,j)|\Z_3) = g_{ij}(\frho(\Z)))$ be the local partial correlation as defined above and expressed as a function $g_{ij}$ of the vector of local correlations according to one of the alternatives (i)', (ii)--(v). Using the notation of Section \@ref(chap:estimation) of this supplement, the correlation vector $\frho(\Z)$ will in general have elements consisting of $\rho_{ij}^{1,2}, i =1,\ldots,d_1, j=1,\ldots,d_2;\;\rho_{ij}^{1,3}, i=1,\ldots,d_1, j=1,\ldots,d_3;\;\rho_{ij}^{2,3}, i=1,\ldots,d_2, j=1,\ldots,d_3$, where $d_1,d_2,d_3$ are chosen according to one of the alternatives (i)', (ii)--(v).

\begin{cor}
Let the assumptions of Theorem \ref{thm:E2} hold and let $\fOmega = \fOmega(\frho)$ be the covariance matrix of the $\widehat{\frho}$ vector. Then,
$$
\sqrt{nb_1b_2}(\alpha_{n,b}^*(i,j)-\alpha_{n,b}(i,j)) \buildrel {d_{\mathbb{P}_{\omega}^{*}}} \over {\longrightarrow} {\cal N}(0,\nabla g(\rho)^T \fOmega(\frho)) \nabla g(\rho)) \,\, \mbox{prob}-\mathbb{P}. 
$$ 
\label{cor:E4}
\end{cor}
*Proof:* The proof of Theorem \ref{thm:E2} is given as the proof of Theorem D2 in the online appendix accompanying @lacal2018estimating. This proof can be adapted to prove Corollary \ref{cor:E4}. The starting point consists in replacing $L_{n,b}^*$ and $L_{n,b}$ by $L^* = \sum_{i,j} a_{ij}L_{ij,n,b}^*$ and $L = \sum_{i,j}a_{ij}L_{ij,n,b}$, respectively, where the sum runs over all possible pairs $(i,j)$ corresponding to $\rho_{ij}^{1,2}$, $\rho_{ij}^{1,3}$, $\rho_{ij}^{2,3}$ that are included in the $\frho$-vector. The derivations in  the proof of Theorem D2 by @lacal2018estimating deal with the more complicated five-parameter case. In the present case, however, the vector $\frho$ consists of correlations $\rho_{ij}$ that have been estimated using the one-parameter pairwise setup, where, cf. @otneim2017locally, the quantities $J_{ij}$ and $M_{ij}$ are scalars. Here $L_{ij,n.b}$ is the log likelihood for the pair $(i,j)$ treated in the proof of Theorem D2, and $a_{ij}$ is a corresponding sequence of real numbers. Because of the linearity of the expressions for $L^*$ and $L$ it is easily checked by going step by step that the proof of Theorem D2 of @lacal2018estimating can be carried through for $L^*$ and $L$. The only difference is that the quantities $J_{n,b}$ and $M_{n,b}$ now have to be replaced by $J_{n.b}(a) = \E(-\sum_{i,j}a_{ij}\nabla^2 L_{ij,n,b}(\rho_{0,b}))$ and $\Var(\sqrt{nb_1b_2}\sum_{i,j} a_{ij}\nabla L_{ij,n,b})$. This results in joint asymptotic bootstrap normality using the Cramer-Wold device. Applying the delta rule now proves the Corollary. $\blacksquare$

Let $\M_{ij}$ and $\J_{ij}$ be the $M$-matrix and $J$-matrix appearing in the proof of Corollary \ref{cor:E4} for a fixed $\rho_{ij}$. Note that in the one-parameter case of $\rho_{ij}$ in Corollary \ref{cor:E4}, to first order the covariance matrix $\fOmega$ is a diagonal matrix having ordered values of $M_{ij}/J_{ij}^2$ along the diagonal. This can be seen using the methods of proofs of Theorem 2 of @otneim2017locally and Theorem 2 @otneim2017conditional. Note also that the 3-coordinate case of alternative (i) can be stated and proved in the same way, but a different scaling factor is required to avoid a singular covariance matrix as seen in the proof of Theorem 3.1 in the main article.

# ASYMPTOTIC THEORY FOR THE TEST STATISTIC

Consider the general case $\Z = (\Z_1, \Z_2, \Z_3)$ where each component is $d_i$-dimensional correspondingly, with $i = 1,2,3$, and $\Z$, naturally, is $d_1+d_2+d_3$-dimensional, and consider also the local partial correlation matrix of $(\Z_1,\Z_2|\Z_3)$. In alternative (iii) in Section \@ref(chap:estimation) this matrix is given by equation \@ref(eq:offdiag3) with
$$
\alpha_{ij} = \rho(Z_{1,i},Z_{2,j}|Z_3), \;\; i=1,\ldots,d_1; j = 1,\dots,d_2.
$$
This forms a $d_1 \times d_2$ matrix ${\falpha}$. One can test for conditional independence between $Z_{1,i}$ and $Z_{2,j}$ given $\Z_3$ by testing whether $\alpha_{ij}=0$, or one can test for conditional independence between $\Z_1$ and $\Z_2$ given $\Z_3$ by testing whether ${\falpha} = \{\alpha_{ij}\} = {\bm{0}}$. The latter case is in a sense the most general, and we will stick to that in the sequel. One can then test for ${\falpha} = {\bm{0}}$ by using a matrix norm, for instance the Frobenius norm, where $||{\falpha}|| = \sqrt{\sum_{i,j} \alpha_{ij}^2} = 0$. A more convenient test statistic is $\sum_{i,j} \alpha_{ij}^2$. To follow the notation of @lacal2018estimating we write, letting $\z=(\z_1,\z_2,\z_3)$,
$$
T = \int_S h(\frho(\z)) \di F(\z)
$$
where $S$ is an a area of integration, $F$ is the cumulative distribution function of $\Z$, and $\frho = \frho(\z)$, as before, is a vector consisting of all possible pairs that can be formed by picking pairwise components from $\Z$. The number of components of this vector depends on which of the alternatives models (i)--(v) in Section \@ref(chap:estimation) one considers. It is reasonable to express $T$ as a function of $\frho$ since all partial correlations can be expressed as a function of this vector (see e.g. again equation \@ref(eq:offdiag3)). One possible choice of the function $h$ is $h(\frho(\z)) = \sum_{i,j} \alpha_{i,j}^2(z_{1,i},z_{2,j}|\ztwo)$, which is a function of $\frho$ since the $\alpha$-s are functions of $\frho$. The corresponding function in @lacal2018estimating is $h(\theta(x))$, where $\theta(x)$ is at most 5-dimensional and $\x$ is bivariate. As in that paper we proceed in two stages. First we let the bandwidth vector $\bb$ be fixed and let $n \to \infty$, and then we let $\bb = (b_1,b_2) = (b,b) \to \bm{0}$.
Accordingly we define
$$
T_b = \int_S h(\frho_b(\z)) \di F(\z) 
$$
where $\frho_b$ is the population quantity for a fixed bandwidth $\bb$, and $T_{n,b}$ by
$$
T_{n,b} = \int_S h(\frho_{b,n}(\z)) \di F_n(\z),
$$
where $\frho_{b,n}(\z)$ is the estimate of $\frho(\z)$ obtained by pairwise likelihood for each component of $\frho(\z)$.

Consistency is straightforward to prove:

\begin{thm}
Let the conditions of Theorem \ref{thm:A1} be fulfilled. Then $T_{n,b} \buildrel {\mathbb{P}} \over {\to} T_b$ as $n \to \infty$ and $T_{n,b} \buildrel {\mathbb{P}} \over {\to} T$ as $n \to \infty$ and $\bb \to \bm{0}$.
\label{thm:F1}
\end{thm}

*Proof.* This result follows directly from Theorem \ref{thm:A1} and the continuous mapping theorem. $\blacksquare$

It is more difficult to obtain the limiting distribution. We will follow the technique in the proof of Theorem 3.1 in @lacal2018estimating. See also @joe1989estimation. The details are a bit more involved in the present case because of the larger dimension and because we need to involve the pairwise log likelihood in the procedure.

The results on asymptotic distribution can be proved on several levels. We consider two cases: a) the pairwise simplification where involving the estimates $\rho_{ij,n,b}(z_i,z_j)$ for all possible pairs $(i,j)$ of the $\frho$ vector, and b) the case where $d_1=d_2=d_3=1$, but where $\rho_{ij} = \rho_{ij}(\z)$ is allowed to depend on all three coordinates $\z=(z_1,z_2,z_3)$. We will put the emphasis on a) and only comment briefly on b).

\begin{thm} 
Let the assumptions A, D--F of Theorem 3 of Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}) and the conditions of Theorem 3.2 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) (which includes conditions A1 and A2 of this supplement) be fulfilled. Then,
as $n \to \infty$,
$$
\sqrt{n}\big(C_{n}(A_b)\big)^{-\frac{1}{2}}(T_{n,b} - T_b) \buildrel {d} \over {\to} {\cal N}(0,1),
$$
where
\begin{align*}
A_b(\z)  & = \int_{S} \left[\sum_{i,j}\left\{\frac{\partial}{\partial \rho_{ij}} h(\frho_b,(\fv)\right\}J_{ij}^{-1}(\rho_{ij,b}(\fv_{ij}))K_b(\z_{ij}- \fv_{ij})u_{ij}(\z_{ij},\rho_{ij,b}(\fv_{ij}))\right] \di F(\fv) \\
& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad + h(\frho_b(\z))1_S(\z)
\end{align*}
and 
\begin{align}
C_n(A_b) &= \int A_b^2(\z) \di F(\z) + \frac{1}{n}\sum_{s \neq t}\int A_b(\z)A_b(\fu) \di F^{s,t}(\z,\fu) \nonumber \\
& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad - n\int A_b(\z)A_b(\fu) \di F(\z) \di F(\fu)
(\#eq:t0)
\end{align}
with $F^{s,t}(\z,\fu)$ being the joint cumulative distribution function of $(\Z_s,\Z_t)$, and where $\{\Z_{t}, t=1,\ldots,n\}$ are the $n$ observations of $\Z$ and $u_{ij}(\Z_{ij}, \rho_{ij,b}(\fv_{ij})) = \frac{\partial}{\partial \rho_{ij,b}}\log\psi(\Z_{ij},\rho_{ij,b}(\fv_{ij}))$.

To ease comparison with Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) we have written and will write $\z_{ij} \stackrel{\textrm{def}}{=} (z_i,z_j)$ and $\fv_{ij} \stackrel{\textrm{def}}{=} (v_i,v_j)$. Further, if $n \to \infty$ and $\bb_n \to \bm{0}$, the relationship between them being as in Assumption 3 of Theorem 3.1 in the main article, then $T_b$ can be replaced by $T$,
$$
\sqrt{n}\big(C_{n}(A_b)\big)^{-\frac{1}{2}}(T_{n,b} - T) \buildrel {d} \over {\to} {\cal N}(0,1).
$$
\label{thm:F2}
\end{thm}
Note that the sum in the second term in the definition of $\fC_n(\fA_b)$ taken together with the third term converges due to the geometric $\alpha$-mixing condition of A1. It is also important to note that conditional independence is not assumed in Theorem \@ref(thm:F2). This will be exploited in Section \@ref(chap:pitman). See also Theorem 3.2 in @lacal2018estimating and its relation to Corollary 3.1 in that paper.

*Proof.* From the proof of Theorem 3.1 in @lacal2017local, one has that asymptotically,
\begin{equation}
T_{n,b}-T_b \sim \frac{1}{\sqrt{n}} \left\{\int_S h(\frho_b(\z)) \di G_n(\z) + \int_S\nabla_{\frho} h(\frho_b(\z)^{T} \fV_n(\z) \di F(\z)\right\}. 
(\#eq:t1)
\end{equation}
Here 
$$
G_n(\z) = \sqrt{n}(F_n(\z)-F(\z)) \quad \mbox{and} \quad \fV_n(\z) = \sqrt{n}((\frho_{n,b}(\z) -\frho_b(\z)),
$$ 
with $F_n$ being the empirical cumulative function of $\Z$, and where, again, $\frho$ is the vector consisting of all possible pairs $\rho_{ij}$. This rewriting of $T_{n,b}-T_b$ is the first step in obtaining an expression in terms over $\Z_{t}$, $t=1,\dots,n$ so that a mixing CLT can be used. We proceed to the pairwise analysis.

Let $(i,j)$ be a fixed pair and $\rho_{ij,n,b}(\z_{ij})$ be the corresponding estimated local correlation for the pair $(i,j)$ chosen from the vector $\frho_{n,b}$. Here the pair of coordinates $\z_{ij} = (z_i,z_j)$ corresponds to the bivariate $\x$ in the proof of Theorem 3.1 by @lacal2017local. We will also omit some indices or parts of indices when it is clear what is meant. Finally, we omit the area of integration $S$. We are using the pairwise local log likelihood given by
$$
L_{ij,n} = \frac{1}{n} \sum_{t=1}^{n} K_b(\Z_{ij,t}-\z_{ij}) \log \psi\left(\Z_{ij}^{k},\rho_{b,ij}(\z_{ij})\right) - \int K_b(\fv_{ij}-\z_{ij}) \psi_{ij}(\fv_{ij},\rho_{b,ij}(\z_{ij})) \di \fv_{ij},
$$
where $K_b = b_1^{-1}K_1(\cdot/b_1) b_2^{-1} K_2(\cdot/b_2)$ is a product kernel function, and $\Z_{ij,t} = (Z_{i,t},Z_{j,t})$ is the $t$-th pair of observations, and $\fv_{ij} = (v_i,v_j)$. Estimates $\rho_{ij,b,n}(\z_{ij})$ are obtained by differentiation of the log likelihood and by introducing the score function $U_{ij}$ and information function $I_{ij}$ as follows
\begin{align*}
U_{ij}(\z_{ij},\rho_{ij}) & = \frac{\partial}{\partial \rho_{ij}}L_{ij,n}(\z_{ij},\rho_{ij}) \\
& = \frac{1}{n}\sum_{t=1}^{n}K_b(\Z_{ij,t}-\z_{ij})u_{ij}(\Z_{ij,t},\rho_{ij}) - 
\int K_b(\fv_{ij}-\z_{ij})u(\fv_{ij},\rho_{ij})\psi(\fv_{ij},\rho_{ij}) \di \fv_{ij}
\end{align*}
where $u_{ij}(\z_{ij},\rho_{ij}) = \partial/\partial \rho_{ij} \log \psi(\z_{ij},\rho_{ij})$. Similarly
$$
I_{ij}(\z_{ij},\rho_{ij}) = \frac{\partial^2}{\partial \rho_{ij}^2} L_{ij,n}(\z_{ij}, \rho_{ij}) \quad \mbox{and} \quad J_{ij,b}(\z_{ij}) = \E(-I_{ij}(\z_{ij},\rho_{ij})),
$$
where Lacal and Tjøstheim [-@lacal2017local; -@lacal2018estimating] may be consulted for more details in the derivation and a more explicit expression for $I_{ij}(\z_{ij},\rho_{ij})$ for the case considered there. In contradistinction to the 5-parameter case considered by @lacal2018estimating, in our present 1-parameter case $J_{b,ij}$ is not singular as $\bb \to \bm{0}$. In fact, it is shown by Otneim and Tjøstheim [-@otneim2017conditional, Theorem 2] that as $\bb \to \bm{0}$, $J_{b,ij} \to u(\z_{ij},\rho_{ij})\psi(\z_{ij},\rho(\z_{ij}))$.

The vector $\fV_n(\z) = \sqrt{n}[\frho_{n,b}(\z)-\frho_b(\z)]$ has the dimension of $\frho(\z)$ and consists of components $\{\rho_{ij,n,b}(\z_{ij})-\rho_{ij,b}(\z_{ij})\}$ where $i$ and $j$ in the most general case run over all possible pairs formed by the components of $\frho(\z)$. To be able to use the mixing central limit theorem we follow @lacal2018estimating and express this more explicitly in terms of the variables $\Z_{t}$, $t=1,\ldots,n$. By Taylor expansion as in @lacal2018estimating,
\begin{align*}
0 &= U_{ij}(\z_{ij},\rho_{ij,n,b}(\z_{ij})) \sim  U_{ij}(\z_{ij},\frho_{ij,b}(\z_{ij})) + I_{ij}(\z_{ij},\rho_{ij,b}(\z_{ij}))(\rho_{ij,n,b}(\z_{ij})-\rho_{ij,b}(\z_{ij})) \\
& \sim U_{ij}(\z_{ij},\rho_{ij,b}(\z_{ij})) - J_{ij,b}(\z_{ij})(\rho_{ij,n,b}(\z_{ij})-\rho_{ij,b}(\z_{ij})),
\end{align*}
which implies

\begin{align}
\rho_{ij,n,b}(\z_{ij}) - \rho_{ij,b}(\z_{ij}) & \sim J_{ij,b}^{-1}(\z_{ij})U_{ij}(\Z_{ij},\rho_{ij,b}(\z_{ij})) \nonumber \\
& = J_{ij,b}^{-1}(\z_{ij})\Big[\frac{1}{n}\sum_{t=1}^{n} K_b(\Z_{ij,t} - \z_{ij})u_{ij}(\Z_{ij,t},\rho_{ij,b}(\z_{ij})) \nonumber \\
& \qquad\qquad - \int K_b(\fv_{ij} - \z_{ij})u_{ij}(\fv_{ij},\rho_{ij,b}(\z_{ij}))\psi(\fv_{ij}, \rho_{ij,b}(\z_{ij})) \di \fv_{ij}\Big].
(\#eq:t2)
\end{align}
The second term in \@ref(eq:t1) can then be expressed as 
\begin{align}
& \int_S \sum_{i,j} \frac{\partial}{\partial \rho_{ij}} h(\rho_{ij,b}(z)) \nonumber \\
& \qquad\qquad \times J_{ij,b}^{-1}(\z_{ij}) \Big[\frac{1}{n}\sum_{t=1}^{n} K_b(\Z_{ij,t}-\z_{ij})u_{ij}(\Z_{ij,t},\rho_{ij,b}(\z_{ij})) \nonumber \\
& \qquad\qquad\qquad\qquad- \int K_b(\fv_{ij}-\z_{ij})u_{ij}(\fv_{ij},\rho_{ij,b}(\z_{ij}))\psi(\fv_{ij},\rho_{ij,b}(\z_{ij})) \di \fv_{ij}\Big] \di F(\z).
(\#eq:t3)
\end{align}
where the sum over $(i,j)$ extends over all possible $\frho_{ij}$ contained in the correlation vector $\frho$. This should be put together with the first term of \@ref(eq:t1);
\begin{equation}
\frac{1}{\sqrt{n}} \int_S h(\frho_b(\z)) \di G_n(\z) = \frac{1}{n}\sum_{t=1}^{n}h\left(\frho_b(\Z_{t})\right) - \int h\left(\frho_b(\z)\right) \di F(\z).
(\#eq:t4)
\end{equation}
We have now reached an expression where one can easily apply the mixing central limit theorem, and an application of that theorem to the sum in \@ref(eq:t3) and \@ref(eq:t4) now yields asymptotic normality of $\sqrt{n}\left(T_{n,b}-T_{b}\right)$.

Using the same argument as in @lacal2018estimating separately on the two main terms yields zero means. Moreover, using the definitions of $G_n$ and $A_b(z)$ it can be seen from \@ref(eq:t3) and \@ref(eq:t4) that asymptotically
$$
\sqrt{n}\left(T_{n,b} - T_b\right) \sim \int_S A_b(\z) \di G_n(\z).
$$

By analogy with the proof of Theorem 3.1 in @lacal2017local it is seen that
\begin{align*}
\Var\left(\int A_b(\z) \di G_n(\z)\right) &\sim \int A_b^2(\z) \di F(\z) + \frac{1}{n}\sum_{s \neq t} \int A_b(\z)A_b(\fu) \di F^{s,t}(\z,\fu) \\ 
& - n\int A_b(\z)A_b(\fu) \di F(\z)\di F(\fu) \\
& = C_n(A_b),
\end{align*}
which is the expression \@ref(eq:t0) stated in the theorem. The sum in the second term taken together with the last term above converges due to the mixing condition. Details of this argument are identical to those of the last part of the proof of Theorem 3.1 in @lacal2018estimating, as is the case of $\bb_n \rightarrow \bm{0}$, completing the proof by referring to Proposition 6.3.9 of @brockwell1991time. $\blacksquare$

It should be carefully noted that it is not entirely obvious what the order of the term $C_n(A_b)$ is as $\bb = (b_i,b_j) = (b,b) \to \bm{0}$ and $n \to \infty$. Looking at the expressions \@ref(eq:t3) and \@ref(eq:t4), it is the term \@ref(eq:t3) that causes potential trouble due to the presence of $b_i^{-1}$ and $b_j^{-1}$ in the two kernel function terms in the product kernel. These are cancelled out by the integration over $(z_i,z_j)$, however. This means that the true convergence rate for $T_{n,b}-T$ is $n^{-1/2}$. This may be somewhat surprising since the convergence rate of $\frho_{n,b}(\z_{ij})-\frho(\z_{ij})$ is $(nb_1b_2)^{-1/2}$, but this is a well-known phenomenon for functionals of type $T_{n,b}$. In @lacal2018estimating the convergence rate is changed from $(nb_1^3b_2^3)^{-1/2}$ to $(nb_1^2b_2^2)^{-1/2}$. These results are consistent with similar results for test functionals for independence in @joe1989estimation.

The variance in \@ref(eq:t4) can be evaluated in more detail, as has been done by @lacal2018estimating, in particular using the mixing inequality and Hall and Heyde [-@hall1980martingale, Corollary A2]. But as shown by @terasvirta2010modelling and in numerous other papers, the asymptotic variance in the distribution of such functionals as $T_{n,b}$ cannot be expected to be accurate unless one has extremely large sample sizes.

The second alternative mentioned prior to stating Theorem \@ref(thm:F2) is the alternative with $d_1=d_2=d_3=1$, but with correlations depending on three coordinates so that $\z=(z_1,z_2,z_3)$. This is easier in the sense that there are only three pairs of correlations, $\rho_{12}, \rho_{13}, \rho_{23}$ and in the above notation $\frho(\z) = (\rho_{12}(\z), \rho_{13}(\z), \rho_{23}(\z))$. The partial correlation is 
$$
\alpha_{12|3}(z) = \frac{\rho_{12}(\z)-\rho_{13}(\z)\rho_{23}(\z)}{\sqrt{1-\rho_{13}^2(\z)}\sqrt{1-\rho_{23}^2(\z)}},
$$
which means that a natural choice of the function $h(\frho(\z))$ might be $h(\frho(\z)) = \alpha_{12|3}^2(\z)$ This is a three-parameter function of the parameters $\rho_{12}, \rho_{13},\rho_{23}$, and an explicit expression for the gradient $\nabla_{\frho}h(\frho)$ can easily be found. More generally again we define 
$$
T_{n,b} = \int_S h(\frho_{b,n}(\z)) \di F_n(\z),\;\; T_b= \int_S h(\frho_b(\z)) \di F(\z) \, \textrm{ and } \, T = \int_S h(\frho(\z)) \di F(\z).
$$

\begin{thm}
Let the assumptions of Theorem \ref{thm:F2} be fulfilled and with a 3-coordinate $\rho$ as defined above. Then,
as $n \to \infty$,
$$
\sqrt{n}\left(C_{n}(A_b)\right)^{-\frac{1}{2}}(T_{n,b} - T_b) \buildrel {d} \over {\to} {\cal N}(0,1).
$$
Here 
$$
A_b(\z) = \int_S \nabla_{\frho}h(\frho_b(\fv))^{T}{\J}^{-1}(\frho_b(\fv))K_b(\z-\fv)\fu(\z,\frho_b(\fv)) \di F(\fv) + h(\frho_b(\z))1_S(\z)
$$
and 
$$
C_n(A_b) = \int A_b^2(\z) \di F(\z) + \frac{1}{n}\sum_{s \neq t}\int A_b(\z)A_b(\fu) \di F^{s,t}(\z,\fu) - n\int A_b(\z)A_b(\fu) \di F(\z) \di F(\fu)
$$
with $F^{s,t}(\z,\fu)$ being the joint cumulative distribution function of $(\Z^s,Z^t)$. Further, if $n \to \infty$ and $\bb_n \to \bm{0}$, the relationship between them being as in Assumption 3 of Theorem 3.1 in the main article, then $T_b$ can be replaced by $T$ such that
$$
\sqrt{n}\left(C_n(A_b)\right)^{-\frac{1}{2}}(T_{n,b} - T) \buildrel {d} \over {\to} {\cal N}(0,1).
$$
\label{thm:F3}
\end{thm}

*Proof.* The proof follows along the lines of the proof of Theorem \ref{thm:F2}. Corresponding to equation \@ref(eq:t1),
$$
T(\frho_{b,n}) - T(\frho_{b}) \sim \frac{1}{\sqrt{n}}\left\{\int_S h(\frho_b(\z)) \di G_n(\z) + \int_S \nabla_{\frho} h(\frho_b(\z)))^T \fV_n(\z)\right\} \di F(\z)
$$
where $G_n(z) = \sqrt{n}(F_n(z)-F(z))$ and
\begin{align*}
\fV_n(z) &= \sqrt{n} (\frho_{n,b}(\z) - \frho_b(\z)) \\
& \sim \sqrt{n}{\J}_b^{-1}(\z)\left\{\frac{1}{n} \sum_{t=1}^{n}K_b(\Z^t-\z)\fu(\Z^t,\frho_b(\z)) - \int K_b(\fv-\z)\fu(\fv,\frho_b(\z))\psi(\fv,\rho_b(\z)) \di \fv \right\}.
\end{align*}
Here $K_b$ consists of a triple product of kernels $\prod_{i=1}^{3}\frac{1}{b_i}K(\cdot/b_i)$, $\fu(\z,\frho_b(\z)) = \nabla_{\frho}\log \psi(\z,\frho_b(\z))$ where $\psi$ is the density function of a zero mean trivariate normal distribution having a covariance matrix equal to the correlation matrix and with off diagonal elements equal to the elements in $\frho_b(\z)$. This is further developed as in the proof of Theorem \ref{thm:F2}, and the asymptotic normality of $T_{n,b}-T_b$ follows as before from the mixing assumptions and the proof is completed by applying proposition 6.3.9 of @brockwell1991time. $\blacksquare$

The order of $C_n(A_b)$ is of interest as $\bb\to \bm{0}$ and $n \to \infty$ since it determines the convergence rate of $T_{n,b}-T$.  From Theorem 3.1 in the main article it follows that $\frho_{n,b}(\z)$ converges at a rate of $(nb^5)^{-1/2}$ to $\frho(z)$ as $n \to \infty$ and $b_1=b_2=b_3 = b \to 0$. In the proof of Theorem \@ref(thm:F3) one integrates over a triple kernel function in $C_n(A_b)$. This cancels out three $b$-s. so that one is left with a convergence rate of $(nb^2)^{-1/2}$ for $T_{n,b}-T$. This is slower than the convergence rate in Theorem \@ref(thm:F3), but this is not unexpected since the faster rate in Theorem \@ref(thm:F3) comes at the price of a 1-parameter pairwise simplification in the model structure.

# ASYMPTOTIC POWER AND PITMAN EFFICIENCY {#chap:pitman}

Since the rates of convergence in Theorems \@ref(thm:F2) and \@ref(thm:F3) are $n^{-1/2}$ and $(nb^2)^{-1/2}$ respectively, the Pitman efficiency will be improved in the case of Theorem \@ref(thm:F2) as compared to that of Theorem \@ref(thm:F3).

We will follow the approach used by @wang2017characteristic, in the proof of their Theorem 2 to derive the results on local asymptotic efficiency. We will explicitly use the results in Theorems \@ref(thm:F2) and \@ref(thm:F3) with $h({\falpha}_{n,b}(\z)) = ||{\falpha}_{n,b}(\z)||^2$, so that $T_{n,b} = \int _S||{\falpha}_{n,b}(\z)||^2 \di F_n(\z)$. The null hypothesis of conditional independence can then be phrased as 
$$
H_0: ||{\falpha(\z)}||^2 =  0 \;\; \mbox{for} \;\;\z \;\; \mbox{almost everywhere in} \;\;S. 
$$
against the Pitman alternative
$$
H[a_n]: ||{\falpha(\z)}||^2 = a_n\beta(\z)
$$
where $a_n$ is a sequence that tends to zero at a certain rate, determining the asymptotic rate at which there is still power in testing $H_0$, and $\beta (\z) =\beta(\rho(\z))$ is a positive function of $\z$ such that $\int_S \beta(\z) \di F(\z) \stackrel{\textrm{def}}{=} \gamma < \infty$. Here $\beta(\z)$ is the result of taking an arbitrary correlation vector $\rho(\z) = \rho_H(\z)$ under the alternative hypothesis such that ${\falpha}(\rho_H(\z)) \neq {\bm{0}}$ and $\beta(\z)$ is the Froebenius norm squared of this ${\falpha}$. Further, let $C_{H,n,b}(A_b)$ be the $C_{n,b}$-function formed as in Theorem \@ref(thm:F2) from $\rho_H(\z)$. It converges towards $C_{H}(\fA_b)$ as depicted in the mixing expression of the proof of Theorem \@ref(thm:F2). See also Theorem 3.2 of @lacal2018estimating which is formulated without using independence.

\begin{thm}
Let the conditions be as in Theorem \ref{thm:F2} and let $H[a_n]$ be as above with $a_n =n^{-1/2}$. Let $C_H(A)$ be the limit of $C_n(A_b)$ as $n \to \infty$ and $\bb \to \bm{0}$ at the rates specified in the assumptions of Theorem \ref{thm:F2}. (This limit exists using the mixing expressions of the proof of Theorem \ref{thm:F3}). Then the power of $n^{1/2}(C_{H,n,b}(A_b))^{-1/2}(T_{n,b}-T)$ ($T = 0$ under our $H_0$, but the procedure below can also be carried through with a general ${\falpha}$ under $H_0$) is given by $\mathbb{P}[T_{n,b} \geq z_{\varepsilon}|H[a_n]) \to 1-\Phi(z_{\varepsilon}-\gamma/[C_{H}(A))]^{1/2})$ as $n \to \infty, \bb \to \bm{0}$, where $\Phi(\cdot)$ is the cumulative standard normal distribution function and $z_{\varepsilon}$ is the one-sided critical value of ${\cal N}(0,1)$ at significance level $\varepsilon$.
\label{thm:G1}
\end{thm}

*Proof.* The test statistic can be written as

\begin{align*}
T_{n,b} & = \int_S ||{\falpha}(\z)_{n,b}||^2 \di F_n(\z) \\
&= \int _S ||{\falpha}(\z)_{n,b}||^2 \di F_n(\z) - a_n \int_S \beta(\z) \di F(\z) + a_n \int \beta(\z) \di F(\z).
\end{align*}
Here, using Theorem \@ref(thm:F2), under $H[a_n]$,
$$
\sqrt{n}[C_{H,n,b}(A_b)]^{-1/2}\left\{\int_S ||{\falpha}(\z)_{n,b}||^2 \di F_n(\z) - a_n \int_S \beta(\z) \di F(\z)\right\} \stackrel{d}{\to} {\cal N}(0,1).
$$ 
and
$$
\sqrt{n}[C_{H,n,b}(A_b)]^{-1/2} a_n \int_S \beta(\z) \di F(\z) \to \gamma/[C_H(A)]^{1/2}
$$
and this completes the proof. $\blacksquare$

The alternative situation that is covered in Theorem \@ref(thm:F3) can now be stated and proved analogously. In this case $b^2\fC_{H,n,b}$ converges towards $\fC^{\prime}(\fA)$, say, where $b$ is the common value of the bandwidths $(b_1,b_2,b_3)$.

\begin{thm}
Let the conditions be as in Theorem \ref{thm:G1} and let $H[a_n]$ be as above with $a_n =(nb^2)^{-1/2}$, with $b$ the common value of the three bandwidths $(b_1,b_2,b_3)$. Then, letting $C_H^{\prime}(A)$ be the limit of $b^2C_{H,n,b}(A_b)$ as $n \to \infty$ and $b \to 0$ at the rates specified in the assumptions of Theorems \ref{thm:F2} and \ref{thm:F3}. (This limit exists using the mixing expressions of the proof of Theorem \ref{thm:F2}). Then the power of $(nb^2)^{1/2}(C_{H,n,b}(A_b))^{-1/2}(T_{n,b}-T)$ ($T = 0$ under our $H_0$, but the procedure can also be carried through with a general ${\falpha}$ under $H_0$.) is given by $\mathbb{P}[T_{n,b} \geq z_{\varepsilon}|H[a_n]) \to 1-\Phi(z_{\varepsilon}-\gamma/[C_{H}(A)^{\prime})]^{1/2})$ as $n \to \infty, b \to 0$, where $\Phi(\cdot)$ is the cumulative standard normal distribution function and $z_{\varepsilon}$ is the one-sided critical value of ${\cal N}(0,1)$ at significance level $\varepsilon$.
\label{thm:G2}
\end{thm}

We conclude that the testing procedure can pick up alternatives that converges at the rate $(nb^2)^{-1/2}$ towards the hypothesis of conditional independence as formulated under the alternative (i) in Section \@ref(chap:estimation).

For a further brief discussion, see the paragraph following eqs. (27) and (28) in Section 5.2 in the main article.

# BOOTSTRAPPING THE TEST STATISTIC

The main principle of this section is that we replace $F_n$-quantities with $F_n^{*}$-quantities and $F$-quantities with $F_n$-quantities, the latter being known under the bootstrap operation. Let
$$T_{n,b}^* = \int_S h(\frho_{b,n}^*(\z))\di F_n^*(\z).$$
We treat the pairwise model case corresponding to Theorem \@ref(thm:F2) first.

\begin{thm}
Let the assumptions of Theorem \ref{thm:E2} be fulfilled. Then
as $n \to \infty$
$$
\sqrt{n}[C_{n}(A_b)]^{-\frac{1}{2}}(T_{n,b}^* - T_{n,b}) \stackrel{d_{\mathbb{P}_{\omega}^{*}}}{\to} {\cal N}(0,1).
$$
Here 
\begin{align*}
A_b(z) = \int_{S} \Big[\sum_{i,j}\frac{\partial}{\partial \rho_{ij,n,b}} h(\rho_{n,b}(\fv)J_{ij}^{-1}(\rho_{ij,n,b}(\fv_{ij}))K_b(\z_{ij}-\fv_{ij})&u_{ij}(\z_{ij},\rho_{ij,n,b}(\fv_{ij})\Big]  \di F(\fv) \\
& + h(\rho_{ij,n,b}(\z)1_S(\z),
\end{align*}
and 
\begin{align}
C_n(A_b) = \int A_b^2(\z)\di F_n(\z) + & \frac{1}{n}\sum_{s \neq t}\int A_b(\z)A_b(\fu)\di F^{s,t}_{n}(\z,\fu) \nonumber \\
& -  n\int A_b(\z)A_b(\fu) \di F_n(\z) \di F_n(\fu), 
(\#eq:t00)
\end{align}
with $F^{s,t}_{n}(\z,\fu)$ being the empirical joint cumulative distribution function of $(\Z_s,\Z_t)$. Note that the sum in the second term in the definition of $C_n(A_b)$ taken together with the third term converges due to the mixing assumption.
\label{thm:H1}
\end{thm}

*Proof.* As in the beginning of the proof of Theorem 4.4, part B, of the supplement of @lacal2017local, one has that asymptotically,
\begin{equation}
T_{n,b}^* - T_{n,b}  \sim \frac{1}{\sqrt{n}} \left\{\int_S h(\frho_b(\z)) \di G_n^*(\z) + \int_S\nabla_{\rho}h(\frho_b(\z)^{T}\tilde{V}_n(\z) \di F_n(\z)\right\},
(\#eq:t11)
\end{equation}
where
$$
G_n(\z) = \sqrt{n}(F_n^*(\z) - F_n(\z)) \quad \mbox{and} \quad \tilde{V}_n(\z) = \sqrt{n}((\frho_{n,b}^{*}(\z) -\frho_{n,b}(\z))
$$ 
with $F_n^*$ being the empirical cumulative function for the bootstrapped time series, and where again $\frho$ is the vector consisting of all possible pairs $\rho_{ij}$. This rewriting of $T_{n,b}^*-T_{n,b}$ is the first step in obtaining an expression in terms over $\Z^*_t$, $t=1,\dots,n$ so that a mixing type CLT can be used as done in the paper by @gonccalves2002bootstrap. We proceed to the pairwise analysis.

Let $(i,j)$ be a fixed pair and $\rho_{ij,n,b}^*(z_i,z_j)$ be the corresponding estimated bootstrap local correlation for the pair $(i,j)$ chosen from the vector $\frho_{n,b}^*$. Here the pair of coordinates $(z_i,z_j)$ corresponds to the bivariate $\x$ in Lacal and Tjøstheim [-@lacal2018estimating, proof of Theorem 3.1]. To ease notation and the comparison with Lacal and Tjøstheim [-@lacal2017local; -@lacal2018estimating], we again write $\z_{ij} = (z_i,z_j)$. We will also omit some indices or parts of indices when it is clear what is meant. Finally, we omit the area of integration $S$. We are using the pairwise log likelihood given by
$$
L_{ij,n}^* = \frac{1}{n} \sum_{t=1}^{n} K_b(\Z_{ij,t}^{*}-\z_{ij}) \log \psi(\Z_{ij,t}^{*},\rho_{ij,n,b}(\z_{ij})) - \int K_b(\fv_{ij}-\z_{ij}) \psi_{ij}(\fv_{ij},\rho_{ij,n,b}(\z_{ij})) \di \fv_{ij},
$$
where $K_b = \frac{1}{b_1}K_1(\cdot/b_1) \frac{1}{b_2} K_2(\cdot/b_2)$ is a product kernel function and $\Z_{ij,t}^{*} = \left(Z_{i,t}^{*},Z_{j,t}^{*}\right)$ is the $t$-th pair of bootstrap observations. Estimates $\rho_{ij,n,b}^*(\z_{ij})$ are obtained by differentiation of the log likelihood and by introducing the score function $U_{ij}^*$ and information function $I_{ij}^*$ as follows
\begin{align*}
U_{ij}^*(\z_{ij},\rho_{ij,n}) &= \frac{\partial}{\partial \rho_{ij,n}}L_{ij,n}^*(\z_{ij},\rho_{ij,n}) \\
& = \frac{1}{n}\sum_{t=1}^{n}K_b(\Z_{ij,t}^{*} - \z_{ij})u_{ij}(\Z_{ij,t}^{*},\rho_{ij,n}) - 
\int K_b(\fv_{ij} - \z_{ij})u_{ij}(\fv_{ij},\rho_{ij,n})\psi(\fv_{ij},\rho_{ij,n}) \di \fv_{ij}
\end{align*}
where $u_{ij}(\z_{ij},\rho_{ij,n}) = \frac{\partial}{\partial \rho_{ij,n}}\log \psi(\z_{ij},\rho_{ij,n})$. Similarly,
$$
I_{ij}^*(\z_{ij},\rho_{ij,n}) = \frac{\partial^2}{\partial \rho_{ij,n}^2} L_{ij,n}^{*}(\z_{ij},\rho_{ij,n}) \quad \mbox{and} \quad J_{ij}^*(\z_{ij}) = \E^*\left(-I_{ij}^*(\z_{ij},\rho_{ij,n})\right),
$$
where Lacal and Tjøstheim [-@lacal2018estimating, proof of Theorem 4.4] may be consulted for more details in the derivation and a more explicit analog expression for $I_{ij}^*(\z_{ij},\rho_{ij,n})$ for the case considered there. In contradistinction to @lacal2017local, $J_{ij,n,b}$ is not singular as $\bb \to \bm{0}$. In fact, it can be shown that, cf. Otneim and Tjøstheim [-@otneim2017conditional, Theorem 2], as $\bb \to \bm{0}$, $J_{ij,n,b} \to u(\z_{ij},\rho_{ij,n})\psi(\z_{ij},\rho(\z_{ij,n}))$.

The vector $\tilde{V}_n(\z) = \sqrt{n}\left(\frho_{n,b}^*(\z)-\frho_{n,b}(\z)\right)$ has the dimension of $\frho(z)$ and consists of components $\{\rho_{ij,n,b}^*(\z_{ij}-\rho_{ij,n,b}(\z_{ij})\}$ where $i$ and $j$ in the most general case run over all possible pairs formed by the components of $\frho(\z)$. To be able to use the mixing CLT analogous to @gonccalves2002bootstrap we follow @lacal2017local and express this more explicitly in terms of the variables $\Z^{*}_t$, $t=1,\ldots,n$. By Taylor expansion as in the proof of Theorem \@ref(thm:F2), and omitting the index $ij$ on $\rho$,
\begin{align*}
0 &= U_{ij}^*(\z_{ij},\rho_{n,b}^*(\z_{ij})) \\
& \sim U_{ij}^*(\z_{ij},\rho_{n,b}(\z_{ij}))+I_{ij}^*(\z_{ij},\rho_{n,b}(\z_{ij}))(\rho_{n,b}^*(z_{ij})-\rho(z_{ij})) \\
& \sim U_{ij}^*(\z_{ij},\rho_{n,b}(\z_{ij})) - J_{ij,n,b}(\z_{ij})(\rho_{n,b}^*(\z_{ij})-\rho_{n,b}(\z_{ij}))
\end{align*}
where we have used that $\E^*(\cdot)$ is the mean value under the measure $\mathbb{P}_{\omega}^{*}$ and again following the reasoning of the proof of Theorem \@ref(thm:F2), this
implies that
\begin{align}
\rho_{n,b}^*(\z_{ij}) - \rho_{n,b}(\z_{ij}) &\sim J_{ij,n,b}^{-1}(\z_{ij})u_{ij}(\z_{ij},\rho_{n,b}(\z_{ij})) \nonumber \\
& = J_{ij,n,b}^{-1}(\z_{ij})\Big[\frac{1}{n}\sum_{t=1}^{n} K_b(\Z_{ij,t}^{*} - \z_{ij})u_{ij}(\Z_{ij,t}^{*},\rho_{n,b}(\z_{ij})) \nonumber \\
 & \qquad\qquad - \int K_b(\fv_{ij} - \z_{ij})u_{ij}(\fv_{ij},\rho_{n,b}(\z_{ij}))\psi(\fv_{ij},\rho_{n,b}(\z_{ij}) \di \fv_{ij}\big].
\end{align}
The second term in \@ref(eq:t11) can then be expressed as
\begin{align}
&\int_S \sum_{i,j} \frac{\partial}{\partial \rho_{ij,n,b}} h(\rho_{n,b}(\z)) \times
 J_{ij,n,b}^{-1}(z_{ij})
\Big[\frac{1}{n}\sum_{t=1}^{n} K_b(\Z_{ij,t}^{*}-\z_{ij})u_{ij}(\Z_{ij,t}^{*},\rho_{n,b}(\z_{ij})  \nonumber \\
& \qquad\qquad\qquad\qquad - \int K_b(\fv_{ij} - \z_{ij})u_{ij}(\fv_{ij},\rho_{n,b}(\z_{ij}))\psi(\fv_{ij},\rho_{n,b}(\z_{ij}) \di \fv_{ij}\Big] \di F_n(\z),
(\#eq:t22)
\end{align}
where the sum over $(i,j)$ extends over all $\rho_{ij}$ contained in the $\rho$-vector. This should be put together with the first term of \@ref(eq:t11),
\begin{equation}
\frac{1}{\sqrt{n}} \int_S h(\rho_{n,b}(\z)) \di G_n^*(\z) = \frac{1}{n}\sum_{t=1}^{n}h(\rho_{n,b}(\Z_t^{*}) - \int h(\rho_{n,b}(\z) \di F_n(\z)
(\#eq:t33)
\end{equation}
and we have thus reached an expression where we can apply the mixing CLT as in @gonccalves2002bootstrap and an application of their result, quoted as Theorems C.1 and E.1 in the supplementary materials of @lacal2017local and @lacal2018estimating respectively, to the expressions in \@ref(eq:t22) and \@ref(eq:t33). This yields asymptotic normality of $\sqrt{n}(T_{n,b}^*-T_{b,n})$.

The rest of the proof now follows along the lines of the end of the proof of Theorem \@ref(thm:F2), including the use of Proposition 6.3.9 of @brockwell1991time. $\blacksquare$

An analogue of Theorem \@ref(thm:F3) can now be phrased and proved similarly.

\begin{thm}
Let the assumptions of Theorem \ref{thm:E2} be fulfilled. Then,
as $n \to \infty$ and $\bb \to \bm{0}$
$$
\sqrt{n}\left(C_{n}(A_b)\right)^{-\frac{1}{2}}\left((T_{n,b}^* - T_{n,b}\right) \buildrel {d_{\mathbb{P}_{\omega}^*}} \over {\to} {\cal N}(0,1).
$$
Here 
$$
A_b(\z) = \int_S \nabla_{\rho}h(\rho_{n,\bb}(\fv))^{T}{\J}^{-1}(\rho_{n,\bb},\fv)K_b(\z-\fv)\fu(\z,\frho_{n,\bb}(\fv) \di F(\fv) + h(\frho_{n,\bb}(\z)1_S(\z)
$$
and 
\begin{align*}
C_n(A_b) = \int A_b^2(\z) \di F_n(\z) &+ \frac{1}{n}\sum_{s \neq t}\int A_b(\z)A_b(\fu) \di F_n^{s,t}(\z,\fu) \\ & \qquad\qquad- n\int A_b(\z)A_b(\fu)\di F_n(\z) \di F_n(\fu),
\end{align*}
with $F_n^{s,t}(\z,\fu)$ being the empirical joint cumulative distribution function of $(\Z_s,\Z_t)$.
\label{thm:H2}
\end{thm}

# THE LOCAL BOOTSTRAP

The limiting results for $\alpha_n(\z)$ (or ${\falpha_n}(\z)$ in the matrix case) and $T_{n,b}$, and the corresponding bootstrap results can be used to construct asymptotic and bootstrap confidence intervals of $T_b$ and $T$. In that sense they give an indication of whether $\alpha(\z) = 0$ locally or $\alpha(\z) \equiv 0$ globally respectively. Such confidence intervals, however, cannot in general be used to test for $\alpha(z) \equiv 0$, i.e. conditional independence, with a given level of the test. To be able to do that one must device a test statistic that is constructed with a probability structure as under the null hypothesis of conditional independence. It is difficult to obtain such a test based on asymptotic arguments for at least two reasons: As mentioned before the asymptotic distribution of test functionals such as $T_{n,b}$ is known to be inaccurate except for extremely large sample sizes. It is also difficult to build the probability null structure into the scaling factor of $T_{n,b}$ since this would amount to building into that structure certain relationships between pairwise local correlations.

An alternative that has been used in the testing theory of conditional independence is the so-called local bootstrap. This is a device that uses testing based on bootstrapping from kernel density estimates. It was used for this purpose  by @paparoditis2000local and much more extensively by @su2008nonparametric. It has also been used (without proof) by @wang2017characteristic and by @huang2010testing. We have used a variation of the local bootstrap in our test of conditional independence with good results in accuracy of level and in power. The primary purpose of the present paper is not to prove in detail properties of the local bootstrap in local Gaussian testing, but we will nevertheless indicate theoretically why a local type bootstrap argument makes sense, and why it can be expected to, and does, work also in our case. To put the local type bootstrap into our test functional context we need first to state some facts and properties of the local bootstrap as used in the literature. Note that our functional is in a sense analogous to that used by @wang2017characteristic.

In our notation $\Z_{3,t}$ is the basic $d_3$-dimensional process corresponding to $X_t$ in @wang2017characteristic. Under $H_0$ we assume that $\Z_{1,t}$ and $\Z_{2,t}$ of dimensions $d_1$ and $d_2$ are conditionally independent given $\Z_{3,t}$, where $\Z_{1,t}$ and $\Z_{2,t}$ correspond to $Y_t$ and $Z_t$ in @wang2017characteristic

The first step consists in constructing a bootstrap realization $\{\Z_{3,t}^*\}$ of $\{\Z_{3,t}\}$ by sampling from the kernel density estimate $\tilde{f}(\z_{3}) = \frac{1}{n}\sum_{t=1}^{n}L_b(\z_3-\Z_{3,t})$, where $L_b$ is a $d_3$ dimensional product kernel function with factors $L_{b_{i}} = \frac{1}{b_{i}}L(\cdot/b_i)$. The bootstrap realizations of $\{\Z_{1,t}^*\}$ and $\{\Z_{2,t}^*\}$ are then created independently of each other by sampling from the conditional kernel estimators $\tilde{f}(\z_1|\Z_{3,t}^*) = \sum_sL_b(\Z_{1,s}-\z_{1})L_b(\Z_{3,s}-\Z_{3,t}^*)/\sum_s L_b(\Z_{3,s}-\Z_{3,t}^*)$ and $\tilde{f}(\z_2|\Z_{3,t}^*) = \sum_sL_b(\Z_{2,s}-\z_2)L_b(\Z_{3,s}-\Z_{3,t}^*)/\sum_sL_b(\Z_{3,s}-\Z_{3,t}^*)$, respectively.

The bootstrap test statistic $T_n^*$ (which is a Hellinger type statistic in the Su-White case. can then be constructed from these bootstrap observations $\{\Z_{1,t}^*, \Z_{2,t}^*, \Z_{3,t}^*\}$  using another kernel function $K$ and a bandwidth $h$ to construct density estimates $\hat{f}^*(\z_3)$, $\hat{f}^*(\z_2,\z_3)$, $\hat{f}^*(\z_1,\z_3)$ and $\hat{f}^*(\z_1,\z_2,\z_3)$ which is inserted in the Hellinger functional to obtain a bootstrap value $T_n^*$ of this statistic. It can then be shown [@su2008nonparametric] that properly normalized $T_n^* \buildrel {d} \over {\to} {\cal N}(0,1)$.

Further, if $\mathbb{P}^*$ denote the probability conditional on the sample $\{\Z_t\}$, then the level-$\varepsilon$ critical values $\tilde{c}_{\varepsilon}$ are computed as approximate solutions of $\mathbb{P}^{*}(T_n^* > \tilde{c}_{\varepsilon}) = \varepsilon$ and the bootstrap $p$-value is given by $p^* = B^{-1}\sum_{j=1}^{B} 1(T_{nj}^* > T_n)$, where $B$ is the number of bootstrap realization. To integrate this procedure into our context is fairly straightforward. The first step is to generate local type bootstrap observations $\{\Z_{1,t}^*,\Z_{2,t}^*,\Z_{3,t}^*\}$. The bootstrap realization $\{\Z_{3,t}^*\}$ can be created using the ordinary bootstrap on $\{\Z_{3,t}\}$, cf. @paparoditis2000local, or the kernel smoothed bootstrap as in @su2008nonparametric. If the dimension of $\Z_{3,t}$ is large, a smoothed bootstrap realization can be created by sampling from the  local Gaussian density estimate of @otneim2017locally. In the next step in creating the conditional bootstrap realizations $\{\Z_{2,t}^*\}$ and $\{\Z_{3,t}^*\}$ one can use the smoothed conditional kernel estimate of @su2008nonparametric, or if the dimensions are large, to avoid the curse of dimensionality, one can sample from the conditional local Gaussian density estimates of @otneim2017conditional using, say, acceptance reject sampling. Again it should be noted that the avoidance of the kernel estimate comes at the cost of a more restrictive model.

Creating bootstrap realizations by the above scheme is quite time consuming since 
for each $\Z_{3,t}^*$, drawings of $\Z_{1,t}^*$ and $\Z_{2,t}^*$ have to be done, and this has to be repeated over the number of bootstrap realizations. We have devised a simplified and more efficient procedure for this as described in the main article.

Once a bootstrap realization $\{\Z_{1,t}^*,\Z_{2,t}^*,\Z_{3,t}^*\}$ is created, following @su2008nonparametric, these realizations are now taken as the original observations in the local likelihood estimation procedure. For a fixed bandwidth $b$ the population values, following the local Gaussian approach, can be taken as the solution of the integral equation (2) in @tjostheim2013local; see also corresponding equations in e.g. Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional]. Note that in the present paper we have largely operated with two alternatives, the general vector case with inference via the pairwise local likelihood, or the the purely scalar case with three coordinates for the local Gaussian correlation. In both cases we have normalized the processes so that they have standard normal marginals.

In the $\{\Z_{1,t}^*,\Z_{2,t}^*,\Z_{3,t}^*\}$-framework we obtain an estimate $\hfrho^*$ of the local Gaussian correlation vector conditional on the bootstrap realizations. An asymptotic theory of this estimate can now be established using the method of proof in Theorem 2 in @otneim2017locally, and subsequently an asymptotic theory for the test statistic $T_n^*$ using the method of proof in Theorem 4.4 in @lacal2017local and Theorem D.4 in @lacal2018estimating. See also the formulation of such proofs in the present paper, theorems \@ref(thm:H1) and \@ref(thm:H2). The scaling factor $C_n(A_b)$ is of the same type as in the formulation of those results with obvious notational changes. Moreover under the present null hypotheses $T = 0$, it may be conjectured that it leads to the asymptotic results that the scaled $T_n^* \buildrel {d} \over {\to} {\cal N}(0,1)$.

# TWO ADDITIONAL EXAMPLES OF ESTIMATION

We will in this section provide two more demonstrations on how the estimated LGPC can be used to reveal nonlinear departures from conditional independence and to measure that conditional dependence. The ordinary partial correlation coefficient fully characterizes conditional dependence between jointly Gaussian variables. It is not hard, though, to construct examples in which strong nonlinear relationships remain completely undetected by the partial correlation. Consider the following structural equation;

\begin{figure}[t]
    \subfloat[Estimated local Gaussian partial correlation \newline between $X_1$ and $X_2$ given $X_3=0$ in the \newline structural model $X_2 = X_1^2 + X_3.$ The color scale \newline goes from $-1$ (dark blue) to $+1$ (dark red).\label{fig:structural-plot1}]{\includegraphics[width=0.49\linewidth]{figures/structural-plot-1} }
    \subfloat[The estimated LGPC along the curve $X_2 = X_1^2$ with 95\% confidence band.\label{fig:structural-plot2}]{\includegraphics[width=0.49\linewidth]{figures/structural-plot-2} }
    \caption{The structural model. The fully trivariate model of Section 3.1.1 is used.}\label{fig:structural-plot}
\end{figure}

\begin{equation}
X_2 = X_1^2 + X_3,
(\#eq:structural)
\end{equation}
where we observe all components in $(X_1, X_2, X_3)$. There is, obviously, a strong dependence between $X_1$ and $X_2$, and furthermore, this dependence is deterministic when conditioning on $X_3 = x_3$. It is, however, well known that $X_1$ and $X_1^2 + x_3$ are uncorrelated if $\E(X_1) = \E(X_1^3) = 0$, but the LGPC easily reveals conditional dependence between $X_1$ and $X_2$ in this case. Let us generate $n = 500$ independent observations each from $X_1 \sim \mathcal{N}(0,1)$ and $X_3 \sim \mathcal{N}(0,1)$ and calculate $X_2$ by \@ref(eq:structural). The sample partial correlation between $X_1$ and $X_2$ is $-0.037$ in this case, but the LGPC, which in this case is calculated using the simplified pairwise structure defined in Section 3.1.2, displayed in Figure \@ref(fig:structural-plot1), indicates strong departures from conditional independence on the plane defined by $X_3 = 0$. Indeed, we identify a region of *negative conditional relationship* on the left hand side of the plot, which makes good sense because small values of $X_2$ is typically observed together with large values of $X_1$ and vice versa. The opposite phenomenon is clearly visible in the right hand half of the plot. In Figure \@ref(fig:structural-plot2) we have plotted the estimated LGPC along the curve $X_2 = X_1^2$, along with an asymptotic 95% confidence band. The full trivariate fit presented in Section 3.1.1 gives a very similar picture.

This points to an important feature of the LGPC, as we also saw in one of the simulated examples in the main article: It is able to *distinguish between positive and negative conditional relationships*, which, to our knowledge, has until now not been possible beyond the linear and jointly Gaussian setting using the ordinary partial correlation.

Let us briefly demonstrate a practical implementation of the LGPC in another situation involving real data. As in the empirical example in the main article, we look at Granger causality,

\begin{figure}
\subfloat[Local partial dependence map of $V_t^*$ vs. $R_{t-1}$ \newline on the plane defined by $V_{t-1}^* = 0$\label{fig:realplots1}]{\includegraphics[width=0.49\linewidth]{figures/realplots-1} }
\subfloat[Local partial dependence map of $R_t$ vs. $V_{t-1}^*$ \newline on the plane defined by $R_{t-1} = 0$\label{fig:realplots2}]{\includegraphics[width=0.49\linewidth]{figures/realplots-2} }\newline
\subfloat[Local partial dependence map of $R_t$ vs. $V_{t-1}^*$ \newline on the  plane defined by $R_{t-1} = -0.05$\label{fig:realplots3}]{\includegraphics[width=0.49\linewidth]{figures/realplots-3} }
\subfloat[Local partial dependence map of $R_t$ vs. $V_{t-1}^*$ \newline on the  plane defined by $R_{t-1} = 0.05$\label{fig:realplots4}]{\includegraphics[width=0.49\linewidth]{figures/realplots-4} }
\caption{Local partial dependence maps for the S\&P 500 data}\label{fig:realplots}
\end{figure}

One fairly recent method for testing for Granger causality is based on the maximal conditional correlation and was introduced by @huang2010testing and later extended to the time series case by @cheng2012conditional. The latter authors use this test to examine whether trading volume Granger caused index value, or vice versa, on the S&P 500 stock index during the first decade of this century. Using the daily price series $\{P_t\}$ and volume $\{V_t\}$ they define the log-differenced series

$$R_t = 100\log\left(\frac{P_t}{P_{t-1}}\right) \,\,\, \textrm{ and } \,\,\, V_t^* = \log\left(\frac{V_t}{V_{t-1}}\right).$$
Further, they denote by $R_{t-1} \not\Rightarrow V_t^*$ the Granger non-causality from $\{R_t\}$ to $\{V_t^*\}$, which they analyze by putting the following null hypothesis to the test:

\begin{equation}
\textrm{H}_0: V_t^* \perp R_{t-1} \,\, | \,\, V_{t-1}^*.
\label{eq:grangernull500}
\end{equation}

Granger non-causality in the opposite direction, $V_{t-1}^* \not\Rightarrow R_t$ is formulated and tested for in the obvious way.

A linear test rejects $R_{t-1} \not\Rightarrow V_t^*$, but not $V_{t-1}^* \not\Rightarrow R_t$. The nonparametric test by @cheng2012conditional rejects both, leading to the natural conclusion that there are nonlinear relationships in the latter situation that remain unseen by linear models. What is it, though, in this particular data set, that leads to such results? Estimates of the local partial Gaussian correlation provide some clues towards answering that question.

We obtain daily observations on price and trading volume on the S&P 500 index from January 1st 2000 until December 31st 2009 from Yahoo Finance [-@yahoo], and plot the estimated trivariate LGPC between $V_t^*$ and $R_{t-1}$ on the plane defined by $V_{t-1}^* = 0$ in Figure \@ref(fig:realplots1). Departures from $\alpha(V_t^*, R_{t-1}) \equiv 0$ provide evidence against $R_{t-1} \not\Rightarrow V_{t}^*$. Similarly, we plot the estimated LGPC  between $R_t$ and $V^*_{t-1}$ on the plane $R_{t-1} = 0$ in Figure \@ref(fig:realplots2) in which departures from $\alpha(R_t, V^*_{t-1}) \equiv 0$ provide evidence against $V^*_{t-1} \not\Rightarrow R_t$. We have used a proportionality constant of $c = 3.5$ to calculate the estimates (see discussion on bandwidth selection in Section 4 of the main article). The observations can be seen in the background of both plots.

The differences between Figures \@ref(fig:realplots1) and \@ref(fig:realplots2) are subtle, but important. In Figure \@ref(fig:realplots1), $\widehat\alpha(V_t^*, R_{t-1})$ is mostly negative, especially in the data rich portions of the sample space (other values of the conditioning variable $V_{t-1}^*$ than zero give similar pictures). Indeed, the *global* partial correlation in this situation is $\widehat{\alpha}_{\textrm{glob}} = -0.086$, which is significantly different from zero. The global partial correlation in the second situation is very small though: $\widehat{\alpha}_{\textrm{glob}} = -0.0018$, but Figure \@ref(fig:realplots2) still reveals departures from conditional independence of similar magnitudes as in Figure \@ref(fig:realplots1). The difference is that the estimated LGPC is positive (primarily in the second and fourth quadrants) as well as negative (in the first and third quadrants), but this pattern collapses to the constant global value $\alpha(R_t, V_{t-1}^*) = \alpha_{\textrm{glob}} \approx 0$ as the bandwidths tend to infinity. In Figures \@ref(fig:realplots3) and \@ref(fig:realplots4) we can explore the conditional dependence between $R_t$ and $V_{t-1}$ at high and low levels of $R_{t-1}$, respectively, and we see even more clear differences in this dimension, especially in the first and second quadrants.

The $p$-values for tests of the hypothesis \@ref(eq:grangernull500) and its opposite counterpart, using our new test for conditional independence as presented in the next section and with $c$-values for the bandwidth as chosen there, are both equal to 0.

# POWER STUDIES

Table 2$^*$ below is the extended version of Table 2 in the main article, sample size is $n=100$:

\vspace{1cm}

\input{tables/power_table_dim3_n100}

\newpage

In Table 3$^*$ we present level and power results for various conditional independence tests applied to the DGPs listed in Table 1 in the paper for sample size $n = 200$:

\input{tables/power_table_dim3_n200}

@su2014testing define two extensions to a subset of the data generating processes defined in Table 1 in which the conditioning variable $\X_3$ is a vector with two and three components respectively. The first extension turns the conditioning variable $\X_{3,t}$ in DGP1-DGP2 and DGP5-DGP10 $X_{3,t}$ into a bivariate vector, where we define DGP1$^{\prime}$ in the same way as before,

\begin{itemize}
\item[1$^{\prime}$.] $(X_{1,t}, X_{2,t}, \X_{3,t}) = (\epsilon_{1,t}, \epsilon_{2,t}, \fepsilon_{3,t})$,
\end{itemize}
but where $\fepsilon_{3,t} \sim \mathcal{N}(0, \bm{I}_2)$, and where we define DGP2$^{\prime}$ and DGP5$^{\prime}$-DGP10$^{\prime}$ by setting $\X_{3,t} = (X_{1, t-1}, X_{1, t-2})$, keeping $X_{2,t}$ as before, and,

\begin{itemize}
\tightlist
\item[2$^{\prime}$.] $X_{1,t} = 0.5X_{1,t-1} + 0.25X_{1, t-2} + \epsilon_{1,t}$,
\item[5$^{\prime}$.] $X_{1,t} = 0.5X_{1,t-1} + 0.25X_{1, t-2} + 0.5X_{2,t} + \epsilon_{1,t}$,
\item[6$^{\prime}$.] $X_{1,t} = 0.5X_{1,t-1} + 0.25X_{1, t-2} + 0.5X_{2,t}^2 + \epsilon_{1,t}$,
\item[7$^{\prime}$.] $X_{1,t} = 0.5X_{1,t-1}X_{2, t} + 0.25X_{1,t-2}+ \epsilon_{1,t}$,
\item[8$^{\prime}$.] $X_{1,t} = 0.5X_{1,t-1} + 0.25X_{1,t-2} + 0.5X_{2, t}\epsilon_{1,t}$,
\item[9$^{\prime}$.] $X_{1,t} = \sqrt{h_t}\epsilon_{1,t}$, $h_t = 0.01 + 0.5X_{1,t-1}^2 + 0.25X_{1,t-2}^2 + 0.25X_{2,t}^2$, and
\item[10$^{\prime}$.] Same as DGP10 above, except for the new definition of $\X_{3,t}$.
\end{itemize}

We report the level and power results obtained by @su2014testing for the methods CHF, CM, KS and SEL on these data by testing the null hypothesis 
$$\textrm{H}_0: X_{t,1} \perp X_{t,2} \,\, | \,\, \X_{t,3},$$
and include results from our new test based on the multivariate simplification of the LGPC as defined in Section 3.1.2 in table 4$^*$. The results are quite acceptable and compares well with other nonparametric methods.

The second extension introduced by @su2014testing increases the dimension of the conditioning variable $\X_{3,t}$ once more, so that $\fepsilon_{3,t} \sim \mathcal{N}(0, \bm{I}_3)$, and where DGP2$^{\prime\prime}$ and DGP5$^{\prime\prime}$-DGP10$^{\prime\prime}$ are defined by by setting $\X_{3,t} = (X_{1, t-1}, X_{1, t-2}, X_{1, t-3})$, and,

\begin{itemize}
\tightlist
\item[2$^{\prime\prime}$.] $X_{1,t} = 0.5X_{1,t-1} + 0.25X_{1, t-2} + 0.125X_{1,t-3} + \epsilon_{1,t}$,
\item[5$^{\prime\prime}$.] $X_{1,t} = 0.5X_{1,t-1} + 0.25X_{1, t-2} + 0.125X_{1,t-3} + 0.5X_{2,t} + \epsilon_{1,t}$,
\item[6$^{\prime\prime}$.] $X_{1,t} = 0.5X_{1,t-1} + 0.25X_{1, t-2} + 0.125X_{1,t-3} + 0.5X_{2,t}^2 + \epsilon_{1,t}$,
\item[7$^{\prime\prime}$.] $X_{1,t} = 0.5X_{1,t-1}X_{2, t} + 0.25X_{1,t-2} + 0.125X_{1,t-3} + \epsilon_{1,t}$,
\item[8$^{\prime\prime}$.] $X_{1,t} = 0.5X_{1,t-1} + 0.25X_{1,t-2} + 0.125X_{1,t-3} + 0.5X_{2, t}\epsilon_{1,t}$,
\item[9$^{\prime\prime}$.] $X_{1,t} = \sqrt{h_t}\epsilon_{1,t}$, $h_t = 0.01 + 0.5X_{1,t-1}^2 + 0.25X_{1,t-2}^2 + 0.125X_{1,t-3}^2 + 0.25X_{2,t}^2$, and
\item[10$^{\prime\prime}$.] Same as DGP10 and DGP10$^{\prime}$ above, except for the new definition of $\X_{3,t}$.
\end{itemize}
\input{tables/power_table_dim4}
\input{tables/power_table_dim5}

Again, we observe simulation results in Table 5$^*$. Contrary to @su2014testing, we have also run our test for $n=100$ which reveals that we can obtain some power also in that case. All in all, the simplified test based on the LGPC performs mostly on par with other non-parametric tests in this setting as well.

\newpage

# REFERENCES

### 2020-09-02 lgpc_JBES_supp.tex
Initial commit, revising for JBES.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
  letterpaper]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Online supplement accompanying `The Locally Gaussian Partial Correlation'},
  pdfauthor={Håkon Otneim and Dag Tjøstheim},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin = 1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{setspace}
\usepackage{sectsty}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\newtheorem{thm}{Theorem}[section]
\newtheorem{cor}[thm]{Corollary}

%\doublespacing
\sectionfont{\centering}
\renewcommand\thesection{\Alph{section}}
\numberwithin{equation}{section}

\newcommand{\X}{\bm{X}}
\newcommand{\Xone}{\bm{X}^{(1)}}
\newcommand{\Xtwo}{\bm{X}^{(2)}}
\newcommand{\xtwo}{\bm{x}^{(2)}}
\newcommand{\x}{\bm{x}}
\newcommand{\hX}{\widehat{X}}
\newcommand{\tX}{\tilde{X}}

\newcommand{\Z}{\bm{Z}}
\newcommand{\z}{\bm{z}}
\newcommand{\hz}{\widehat{z}}
\newcommand{\hfz}{\widehat{\bm{z}}}
\newcommand{\hZ}{\widehat{\bm{Z}}}
\newcommand{\hnZ}{\widehat{Z}}
\newcommand{\Zone}{\bm{Z}^{(1)}}
\newcommand{\Ztwo}{\bm{Z}^{(2)}}
\newcommand{\zone}{\bm{z}^{(1)}}
\newcommand{\ztwo}{\bm{z}^{(2)}}

\newcommand{\D}{\bm{D}}
\newcommand{\hD}{\widehat{\bm{D}}}

\newcommand{\y}{\bm{y}}
\newcommand{\Y}{\bm{Y}}
\newcommand{\hY}{\widehat{Y}}

\newcommand{\fv}{\bm{v}}
\newcommand{\fu}{\bm{u}}

\newcommand{\fC}{\bm{C}}
\newcommand{\fA}{\bm{A}}
\newcommand{\fV}{\bm{V}}
\newcommand{\M}{\bm{M}}
\newcommand{\J}{\bm{J}}

\newcommand{\R}{\bm{R}}
\newcommand{\hR}{\widehat{\bm{R}}}

\newcommand{\hF}{\widehat{F}}
\newcommand{\hf}{\widehat{f}}

\newcommand{\hrho}{\widehat{\rho}}
\newcommand{\frho}{\bm{\rho}}
\newcommand{\hfrho}{\widehat{\bm{\rho}}}

\newcommand{\hh}{\bm{b}}
\newcommand{\bb}{\bm{b}}

\newcommand{\fmu}{\bm{\mu}}
\newcommand{\hfmu}{\widehat{\bm{\mu}}}
\newcommand{\hsigma}{\widehat{\sigma}}
\newcommand{\fSigma}{\bm{\Sigma}}
\newcommand{\hfSigma}{\widehat{\bm{\Sigma}}}
\newcommand{\halpha}{\widehat{\alpha}}
\newcommand{\falpha}{\bm{\alpha}}
\newcommand{\fOmega}{\bm{\Omega}}
\newcommand{\fLambda}{\bm{\Lambda}}
\newcommand{\fepsilon}{\bm{\epsilon}}

\newcommand{\Jb}{\bm{J}_{\hh}}
\newcommand{\Mb}{\bm{M}_{\hh}}

\newcommand{\Cov}{\textrm{Cov}}
\newcommand{\E}{\textrm{E}}
\newcommand{\Var}{\textrm{Var}} 
\newcommand{\di}{\,\textrm{d}}
\newcommand{\Corr}{\textrm{Corr}}

\usepackage{color}
\usepackage{colortbl}

\definecolor{Gray}{gray}{0.9}

\title{Online supplement accompanying `The Locally Gaussian Partial Correlation'}
\author{Håkon Otneim\footnote{Corresponding author. Department of Business and Management Science, NHH Norwegian School of Economics. Helleveien 30, 5045 Bergen, Norway. \texttt{hakon.otneim@nhh.no}} and Dag Tjøstheim\footnote{University of Bergen}}
\date{}

\begin{document}
\maketitle

\hypertarget{the-multivariate-lgpc}{%
\section{THE MULTIVARIATE LGPC}\label{the-multivariate-lgpc}}

\hypertarget{definition}{%
\subsection{Definition}\label{definition}}

The literature on nonlinear conditional dependence modeling is mostly concerned with the simple case where the vector of conditionally independent variables consists of two scalar random variables, \(\Zone = (Z_1, Z_2)\), and this case is therefore the focus of our treatment of the LGPC in the main article. This means that we estimate a single local Gaussian partial correlation function \(\widehat \alpha(\z)\). One of the most attractive properties of the LGPC, however, is the close parallel to the multivariate Gaussian distribution, and this is also the case when describing conditional dependence among more than two variables, which in our case leads to the estimation of a \emph{matrix} of local Gaussian partial correlations. Besides needing a slightly more involved notation, this is a straightforward exercise in our framework as we will see below.

Let \(\Z\) be a vector of continuous random variables, that we split in two components \(\Z = (\Zone, \Ztwo)\), but this time \(\Zone = (Z_1, \ldots, Z_{d_1 + d_2})\) is itself a vector with dimension \(d_1 + d_2\) that is split in two groups; the \(d_1\)-dimensional vector \(\Z_1\) and the \(d_2\)-dimensional vector \(\Z_2\). The vector \(\Ztwo = \Z_3\) is a vector of \(d_3\) random variables. Denote by \(\fmu\) and \(\fSigma\) the expectation vector and the covariance matrix of \(\Z\), now partitioned according to our partition of \(\Z = (\Z_1, \Z_2, \Z_3)\) as

\[\fmu = \begin{pmatrix} \fmu_1 \\ \fmu_2 \\ \fmu_3 \end{pmatrix}, \qquad \fSigma = \begin{pmatrix} \fSigma_{11} & \fSigma_{12} & \fSigma_{13} \\ \fSigma_{21} & \fSigma_{22} & \fSigma_{23} \\ \fSigma_{31} & \fSigma_{32} & \fSigma_{33} \end{pmatrix},\]
where \(\fSigma_{11}\) is the covariance matrix of \(\Z_1\), \(\fSigma_{23}\) contains the covariances between \(\Z_2\) and \(\Z_3\), and so forth, and where naturally \(\fSigma^T = \fSigma\), \(\fSigma_{12}^T = \fSigma_{21}\), \(\fSigma_{13}^T = \fSigma_{31}\), and \(\fSigma_{23}^T = \fSigma_{32}\).

All of these parameters can be described and estimated locally in the exact same way as laid out in the main article, resulting from the locally Gaussian joint density

\begin{equation}
\psi(\z) = \frac{1}{(2\pi)^{(d_1+d_2+d_3)/2}|\fSigma(\z)|^{1/2}} \exp \left\{-\frac{1}{2}(\z - \fmu(\z))^T\fSigma^{-1}(\z)(\z - \fmu(\z))\right\}
\label{eq:localgaussian0}
\end{equation}
and the locally Gaussian \emph{conditional} density

\begin{align*}
\psi(\z_1, \z_2 | \Z_3 = \z_3) & = \frac{1}{(2\pi)^{(d_1+d_2)/2}|\fSigma_{12|3}(\z)|^{1/2}} \nonumber  \\ 
\qquad\qquad\qquad & \times \exp \left\{-\frac{1}{2}(\z - \fmu_{12|3}(\z))^T\fSigma_{12|3}^{-1}(\z)(\z - \fmu_{12|3}(\z))\right\}.
\end{align*}
In order to simplify notation, in the following we will drop the \(z\)-dependence in the parameters. In order to define \(\fmu_{12|3}\) and \(\fSigma_{12|3}\) let \(\fSigma^{11}\), \(\fSigma^{12}\), \(\fSigma^{21}\) and \(\fSigma^{22}\) be matrices of dimension \((d_1 + d_2) \times (d_1 + d_2)\), \((d_1 + d_2) \times d_3\), \(d_3 \times (d_1 + d_2)\) and \(d_3 \times d_3\) respectively given by

\[\fSigma^{11} = \begin{pmatrix} \fSigma_{11} & \fSigma_{12} \\ \fSigma_{21} & \fSigma_{22}\end{pmatrix}, \fSigma^{12} = \begin{pmatrix} \fSigma_{13} \\ \fSigma_{23}\end{pmatrix}, \fSigma^{21} = \begin{pmatrix} \fSigma_{31} & \fSigma_{32}\end{pmatrix}, \textrm{ and } \, \fSigma^{22} = \fSigma_{33}\]
such that
\[\fSigma = \begin{pmatrix} \fSigma^{11} & \fSigma^{12} \\ \fSigma^{21} & \fSigma^{22}\end{pmatrix},\]
It follows from standard arguments (cf.~Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018})) that
\begin{equation}
\fmu_{12|3} = \begin{pmatrix} \fmu_1 \\ \fmu_2 \end{pmatrix} + \fSigma^{12}\left(\fSigma^{22}\right)^{-1}\left(\ztwo - \fmu_3\right) = \begin{pmatrix} \fmu_1 \\ \fmu_2 \end{pmatrix} + \begin{pmatrix} \fSigma_{13}\fSigma_{33}^{-1} \left(\ztwo - \fmu_3\right) \\  \fSigma_{23}\fSigma_{33}^{-1} \left(\ztwo - \fmu_3\right)\end{pmatrix}
\end{equation}
and

\begin{align}
\fSigma_{12|3} &= \fSigma^{11} - \fSigma^{12}\left(\fSigma^{22}\right)^{-1}\fSigma^{21} = \fSigma^{11} - \begin{pmatrix} \fSigma_{13} \\ \fSigma_{23} \end{pmatrix} \fSigma_{33}^{-1} \begin{pmatrix} \fSigma_{31} & \fSigma_{32} \end{pmatrix}\nonumber \\
&= \begin{pmatrix} \fSigma_{11} - \fSigma_{13}\fSigma_{33}^{-1} \fSigma_{31} & \fSigma_{12} - \fSigma_{13}\fSigma_{33}^{-1} \fSigma_{32} \\ \fSigma_{21} - \fSigma_{23}\fSigma_{33}^{-1} \fSigma_{31} & \fSigma_{22} - \fSigma_{23}\fSigma_{33}^{-1} \fSigma_{32}\end{pmatrix}. 
\label{eq:blocksigma}
\end{align}

We proceed now to establish the notion of conditional independence between \(\Z_1\) and \(\Z_2\) using the LGPC in the multivariate case. The conditional densities of \(\Z_1|\Z_3\) and \(\Z_2|\Z_3\) in terms of the local Gaussian parameters are given by

\begin{align*}
f(\z_1 |\z_3) &= \frac{1}{(2\pi)^{d_1/2}|\fSigma_{1|3}|^{1/2}} \exp \left\{-\frac{1}{2}(\z_1 - \fmu_{1|3})^T\fSigma_{1|3}^{-1}(\z_1 - \fmu_{1|3})\right\}, \\
f(\z_2 |\z_3) &= \frac{1}{(2\pi)^{d_2/2}|\fSigma_{2|3}|^{1/2}} \exp \left\{-\frac{1}{2}(\z_2 - \fmu_{2|3})^T\fSigma_{2|3}^{-1}(\z_2 - \fmu_{2|3})\right\} 
\end{align*}
where

\begin{equation}
\fmu_{1|3} = \fmu_1 + \fSigma^{13}\left(\fSigma^{33}\right)^{-1}\left(\z_3 - \fmu_3\right), \,\, \fmu_{2|3} = \fmu_2 + \fSigma^{23}\left(\fSigma^{33}\right)^{-1}\left(\z_3 - \fmu_3\right)  
\end{equation}
and

\begin{equation}
\fSigma_{1|3} = \fSigma_{11}-\fSigma_{13}\fSigma_{33}^{-1}\fSigma_{31}, \,\, \fSigma_{2|3} = \fSigma_{22}-\fSigma_{23}\fSigma_{33}^{-1}\fSigma_{32}.
\label{eq:localsigma}
\end{equation}
Under conditional independence between \(\Z_1\) and \(\Z_2\) given \(\Z_3\), the joint conditional density of \((\Z_1, \Z_2|\Z_3)\) is given by

\begin{align*}
f(\z_1, \z_2|\z_3) &= f(\z_1|\z_3)f(\z_2|\z_3) \\
&= \frac{1}{(2\pi)^{(d_1+d_2)/2}|\fSigma_{1|3}|^{1/2}|\fSigma_{2|3}|^{1/2}} \\
& \qquad \times \exp \Big\{-\frac{1}{2}\big[(\z_1 - \fmu_{1|3})^T\fSigma_{1|3}^{-1}(\z_1 - \fmu_{1|3}) \,\, + \\
& \qquad\qquad\qquad\qquad\qquad\qquad(\z_2 - \fmu_{2|3})^T\fSigma_{2|3}^{-1}(\z_2 - \fmu_{2|3})\big]\Big\}.
\end{align*}
Comparing expressions \eqref{eq:localgaussian0} -- \eqref{eq:localsigma} we see that we have conditional independence if and only if the matrix \(\fSigma_{12|3}\) in \eqref{eq:blocksigma} is block diagonal, i.e.~if and only if
\[\fSigma_{12} - \fSigma_{13}\fSigma_{33}^{-1} \fSigma_{32} = \fSigma_{21} - \fSigma_{23}\fSigma_{33}^{-1} \fSigma_{31} = \bm{0}.\]

\hypertarget{special-cases}{%
\subsection{Some special cases}\label{special-cases}}

Let \(\D = \textrm{diag}\left\{\fSigma_{12|3}\right\}\). The local Gaussian partial correlation matrix can be expressed as the lower left hand block matrix of \(\D^{-1/2}\fSigma_{12|3}\D^{-1/2}\), and so the LGPC matrix is thus given as the upper right hand block of the matrix
\begin{equation}
\D^{-1/2}\fSigma_{12|3}\D^{-1/2} = \D^{-1/2}\begin{pmatrix} \fSigma_{11} - \fSigma_{13}\fSigma_{33}^{-1} \fSigma_{31} & \fSigma_{12} - \fSigma_{13}\fSigma_{33}^{-1} \fSigma_{32} \\ \fSigma_{21} - \fSigma_{23}\fSigma_{33}^{-1} \fSigma_{31} & \fSigma_{22} - \fSigma_{23}\fSigma_{33}^{-1} \fSigma_{32}\end{pmatrix}\D^{-1/2}.
\label{eq:matrixdef}
\end{equation}
Let us consider this expression for a few special cases.

\begin{itemize}
\item[\textbf{a)}] \textbf{Let} $\bm{d_1 = d_2 = d_3 = 1}$. Consider the partial covariance matrix, where we again drop the $z$-dependence in all parameters: 
$$\fSigma_{12|3} = \fSigma^{11} - \fSigma^{12}\left(\fSigma^{22}\right)^{-1}\fSigma^{21}.$$
Here, since all variables are normalized with variance one, we have that
$$\fSigma = \begin{pmatrix} 1 & \rho_{12} & \rho_{13} \\ \rho_{12} & 1 & \rho_{23} \\ \rho_{31} & \rho_{32} & 1 \end{pmatrix},$$
and 
$$\fSigma^{11} = \begin{pmatrix} 1 & \rho_{12} \\ \rho_{21} & 1 \end{pmatrix}, \,\, \fSigma^{12} = \begin{pmatrix} \rho_{13} \\ \rho_{23} \end{pmatrix},\,\, \left(\fSigma^{22}\right)^{-1} = 1, \,\, \fSigma^{21} = \begin{pmatrix} \rho_{31} & \rho_{32} \end{pmatrix}.$$
Hence,
\begin{align*}
\fSigma^{11} - \fSigma^{12}\left(\fSigma^{22}\right)^{-1}\fSigma^{21} &=  \begin{pmatrix} 1 & \rho_{12} \\ \rho_{21} & 1 \end{pmatrix} - \begin{pmatrix} \rho_{13} \\ \rho_{23} \end{pmatrix} \begin{pmatrix} \rho_{31} & \rho_{32} \end{pmatrix} \\
&= \begin{pmatrix} 1 - \rho_{13}^2 & \rho_{12} - \rho_{13}\rho_{23} \\ \rho_{12} - \rho_{13}\rho_{23} & 1 - \rho_{23}^2 \end{pmatrix}.
\end{align*}
This is a covariance matrix, and not a correlation matrix, so normalizing with standard deviations results in the partial correlation matrix
\begin{equation}
\begin{pmatrix}
1 & \frac{\rho_{12} - \rho_{13}\rho_{23}}{\sqrt{1-\rho_{13}^2}\sqrt{1-\rho_{23}^2}} \\  \frac{\rho_{12} - \rho_{13}\rho_{23}}{\sqrt{1-\rho_{13}^2}\sqrt{1-\rho_{23}^2}} & 1 
\end{pmatrix}.
\label{eq:partialmatrix}
\end{equation}
This result can also be obtained by direct application of \eqref{eq:matrixdef}:
$$\begin{pmatrix}1-\rho_{13}^2 & 0 \\ 0 & 1-\rho_ {23}^2\end{pmatrix}^{-1/2} \begin{pmatrix}  1 - \rho_{13}^2 & \rho_{12} - \rho_{13}\rho_{23} \\ \rho_{12} - \rho_{13}\rho_{23} & 1 - \rho_{23}^2  \end{pmatrix} \begin{pmatrix}1-\rho_{13}^2 & 0 \\ 0 & 1-\rho_ {23}^2\end{pmatrix}^{-1/2}.$$

\item[\textbf{b)}] \textbf{Let} $\bm{d_3=1}$ \textbf{and let} $\bm{d_1}$ \textbf{and} $\bm{d_2}$ \textbf{be arbitrary}. In this case 
$$\fSigma^{11} = \begin{pmatrix} \fSigma_{11} & \fSigma_{12} \\ \fSigma_{21} & \fSigma_{22}\end{pmatrix},$$
where
\begin{align*}
&\fSigma_{11} = \left\{ \rho_{ij}^{1,1} \right\}, i = 1, \ldots, d_1; j = 1, \ldots, d_1,   &\fSigma_{12} = \left\{ \rho_{ij}^{1,2} \right\}, i = 1, \ldots, d_1; j = 1, \ldots, d_2 \\
&\fSigma_{21} = \left\{ \rho_{ij}^{2,1} \right\}, i = 1, \ldots, d_2; j = 1, \ldots, d_1,   &\fSigma_{22} = \left\{ \rho_{ij}^{2,2} \right\}, i = 1, \ldots, d_2; j = 1, \ldots, d_2.
\end{align*}
Since $\fSigma^{22} = \fSigma_{33} = 1,$
$$\fSigma^{12}\left(\fSigma^{22}\right)^{-1}\fSigma^{21} = \fSigma^{12}\fSigma^{21} = \begin{pmatrix} \fSigma_{13} \\ \fSigma_{23} \end{pmatrix} \begin{pmatrix}\fSigma_{31} \fSigma_{32} \end{pmatrix},$$
where 
\begin{align*}
&\fSigma_{13} = \left\{ \rho_{i3}^{1,3} \right\}, i = 1, \ldots, d_1;   &\fSigma_{23} = \left\{ \rho_{j3}^{2,3} \right\}, j = 1, \ldots, d_2 \\
&\fSigma_{31} = \left\{ \rho_{3i}^{3,1} \right\}, i = 1, \ldots, d_1;   &\fSigma_{32} = \left\{ \rho_{3j}^{3,2} \right\}, j = 1, \ldots, d_2.
\end{align*}
Furthermore, 
$$\fSigma_{12|3} = \begin{pmatrix}\fSigma_{11} & \fSigma_{12} \\ \fSigma_{21} & \fSigma_{22}\end{pmatrix} - \begin{pmatrix} \fSigma_{13} \\ \fSigma_{23}\end{pmatrix}\begin{pmatrix} \fSigma_{31} & \fSigma_{32}\end{pmatrix}.$$
The diagonal of $\fSigma_{11}$ and $\fSigma_{22}$ are populated with ones, whereas, using symmetry of correlations, $\begin{pmatrix} \fSigma_{13} & \fSigma_{23}\end{pmatrix}^T\begin{pmatrix} \fSigma_{31} & \fSigma_{32}\end{pmatrix}$ has $\left(\rho_{i3}^{1,2}\right)^2$, $i=1,\ldots,d_1$ and $\left(\rho_{3j}^{3,2}\right)^2$, $j=1,\ldots,d_2$ along the diagonal, such that $\D$ is the $(d_1+d_2)\times(d_1+d_2)$ diagonal matrix with the concatenated sequence $1-\left(\rho_{i3}^{1,2}\right)^2, i=1,\ldots,d_1; 1- \left(\rho_{3j}^{3,2}\right)^2, j=1,\ldots,d_2$ along the diagonal.

Pre- and post multiplying $\fSigma_{12|3}$ by $\D^{-1/2}$ converts $\fSigma_{12|3}$ into a correlation matrix. All of the off-diagonal elements in this matrix has the same form as the off-diagonal elements in \eqref{eq:partialmatrix} so that
\begin{align}
\rho(Z_{1,i},Z_{1,j}|Z_3) &= \frac{\rho(Z_{1,i},Z_{1,j}) - \rho(Z_{1,i},Z_3)\rho(Z_{1,j},Z_3)}{\sqrt{1-\rho^2(Z_{1,i},Z_3)} \sqrt{1-\rho^2(Z_{1,j},Z_3)}} 
, \; i,j = 1,\ldots,d_1, \label{eq:offdiag1}\\
\rho(Z_{2,i},Z_{2,j}|Z_3) &= \frac{\rho(Z_{2,i},Z_{2,j}) - \rho(Z_{2,i},Z_3)\rho(Z_{2,j},Z_3)}{\sqrt{1-\rho^2(Z_{2,i},Z_3)} \sqrt{1-\rho^2(Z_{2,j},Z_3)}}
, \; i,j = 1,\ldots,d_2; \label{eq:offdiag2} \\
\rho(Z_{1,i},Z_{2,j}|Z_3) &= \frac{\rho(Z_{1,i},Z_{2,j}) - \rho(Z_{1,i},Z_3)\rho(Z_{2,j},Z_3)}{\sqrt{1-\rho^2(Z_{1,i},Z_3)} \sqrt{1-\rho^2(Z_{2,j},Z_3)}}
, \; i = 1,\ldots,d_1; j=1,\ldots,d_2. \label{eq:offdiag3}
\end{align}

Testing for conditional independence between $\Z_1$ and $\Z_2$ given $\Z_3$ is in our context the same as testing for zero local partial correlation between all pairs of variables, a test for which is presented in the main article and discussed in technical detail later in this supplement. One can also, based on \eqref{eq:offdiag1} and \eqref{eq:offdiag2}, test for conditional independence between combinations of components given $\Z_3$. Equations \eqref{eq:offdiag1}--\eqref{eq:offdiag3} above give explicit expressions for the conditional correlation matrix of $\Z_1$ and $\Z_2$ given $\Z_3$.

\item[\textbf{c)}] \textbf{Let} $\bm{d_1 = d_2 = 1}$ \textbf{and let} $\bm{d_3}$ \textbf{be arbitrary}. This is the case usually treated in the literature, also the focus of the discussions in the main article. In this case
$$\fSigma^{11} = \begin{pmatrix} 1 & \rho_{12} \\ \rho_{21} & 1 \end{pmatrix}, \fSigma_{13} = \left\{\rho_{1i}^{1,3}\right\}, i=1,\ldots,d_3; \fSigma_{23} = \left\{\rho_{2i}^{2,3}\right\}; i=1,\ldots,d_3,$$
hence $\begin{pmatrix} \fSigma_{13} \fSigma_{23}\end{pmatrix}^T$ is of dimension $2\times d_3$, whereas $\begin{pmatrix} \fSigma_{31} \fSigma_{32}\end{pmatrix}$ is of dimension $d_3\times 2$ such that
$$\begin{pmatrix} \fSigma_{13} \\ \fSigma_{23} \end{pmatrix} \fSigma_{33}^{-1} \begin{pmatrix} \fSigma_{31} & \fSigma_32 \end{pmatrix}$$
is of dimension $2\times2$. However, as is immediately seen, no formulas like \eqref{eq:offdiag1}--\eqref{eq:offdiag3} will appear; they will be much more complicated and include sums and products of various correlations.

\item[\textbf{d)}] \textbf{Let} $\bm{d_1, d_2}$ \textbf{and} $\bm{d_3}$ \textbf{be arbitrary}. This is the general case, and it seems impossible to obtain explicit formulas for the local Gaussian partial correlations, but the general fact still holds of course: The local partial correlation matrix of $\Z_1$and $\Z_2$ given $\Z_3$ can be identified by the lower left hand block matrix of $\D^{-1/2}\fSigma_{12|3}\D^{-1/2}$, and can be estimated using pairwise local correlations in \eqref{eq:matrixdef}.
\end{itemize}

\hypertarget{chap:estimation}{%
\subsection{Estimation and estimation theory in more general situations}\label{chap:estimation}}

Sections \ref{app-loccor} -- \ref{proofpseudo} of this supplement deal with the results stated in Section 3 of the main article where \(X_1\) and \(X_2\) are scalars. The present subsection looks briefly at more general situations.

Recall that the matrix \(D\) defined in the beginning of Section \ref{special-cases} is given by \(D = \textrm{diag}(\fSigma_{12|3})\). Let us block divide the \((d_1+d_2)\times (d_1+d_2)\) diagonal matrix \(\D\) into
\[\D = \begin{pmatrix} \D_{11} & \bm{0} \\ \bm{0} & \D_{22}\end{pmatrix}.\]
The estimated partial correlation matrix between \(\X_1\) and \(\X_2\) given \(\X_3\) is then given as
\begin{equation}
\hD_{11}^{-1/2}\left(\hfSigma_{21} - \hfSigma_{23}\hfSigma_{33}^{-1}\hfSigma_{31}\right)\hD_{22}^{-1/2} \stackrel{\textrm{def}}{=} g(\hfSigma).
\label{eq:lgpc-matrix}
\end{equation}

We present the asymptotic theory for the local likelihood estimator of the local partial correlations in two situations in the main article with details of proof in Sections \ref{app-loccor} -- \ref{proofpseudo}:

\begin{itemize}
\item[\textbf{(i)}] If $\bm{d_1 = d_2 = d_3=1}$, then we can estimate the full trivariate local likelihood estimate, meaning that all the local correlations depend on all three coordinates $z=(z_1,z_2,z_3)$. The asymptotic theory for this estimator is covered by Theorem 3.1 in the main article.

\item[\textbf{(ii)}] If $\bm{d_1 = d_2 = 1}$ \textbf{and} $\bm{d_3 > 1}$, we estimate all correlations in a pairwise manner as laid out in the main article. The limiting distribution of the (scalar) local Gaussian partial correlation is in general given by 
\begin{equation}
\sqrt{nb_n^2}\left(\widehat{\alpha}(z))-\alpha(z)\right)  \stackrel {{\cal L}}{\to} {\cal N}(0,\nabla g(\frho)^{T} {\fSigma} \nabla g (\frho)),
\end{equation}
see eq. (22) in the main article, and see Theorem 2 in Otneim and Tjøstheim (2018) for a definition of $\fSigma$. The computation of $\nabla g$ is outlined in Section \ref{app-asvar}.
\end{itemize}

These results can be extended to the multivariate cases that we treat in this supplement in Section \ref{special-cases} as well. The assumptions remain identical to the assumptions in Theorem 3.1 in the main article and Theorem 3 in Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}).

\begin{itemize}
\item[\textbf{(iii)}] If $\bm{d_1}$ \textbf{and} $\bm{d_2}$ \textbf{are arbitrary, and} $\bm{d_3 = 1}$, the local partial correlations take the form \eqref{eq:offdiag3}. The limiting distribution of $\widehat\rho(Z_{1i}, Z_{2j}|Z_3)$ follows from the delta rule applied to the joint distribution of $\{\widehat\rho(Z_{1i}, Z_{2j}), \widehat\rho(Z_{1i}, Z_3), \widehat\rho(Z_{2j}, Z_3)\}$, the joint distribution of which again follows directly from Theorem 2 in Otneim and Tjøstheim (2018), who also note that the asymptotic correlation between two local correlation estimates is of smaller order if the two index pairs are distinct. The computation of $\nabla g$ is fairly straightforward in this case, by taking the appropriate derivatives in \eqref{eq:offdiag3}.

\item[(\textbf{iv})] If $\bm{d_1, d_2}$ \textbf{and} $\bm{d_3}$ \textbf{are all arbitrary}, we have the general case, where, in principle, for each $i=1,\ldots,d_1$ and $j=1,\ldots,d_2$, the local partial correlation $\rho(Z_{1i}, Z_{2j}|\Z_3)$ can be expressed as a deterministic function $g_{ij}$ of a local correlation vector $\frho$ using equation \eqref{eq:lgpc-matrix}, the asymptotic normality for which is proved as before using the delta rule and Theorem 2 of Otneim and Tjøstheim (2018). It is, however, difficult to obtain explicit expressions of the asymptotic covariance matrix in this case.
\end{itemize}

The joint asymptotic normality of the pairwise partial correlation estimates can be proved using the Cramer-Wold device. But, again, obtaining an expression for the asymptotic covariance matrix is hard in the general case.

Observe, finally, that a more general pairwise scheme can be devised, as treated in detail by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}), Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) and Jordanger and Tjøstheim (\protect\hyperlink{ref-jordanger2017nonlinear}{2020}):

\begin{itemize}
\item[\textbf{(v)}] In this scheme, we do not use a pairwise set-up for the local correlation only (the one-parameter case), but for all parameters in a bivariate normal distribution; the correlation $\rho(x_1, x_2)$, the expectations $\mu_1(x_1,x_2), \mu_2(x_1,x_2)$ and the variances $\sigma_1(x_1,x_2),\sigma_2(x_1,x_2)$ (the five-parameter case). This setup can also be simplified and fitted to a $p$-variate distribution by estimating $\mu_i = \mu_i(x_1)$ and $\sigma_i = \sigma_i(x_i)$, $i=1,\ldots,p$, marginally. Tjøstheim and Hufthammer (2013) give the asymptotic theory in this case. 
\end{itemize}

Below is a brief summary of the relevant local likelihood theory for the general pairwise case of (ii)--(iv). We also state a general consistency result in Theorem \ref{thm:A1} at the end of this section.

The pairwise log-likelihood in cases (ii) -- (iv) is given by

\begin{align}
L_{n,b}(\Z_{in}, \Z_{jn}; \rho_b(z_i, z_j)) &= \frac{1}{n}\sum_{t=1}^n K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
& \qquad\qquad\qquad - \int K_b(v-z_i, w- z_j)\psi(v,w;\rho_b(z_i, z_j)) \, \di v \di w,
\label{eq:loglikelihood}
\end{align}
where \(\Z_{in} = (Z_{i,1}, Z_{i,2}, \ldots, Z_{i,n})\) and \(\Z_{jn} = (Z_{j,1}, Z_{j,2}, \ldots, Z_{j,n})\) are the observation vectors for components \(i\) and \(j\) respectively, \(n\) is the number of observations, \(z_i, z_j, v, w \in \mathbb{R}\), \(\rho_b = \rho_{ij,b}(z_i,z_j)\) is the local Gaussian correlation parameter, \(K_b(v-z_i, w-z_j) = (b_1b_2)^{-1}K(b_1^{-1}(v-z_i))K(b_2^{-1}(w-z_j))\) is a product kernel with bandwidth \(b=b_n=(b_1,b_2) = (b_{1n}, b_{2n})\), and \(K\) is an non-negative kernel function with compact support such that \(\int K(v,w) = 1\). For a fixed bandwidth \(b\), \(\rho_{0,b}\) is the true local parameter, that is, the minimizer of the local penalty function

\begin{equation}
\int K_b(v-z_i, w-z_j)\big[\psi(v,w; \rho_b(z_i, z_j)) - \log(\psi(v,w; \rho_b(z_i, z_j)))f(v,w)\big] \di v \di w,
\label{eq:penalty}
\end{equation}
where \(f\) is the density function of \((Z_i, Z_j)\). Note in particular that \(\rho_{0,b} \rightarrow \rho_0\) as \(b\rightarrow0\), which is the true local correlation parameter. The reader may consult Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) for more details and regularity conditions in this regard. The corresponding estimate \(\rho_{n, b}(z_i,z_j)\) is obtained by maximizing the log-likelihood in \eqref{eq:loglikelihood}. Note that, writing \(\rho\) for \(\rho_b\),
\begin{equation}
L_{n,b}(\rho) = \frac{1}{n}\sum_{t=1}^n L^{(b)}(Z_{i,t}, Z_{j,t}; \rho),
\end{equation}
where
\begin{align}
L^{(b)}(Z_{i,t}, Z_{j,t}; \rho) &= K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
& \qquad\qquad\qquad \int K_b(v-z_i, w- z_j)\rho_b(v,w;\psi_b(z_i, z_j)) \, \di v \di w.
\label{eq:A16}
\end{align}

We must next proceed by introducing some new concepts before stating the equivalents of assumptions A1 -- A2 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}). These assumptions are similar but distinct from those used in Theorem 3.1 in the main article and Theorem 2 in Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}). They will also be used to prove the consistency result in Theorem \ref{thm:A1} at the end of this section, and they will be part of the assumptions used to prove the validity of the bootstrap in Section \ref{validity-bootstrap}.

For a given function \(g\), the process \(\{g(Z_{i,t}, Z_{j,t}; \rho)\}\) is said to be \(r\)-dominated on \(\Theta\) uniformly in \(t\), where \(t = 1,\ldots,n\), if there exists a function \(D_t: \mathbb{R} \rightarrow \mathbb{R}\) such that \(|g(Z_{i,t}, Z_{j,t}; \rho)| \leq D_t\) for all \(\rho \in \Theta\), \(t = 1,\ldots, n\) and \(D_t\) is \(\mathcal{B}\)-measurable such that for some \(\Delta > 0\), \(||D_t||_r \stackrel{\textrm{def}}{=} \left(\int |D_t(u)|^r\di u\right)^{1/r} \leq \Delta < \infty\) for all \(t = 1,\ldots,n\), where \(\mathcal{B}\) is the Borel \(\sigma\)-algebra. In this particular case the location \((z_i, z_j)\) is fixed, and \(\Theta = [-1,1]\). Moreover, the process \(\{g(Z_{i,t}, Z_{j,t}; \rho)\}\) is said to be Lipshitz continuous on \(\Theta\) if

\[|\{g(Z_{i,t}, Z_{j,t}; \rho)\} - \{g(Z_{i,t}, Z_{j,t}; \rho')\}| \leq C_t|\rho - \rho'|\]
almost surely in \(\mathbb{P}\) for all \(\rho,\rho'\in\Theta\) and with a sufficiently large constant \(M\) such that \(C_t\leq M\). Here \(\mathbb{P}\) is the probability measure on the probability space in which \(\{Z_{i,t}, Z_{j,t}\}\) is defined. We can now state conditions A1 -- A2 which are basic assumptions for proving consistency and asymptotic normality of \(\rho_{n,b}\):

\begin{itemize}
\item[\textbf{A1}] $(\Omega, \mathcal{F}, \mathbb{P})$ is a complete probability space on which is defined a sequence of random functions $\{L_{n,b}: \Omega\rightarrow\overline{\mathbb{R}}\}$, where $\overline{\mathbb{R}}$ is the extended real line. The parameter set $\Theta$ is compact (trivially fulfilled in our case because $\Theta = [-1,1]$). The random process $\{Z_{i,t}, Z_{j,t}\}$ is $\mathcal{F}_t$-measurable, stationary and $\alpha$-mixing with mixing coefficients $\alpha_k = O(\alpha^k)$, $\alpha \in (0,1)$ for all $t$, and with a bivariate density having support in all of $\mathbb{R}^2$. Here $\mathcal{F}_t$ is the $\sigma$-algebra generated by $\{Z_{i,s}, Z_{j,s}, s \leq t\}$.
\item[\textbf{A2}] The parameter value $\rho_{0,b}$ is the unique maximizer of $\E(L_{n,b}(\rho)): \Theta \rightarrow \overline{\mathbb{R}}$, with $\E(L_{n,b}(\rho))$ is defined in \eqref{eq:penalty}.
\end{itemize}

The conditions A1-- A2 are quite mild and discussed by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}). Note that they have been formulated for a scalar parameter \(\rho = \rho_{ij}\). To have them fulfilled for all possible combinations of \(i,j\), the assumption A1, for example can be easily formulated for the \(\{\Z_t\}\) process. Consistency for the local parameter estimates is required for consistency of the bootstrap:

\begin{thm} Let assumptions A1 and A2 hold, and let $\{L^{(b)}(Z_{i,t}, Z_{j,t};\rho)\}, \rho \in \Theta$ be Lipshitz continuous on $\Theta$ and 4-dominated on $\Theta$ uniformly in $t$ and $n$. Moreover, assume that as $b=b_n\rightarrow 0$, $\sigma^2_b \stackrel{\textrm{def}}{=} \Var\left(L^{(b)}(Z_{i,t}, Z_{j,t};\rho)\right) \rightarrow \sigma^2 < \infty$ and $\mu_b \stackrel{\textrm{def}}{=} \E\left(L^{(b)}(Z_{i,t}, Z_{j,t};\rho)\right) \rightarrow \mu < \infty$. Then $\rho_{n,b} - \rho_{0,b} \stackrel{\mathbb{P}}{\rightarrow} 0$ for a fixed $b$ and $\rho_{n,b} - \rho_{0} \stackrel{\mathbb{P}}{\rightarrow} 0$ as $b \rightarrow 0$.
\label{thm:A1}
\end{thm}

The proof of Theorem \ref{thm:A1} is mostly identical to the proof of Theorem 2 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) with trivial modifications to take into account the scalar nature of the parameter \(\rho_{ij}\) in the present case. A distributional limit theorem analogous to Theorem 2.2 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) can also be rephrased and proved as in that paper, but we refrain from doing this given the distributional results already stated in Section 3 of the main part of the paper.

\hypertarget{app-loccor}{%
\section{DETAILS REGARDING THE PROOF OF THEOREM 3.1}\label{app-loccor}}

For fixed bandwidths we have from standard arguments, such as those provided by Hjort and Jones (\protect\hyperlink{ref-hjort1996locally}{1996}), that
\begin{equation}
\sqrt{nb^3}\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(0, \Jb^{-1}\Mb(\Jb^{-1})^T),
\label{eq:fixed-h}
\end{equation}
which corresponds to the usual rate for nonparametric density estimation. As \(\hh\rightarrow0\), however, we must take extra care when considering the asymptotic covariance matrix. Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}), Section 4, write the Taylor expansions of \(\Mb\) and \(\Jb\), being functionals of three variables \((x_1, x_2, x_3)\) or \((z_1,z_2,z_3)\), as
\[\Mb = I_{\bm{M}} + II_{\bm{M}} + o(b^2) \,\,\, \textrm{ and } \Jb = I_{\bm{J}} + II_{\bm{J}} + o(b^2),\]
and consider each term. For example, we write their first term in the expansion of \(\Mb\) in the three-variate case as follows:
\[I_{\bm{M}} \sim \int K^2(w_1, w_2, w_3)Ab_wb_w^TA^Tf(\z + b_1w_1 + b_2w_2 + b_3w_3)\,\textrm{d}w_1\,\textrm{d}w_2\,\textrm{d}w_3,\]
where \(K\) is a product kernel, and, in our case, \(b_w\) is the vector defined by \(b_w^T = \begin{pmatrix} 1 & b_1w_1 & b_2w_2 & b_3w_3 \end{pmatrix}\), and \(A\) is the \(3\times 4\) matrix \(A = \begin{pmatrix} v & v_{z_1} & v_{z_2} & v_{z_3}\end{pmatrix}\) with \(\fv^T = \begin{pmatrix} v_1 & v_2 & v_3 \end{pmatrix}\), \(v_i = \partial L_n/\partial \rho_i\), where the index \(i=1,2,3\) represents the three local correlations, and \(v_{z_i} = \partial \fv/\partial z_i\). In the next step, they compute the matrix

\noindent \(\int K^2(w_1, w_2, w_3)b_wb_w^T\,\textrm{d}w_1\,\textrm{d}w_2\,\textrm{d}w_3\), which, by omitting all constant factors and exploiting that \(\int w^k K^2(w)\,\textrm{d}w = 0\) for \(k = 1,3\), in our case becomes the diagonal matrix
\[H = \begin{pmatrix} 1 & 0 & 0 & 0 \\ 0 & b_1^2w_1^2 & 0 & 0 \\ 0 & 0 & b_2^2w_2^2 & 0 \\ 0 & 0 & 0 & b_3^2w_3^2 \end{pmatrix}.\]

The second term defining \(\Mb\) has smaller order than \(I_{\bm{M}}\) as \(b \rightarrow 0\), while the first term of \(\Jb\) can be treated exactly as \(I_{\bm{M}}\) above resulting in a matrix of order \(b^2\).

The second term in \(\Jb\) can similarly be written as \(II_J = B(\psi(\z, \theta_h(\z)) - f(\z)) + o(b^2)\), where \(B = \nabla \fv(\z, \theta_b(\z))\) are second derivatives of the local likelihood function with respect to the local correlations, and \(\psi(\z, \theta_b(\z)) - f(\z) = O(b^2)\) according to Hjort and Jones (\protect\hyperlink{ref-hjort1996locally}{1996}). Hence, it follows that the leading term of \(\Jb^{-1}\Mb(\Jb^{-1})^T\) is \(O(b^{-2})\), which again means that the convergence rate in eq. \eqref{eq:fixed-h} above must be modified accordingly in order to balance the convergence in eq. (20) in the main article. Extracting an analytic expression for the leading term of the asymptotic covariance matrix is, as mentioned by Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}), possible by means of symbolic manipulation software, but the expression is very complicated and of little practical use. See also the corresponding proof of Theorem 3 in Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}).

\hypertarget{app-asvar}{%
\section{THE ASYMPTOTIC VARIANCE OF THE LGPC}\label{app-asvar}}

The limiting distribution of the LGPC is given in eq. (22) in the main article. In this section, we will present the basic steps needed to work out the value of \(\nabla g(\frho)\) in the simplest case where \(d_1 = d_2=1\). In the more general situation it is not feasible to obtain analytic expressions for the asymptotic variance, and the bootstrap should be used. For the sake of this particular argument, we may simplify notation quite a bit, leaving out what we do not need: We drop the \(\z\)-dependence, and simply write \(\fSigma = \fSigma_{12|3}(\z)\) for the \(2\times2\) local partial covariance matrix between \(Z_1\) and \(Z_2\) given \((Z_3,\ldots,Z_p)\). The vector of local correlations \(\frho\) is indexed by \(k\), which in turn means that \(k\) corresponds to a specific \emph{pair} \((i,j)\): \(\rho_k = \rho_{ij}\), with \(\rho_{ij} = \rho_{ji}\). Let \(\fSigma^{(k)}\) be the element-wise matrix of partial derivatives
\[\fSigma^{(k)} = \frac{\partial \fSigma}{\partial \rho_k} = \frac{\partial \fSigma}{\partial \rho_{ij}},\]
which means that element \(k\) in the gradient \(\nabla g(\frho)\) is given by
\begin{equation}
\left\{\nabla g(\frho) \right\}_k = \frac{\fSigma_{11}\fSigma_{22}\fSigma^{(k)}_{12} - \fSigma_{12}\left(\fSigma_{11}^{(k)}\fSigma_{22} + \fSigma_{11}\fSigma_{22}^{(k)}\right)}{\left(\fSigma_{11}\fSigma_{22}\right)^2},
\label{eq:gradientelement}
\end{equation}\\
where the double subscript to \(\fSigma\) and \(\fSigma^{(k)}\) here means matrix elements. In order to avoid confusion with the double subscripts defining matrix \emph{partitions} in the preceding sections, we re-label the matrix blocks \(\R^{11}\), \(\R^{12}\) and \(\R^{22}\) to \(\R_1\), \(\R_2\) and \(\R_3\) respectively, with \(\R^{21} = \R_2^T\). Following basic differentiation rules for matrices (see e.g.~Van den Bos (\protect\hyperlink{ref-van2007parameter}{2007})), we have that, element-wise,
\begin{equation}
\frac{\partial\left(\R_2\R_3^{-1}\R_2^T\right)}{\partial\rho_k} = \frac{\partial\R_2}{\partial\rho_k}\R_3^{-1}\R_3^T - \R_2\R_3^{-1}\frac{\partial \R_3}{\partial \rho_k}\R_3^{-1}\R_2^T + \R_2\R_3^{-1}\frac{\partial\R_2^T}{\partial\rho_k}.
\label{eq:derivative}
\end{equation}
The value of the gradient \eqref{eq:gradientelement} depends on \(k\), and when differentiating (7) with respect to \(\rho_k\) there are three different cases we must consider:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  If \(\rho_k = \rho_{12}\) the second term in (7) vanishes, because \(\rho_{12}\) is only present in \(\R_1\). Indeed,
  \[\R^{11} = \begin{pmatrix}1 & \rho_{12} \\ \rho_{12} & 1 \end{pmatrix},\]
  which means that
  \[\fSigma^{(k)} = \begin{pmatrix}0&1\\1&0\end{pmatrix} - \begin{pmatrix}0&0\\0&0\end{pmatrix},\]
  and we easily see from \eqref{eq:gradientelement} that in this particular case, \(\left\{\nabla g(\frho) \right\}_k = 1\).
\item
  If \(\rho_k = \rho_{ij}\) where \(i \in \{1,2\}\) and \(j \in \{3, \ldots,p\}\), then \(\rho_{ij}\) is an element in \(\R_2\), but not \(\R_1\) or \(\R_3\). For that reason, we are left with the second term in (7), and furthermore, the middle term in the derivative \eqref{eq:derivative} above will also vanish. The \(2\times(p-2)\) matrix \(\partial\R_{12}/\partial\rho_{ij}\) will contain only zeros except for a \(1\) in the position of \(\rho_{ij}\); and \(\partial\R^{21}/\partial\rho_{ij}\) is its transposed. Write \(\fC = \left(\R^{22}\right)^{-1}\R^{21}\), and we see that \(\fSigma^{(k)}\) in this case is a simple linear combination of the elements in \(\fC\):
  \[\fSigma^{(k)} = \frac{\partial\R^{12}}{\partial\rho_{ij}}\fC + \fC^T\frac{\partial\R^{21}}{\partial\rho_{ij}},\]
  from which we select the elements needed to calculate \(\left\{\nabla g(\frho) \right\}_k\) through \eqref{eq:gradientelement}.\\
\item
  In this case, we look at the derivative of \(g(\frho)\) with respect to the local correlations located in \(\R_3\), that is, \(\rho_{ij}\) for \(i,j \in \{3,\ldots,p\}\). We only need to keep the middle term in \eqref{eq:derivative}. We have that
  \[\fSigma^{(k)} = -\fC^T\frac{\partial\R_3}{\partial\rho_j}\fC,\]
  where \(\partial\R_3/\partial\rho_k\) is a symmetric \((p-2)\times(p-2)\) matrix of zeros everywhere, except for 1s in the positions corresponding to \(\rho_k\) with respect to which we perform the differentiation. Finally, equation \eqref{eq:derivative} provides the final expression for \(\left\{\nabla g(\frho) \right\}_k\).
\end{enumerate}

\hypertarget{proofpseudo}{%
\section{PROOF OF THEOREM 3.2}\label{proofpseudo}}

Theorem 3.2 states that using marginally standard normal pseudo-observations instead of exactly standard normally distributed observations for estimating the LGPC does not change the conclusions in Theorem 3.1 in the main article and Theorem 2 in Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}). The result may be expected, as the empirical distribution functions \(\widehat{F}_j\), \(j = 1,\ldots,p\) converge at a rate faster than the local correlations (\(n^{-\frac{1}{2}}\) vs.~\((nb^5)^{-\frac{1}{2}}\) or \((nb^2)^{-\frac{1}{2}}\)) depending on which theorem we refer to), but we must attend to some details nevertheless.

Proving joint asymptotic normality of the local Gaussian correlations with marginally normally distributed variables \((Z_1, \ldots, Z_p) = \left(\Phi^{-1}(F_1(X_1)), \ldots, \Phi^{-1}(F_p(X_p))\right)\), where \(F_j\) is the cdf of \(X_j\) for \(i = 1,\ldots,p\), relies on proving asymptotic normality for the variables
\[Y_n(\z) = \frac{1}{n}\sum_{t=1}^nK_b(\Z_t - \z)\fv(\Z_t, \frho_0),\]
where \(\fv(\cdot) = \partial L(\cdot)/\partial\frho\) in the same way as in Section \ref{app-loccor} in this supplement. This has been done by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}) in the iid case using arguments from Schervish (\protect\hyperlink{ref-schervish1995theory}{2012}), and by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}) in the \(\alpha\)-mixing case, using arguments from Fan and Yao (\protect\hyperlink{ref-fan2008nonlinear}{2008}). We need here to prove asymptotic normality of the variables

\[\hY_n(\z) = \frac{1}{n}\sum_{t=1}^nK_b(\hZ_t - \z)\fv(\hZ_t,\frho_0),\]
with \(\hnZ_t = \Phi^{-1}(F_n(\X_t))\), and \(F_n(\X_t)\) is the application of the marginal empirical distribution functions to the corresponding components of the vector \(\X_t\).

In order to ease notation we state the proof in the simplified case only, where the local correlations are estimated using \emph{pairs} \((i,j)\) of variables. It can be extended to the trivariate case by adding the appropriate terms in the derivations below, without changing the final conclusion. Write \(\hY_n(\z) = Y_n(\z) - (Y_n(\z) - \hY_n(\z))\) and do a Taylor expansion of \(Y_n(\z)\) around \(F_n(\X_t)\). Since we use product kernels, we then need derivatives of (assuming without loss of generality that \(K(x) = K(-x)\))

\[H_{\z,\hh} = K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right)K\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right)v(\Phi^{-1}(\y)),\]
where \(\y = F(\x)\), \(y_i = F(x_i)\), \(i=1,2\), and we write \(K(x_1,x_2) = K(x_1)K(x_2)\). Writing \(k(z) = K'(z)\) and \(v_{y_i} = \partial v_i/\partial y_i,\)
\begin{align*}
\frac{\partial H_{\bm{z},b}}{\partial y_1} 
&= k\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right) \frac{v(\Phi^{-1}(\y))}{b\phi(\Phi^{-1}(y_1))} \\
& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad +  K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right) \frac{v_1(\Phi^{-1}(\y))}{\phi(\Phi^{-1}(y_1))}, \\
\frac{\partial H_{\bm{z},b}}{\partial y_2} 
&= K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right) k\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right) \frac{v(\Phi^{-1}(\y))}{b\phi(\Phi^{-1}(y_2))} \\
& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad+  K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right) \frac{v_2(\Phi^{-1}(\y))}{\phi(\Phi^{-1}(y_1))}.
\end{align*}
A typical term in the Taylor expansion of \(Y_n(\z)\) takes the following form:

\begin{align*}
& K_b\left(\Phi^{-1}(F(\X_t)) - \z\right)v\left(\Phi^{-1}(F(\X_t)), \rho_0\right) = \\
& \qquad K_b\left(\Phi^{-1}(F_n(\X_t)) - \z\right)v\left(\Phi^{-1}(F_n(\X_t)), \rho_0\right) \\
& \qquad\qquad + k\left(\frac{z_1 - \Phi^{-1}(F_n^*(X_{1,t}))}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(F_n^*(X_{2,t}))}{b}\right) \frac{v(\Phi^{-1}(F_n^*(\X_t)))}{b\phi(\Phi^{-1}(F_n^*(X_{1,t})))}\left(F(X_{1,t}) - F_n(X_{1,t})\right) \\
& \qquad\qquad + K\left(\frac{z_1 - \Phi^{-1}(F_n^*(X_{1,t}))}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(F_n^*(X_{2,t}))}{b}\right) \frac{v_1(\Phi^{-1}(F_n^*(\bm{X}_t)))}{\phi(\Phi^{-1}(F_n^*(X_{1,t})))}\left(F(X_{1,t}) - F_n(X_{1,t})\right) \\
&\qquad + \textrm{ analogous terms for the second (and possibly third) index involving } \left(F(X_{2,t}) - F_n(X_{2,t})\right), 
\end{align*}
where \(F_n^*\) comes from the mean value theorem. The challenge in proving the desired result is to control the behavior of quantities on the form \(1/\phi(\Phi^{-1}(F(x)))\), because this fraction tends to infinity as \(|x| \rightarrow \infty\). The key to this problem is the assumption that the kernel function \(K(\cdot)\), and thus its derivative \(k\), have compact support. This means, that in the expressions above, \(K = k \equiv 0\) if for some \(M > 0\)
\[\frac{|z_1 - \Phi^{-1}(F_n^*(X_{1,t}))|}{b} \geq M,\]
or
\[|z_1 - \Phi^{-1}(F_n^*(X_{1,t}))| \geq Mb,\]
or, by removing the absolute value signs,
\[ \Phi^{-1}(F_n^*(X_{1,t})) \leq z_1 - Mb\,\, \textrm{ or }\,\, \Phi^{-1}(F_n^*(X_{1,t})) \geq z_1 + Mb,\]
where \(M\) may be large. The same reasoning applies of course to the second index as well. Letting \(n\rightarrow\infty\) and using consistency of the empirical distribution function, it follows that the kernel function term is zero if

\[X_{1,t} \leq F^{-1}(\Phi(z_1 - Mb)) \,\,\, \textrm{ or } \,\,\, X_{1,t} \geq F^{-1}(\Phi(z_1 + Mb)).\]
The same reasoning applies to the derivative of the kernel function, where we consider the function \(k/b\) instead of \(K\). All of this implies that the challenge of controlling the magnitude of the Taylor terms above as \(x\rightarrow\infty\) disappears, since, taking boundedness of the other terms into account,

\[ k\left(\frac{z_1 - \Phi^{-1}(F_n^*(X_{t,1}))}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(F_n^*(X_{t, 2}))}{b}\right) \frac{v(\Phi^{-1}(F_n^*(\X_t)))}{b\phi(\Phi^{-1}(F_n^*(X_{t,1})))}\]
and
\[K\left(\frac{z_1 - \Phi^{-1}(F_n^*(X_{t,1}))}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(F_n^*(X_{t, 2}))}{b}\right) \frac{v_1(\Phi^{-1}(F_n^*(\X_t)))}{\phi(\Phi^{-1}(F_n^*(X_{t,1})))}\]
are bounded almost surely as \(n\rightarrow\infty\) and \(b\rightarrow 0\).

All this means that if we can prove that for \(i=1,2\),
\[\frac{1}{n}\sum_{t=1}^n(F_n(X_{i,t}) - F(X_{i,t})) \stackrel{\mathbb{P}}{\rightarrow} 0\]
at an appropriate rate, then we are done by Slutsky's Theorem.

In the following we omit the index \(i=1,2\). We need a bound on the supremum of \(|F_n(X_t) - F(X_t)|\) and \(|F_n(x) - F(x)|\) as \(n \rightarrow \infty\). According to Arcones (\protect\hyperlink{ref-arcones1995law}{1995}), the best law of the iterated logarithm has been obtained by Rio (\protect\hyperlink{ref-rio1995functional}{1995}). One statement of this is as follows:

Let \(f(x) \in L_p\), \(2<p\leq\infty\), \(\sum\alpha_k k^{2/(p-2)} < \infty\), then the sequence \(f(X_t)\) satisfies the compact LIL (\emph{law of the iterated logarithms}), i.e.
\[\left\{\left(n2\log\log n\right)^{-\frac{1}{2}} \sum_{t=1}^n(f(X_t) - E[f(X_t)]) \right\}_{n=1}^{\infty}\]
is relatively compact and its limit set is \([-\sigma, \sigma]\) where
\[\sigma^2 = \textrm{Var}\Big(f(X_t)\Big) + 2\sum_{s=1}^{\infty}\textrm{Cov}\Big(f(X_t), f(X_s)\Big).\]
For a fixed \(x\), take \(f(X_t) = I(X_t\leq x)\). Then E\((f(X_t)) = F(x)\) and \(f(x)\in L_p\) for all \(p\), and
\[\left\{\left(n2\log\log n\right)^{-\frac{1}{2}} \sum_{s=1}^n \left(I(X_s \leq x) - F(x) \right) \right\}\]
has its limit set in \([-\sigma, \sigma]\). Here
\[\textrm{Var}(I(X_t \leq x)) = F(x)(1-F(x)) \leq \frac{1}{4},\]
and according to Davydov's Lemma (see e.g.~Fan and Yao (\protect\hyperlink{ref-fan2008nonlinear}{2008}) p.~278),
\[\big| \textrm{Cov}(I(X_t \leq x), I(X_s \leq x))\big| \leq 8 \|I(X_t \leq x)\|_p \|I(X_s \leq x)\|_q \alpha(|s-t)^{1-p^{-1}-q^{-1}}, \,\,\, p,q>1, \,\,\, \frac{1}{p}+\frac{1}{q} < 1,\]
with \(\| I(X_t \leq x)\|_p = \left(\textrm{E}|I(X_t \leq x)|^p\right)^{1/p}\). We may for example take \(p=q=4\), leading to \(\| I(X_t \leq x) \|_4 = F(x)^{1/4} \leq 1\).

Since we have assumed exponential mixing (much weaker conditions suffice), then Rio's condition \(\sum \alpha_k k^{2/(p-2)} < \infty\) is satisfied, and the bounds of \(-\sigma\) and \(\sigma\) can be made independent of \(x\).

Having
\[F_n(x) = \frac{1}{n}\sum_{t=1}^n I(X_t \leq x)\]
means that
\[n^{\frac{1}{2}}[2\log \log n]^{-\frac{1}{2}} (F_n(x) - F(x))\]
has its limits in \([-\sigma, \sigma]\) or that the supremum of \(|F_n(x) - F(x)|\) is almost surely bounded by \(n^{-1/2}(2\log\log n)^{1/2}\).

This immediately implies that
\[\frac{1}{n}\sum_{t=1}^n\left(F_n(X_t) - F(X_t)\right)\]
is almost surely (and consequentially in probability) bounded by \(n^{-\frac{1}{2}}(2 \log \log n)^{\frac{1}{2}}\) which for all conceivable bandwidths is faster than the convergence rates of the CLT for \(Y_n(\z)\), which is \((nb^2)^{-\frac{1}{2}}\).

\hypertarget{validity-bootstrap}{%
\section{THE VALIDITY OF THE BOOTSTRAP FOR ESTIMATION}\label{validity-bootstrap}}

The asymptotic theory for the LGPC involves complicated expressions that may be difficult to implement in practice. It is then natural to use the bootstrap instead to construct confidence intervals. In order to accommodate the time series case, we must use the block bootstrap, and we are thus faced with the problem of proving the validity of the block bootstrap.

We have seen that the building blocks of the asymptotic estimation theory of the LGPC between the vectors \(\Z_1\) and \(\Z_2\) given \(\Z_3\) are the pairwise local Gaussian correlations contained in the local correlation matrix \(\fSigma\). Let us simply use the notation \(\rho_{ij} = \rho_{ij}(z_i, z_j)\) for one of these local correlation and \(\hrho_{ij}\) for its estimate as outlined in the main article. If we can prove that the block bootstrap is valid for such a pairwise local likelihood estimate, then it follows fairly straightforwardly that it is also valid for the appropriate vector of such pairwise estimates used in the asymptotic theory laid out above. Note that the 3-coordinate case (i) in Section \ref{chap:estimation} requires essentially the same arguments.

We can use the bootstrap theory by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) more or less directly, which in turn is based on Gon\c{c}alves and White (\protect\hyperlink{ref-gonccalves2002bootstrap}{2002}, \protect\hyperlink{ref-gonccalves2004maximum}{2004}), but we spell out some details below. First, we need to specify the analogue of the conditions A1--A5 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}), where we adapt their notation to our situation by replacing the parameter \(\theta\) with \(\rho_{ij}\), and the time series \(\{X_t, Y_t\}\) by our multivariate observations \(\{\Z_t\}\), possibly a multivariate time series. Since \(\rho = \rho_{ij}\) is estimated using the two corresponding data vectors only, we can, for the time being and without loss of generality, assume that the observation series is bivariate: \({\Z_t} = \{Z_{i,t}, Z_{j,t}\}\). Note further that it suffices to consider the true pseudo-observations \(\{\Z_t\}\) instead of the empirical pseudo observations \(\{\hZ_t\}\) in the following because the difference between them is asymptotically negligible according to Theorem 3.2 in the main article.

To state the bootstrap results we need a few more concepts and three additional conditions. We block bootstrap the entire vector series \(\{\Z_t\}\). This automatically leads to the same block-bootstrap for any pair of components which we have denoted by \(\{Z_{i,t}, Z_{j,t}\}\) above. Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) use both the block bootstrap as well as the stationary bootstrap, and the results below hold irrespective of this particular choice. We will therefore not distinguish between them here, but rather refer the reader to Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) for further details on this distinction. Bootstrapped quantities will be denoted with a star (\(*\)) in the following.

Before we introduce the specific bootstrap conditions \textbf{A4} and \textbf{A5} below, we need to state the following condition. The notation of equations \eqref{eq:loglikelihood}--\eqref{eq:A16} is used.

\begin{itemize}
\item[\textbf{A3}] For a fixed bandwidth $b$, and using stationarity,
$$J_{n,b} = \E\left(-\nabla^2L_{n,b}(\rho_{0,b})\right) = J_b = \E\left(-\nabla^2 L^{(b)}(Z_{i,t}, Z_{j,t};\rho_{0,b})\right)$$
is of order $O(1)$, and continuous on $\Theta$ uniformly in $n$. Further,
$$M_{n,b} = \Var\left(\sqrt{nb_1b_2}\nabla L_{n,b}(\rho_0,b)\right)$$
is $O(1)$ and uniformly positive. Here $\nabla$ is the gradient with respect to $\rho$, which in the present scalar parameter case is simply the differential operator $\di/\di\rho$.
\end{itemize}

The bootstrap lives on the probability space \((\Lambda, \mathcal{G}, \mathbb{P}^*)\). The outcomes \(\lambda \in \Lambda\) depend on the outcomes \(\omega \in \Omega\) in the original probability space, and we can now state two more assumptions:

\begin{itemize}
\item[\textbf{A4}] The bootstrap probability space $(\Lambda, \mathcal{G}, \mathbb{P}^*)$ is a complete probability space for all $\omega \in \Omega$. Let $L_{n,b}$ be defined as in \eqref{eq:loglikelihood} with $\rho \in \Theta$. Let $\{L^*_{n,b}: \Lambda\times\Omega\times\Theta \rightarrow \overline{\mathbb{R}}\}$ be a sequence of random functions such that $L_{n,b}^*(\rho) = L_{n,b}((\Z_{in}^{*}(\lambda,\omega),\Z_{jn}^*(\lambda,\omega)),\rho)$ where $Z_{i,t}^*(\lambda,\omega) = Z_{i\tau_t^{Z}(\lambda)}(\omega)$, $Z_{j,t}^*(\lambda,\omega)= Z_{j\tau_t^{Z}(\lambda)}(\omega)$, where $\tau_t^{Z}: \Lambda \to \mathbb{N},\; \omega \in \Omega, \; \lambda \in \Lambda$ are vectors of random indices representing the block bootstrap operation on $\{\Z_t\}$. Moreover, the block length $l_n$ is such that $l_n=o(\sqrt{n})$ as $n \to \infty$. 
\item[\textbf{A5}] For $\rho \in \Theta$ and for every $t=1,\ldots,n$, $\{L^{(b)}(Z_{i,t},Z_{j,t}),\rho\}$ is Lipschitz continuous on $\Theta$, $\{\nabla L^{(b)}((Z_{i,t},Z_{j,t}),\rho)\}$ is 6-dominated  on $\Theta$ uniformly in $t,n$, and $\{\nabla^2 L^{(b)}((Z_{i,t},Z_{j,t}),\rho)\}$ is Lipschitz continuous on $\Theta$ and 2-dominated on $\Theta$ uniformly in $t,n$ ($r$-dominated is defined after \eqref{eq:A16}).
\end{itemize}

Assumption A4 sets the stage for the block bootstrapping, whereas A5 is needed in proving the validity of the bootstrap central limit theorem. The assumption A4 is not restrictive, and again we refer to Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) for more discussion.

In order to state the bootstrap validity theorems, we need to specify convergence in the space \((\Lambda, {\mathcal G}, \mathbb{P}_{\omega}^*)\) . This is done as by Gon\c{c}alves and White (\protect\hyperlink{ref-gonccalves2002bootstrap}{2002}, \protect\hyperlink{ref-gonccalves2004maximum}{2004}). First, for a random variable \(Y_n^*\), we write \(Y_n^* \stackrel {\mathbb{P}_{\omega}^*,\mathbb{P}}{\rightarrow} 0\) if for any \(\varepsilon, \delta > 0\),
\[
\lim_{n \to \infty} \mathbb{P}(\omega:\mathbb{P}_{\omega}^*(\lambda:|Y_n^*(\lambda,\omega)| > \varepsilon) > \delta) = 0.
\]
Further, we write \(Y_n^* \stackrel {d_{\mathbb{P}_{\omega}^*}}{\longrightarrow} {\mathcal N}(0,1)\;\mbox{prob}-\mathbb{P}\), if for every sub-sequence \(\{n^{\prime}\}\), there exists a further sub-sequence \(\{n^{\prime \prime}\}\) such that \(Y_{n^{\prime \prime}} \stackrel {d_{\mathbb{P}_{\omega}^{*}}} {\longrightarrow} {\cal N}(0,1)\) almost surely , see Gon\c{c}alves and White (2004, page 210). This definition is based on the fact that convergence in probability implies almost sure convergence for such kinds of sub-sequences (see Theorem 20.5 of Billingsley (\protect\hyperlink{ref-bill:2008}{2008})).

We proceed by stating the validity of the bootstrap in terms of consistency and a central limit theorem:

\begin{thm}
Let assumptions A1, A2 and A4 hold, and  let $\{L^{(b)}((Z_{i,t},Z_{j,t}),\rho)\}$, $\rho \in \Theta$ be Lipschitz continuous and 4-dominated on $\Theta$ uniformly in $t,n$. Moreover, assume that, as $b=b_n \to 0, \sigma_b^2 \stackrel{\textrm{def}}{=} \Var(L^{(b)}((Z_{i,t},Z_{j,t}),\rho) \to \sigma^2 < \infty$ and $\mu_b \stackrel{\textrm{def}}{=} \E(L^{(b)}(Z_{i,t},Z_{j,t}),\rho)) \to \mu < \infty$, for all $t=1,\ldots,n$ and $\rho \in \Theta$. 
Then,
\begin{equation}
\rho_{n,b}^* - \rho_{b,n} \stackrel {\mathbb{P}_{\omega}^{*},\mathbb{P}}{\to} 0.
\label{eq:thmB}
\end{equation}
\label{thm:E1}
\end{thm}

\begin{thm}
Under the conditions 1--5 of Theorems 3.1 and 3.2 in the  main article and assumptions A1--A5 stated above,

\begin{equation}
\sqrt{nb_1b_2} M_{n,b}^{-1/2}J_{n,b}(\rho_{n,b}^{*} - \rho_{n,b}) \stackrel {d_{\mathbb{P}_{\omega}^{*}}}{\to} {\cal N}(0,1)\mbox{prob}-\mathbb{P}.
\end{equation}
\label{thm:E2}
\end{thm}

Theorems \ref{thm:E1} and \ref{thm:E2} are formulated in their simplest case here, which concerns only one local parameter. A more complicated version is given by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) concerning the five parameter case. The proofs are essentially the same and are spelled out in some detail in the supplementary material of that paper as Theorems D1 and D2.

Theorems \ref{thm:E1} and \ref{thm:E2} concern only one local correlation. The local \emph{partial} correlation, however, is a function of a vector of local correlations. The form of this function and the vector of local correlations depend on the generality and the type of model considered as seen from the alternatives (i) -- (v) in Section \ref{chap:estimation} of this supplement. Corollaries \ref{cor:E3} and \ref{cor:E4} cover the general situation with pairwise model and pairwise coordinates, including alternative (i)', which is alternative (i) but restricted to pairwise coordinates. Corollary \ref{cor:E3} treats bootstrap consistency and Corollary \ref{cor:E4} gives a bootstrap CLT:

\begin{cor}
Assume the conditions of Theorem \ref{thm:E1}. Let $\alpha_{n,b}(i,j|\Z_3) \stackrel{\textrm{def}}{=} \widehat \rho(Z_{1i}, Z_{2j}|\Z_3)$ be the $i,j$th element of the estimated partial correlation matrix of $(\Z_1, \Z_2)$ given $\Z_3$. This depends on scalar coordinates $(z_1, z_2)$ and a vector coordinate $\z_3$. Note that $\alpha_{n,b}(i,j|\Z_3)$ is a continuous function of a vector of local correlations according to one of the alternatives (i)', (ii)--(v) in Section \ref{chap:estimation}. Then $\alpha_{n,b}(i,j|\Z_3) - \alpha^*_{n,b}(i,j|\Z_3) \buildrel {\mathbb{P}_{\omega}^{*},\mathbb{P}} \over {\longrightarrow} 0$. 
\label{cor:E3}
\end{cor}

\emph{Proof.} Denote by \(\frho = \frho(\Z)\) the vector of local correlations according to one of the alternatives (i)', (ii)--(v). This vector has a finite number of components and equation \eqref{eq:thmB} guarantees the bootstrap consistency of each component. The bootstrap consistency for the vector follows immediately and the proof is concluded by the continuous mapping theorem. \(\blacksquare\)

The next results gives the asymptotic normality of the bootstrapped local partial correlation. We let \(\alpha(i,j)|\Z_3) = g_{ij}(\frho(\Z)))\) be the local partial correlation as defined above and expressed as a function \(g_{ij}\) of the vector of local correlations according to one of the alternatives (i)`, (ii)--(v). Using the notation of Section \ref{chap:estimation} of this supplement, the correlation vector \(\frho(\Z)\) will in general have elements consisting of \(\rho_{ij}^{1,2}, i =1,\ldots,d_1, j=1,\ldots,d_2;\;\rho_{ij}^{1,3}, i=1,\ldots,d_1, j=1,\ldots,d_3;\;\rho_{ij}^{2,3}, i=1,\ldots,d_2, j=1,\ldots,d_3\), where \(d_1,d_2,d_3\) are chosen according to one of the alternatives (i)', (ii)--(v).

\begin{cor}
Let the assumptions of Theorem \ref{thm:E2} hold and let $\fOmega = \fOmega(\frho)$ be the covariance matrix of the $\widehat{\frho}$ vector. Then,
$$
\sqrt{nb_1b_2}(\alpha_{n,b}^*(i,j)-\alpha_{n,b}(i,j)) \buildrel {d_{\mathbb{P}_{\omega}^{*}}} \over {\longrightarrow} {\cal N}(0,\nabla g(\rho)^T \fOmega(\frho)) \nabla g(\rho)) \,\, \mbox{prob}-\mathbb{P}. 
$$ 
\label{cor:E4}
\end{cor}

\emph{Proof:} The proof of Theorem \ref{thm:E2} is given as the proof of Theorem D2 in the online appendix accompanying Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}). This proof can be adapted to prove Corollary \ref{cor:E4}. The starting point consists in replacing \(L_{n,b}^*\) and \(L_{n,b}\) by \(L^* = \sum_{i,j} a_{ij}L_{ij,n,b}^*\) and \(L = \sum_{i,j}a_{ij}L_{ij,n,b}\), respectively, where the sum runs over all possible pairs \((i,j)\) corresponding to \(\rho_{ij}^{1,2}\), \(\rho_{ij}^{1,3}\), \(\rho_{ij}^{2,3}\) that are included in the \(\frho\)-vector. The derivations in the proof of Theorem D2 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) deal with the more complicated five-parameter case. In the present case, however, the vector \(\frho\) consists of correlations \(\rho_{ij}\) that have been estimated using the one-parameter pairwise setup, where, cf.~Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}), the quantities \(J_{ij}\) and \(M_{ij}\) are scalars. Here \(L_{ij,n.b}\) is the log likelihood for the pair \((i,j)\) treated in the proof of Theorem D2, and \(a_{ij}\) is a corresponding sequence of real numbers. Because of the linearity of the expressions for \(L^*\) and \(L\) it is easily checked by going step by step that the proof of Theorem D2 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) can be carried through for \(L^*\) and \(L\). The only difference is that the quantities \(J_{n,b}\) and \(M_{n,b}\) now have to be replaced by \(J_{n.b}(a) = \E(-\sum_{i,j}a_{ij}\nabla^2 L_{ij,n,b}(\rho_{0,b}))\) and \(\Var(\sqrt{nb_1b_2}\sum_{i,j} a_{ij}\nabla L_{ij,n,b})\). This results in joint asymptotic bootstrap normality using the Cramer-Wold device. Applying the delta rule now proves the Corollary. \(\blacksquare\)

Let \(\M_{ij}\) and \(\J_{ij}\) be the \(M\)-matrix and \(J\)-matrix appearing in the proof of Corollary \ref{cor:E4} for a fixed \(\rho_{ij}\). Note that in the one-parameter case of \(\rho_{ij}\) in Corollary \ref{cor:E4}, to first order the covariance matrix \(\fOmega\) is a diagonal matrix having ordered values of \(M_{ij}/J_{ij}^2\) along the diagonal. This can be seen using the methods of proofs of Theorem 2 of Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}) and Theorem 2 Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}). Note also that the 3-coordinate case of alternative (i) can be stated and proved in the same way, but a different scaling factor is required to avoid a singular covariance matrix as seen in the proof of Theorem 3.1 in the main article.

\hypertarget{asymptotic-theory-for-the-test-statistic}{%
\section{ASYMPTOTIC THEORY FOR THE TEST STATISTIC}\label{asymptotic-theory-for-the-test-statistic}}

Consider the general case \(\Z = (\Z_1, \Z_2, \Z_3)\) where each component is \(d_i\)-dimensional correspondingly, with \(i = 1,2,3\), and \(\Z\), naturally, is \(d_1+d_2+d_3\)-dimensional, and consider also the local partial correlation matrix of \((\Z_1,\Z_2|\Z_3)\). In alternative (iii) in Section \ref{chap:estimation} this matrix is given by equation \eqref{eq:offdiag3} with
\[
\alpha_{ij} = \rho(Z_{1,i},Z_{2,j}|Z_3), \;\; i=1,\ldots,d_1; j = 1,\dots,d_2.
\]
This forms a \(d_1 \times d_2\) matrix \({\falpha}\). One can test for conditional independence between \(Z_{1,i}\) and \(Z_{2,j}\) given \(\Z_3\) by testing whether \(\alpha_{ij}=0\), or one can test for conditional independence between \(\Z_1\) and \(\Z_2\) given \(\Z_3\) by testing whether \({\falpha} = \{\alpha_{ij}\} = {\bm{0}}\). The latter case is in a sense the most general, and we will stick to that in the sequel. One can then test for \({\falpha} = {\bm{0}}\) by using a matrix norm, for instance the Frobenius norm, where \(||{\falpha}|| = \sqrt{\sum_{i,j} \alpha_{ij}^2} = 0\). A more convenient test statistic is \(\sum_{i,j} \alpha_{ij}^2\). To follow the notation of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) we write, letting \(\z=(\z_1,\z_2,\z_3)\),
\[
T = \int_S h(\frho(\z)) \di F(\z)
\]
where \(S\) is an a area of integration, \(F\) is the cumulative distribution function of \(\Z\), and \(\frho = \frho(\z)\), as before, is a vector consisting of all possible pairs that can be formed by picking pairwise components from \(\Z\). The number of components of this vector depends on which of the alternatives models (i)--(v) in Section \ref{chap:estimation} one considers. It is reasonable to express \(T\) as a function of \(\frho\) since all partial correlations can be expressed as a function of this vector (see e.g.~again equation \eqref{eq:offdiag3}). One possible choice of the function \(h\) is \(h(\frho(\z)) = \sum_{i,j} \alpha_{i,j}^2(z_{1,i},z_{2,j}|\ztwo)\), which is a function of \(\frho\) since the \(\alpha\)-s are functions of \(\frho\). The corresponding function in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) is \(h(\theta(x))\), where \(\theta(x)\) is at most 5-dimensional and \(\x\) is bivariate. As in that paper we proceed in two stages. First we let the bandwidth vector \(\bb\) be fixed and let \(n \to \infty\), and then we let \(\bb = (b_1,b_2) = (b,b) \to \bm{0}\).
Accordingly we define
\[
T_b = \int_S h(\frho_b(\z)) \di F(\z) 
\]
where \(\frho_b\) is the population quantity for a fixed bandwidth \(\bb\), and \(T_{n,b}\) by
\[
T_{n,b} = \int_S h(\frho_{b,n}(\z)) \di F_n(\z),
\]
where \(\frho_{b,n}(\z)\) is the estimate of \(\frho(\z)\) obtained by pairwise likelihood for each component of \(\frho(\z)\).

Consistency is straightforward to prove:

\begin{thm}
Let the conditions of Theorem \ref{thm:A1} be fulfilled. Then $T_{n,b} \buildrel {\mathbb{P}} \over {\to} T_b$ as $n \to \infty$ and $T_{n,b} \buildrel {\mathbb{P}} \over {\to} T$ as $n \to \infty$ and $\bb \to \bm{0}$.
\label{thm:F1}
\end{thm}

\emph{Proof.} This result follows directly from Theorem \ref{thm:A1} and the continuous mapping theorem. \(\blacksquare\)

It is more difficult to obtain the limiting distribution. We will follow the technique in the proof of Theorem 3.1 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}). See also Joe (\protect\hyperlink{ref-joe1989estimation}{1989}). The details are a bit more involved in the present case because of the larger dimension and because we need to involve the pairwise log likelihood in the procedure.

The results on asymptotic distribution can be proved on several levels. We consider two cases: a) the pairwise simplification where involving the estimates \(\rho_{ij,n,b}(z_i,z_j)\) for all possible pairs \((i,j)\) of the \(\frho\) vector, and b) the case where \(d_1=d_2=d_3=1\), but where \(\rho_{ij} = \rho_{ij}(\z)\) is allowed to depend on all three coordinates \(\z=(z_1,z_2,z_3)\). We will put the emphasis on a) and only comment briefly on b).

\begin{thm} 
Let the assumptions A, D--F of Theorem 3 of Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}) and the conditions of Theorem 3.2 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) (which includes conditions A1 and A2 of this supplement) be fulfilled. Then,
as $n \to \infty$,
$$
\sqrt{n}\big(C_{n}(A_b)\big)^{-\frac{1}{2}}(T_{n,b} - T_b) \buildrel {d} \over {\to} {\cal N}(0,1),
$$
where
\begin{align*}
A_b(\z)  & = \int_{S} \left[\sum_{i,j}\left\{\frac{\partial}{\partial \rho_{ij}} h(\frho_b,(\fv)\right\}J_{ij}^{-1}(\rho_{ij,b}(\fv_{ij}))K_b(\z_{ij}- \fv_{ij})u_{ij}(\z_{ij},\rho_{ij,b}(\fv_{ij}))\right] \di F(\fv) \\
& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad + h(\frho_b(\z))1_S(\z)
\end{align*}
and 
\begin{align}
C_n(A_b) &= \int A_b^2(\z) \di F(\z) + \frac{1}{n}\sum_{s \neq t}\int A_b(\z)A_b(\fu) \di F^{s,t}(\z,\fu) \nonumber \\
& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad - n\int A_b(\z)A_b(\fu) \di F(\z) \di F(\fu)
\label{eq:t0}
\end{align}
with $F^{s,t}(\z,\fu)$ being the joint cumulative distribution function of $(\Z_s,\Z_t)$, and where $\{\Z_{t}, t=1,\ldots,n\}$ are the $n$ observations of $\Z$ and $u_{ij}(\Z_{ij}, \rho_{ij,b}(\fv_{ij})) = \frac{\partial}{\partial \rho_{ij,b}}\log\psi(\Z_{ij},\rho_{ij,b}(\fv_{ij}))$.

To ease comparison with Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) we have written and will write $\z_{ij} \stackrel{\textrm{def}}{=} (z_i,z_j)$ and $\fv_{ij} \stackrel{\textrm{def}}{=} (v_i,v_j)$. Further, if $n \to \infty$ and $\bb_n \to \bm{0}$, the relationship between them being as in Assumption 3 of Theorem 3.1 in the main article, then $T_b$ can be replaced by $T$,
$$
\sqrt{n}\big(C_{n}(A_b)\big)^{-\frac{1}{2}}(T_{n,b} - T) \buildrel {d} \over {\to} {\cal N}(0,1).
$$
\label{thm:F2}
\end{thm}

Note that the sum in the second term in the definition of \(\fC_n(\fA_b)\) taken together with the third term converges due to the geometric \(\alpha\)-mixing condition of A1. It is also important to note that conditional independence is not assumed in Theorem \ref{thm:F2}. This will be exploited in Section \ref{chap:pitman}. See also Theorem 3.2 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) and its relation to Corollary 3.1 in that paper.

\emph{Proof.} From the proof of Theorem 3.1 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}), one has that asymptotically,
\begin{equation}
T_{n,b}-T_b \sim \frac{1}{\sqrt{n}} \left\{\int_S h(\frho_b(\z)) \di G_n(\z) + \int_S\nabla_{\frho} h(\frho_b(\z)^{T} \fV_n(\z) \di F(\z)\right\}. 
\label{eq:t1}
\end{equation}
Here
\[
G_n(\z) = \sqrt{n}(F_n(\z)-F(\z)) \quad \mbox{and} \quad \fV_n(\z) = \sqrt{n}((\frho_{n,b}(\z) -\frho_b(\z)),
\]
with \(F_n\) being the empirical cumulative function of \(\Z\), and where, again, \(\frho\) is the vector consisting of all possible pairs \(\rho_{ij}\). This rewriting of \(T_{n,b}-T_b\) is the first step in obtaining an expression in terms over \(\Z_{t}\), \(t=1,\dots,n\) so that a mixing CLT can be used. We proceed to the pairwise analysis.

Let \((i,j)\) be a fixed pair and \(\rho_{ij,n,b}(\z_{ij})\) be the corresponding estimated local correlation for the pair \((i,j)\) chosen from the vector \(\frho_{n,b}\). Here the pair of coordinates \(\z_{ij} = (z_i,z_j)\) corresponds to the bivariate \(\x\) in the proof of Theorem 3.1 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}). We will also omit some indices or parts of indices when it is clear what is meant. Finally, we omit the area of integration \(S\). We are using the pairwise local log likelihood given by
\[
L_{ij,n} = \frac{1}{n} \sum_{t=1}^{n} K_b(\Z_{ij,t}-\z_{ij}) \log \psi\left(\Z_{ij}^{k},\rho_{b,ij}(\z_{ij})\right) - \int K_b(\fv_{ij}-\z_{ij}) \psi_{ij}(\fv_{ij},\rho_{b,ij}(\z_{ij})) \di \fv_{ij},
\]
where \(K_b = b_1^{-1}K_1(\cdot/b_1) b_2^{-1} K_2(\cdot/b_2)\) is a product kernel function, and \(\Z_{ij,t} = (Z_{i,t},Z_{j,t})\) is the \(t\)-th pair of observations, and \(\fv_{ij} = (v_i,v_j)\). Estimates \(\rho_{ij,b,n}(\z_{ij})\) are obtained by differentiation of the log likelihood and by introducing the score function \(U_{ij}\) and information function \(I_{ij}\) as follows
\begin{align*}
U_{ij}(\z_{ij},\rho_{ij}) & = \frac{\partial}{\partial \rho_{ij}}L_{ij,n}(\z_{ij},\rho_{ij}) \\
& = \frac{1}{n}\sum_{t=1}^{n}K_b(\Z_{ij,t}-\z_{ij})u_{ij}(\Z_{ij,t},\rho_{ij}) - 
\int K_b(\fv_{ij}-\z_{ij})u(\fv_{ij},\rho_{ij})\psi(\fv_{ij},\rho_{ij}) \di \fv_{ij}
\end{align*}
where \(u_{ij}(\z_{ij},\rho_{ij}) = \partial/\partial \rho_{ij} \log \psi(\z_{ij},\rho_{ij})\). Similarly
\[
I_{ij}(\z_{ij},\rho_{ij}) = \frac{\partial^2}{\partial \rho_{ij}^2} L_{ij,n}(\z_{ij}, \rho_{ij}) \quad \mbox{and} \quad J_{ij,b}(\z_{ij}) = \E(-I_{ij}(\z_{ij},\rho_{ij})),
\]
where Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}, \protect\hyperlink{ref-lacal2018estimating}{2018}) may be consulted for more details in the derivation and a more explicit expression for \(I_{ij}(\z_{ij},\rho_{ij})\) for the case considered there. In contradistinction to the 5-parameter case considered by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}), in our present 1-parameter case \(J_{b,ij}\) is not singular as \(\bb \to \bm{0}\). In fact, it is shown by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}, Theorem 2) that as \(\bb \to \bm{0}\), \(J_{b,ij} \to u(\z_{ij},\rho_{ij})\psi(\z_{ij},\rho(\z_{ij}))\).

The vector \(\fV_n(\z) = \sqrt{n}[\frho_{n,b}(\z)-\frho_b(\z)]\) has the dimension of \(\frho(\z)\) and consists of components \(\{\rho_{ij,n,b}(\z_{ij})-\rho_{ij,b}(\z_{ij})\}\) where \(i\) and \(j\) in the most general case run over all possible pairs formed by the components of \(\frho(\z)\). To be able to use the mixing central limit theorem we follow Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) and express this more explicitly in terms of the variables \(\Z_{t}\), \(t=1,\ldots,n\). By Taylor expansion as in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}),
\begin{align*}
0 &= U_{ij}(\z_{ij},\rho_{ij,n,b}(\z_{ij})) \sim  U_{ij}(\z_{ij},\frho_{ij,b}(\z_{ij})) + I_{ij}(\z_{ij},\rho_{ij,b}(\z_{ij}))(\rho_{ij,n,b}(\z_{ij})-\rho_{ij,b}(\z_{ij})) \\
& \sim U_{ij}(\z_{ij},\rho_{ij,b}(\z_{ij})) - J_{ij,b}(\z_{ij})(\rho_{ij,n,b}(\z_{ij})-\rho_{ij,b}(\z_{ij})),
\end{align*}
which implies

\begin{align}
\rho_{ij,n,b}(\z_{ij}) - \rho_{ij,b}(\z_{ij}) & \sim J_{ij,b}^{-1}(\z_{ij})U_{ij}(\Z_{ij},\rho_{ij,b}(\z_{ij})) \nonumber \\
& = J_{ij,b}^{-1}(\z_{ij})\Big[\frac{1}{n}\sum_{t=1}^{n} K_b(\Z_{ij,t} - \z_{ij})u_{ij}(\Z_{ij,t},\rho_{ij,b}(\z_{ij})) \nonumber \\
& \qquad\qquad - \int K_b(\fv_{ij} - \z_{ij})u_{ij}(\fv_{ij},\rho_{ij,b}(\z_{ij}))\psi(\fv_{ij}, \rho_{ij,b}(\z_{ij})) \di \fv_{ij}\Big].
\label{eq:t2}
\end{align}
The second term in \eqref{eq:t1} can then be expressed as
\begin{align}
& \int_S \sum_{i,j} \frac{\partial}{\partial \rho_{ij}} h(\rho_{ij,b}(z)) \nonumber \\
& \qquad\qquad \times J_{ij,b}^{-1}(\z_{ij}) \Big[\frac{1}{n}\sum_{t=1}^{n} K_b(\Z_{ij,t}-\z_{ij})u_{ij}(\Z_{ij,t},\rho_{ij,b}(\z_{ij})) \nonumber \\
& \qquad\qquad\qquad\qquad- \int K_b(\fv_{ij}-\z_{ij})u_{ij}(\fv_{ij},\rho_{ij,b}(\z_{ij}))\psi(\fv_{ij},\rho_{ij,b}(\z_{ij})) \di \fv_{ij}\Big] \di F(\z).
\label{eq:t3}
\end{align}
where the sum over \((i,j)\) extends over all possible \(\frho_{ij}\) contained in the correlation vector \(\frho\). This should be put together with the first term of \eqref{eq:t1};
\begin{equation}
\frac{1}{\sqrt{n}} \int_S h(\frho_b(\z)) \di G_n(\z) = \frac{1}{n}\sum_{t=1}^{n}h\left(\frho_b(\Z_{t})\right) - \int h\left(\frho_b(\z)\right) \di F(\z).
\label{eq:t4}
\end{equation}
We have now reached an expression where one can easily apply the mixing central limit theorem, and an application of that theorem to the sum in \eqref{eq:t3} and \eqref{eq:t4} now yields asymptotic normality of \(\sqrt{n}\left(T_{n,b}-T_{b}\right)\).

Using the same argument as in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) separately on the two main terms yields zero means. Moreover, using the definitions of \(G_n\) and \(A_b(z)\) it can be seen from \eqref{eq:t3} and \eqref{eq:t4} that asymptotically
\[
\sqrt{n}\left(T_{n,b} - T_b\right) \sim \int_S A_b(\z) \di G_n(\z).
\]

By analogy with the proof of Theorem 3.1 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}) it is seen that
\begin{align*}
\Var\left(\int A_b(\z) \di G_n(\z)\right) &\sim \int A_b^2(\z) \di F(\z) + \frac{1}{n}\sum_{s \neq t} \int A_b(\z)A_b(\fu) \di F^{s,t}(\z,\fu) \\ 
& - n\int A_b(\z)A_b(\fu) \di F(\z)\di F(\fu) \\
& = C_n(A_b),
\end{align*}
which is the expression \eqref{eq:t0} stated in the theorem. The sum in the second term taken together with the last term above converges due to the mixing condition. Details of this argument are identical to those of the last part of the proof of Theorem 3.1 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}), as is the case of \(\bb_n \rightarrow \bm{0}\), completing the proof by referring to Proposition 6.3.9 of Brockwell and Davis (\protect\hyperlink{ref-brockwell1991time}{2006}). \(\blacksquare\)

It should be carefully noted that it is not entirely obvious what the order of the term \(C_n(A_b)\) is as \(\bb = (b_i,b_j) = (b,b) \to \bm{0}\) and \(n \to \infty\). Looking at the expressions \eqref{eq:t3} and \eqref{eq:t4}, it is the term \eqref{eq:t3} that causes potential trouble due to the presence of \(b_i^{-1}\) and \(b_j^{-1}\) in the two kernel function terms in the product kernel. These are cancelled out by the integration over \((z_i,z_j)\), however. This means that the true convergence rate for \(T_{n,b}-T\) is \(n^{-1/2}\). This may be somewhat surprising since the convergence rate of \(\frho_{n,b}(\z_{ij})-\frho(\z_{ij})\) is \((nb_1b_2)^{-1/2}\), but this is a well-known phenomenon for functionals of type \(T_{n,b}\). In Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) the convergence rate is changed from \((nb_1^3b_2^3)^{-1/2}\) to \((nb_1^2b_2^2)^{-1/2}\). These results are consistent with similar results for test functionals for independence in Joe (\protect\hyperlink{ref-joe1989estimation}{1989}).

The variance in \eqref{eq:t4} can be evaluated in more detail, as has been done by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}), in particular using the mixing inequality and Hall and Heyde (\protect\hyperlink{ref-hall1980martingale}{1980}, Corollary A2). But as shown by Teräsvirta, Tjøstheim, and Granger (\protect\hyperlink{ref-terasvirta2010modelling}{2010}) and in numerous other papers, the asymptotic variance in the distribution of such functionals as \(T_{n,b}\) cannot be expected to be accurate unless one has extremely large sample sizes.

The second alternative mentioned prior to stating Theorem \ref{thm:F2} is the alternative with \(d_1=d_2=d_3=1\), but with correlations depending on three coordinates so that \(\z=(z_1,z_2,z_3)\). This is easier in the sense that there are only three pairs of correlations, \(\rho_{12}, \rho_{13}, \rho_{23}\) and in the above notation \(\frho(\z) = (\rho_{12}(\z), \rho_{13}(\z), \rho_{23}(\z))\). The partial correlation is
\[
\alpha_{12|3}(z) = \frac{\rho_{12}(\z)-\rho_{13}(\z)\rho_{23}(\z)}{\sqrt{1-\rho_{13}^2(\z)}\sqrt{1-\rho_{23}^2(\z)}},
\]
which means that a natural choice of the function \(h(\frho(\z))\) might be \(h(\frho(\z)) = \alpha_{12|3}^2(\z)\) This is a three-parameter function of the parameters \(\rho_{12}, \rho_{13},\rho_{23}\), and an explicit expression for the gradient \(\nabla_{\frho}h(\frho)\) can easily be found. More generally again we define
\[
T_{n,b} = \int_S h(\frho_{b,n}(\z)) \di F_n(\z),\;\; T_b= \int_S h(\frho_b(\z)) \di F(\z) \, \textrm{ and } \, T = \int_S h(\frho(\z)) \di F(\z).
\]

\begin{thm}
Let the assumptions of Theorem \ref{thm:F2} be fulfilled and with a 3-coordinate $\rho$ as defined above. Then,
as $n \to \infty$,
$$
\sqrt{n}\left(C_{n}(A_b)\right)^{-\frac{1}{2}}(T_{n,b} - T_b) \buildrel {d} \over {\to} {\cal N}(0,1).
$$
Here 
$$
A_b(\z) = \int_S \nabla_{\frho}h(\frho_b(\fv))^{T}{\J}^{-1}(\frho_b(\fv))K_b(\z-\fv)\fu(\z,\frho_b(\fv)) \di F(\fv) + h(\frho_b(\z))1_S(\z)
$$
and 
$$
C_n(A_b) = \int A_b^2(\z) \di F(\z) + \frac{1}{n}\sum_{s \neq t}\int A_b(\z)A_b(\fu) \di F^{s,t}(\z,\fu) - n\int A_b(\z)A_b(\fu) \di F(\z) \di F(\fu)
$$
with $F^{s,t}(\z,\fu)$ being the joint cumulative distribution function of $(\Z^s,Z^t)$. Further, if $n \to \infty$ and $\bb_n \to \bm{0}$, the relationship between them being as in Assumption 3 of Theorem 3.1 in the main article, then $T_b$ can be replaced by $T$ such that
$$
\sqrt{n}\left(C_n(A_b)\right)^{-\frac{1}{2}}(T_{n,b} - T) \buildrel {d} \over {\to} {\cal N}(0,1).
$$
\label{thm:F3}
\end{thm}

\emph{Proof.} The proof follows along the lines of the proof of Theorem \ref{thm:F2}. Corresponding to equation \eqref{eq:t1},
\[
T(\frho_{b,n}) - T(\frho_{b}) \sim \frac{1}{\sqrt{n}}\left\{\int_S h(\frho_b(\z)) \di G_n(\z) + \int_S \nabla_{\frho} h(\frho_b(\z)))^T \fV_n(\z)\right\} \di F(\z)
\]
where \(G_n(z) = \sqrt{n}(F_n(z)-F(z))\) and
\begin{align*}
\fV_n(z) &= \sqrt{n} (\frho_{n,b}(\z) - \frho_b(\z)) \\
& \sim \sqrt{n}{\J}_b^{-1}(\z)\left\{\frac{1}{n} \sum_{t=1}^{n}K_b(\Z^t-\z)\fu(\Z^t,\frho_b(\z)) - \int K_b(\fv-\z)\fu(\fv,\frho_b(\z))\psi(\fv,\rho_b(\z)) \di \fv \right\}.
\end{align*}
Here \(K_b\) consists of a triple product of kernels \(\prod_{i=1}^{3}\frac{1}{b_i}K(\cdot/b_i)\), \(\fu(\z,\frho_b(\z)) = \nabla_{\frho}\log \psi(\z,\frho_b(\z))\) where \(\psi\) is the density function of a zero mean trivariate normal distribution having a covariance matrix equal to the correlation matrix and with off diagonal elements equal to the elements in \(\frho_b(\z)\). This is further developed as in the proof of Theorem \ref{thm:F2}, and the asymptotic normality of \(T_{n,b}-T_b\) follows as before from the mixing assumptions and the proof is completed by applying proposition 6.3.9 of Brockwell and Davis (\protect\hyperlink{ref-brockwell1991time}{2006}). \(\blacksquare\)

The order of \(C_n(A_b)\) is of interest as \(\bb\to \bm{0}\) and \(n \to \infty\) since it determines the convergence rate of \(T_{n,b}-T\). From Theorem 3.1 in the main article it follows that \(\frho_{n,b}(\z)\) converges at a rate of \((nb^5)^{-1/2}\) to \(\frho(z)\) as \(n \to \infty\) and \(b_1=b_2=b_3 = b \to 0\). In the proof of Theorem \ref{thm:F3} one integrates over a triple kernel function in \(C_n(A_b)\). This cancels out three \(b\)-s. so that one is left with a convergence rate of \((nb^2)^{-1/2}\) for \(T_{n,b}-T\). This is slower than the convergence rate in Theorem \ref{thm:F3}, but this is not unexpected since the faster rate in Theorem \ref{thm:F3} comes at the price of a 1-parameter pairwise simplification in the model structure.

\hypertarget{chap:pitman}{%
\section{ASYMPTOTIC POWER AND PITMAN EFFICIENCY}\label{chap:pitman}}

Since the rates of convergence in Theorems \ref{thm:F2} and \ref{thm:F3} are \(n^{-1/2}\) and \((nb^2)^{-1/2}\) respectively, the Pitman efficiency will be improved in the case of Theorem \ref{thm:F2} as compared to that of Theorem \ref{thm:F3}.

We will follow the approach used by Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2017}), in the proof of their Theorem 2 to derive the results on local asymptotic efficiency. We will explicitly use the results in Theorems \ref{thm:F2} and \ref{thm:F3} with \(h({\falpha}_{n,b}(\z)) = ||{\falpha}_{n,b}(\z)||^2\), so that \(T_{n,b} = \int _S||{\falpha}_{n,b}(\z)||^2 \di F_n(\z)\). The null hypothesis of conditional independence can then be phrased as
\[
H_0: ||{\falpha(\z)}||^2 =  0 \;\; \mbox{for} \;\;\z \;\; \mbox{almost everywhere in} \;\;S. 
\]
against the Pitman alternative
\[
H[a_n]: ||{\falpha(\z)}||^2 = a_n\beta(\z)
\]
where \(a_n\) is a sequence that tends to zero at a certain rate, determining the asymptotic rate at which there is still power in testing \(H_0\), and \(\beta (\z) =\beta(\rho(\z))\) is a positive function of \(\z\) such that \(\int_S \beta(\z) \di F(\z) \stackrel{\textrm{def}}{=} \gamma < \infty\). Here \(\beta(\z)\) is the result of taking an arbitrary correlation vector \(\rho(\z) = \rho_H(\z)\) under the alternative hypothesis such that \({\falpha}(\rho_H(\z)) \neq {\bm{0}}\) and \(\beta(\z)\) is the Froebenius norm squared of this \({\falpha}\). Further, let \(C_{H,n,b}(A_b)\) be the \(C_{n,b}\)-function formed as in Theorem \ref{thm:F2} from \(\rho_H(\z)\). It converges towards \(C_{H}(\fA_b)\) as depicted in the mixing expression of the proof of Theorem \ref{thm:F2}. See also Theorem 3.2 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) which is formulated without using independence.

\begin{thm}
Let the conditions be as in Theorem \ref{thm:F2} and let $H[a_n]$ be as above with $a_n =n^{-1/2}$. Let $C_H(A)$ be the limit of $C_n(A_b)$ as $n \to \infty$ and $\bb \to \bm{0}$ at the rates specified in the assumptions of Theorem \ref{thm:F2}. (This limit exists using the mixing expressions of the proof of Theorem \ref{thm:F3}). Then the power of $n^{1/2}(C_{H,n,b}(A_b))^{-1/2}(T_{n,b}-T)$ ($T = 0$ under our $H_0$, but the procedure below can also be carried through with a general ${\falpha}$ under $H_0$) is given by $\mathbb{P}[T_{n,b} \geq z_{\varepsilon}|H[a_n]) \to 1-\Phi(z_{\varepsilon}-\gamma/[C_{H}(A))]^{1/2})$ as $n \to \infty, \bb \to \bm{0}$, where $\Phi(\cdot)$ is the cumulative standard normal distribution function and $z_{\varepsilon}$ is the one-sided critical value of ${\cal N}(0,1)$ at significance level $\varepsilon$.
\label{thm:G1}
\end{thm}

\emph{Proof.} The test statistic can be written as

\begin{align*}
T_{n,b} & = \int_S ||{\falpha}(\z)_{n,b}||^2 \di F_n(\z) \\
&= \int _S ||{\falpha}(\z)_{n,b}||^2 \di F_n(\z) - a_n \int_S \beta(\z) \di F(\z) + a_n \int \beta(\z) \di F(\z).
\end{align*}
Here, using Theorem \ref{thm:F2}, under \(H[a_n]\),
\[
\sqrt{n}[C_{H,n,b}(A_b)]^{-1/2}\left\{\int_S ||{\falpha}(\z)_{n,b}||^2 \di F_n(\z) - a_n \int_S \beta(\z) \di F(\z)\right\} \stackrel{d}{\to} {\cal N}(0,1).
\]
and
\[
\sqrt{n}[C_{H,n,b}(A_b)]^{-1/2} a_n \int_S \beta(\z) \di F(\z) \to \gamma/[C_H(A)]^{1/2}
\]
and this completes the proof. \(\blacksquare\)

The alternative situation that is covered in Theorem \ref{thm:F3} can now be stated and proved analogously. In this case \(b^2\fC_{H,n,b}\) converges towards \(\fC^{\prime}(\fA)\), say, where \(b\) is the common value of the bandwidths \((b_1,b_2,b_3)\).

\begin{thm}
Let the conditions be as in Theorem \ref{thm:G1} and let $H[a_n]$ be as above with $a_n =(nb^2)^{-1/2}$, with $b$ the common value of the three bandwidths $(b_1,b_2,b_3)$. Then, letting $C_H^{\prime}(A)$ be the limit of $b^2C_{H,n,b}(A_b)$ as $n \to \infty$ and $b \to 0$ at the rates specified in the assumptions of Theorems \ref{thm:F2} and \ref{thm:F3}. (This limit exists using the mixing expressions of the proof of Theorem \ref{thm:F2}). Then the power of $(nb^2)^{1/2}(C_{H,n,b}(A_b))^{-1/2}(T_{n,b}-T)$ ($T = 0$ under our $H_0$, but the procedure can also be carried through with a general ${\falpha}$ under $H_0$.) is given by $\mathbb{P}[T_{n,b} \geq z_{\varepsilon}|H[a_n]) \to 1-\Phi(z_{\varepsilon}-\gamma/[C_{H}(A)^{\prime})]^{1/2})$ as $n \to \infty, b \to 0$, where $\Phi(\cdot)$ is the cumulative standard normal distribution function and $z_{\varepsilon}$ is the one-sided critical value of ${\cal N}(0,1)$ at significance level $\varepsilon$.
\label{thm:G2}
\end{thm}

We conclude that the testing procedure can pick up alternatives that converges at the rate \((nb^2)^{-1/2}\) towards the hypothesis of conditional independence as formulated under the alternative (i) in Section \ref{chap:estimation}.

For a further brief discussion, see the paragraph following eqs. (27) and (28) in Section 5.2 in the main article.

\hypertarget{bootstrapping-the-test-statistic}{%
\section{BOOTSTRAPPING THE TEST STATISTIC}\label{bootstrapping-the-test-statistic}}

The main principle of this section is that we replace \(F_n\)-quantities with \(F_n^{*}\)-quantities and \(F\)-quantities with \(F_n\)-quantities, the latter being known under the bootstrap operation. Let
\[T_{n,b}^* = \int_S h(\frho_{b,n}^*(\z))\di F_n^*(\z).\]
We treat the pairwise model case corresponding to Theorem \ref{thm:F2} first.

\begin{thm}
Let the assumptions of Theorem \ref{thm:E2} be fulfilled. Then
as $n \to \infty$
$$
\sqrt{n}[C_{n}(A_b)]^{-\frac{1}{2}}(T_{n,b}^* - T_{n,b}) \stackrel{d_{\mathbb{P}_{\omega}^{*}}}{\to} {\cal N}(0,1).
$$
Here 
\begin{align*}
A_b(z) = \int_{S} \Big[\sum_{i,j}\frac{\partial}{\partial \rho_{ij,n,b}} h(\rho_{n,b}(\fv)J_{ij}^{-1}(\rho_{ij,n,b}(\fv_{ij}))K_b(\z_{ij}-\fv_{ij})&u_{ij}(\z_{ij},\rho_{ij,n,b}(\fv_{ij})\Big]  \di F(\fv) \\
& + h(\rho_{ij,n,b}(\z)1_S(\z),
\end{align*}
and 
\begin{align}
C_n(A_b) = \int A_b^2(\z)\di F_n(\z) + & \frac{1}{n}\sum_{s \neq t}\int A_b(\z)A_b(\fu)\di F^{s,t}_{n}(\z,\fu) \nonumber \\
& -  n\int A_b(\z)A_b(\fu) \di F_n(\z) \di F_n(\fu), 
\label{eq:t00}
\end{align}
with $F^{s,t}_{n}(\z,\fu)$ being the empirical joint cumulative distribution function of $(\Z_s,\Z_t)$. Note that the sum in the second term in the definition of $C_n(A_b)$ taken together with the third term converges due to the mixing assumption.
\label{thm:H1}
\end{thm}

\emph{Proof.} As in the beginning of the proof of Theorem 4.4, part B, of the supplement of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}), one has that asymptotically,
\begin{equation}
T_{n,b}^* - T_{n,b}  \sim \frac{1}{\sqrt{n}} \left\{\int_S h(\frho_b(\z)) \di G_n^*(\z) + \int_S\nabla_{\rho}h(\frho_b(\z)^{T}\tilde{V}_n(\z) \di F_n(\z)\right\},
\label{eq:t11}
\end{equation}
where
\[
G_n(\z) = \sqrt{n}(F_n^*(\z) - F_n(\z)) \quad \mbox{and} \quad \tilde{V}_n(\z) = \sqrt{n}((\frho_{n,b}^{*}(\z) -\frho_{n,b}(\z))
\]
with \(F_n^*\) being the empirical cumulative function for the bootstrapped time series, and where again \(\frho\) is the vector consisting of all possible pairs \(\rho_{ij}\). This rewriting of \(T_{n,b}^*-T_{n,b}\) is the first step in obtaining an expression in terms over \(\Z^*_t\), \(t=1,\dots,n\) so that a mixing type CLT can be used as done in the paper by Gonçalves and White (\protect\hyperlink{ref-gonccalves2002bootstrap}{2002}). We proceed to the pairwise analysis.

Let \((i,j)\) be a fixed pair and \(\rho_{ij,n,b}^*(z_i,z_j)\) be the corresponding estimated bootstrap local correlation for the pair \((i,j)\) chosen from the vector \(\frho_{n,b}^*\). Here the pair of coordinates \((z_i,z_j)\) corresponds to the bivariate \(\x\) in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}, proof of Theorem 3.1). To ease notation and the comparison with Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}, \protect\hyperlink{ref-lacal2018estimating}{2018}), we again write \(\z_{ij} = (z_i,z_j)\). We will also omit some indices or parts of indices when it is clear what is meant. Finally, we omit the area of integration \(S\). We are using the pairwise log likelihood given by
\[
L_{ij,n}^* = \frac{1}{n} \sum_{t=1}^{n} K_b(\Z_{ij,t}^{*}-\z_{ij}) \log \psi(\Z_{ij,t}^{*},\rho_{ij,n,b}(\z_{ij})) - \int K_b(\fv_{ij}-\z_{ij}) \psi_{ij}(\fv_{ij},\rho_{ij,n,b}(\z_{ij})) \di \fv_{ij},
\]
where \(K_b = \frac{1}{b_1}K_1(\cdot/b_1) \frac{1}{b_2} K_2(\cdot/b_2)\) is a product kernel function and \(\Z_{ij,t}^{*} = \left(Z_{i,t}^{*},Z_{j,t}^{*}\right)\) is the \(t\)-th pair of bootstrap observations. Estimates \(\rho_{ij,n,b}^*(\z_{ij})\) are obtained by differentiation of the log likelihood and by introducing the score function \(U_{ij}^*\) and information function \(I_{ij}^*\) as follows
\begin{align*}
U_{ij}^*(\z_{ij},\rho_{ij,n}) &= \frac{\partial}{\partial \rho_{ij,n}}L_{ij,n}^*(\z_{ij},\rho_{ij,n}) \\
& = \frac{1}{n}\sum_{t=1}^{n}K_b(\Z_{ij,t}^{*} - \z_{ij})u_{ij}(\Z_{ij,t}^{*},\rho_{ij,n}) - 
\int K_b(\fv_{ij} - \z_{ij})u_{ij}(\fv_{ij},\rho_{ij,n})\psi(\fv_{ij},\rho_{ij,n}) \di \fv_{ij}
\end{align*}
where \(u_{ij}(\z_{ij},\rho_{ij,n}) = \frac{\partial}{\partial \rho_{ij,n}}\log \psi(\z_{ij},\rho_{ij,n})\). Similarly,
\[
I_{ij}^*(\z_{ij},\rho_{ij,n}) = \frac{\partial^2}{\partial \rho_{ij,n}^2} L_{ij,n}^{*}(\z_{ij},\rho_{ij,n}) \quad \mbox{and} \quad J_{ij}^*(\z_{ij}) = \E^*\left(-I_{ij}^*(\z_{ij},\rho_{ij,n})\right),
\]
where Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}, proof of Theorem 4.4) may be consulted for more details in the derivation and a more explicit analog expression for \(I_{ij}^*(\z_{ij},\rho_{ij,n})\) for the case considered there. In contradistinction to Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}), \(J_{ij,n,b}\) is not singular as \(\bb \to \bm{0}\). In fact, it can be shown that, cf.~Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}, Theorem 2), as \(\bb \to \bm{0}\), \(J_{ij,n,b} \to u(\z_{ij},\rho_{ij,n})\psi(\z_{ij},\rho(\z_{ij,n}))\).

The vector \(\tilde{V}_n(\z) = \sqrt{n}\left(\frho_{n,b}^*(\z)-\frho_{n,b}(\z)\right)\) has the dimension of \(\frho(z)\) and consists of components \(\{\rho_{ij,n,b}^*(\z_{ij}-\rho_{ij,n,b}(\z_{ij})\}\) where \(i\) and \(j\) in the most general case run over all possible pairs formed by the components of \(\frho(\z)\). To be able to use the mixing CLT analogous to Gonçalves and White (\protect\hyperlink{ref-gonccalves2002bootstrap}{2002}) we follow Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}) and express this more explicitly in terms of the variables \(\Z^{*}_t\), \(t=1,\ldots,n\). By Taylor expansion as in the proof of Theorem \ref{thm:F2}, and omitting the index \(ij\) on \(\rho\),
\begin{align*}
0 &= U_{ij}^*(\z_{ij},\rho_{n,b}^*(\z_{ij})) \\
& \sim U_{ij}^*(\z_{ij},\rho_{n,b}(\z_{ij}))+I_{ij}^*(\z_{ij},\rho_{n,b}(\z_{ij}))(\rho_{n,b}^*(z_{ij})-\rho(z_{ij})) \\
& \sim U_{ij}^*(\z_{ij},\rho_{n,b}(\z_{ij})) - J_{ij,n,b}(\z_{ij})(\rho_{n,b}^*(\z_{ij})-\rho_{n,b}(\z_{ij}))
\end{align*}
where we have used that \(\E^*(\cdot)\) is the mean value under the measure \(\mathbb{P}_{\omega}^{*}\) and again following the reasoning of the proof of Theorem \ref{thm:F2}, this
implies that
\begin{align}
\rho_{n,b}^*(\z_{ij}) - \rho_{n,b}(\z_{ij}) &\sim J_{ij,n,b}^{-1}(\z_{ij})u_{ij}(\z_{ij},\rho_{n,b}(\z_{ij})) \nonumber \\
& = J_{ij,n,b}^{-1}(\z_{ij})\Big[\frac{1}{n}\sum_{t=1}^{n} K_b(\Z_{ij,t}^{*} - \z_{ij})u_{ij}(\Z_{ij,t}^{*},\rho_{n,b}(\z_{ij})) \nonumber \\
 & \qquad\qquad - \int K_b(\fv_{ij} - \z_{ij})u_{ij}(\fv_{ij},\rho_{n,b}(\z_{ij}))\psi(\fv_{ij},\rho_{n,b}(\z_{ij}) \di \fv_{ij}\big].
\end{align}
The second term in \eqref{eq:t11} can then be expressed as
\begin{align}
&\int_S \sum_{i,j} \frac{\partial}{\partial \rho_{ij,n,b}} h(\rho_{n,b}(\z)) \times
 J_{ij,n,b}^{-1}(z_{ij})
\Big[\frac{1}{n}\sum_{t=1}^{n} K_b(\Z_{ij,t}^{*}-\z_{ij})u_{ij}(\Z_{ij,t}^{*},\rho_{n,b}(\z_{ij})  \nonumber \\
& \qquad\qquad\qquad\qquad - \int K_b(\fv_{ij} - \z_{ij})u_{ij}(\fv_{ij},\rho_{n,b}(\z_{ij}))\psi(\fv_{ij},\rho_{n,b}(\z_{ij}) \di \fv_{ij}\Big] \di F_n(\z),
\label{eq:t22}
\end{align}
where the sum over \((i,j)\) extends over all \(\rho_{ij}\) contained in the \(\rho\)-vector. This should be put together with the first term of \eqref{eq:t11},
\begin{equation}
\frac{1}{\sqrt{n}} \int_S h(\rho_{n,b}(\z)) \di G_n^*(\z) = \frac{1}{n}\sum_{t=1}^{n}h(\rho_{n,b}(\Z_t^{*}) - \int h(\rho_{n,b}(\z) \di F_n(\z)
\label{eq:t33}
\end{equation}
and we have thus reached an expression where we can apply the mixing CLT as in Gonçalves and White (\protect\hyperlink{ref-gonccalves2002bootstrap}{2002}) and an application of their result, quoted as Theorems C.1 and E.1 in the supplementary materials of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}) and Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) respectively, to the expressions in \eqref{eq:t22} and \eqref{eq:t33}. This yields asymptotic normality of \(\sqrt{n}(T_{n,b}^*-T_{b,n})\).

The rest of the proof now follows along the lines of the end of the proof of Theorem \ref{thm:F2}, including the use of Proposition 6.3.9 of Brockwell and Davis (\protect\hyperlink{ref-brockwell1991time}{2006}). \(\blacksquare\)

An analogue of Theorem \ref{thm:F3} can now be phrased and proved similarly.

\begin{thm}
Let the assumptions of Theorem \ref{thm:E2} be fulfilled. Then,
as $n \to \infty$ and $\bb \to \bm{0}$
$$
\sqrt{n}\left(C_{n}(A_b)\right)^{-\frac{1}{2}}\left((T_{n,b}^* - T_{n,b}\right) \buildrel {d_{\mathbb{P}_{\omega}^*}} \over {\to} {\cal N}(0,1).
$$
Here 
$$
A_b(\z) = \int_S \nabla_{\rho}h(\rho_{n,\bb}(\fv))^{T}{\J}^{-1}(\rho_{n,\bb},\fv)K_b(\z-\fv)\fu(\z,\frho_{n,\bb}(\fv) \di F(\fv) + h(\frho_{n,\bb}(\z)1_S(\z)
$$
and 
\begin{align*}
C_n(A_b) = \int A_b^2(\z) \di F_n(\z) &+ \frac{1}{n}\sum_{s \neq t}\int A_b(\z)A_b(\fu) \di F_n^{s,t}(\z,\fu) \\ & \qquad\qquad- n\int A_b(\z)A_b(\fu)\di F_n(\z) \di F_n(\fu),
\end{align*}
with $F_n^{s,t}(\z,\fu)$ being the empirical joint cumulative distribution function of $(\Z_s,\Z_t)$.
\label{thm:H2}
\end{thm}

\hypertarget{the-local-bootstrap}{%
\section{THE LOCAL BOOTSTRAP}\label{the-local-bootstrap}}

The limiting results for \(\alpha_n(\z)\) (or \({\falpha_n}(\z)\) in the matrix case) and \(T_{n,b}\), and the corresponding bootstrap results can be used to construct asymptotic and bootstrap confidence intervals of \(T_b\) and \(T\). In that sense they give an indication of whether \(\alpha(\z) = 0\) locally or \(\alpha(\z) \equiv 0\) globally respectively. Such confidence intervals, however, cannot in general be used to test for \(\alpha(z) \equiv 0\), i.e.~conditional independence, with a given level of the test. To be able to do that one must device a test statistic that is constructed with a probability structure as under the null hypothesis of conditional independence. It is difficult to obtain such a test based on asymptotic arguments for at least two reasons: As mentioned before the asymptotic distribution of test functionals such as \(T_{n,b}\) is known to be inaccurate except for extremely large sample sizes. It is also difficult to build the probability null structure into the scaling factor of \(T_{n,b}\) since this would amount to building into that structure certain relationships between pairwise local correlations.

An alternative that has been used in the testing theory of conditional independence is the so-called local bootstrap. This is a device that uses testing based on bootstrapping from kernel density estimates. It was used for this purpose by Paparoditis and Politis (\protect\hyperlink{ref-paparoditis2000local}{2000}) and much more extensively by Su and White (\protect\hyperlink{ref-su2008nonparametric}{2008}). It has also been used (without proof) by Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2017}) and by Huang (\protect\hyperlink{ref-huang2010testing}{2010}). We have used a variation of the local bootstrap in our test of conditional independence with good results in accuracy of level and in power. The primary purpose of the present paper is not to prove in detail properties of the local bootstrap in local Gaussian testing, but we will nevertheless indicate theoretically why a local type bootstrap argument makes sense, and why it can be expected to, and does, work also in our case. To put the local type bootstrap into our test functional context we need first to state some facts and properties of the local bootstrap as used in the literature. Note that our functional is in a sense analogous to that used by Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2017}).

In our notation \(\Z_{3,t}\) is the basic \(d_3\)-dimensional process corresponding to \(X_t\) in Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2017}). Under \(H_0\) we assume that \(\Z_{1,t}\) and \(\Z_{2,t}\) of dimensions \(d_1\) and \(d_2\) are conditionally independent given \(\Z_{3,t}\), where \(\Z_{1,t}\) and \(\Z_{2,t}\) correspond to \(Y_t\) and \(Z_t\) in Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2017})

The first step consists in constructing a bootstrap realization \(\{\Z_{3,t}^*\}\) of \(\{\Z_{3,t}\}\) by sampling from the kernel density estimate \(\tilde{f}(\z_{3}) = \frac{1}{n}\sum_{t=1}^{n}L_b(\z_3-\Z_{3,t})\), where \(L_b\) is a \(d_3\) dimensional product kernel function with factors \(L_{b_{i}} = \frac{1}{b_{i}}L(\cdot/b_i)\). The bootstrap realizations of \(\{\Z_{1,t}^*\}\) and \(\{\Z_{2,t}^*\}\) are then created independently of each other by sampling from the conditional kernel estimators \(\tilde{f}(\z_1|\Z_{3,t}^*) = \sum_sL_b(\Z_{1,s}-\z_{1})L_b(\Z_{3,s}-\Z_{3,t}^*)/\sum_s L_b(\Z_{3,s}-\Z_{3,t}^*)\) and \(\tilde{f}(\z_2|\Z_{3,t}^*) = \sum_sL_b(\Z_{2,s}-\z_2)L_b(\Z_{3,s}-\Z_{3,t}^*)/\sum_sL_b(\Z_{3,s}-\Z_{3,t}^*)\), respectively.

The bootstrap test statistic \(T_n^*\) (which is a Hellinger type statistic in the Su-White case. can then be constructed from these bootstrap observations \(\{\Z_{1,t}^*, \Z_{2,t}^*, \Z_{3,t}^*\}\) using another kernel function \(K\) and a bandwidth \(h\) to construct density estimates \(\hat{f}^*(\z_3)\), \(\hat{f}^*(\z_2,\z_3)\), \(\hat{f}^*(\z_1,\z_3)\) and \(\hat{f}^*(\z_1,\z_2,\z_3)\) which is inserted in the Hellinger functional to obtain a bootstrap value \(T_n^*\) of this statistic. It can then be shown (Su and White \protect\hyperlink{ref-su2008nonparametric}{2008}) that properly normalized \(T_n^* \buildrel {d} \over {\to} {\cal N}(0,1)\).

Further, if \(\mathbb{P}^*\) denote the probability conditional on the sample \(\{\Z_t\}\), then the level-\(\varepsilon\) critical values \(\tilde{c}_{\varepsilon}\) are computed as approximate solutions of \(\mathbb{P}^{*}(T_n^* > \tilde{c}_{\varepsilon}) = \varepsilon\) and the bootstrap \(p\)-value is given by \(p^* = B^{-1}\sum_{j=1}^{B} 1(T_{nj}^* > T_n)\), where \(B\) is the number of bootstrap realization. To integrate this procedure into our context is fairly straightforward. The first step is to generate local type bootstrap observations \(\{\Z_{1,t}^*,\Z_{2,t}^*,\Z_{3,t}^*\}\). The bootstrap realization \(\{\Z_{3,t}^*\}\) can be created using the ordinary bootstrap on \(\{\Z_{3,t}\}\), cf.~Paparoditis and Politis (\protect\hyperlink{ref-paparoditis2000local}{2000}), or the kernel smoothed bootstrap as in Su and White (\protect\hyperlink{ref-su2008nonparametric}{2008}). If the dimension of \(\Z_{3,t}\) is large, a smoothed bootstrap realization can be created by sampling from the local Gaussian density estimate of Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}). In the next step in creating the conditional bootstrap realizations \(\{\Z_{2,t}^*\}\) and \(\{\Z_{3,t}^*\}\) one can use the smoothed conditional kernel estimate of Su and White (\protect\hyperlink{ref-su2008nonparametric}{2008}), or if the dimensions are large, to avoid the curse of dimensionality, one can sample from the conditional local Gaussian density estimates of Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}) using, say, acceptance reject sampling. Again it should be noted that the avoidance of the kernel estimate comes at the cost of a more restrictive model.

Creating bootstrap realizations by the above scheme is quite time consuming since
for each \(\Z_{3,t}^*\), drawings of \(\Z_{1,t}^*\) and \(\Z_{2,t}^*\) have to be done, and this has to be repeated over the number of bootstrap realizations. We have devised a simplified and more efficient procedure for this as described in the main article.

Once a bootstrap realization \(\{\Z_{1,t}^*,\Z_{2,t}^*,\Z_{3,t}^*\}\) is created, following Su and White (\protect\hyperlink{ref-su2008nonparametric}{2008}), these realizations are now taken as the original observations in the local likelihood estimation procedure. For a fixed bandwidth \(b\) the population values, following the local Gaussian approach, can be taken as the solution of the integral equation (2) in Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}); see also corresponding equations in e.g.~Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, \protect\hyperlink{ref-otneim2017conditional}{2018}). Note that in the present paper we have largely operated with two alternatives, the general vector case with inference via the pairwise local likelihood, or the the purely scalar case with three coordinates for the local Gaussian correlation. In both cases we have normalized the processes so that they have standard normal marginals.

In the \(\{\Z_{1,t}^*,\Z_{2,t}^*,\Z_{3,t}^*\}\)-framework we obtain an estimate \(\hfrho^*\) of the local Gaussian correlation vector conditional on the bootstrap realizations. An asymptotic theory of this estimate can now be established using the method of proof in Theorem 2 in Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}), and subsequently an asymptotic theory for the test statistic \(T_n^*\) using the method of proof in Theorem 4.4 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}) and Theorem D.4 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}). See also the formulation of such proofs in the present paper, theorems \ref{thm:H1} and \ref{thm:H2}. The scaling factor \(C_n(A_b)\) is of the same type as in the formulation of those results with obvious notational changes. Moreover under the present null hypotheses \(T = 0\), it may be conjectured that it leads to the asymptotic results that the scaled \(T_n^* \buildrel {d} \over {\to} {\cal N}(0,1)\).

\hypertarget{two-additional-examples-of-estimation}{%
\section{TWO ADDITIONAL EXAMPLES OF ESTIMATION}\label{two-additional-examples-of-estimation}}

We will in this section provide two more demonstrations on how the estimated LGPC can be used to reveal nonlinear departures from conditional independence and to measure that conditional dependence. The ordinary partial correlation coefficient fully characterizes conditional dependence between jointly Gaussian variables. It is not hard, though, to construct examples in which strong nonlinear relationships remain completely undetected by the partial correlation. Consider the following structural equation;

\begin{figure}[t]
    \subfloat[Estimated local Gaussian partial correlation \newline between $X_1$ and $X_2$ given $X_3=0$ in the \newline structural model $X_2 = X_1^2 + X_3.$ The color scale \newline goes from $-1$ (dark blue) to $+1$ (dark red).\label{fig:structural-plot1}]{\includegraphics[width=0.49\linewidth]{figures/structural-plot-1} }
    \subfloat[The estimated LGPC along the curve $X_2 = X_1^2$ with 95\% confidence band.\label{fig:structural-plot2}]{\includegraphics[width=0.49\linewidth]{figures/structural-plot-2} }
    \caption{The structural model. The fully trivariate model of Section 3.1.1 is used.}\label{fig:structural-plot}
\end{figure}

\begin{equation}
X_2 = X_1^2 + X_3,
\label{eq:structural}
\end{equation}
where we observe all components in \((X_1, X_2, X_3)\). There is, obviously, a strong dependence between \(X_1\) and \(X_2\), and furthermore, this dependence is deterministic when conditioning on \(X_3 = x_3\). It is, however, well known that \(X_1\) and \(X_1^2 + x_3\) are uncorrelated if \(\E(X_1) = \E(X_1^3) = 0\), but the LGPC easily reveals conditional dependence between \(X_1\) and \(X_2\) in this case. Let us generate \(n = 500\) independent observations each from \(X_1 \sim \mathcal{N}(0,1)\) and \(X_3 \sim \mathcal{N}(0,1)\) and calculate \(X_2\) by \eqref{eq:structural}. The sample partial correlation between \(X_1\) and \(X_2\) is \(-0.037\) in this case, but the LGPC, which in this case is calculated using the simplified pairwise structure defined in Section 3.1.2, displayed in Figure \ref{fig:structural-plot1}, indicates strong departures from conditional independence on the plane defined by \(X_3 = 0\). Indeed, we identify a region of \emph{negative conditional relationship} on the left hand side of the plot, which makes good sense because small values of \(X_2\) is typically observed together with large values of \(X_1\) and vice versa. The opposite phenomenon is clearly visible in the right hand half of the plot. In Figure \ref{fig:structural-plot2} we have plotted the estimated LGPC along the curve \(X_2 = X_1^2\), along with an asymptotic 95\% confidence band. The full trivariate fit presented in Section 3.1.1 gives a very similar picture.

This points to an important feature of the LGPC, as we also saw in one of the simulated examples in the main article: It is able to \emph{distinguish between positive and negative conditional relationships}, which, to our knowledge, has until now not been possible beyond the linear and jointly Gaussian setting using the ordinary partial correlation.

Let us briefly demonstrate a practical implementation of the LGPC in another situation involving real data. As in the empirical example in the main article, we look at Granger causality,

\begin{figure}
\subfloat[Local partial dependence map of $V_t^*$ vs. $R_{t-1}$ \newline on the plane defined by $V_{t-1}^* = 0$\label{fig:realplots1}]{\includegraphics[width=0.49\linewidth]{figures/realplots-1} }
\subfloat[Local partial dependence map of $R_t$ vs. $V_{t-1}^*$ \newline on the plane defined by $R_{t-1} = 0$\label{fig:realplots2}]{\includegraphics[width=0.49\linewidth]{figures/realplots-2} }\newline
\subfloat[Local partial dependence map of $R_t$ vs. $V_{t-1}^*$ \newline on the  plane defined by $R_{t-1} = -0.05$\label{fig:realplots3}]{\includegraphics[width=0.49\linewidth]{figures/realplots-3} }
\subfloat[Local partial dependence map of $R_t$ vs. $V_{t-1}^*$ \newline on the  plane defined by $R_{t-1} = 0.05$\label{fig:realplots4}]{\includegraphics[width=0.49\linewidth]{figures/realplots-4} }
\caption{Local partial dependence maps for the S\&P 500 data}\label{fig:realplots}
\end{figure}

One fairly recent method for testing for Granger causality is based on the maximal conditional correlation and was introduced by Huang (\protect\hyperlink{ref-huang2010testing}{2010}) and later extended to the time series case by Cheng and Huang (\protect\hyperlink{ref-cheng2012conditional}{2012}). The latter authors use this test to examine whether trading volume Granger caused index value, or vice versa, on the S\&P 500 stock index during the first decade of this century. Using the daily price series \(\{P_t\}\) and volume \(\{V_t\}\) they define the log-differenced series

\[R_t = 100\log\left(\frac{P_t}{P_{t-1}}\right) \,\,\, \textrm{ and } \,\,\, V_t^* = \log\left(\frac{V_t}{V_{t-1}}\right).\]
Further, they denote by \(R_{t-1} \not\Rightarrow V_t^*\) the Granger non-causality from \(\{R_t\}\) to \(\{V_t^*\}\), which they analyze by putting the following null hypothesis to the test:

\begin{equation}
\textrm{H}_0: V_t^* \perp R_{t-1} \,\, | \,\, V_{t-1}^*.
\label{eq:grangernull500}
\end{equation}

Granger non-causality in the opposite direction, \(V_{t-1}^* \not\Rightarrow R_t\) is formulated and tested for in the obvious way.

A linear test rejects \(R_{t-1} \not\Rightarrow V_t^*\), but not \(V_{t-1}^* \not\Rightarrow R_t\). The nonparametric test by Cheng and Huang (\protect\hyperlink{ref-cheng2012conditional}{2012}) rejects both, leading to the natural conclusion that there are nonlinear relationships in the latter situation that remain unseen by linear models. What is it, though, in this particular data set, that leads to such results? Estimates of the local partial Gaussian correlation provide some clues towards answering that question.

We obtain daily observations on price and trading volume on the S\&P 500 index from January 1st 2000 until December 31st 2009 from Yahoo Finance (\protect\hyperlink{ref-yahoo}{2018}), and plot the estimated trivariate LGPC between \(V_t^*\) and \(R_{t-1}\) on the plane defined by \(V_{t-1}^* = 0\) in Figure \ref{fig:realplots1}. Departures from \(\alpha(V_t^*, R_{t-1}) \equiv 0\) provide evidence against \(R_{t-1} \not\Rightarrow V_{t}^*\). Similarly, we plot the estimated LGPC between \(R_t\) and \(V^*_{t-1}\) on the plane \(R_{t-1} = 0\) in Figure \ref{fig:realplots2} in which departures from \(\alpha(R_t, V^*_{t-1}) \equiv 0\) provide evidence against \(V^*_{t-1} \not\Rightarrow R_t\). We have used a proportionality constant of \(c = 3.5\) to calculate the estimates (see discussion on bandwidth selection in Section 4 of the main article). The observations can be seen in the background of both plots.

The differences between Figures \ref{fig:realplots1} and \ref{fig:realplots2} are subtle, but important. In Figure \ref{fig:realplots1}, \(\widehat\alpha(V_t^*, R_{t-1})\) is mostly negative, especially in the data rich portions of the sample space (other values of the conditioning variable \(V_{t-1}^*\) than zero give similar pictures). Indeed, the \emph{global} partial correlation in this situation is \(\widehat{\alpha}_{\textrm{glob}} = -0.086\), which is significantly different from zero. The global partial correlation in the second situation is very small though: \(\widehat{\alpha}_{\textrm{glob}} = -0.0018\), but Figure \ref{fig:realplots2} still reveals departures from conditional independence of similar magnitudes as in Figure \ref{fig:realplots1}. The difference is that the estimated LGPC is positive (primarily in the second and fourth quadrants) as well as negative (in the first and third quadrants), but this pattern collapses to the constant global value \(\alpha(R_t, V_{t-1}^*) = \alpha_{\textrm{glob}} \approx 0\) as the bandwidths tend to infinity. In Figures \ref{fig:realplots3} and \ref{fig:realplots4} we can explore the conditional dependence between \(R_t\) and \(V_{t-1}\) at high and low levels of \(R_{t-1}\), respectively, and we see even more clear differences in this dimension, especially in the first and second quadrants.

The \(p\)-values for tests of the hypothesis \eqref{eq:grangernull500} and its opposite counterpart, using our new test for conditional independence as presented in the next section and with \(c\)-values for the bandwidth as chosen there, are both equal to 0.

\hypertarget{power-studies}{%
\section{POWER STUDIES}\label{power-studies}}

Table 2\(^*\) below is the extended version of Table 2 in the main article, sample size is \(n=100\):

\vspace{1cm}

\input{tables/power_table_dim3_n100}

\newpage

In Table 3\(^*\) we present level and power results for various conditional independence tests applied to the DGPs listed in Table 1 in the paper for sample size \(n = 200\):

\input{tables/power_table_dim3_n200}

Su and White (\protect\hyperlink{ref-su2014testing}{2014}) define two extensions to a subset of the data generating processes defined in Table 1 in which the conditioning variable \(\X_3\) is a vector with two and three components respectively. The first extension turns the conditioning variable \(\X_{3,t}\) in DGP1-DGP2 and DGP5-DGP10 \(X_{3,t}\) into a bivariate vector, where we define DGP1\(^{\prime}\) in the same way as before,

\begin{itemize}
\item[1$^{\prime}$.] $(X_{1,t}, X_{2,t}, \X_{3,t}) = (\epsilon_{1,t}, \epsilon_{2,t}, \fepsilon_{3,t})$,
\end{itemize}

but where \(\fepsilon_{3,t} \sim \mathcal{N}(0, \bm{I}_2)\), and where we define DGP2\(^{\prime}\) and DGP5\(^{\prime}\)-DGP10\(^{\prime}\) by setting \(\X_{3,t} = (X_{1, t-1}, X_{1, t-2})\), keeping \(X_{2,t}\) as before, and,

\begin{itemize}
\tightlist
\item[2$^{\prime}$.] $X_{1,t} = 0.5X_{1,t-1} + 0.25X_{1, t-2} + \epsilon_{1,t}$,
\item[5$^{\prime}$.] $X_{1,t} = 0.5X_{1,t-1} + 0.25X_{1, t-2} + 0.5X_{2,t} + \epsilon_{1,t}$,
\item[6$^{\prime}$.] $X_{1,t} = 0.5X_{1,t-1} + 0.25X_{1, t-2} + 0.5X_{2,t}^2 + \epsilon_{1,t}$,
\item[7$^{\prime}$.] $X_{1,t} = 0.5X_{1,t-1}X_{2, t} + 0.25X_{1,t-2}+ \epsilon_{1,t}$,
\item[8$^{\prime}$.] $X_{1,t} = 0.5X_{1,t-1} + 0.25X_{1,t-2} + 0.5X_{2, t}\epsilon_{1,t}$,
\item[9$^{\prime}$.] $X_{1,t} = \sqrt{h_t}\epsilon_{1,t}$, $h_t = 0.01 + 0.5X_{1,t-1}^2 + 0.25X_{1,t-2}^2 + 0.25X_{2,t}^2$, and
\item[10$^{\prime}$.] Same as DGP10 above, except for the new definition of $\X_{3,t}$.
\end{itemize}

We report the level and power results obtained by Su and White (\protect\hyperlink{ref-su2014testing}{2014}) for the methods CHF, CM, KS and SEL on these data by testing the null hypothesis
\[\textrm{H}_0: X_{t,1} \perp X_{t,2} \,\, | \,\, \X_{t,3},\]
and include results from our new test based on the multivariate simplification of the LGPC as defined in Section 3.1.2 in table 4\(^*\). The results are quite acceptable and compares well with other nonparametric methods.

The second extension introduced by Su and White (\protect\hyperlink{ref-su2014testing}{2014}) increases the dimension of the conditioning variable \(\X_{3,t}\) once more, so that \(\fepsilon_{3,t} \sim \mathcal{N}(0, \bm{I}_3)\), and where DGP2\(^{\prime\prime}\) and DGP5\(^{\prime\prime}\)-DGP10\(^{\prime\prime}\) are defined by by setting \(\X_{3,t} = (X_{1, t-1}, X_{1, t-2}, X_{1, t-3})\), and,

\begin{itemize}
\tightlist
\item[2$^{\prime\prime}$.] $X_{1,t} = 0.5X_{1,t-1} + 0.25X_{1, t-2} + 0.125X_{1,t-3} + \epsilon_{1,t}$,
\item[5$^{\prime\prime}$.] $X_{1,t} = 0.5X_{1,t-1} + 0.25X_{1, t-2} + 0.125X_{1,t-3} + 0.5X_{2,t} + \epsilon_{1,t}$,
\item[6$^{\prime\prime}$.] $X_{1,t} = 0.5X_{1,t-1} + 0.25X_{1, t-2} + 0.125X_{1,t-3} + 0.5X_{2,t}^2 + \epsilon_{1,t}$,
\item[7$^{\prime\prime}$.] $X_{1,t} = 0.5X_{1,t-1}X_{2, t} + 0.25X_{1,t-2} + 0.125X_{1,t-3} + \epsilon_{1,t}$,
\item[8$^{\prime\prime}$.] $X_{1,t} = 0.5X_{1,t-1} + 0.25X_{1,t-2} + 0.125X_{1,t-3} + 0.5X_{2, t}\epsilon_{1,t}$,
\item[9$^{\prime\prime}$.] $X_{1,t} = \sqrt{h_t}\epsilon_{1,t}$, $h_t = 0.01 + 0.5X_{1,t-1}^2 + 0.25X_{1,t-2}^2 + 0.125X_{1,t-3}^2 + 0.25X_{2,t}^2$, and
\item[10$^{\prime\prime}$.] Same as DGP10 and DGP10$^{\prime}$ above, except for the new definition of $\X_{3,t}$.
\end{itemize}
\input{tables/power_table_dim4}
\input{tables/power_table_dim5}

Again, we observe simulation results in Table 5\(^*\). Contrary to Su and White (\protect\hyperlink{ref-su2014testing}{2014}), we have also run our test for \(n=100\) which reveals that we can obtain some power also in that case. All in all, the simplified test based on the LGPC performs mostly on par with other non-parametric tests in this setting as well.

\newpage

\hypertarget{references}{%
\section*{REFERENCES}\label{references}}
\addcontentsline{toc}{section}{REFERENCES}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-arcones1995law}{}%
Arcones, Miguel A. 1995. ``The Law of the Iterated Logarithm for Empirical Processes Under Absolute Regularity.'' \emph{Journal of Theoretical Probability} 8 (2): 433--51. \url{https://doi.org/10.1007/BF02212887}.

\leavevmode\hypertarget{ref-bill:2008}{}%
Billingsley, Patrick. 2008. \emph{Probability and Measure}. John Wiley \& Sons.

\leavevmode\hypertarget{ref-brockwell1991time}{}%
Brockwell, Peter J, and Richard A Davis. 2006. \emph{Time Series: Theory and Methods: Theory and Methods}. Springer.

\leavevmode\hypertarget{ref-cheng2012conditional}{}%
Cheng, Yu-Hsiang, and Tzee-Ming Huang. 2012. ``A Conditional Independence Test for Dependent Data Based on Maximal Conditional Correlation.'' \emph{Journal of Multivariate Analysis} 107: 210--26.

\leavevmode\hypertarget{ref-fan2008nonlinear}{}%
Fan, Jianqing, and Qiwei Yao. 2008. \emph{Nonlinear Time Series: Nonparametric and Parametric Methods}. Springer Science \& Business Media.

\leavevmode\hypertarget{ref-gonccalves2002bootstrap}{}%
Gonçalves, Si@articlelacal2017local, title=Local Gaussian autocorrelation and tests for serial independence, author=Lacal, Virginia and TjØstheim, Dag, journal=Journal of Time Series Analysis, volume=38, number=1, pages=51--71, year=2017, publisher=Wiley Online Library lvia, and Halbert White. 2002. ``The Bootstrap of the Mean for Dependent Heterogeneous Arrays.'' \emph{Econometric Theory} 18 (6): 1367--84.

\leavevmode\hypertarget{ref-gonccalves2004maximum}{}%
Gonçalves, Silvia, and Halbert White. 2004. ``Maximum Likelihood and the Bootstrap for Nonlinear Dynamic Models.'' \emph{Journal of Econometrics} 119 (1): 199--219.

\leavevmode\hypertarget{ref-hall1980martingale}{}%
Hall, P, and CC Heyde. 1980. ``Martingale Limit Theory and Its Applications, Acad.'' \emph{Press, New York}.

\leavevmode\hypertarget{ref-hjort1996locally}{}%
Hjort, Nils Lid, and MC Jones. 1996. ``Locally Parametric Nonparametric Density Estimation.'' \emph{The Annals of Statistics}, 1619--47.

\leavevmode\hypertarget{ref-huang2010testing}{}%
Huang, Tzee-Ming. 2010. ``Testing Conditional Independence Using Maximal Nonlinear Conditional Correlation.'' \emph{The Annals of Statistics} 38 (4): 2047--91.

\leavevmode\hypertarget{ref-joe1989estimation}{}%
Joe, Harry. 1989. ``Estimation of Entropy and Other Functionals of a Multivariate Density.'' \emph{Annals of the Institute of Statistical Mathematics} 41 (4): 683--97.

\leavevmode\hypertarget{ref-jordanger2017nonlinear}{}%
Jordanger, Lars Arne, and Dag Tjøstheim. 2020. ``Nonlinear Spectral Analysis: A Local Gaussian Approach.'' \emph{arXiv Preprint arXiv:1708.02166v4, Revised version submitted to the Journal of the American Statistical Association}.

\leavevmode\hypertarget{ref-lacal2017local}{}%
Lacal, Virginia, and Dag Tjøstheim. 2017. ``Local Gaussian Autocorrelation and Tests for Serial Independence.'' \emph{Journal of Time Series Analysis} 38 (1): 51--71.

\leavevmode\hypertarget{ref-lacal2018estimating}{}%
---------. 2018. ``Estimating and Testing Nonlinear Local Dependence Between Two Time Series.'' \emph{Journal of Business \& Economic Statistics}, 1--13.

\leavevmode\hypertarget{ref-otneim2017locally}{}%
Otneim, Håkon, and Dag Tjøstheim. 2017. ``The Locally Gaussian Density Estimator for Multivariate Data.'' \emph{Statistics and Computing} 27 (6): 1595--1616.

\leavevmode\hypertarget{ref-otneim2017conditional}{}%
---------. 2018. ``Conditional Density Estimation Using the Local Gaussian Correlation.'' \emph{Statistics and Computing} 28 (2): 303--21.

\leavevmode\hypertarget{ref-paparoditis2000local}{}%
Paparoditis, Efstathios, and Dimitris N Politis. 2000. ``The Local Bootstrap for Kernel Estimators Under General Dependence Conditions.'' \emph{Annals of the Institute of Statistical Mathematics} 52 (1): 139--59.

\leavevmode\hypertarget{ref-rio1995functional}{}%
Rio, Emmanuel. 1995. ``The Functional Law of the Iterated Logarithm for Stationary Strongly Mixing Sequences.'' \emph{The Annals of Probability}, 1188--1203.

\leavevmode\hypertarget{ref-schervish1995theory}{}%
Schervish, Mark J. 2012. \emph{Theory of Statistics}. Springer Science \& Business Media.

\leavevmode\hypertarget{ref-su2008nonparametric}{}%
Su, Liangjun, and Halbert White. 2008. ``A Nonparametric Hellinger Metric Test for Conditional Independence.'' \emph{Econometric Theory} 24 (4): 829--64.

\leavevmode\hypertarget{ref-su2014testing}{}%
---------. 2014. ``Testing Conditional Independence via Empirical Likelihood.'' \emph{Journal of Econometrics} 182 (1): 27--44.

\leavevmode\hypertarget{ref-terasvirta2010modelling}{}%
Teräsvirta, Timo, Dag Tjøstheim, and Clive W. J. Granger. 2010. \emph{Modelling Nonlinear Economic Time Series}. Oxford University Press.

\leavevmode\hypertarget{ref-tjostheim2013local}{}%
Tjøstheim, Dag, and Karl Ove Hufthammer. 2013. ``Local Gaussian Correlation: A New Measure of Dependence.'' \emph{Journal of Econometrics} 172 (1): 33--48.

\leavevmode\hypertarget{ref-van2007parameter}{}%
Van den Bos, Adriaan. 2007. \emph{Parameter Estimation for Scientists and Engineers}. John Wiley \& Sons.

\leavevmode\hypertarget{ref-wang2017characteristic}{}%
Wang, Xia, and Yongmiao Hong. 2017. ``Characteristic Function Based Testing for Conditional Independence: A Nonparametric Regression Approach.'' \emph{Econometric Theory}, 1--35.

\leavevmode\hypertarget{ref-yahoo}{}%
Yahoo Finance. 2018. \emph{Accessed June 2018}.

\end{document}

### 2020-09-02 lgpc_JBES.Rmd
Implementing comments from Dag and one referee

@@
 \begin{equation}
-\fSigma_{12|3} = \fSigma^{11} - \fSigma^{12}\left(\fSigma^{-1}\right)^{22}\fSigma^{21},
+\fSigma_{12|3} = \fSigma^{11} - \fSigma^{12}\left(\fSigma^{22}\right)^{-1}\fSigma^{21},
 (\#eq:matrixpartial)
@@
 \end{equation}
-The transformation enables us to simplify the local Gaussian representation \@ref(eq:localgaussian) by writing the density $f_{\Z}$ of $\Z$ as 
+The transformation enables us to simplify the local Gaussian approximation \@ref(eq:localgaussian) by writing the density $f_{\Z}$ of $\Z$  at the point $\fv = \z$ as 
 \begin{equation}
@@
 \end{equation}  
-where, as in Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional] in a simplified representation, we have fixed local means and standard deviations $\mu_j(\z) = 0$ and $\sigma_j^2(\z) = 1$, $j = 1,\ldots,p$, and where $\R(\z) = \{\rho_{jk}(\z)\}$ is the *local correlation matrix*.
+where, as in Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional] in a further simplified representation, we have fixed local means and standard deviations $\mu_j(\z) = 0$ and $\sigma_j^2(\z) = 1$, $j = 1,\ldots,p$, and where $\R(\z) = \{\rho_{jk}(\z)\}$ is the *local correlation matrix*.
 
@@
 
-$$f_{z_1|\z_3}(z_1) = \frac{1}{|2\pi\sigma_{1|3}(z_1)|^{1/2}}\times\exp \left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1)}{\sigma_{1|3}(z_1)}\right)^2\right\},$$
+$$f_{z_1|\z_3}(z_1|\z_3) = \frac{1}{|2\pi\sigma_{1|3}(z_1)|^{1/2}}\times\exp \left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1)}{\sigma_{1|3}(z_1)}\right)^2\right\},$$
 where
@@
 \end{equation}
-in each point $\z$, where $K_{\hh}(\x) = |\hh|^{-1/2}K(\hh^{-1/2}\x)$, $K$ is a non-negative and radially symmetric kernel function that satisfies $\int K(\x)\,\textrm{d}\x=1$, and $\hh$ is a diagonal $3\times3$ matrix of bandwidths that serve as smoothing parameters.
+in each point $\z$, where $K_{\hh}(\x) = |\hh|^{-1}K(\hh^{-1}\x)$, $K$ is a non-negative and radially symmetric kernel function that satisfies $\int K(\x)\,\textrm{d}\x=1$, and $\hh$ is a diagonal $3\times3$ matrix of bandwidths that serve as smoothing parameters.
 The estimate $\hf_{\Z}(\z) = \psi(\z, \hR(\z))$ aims at minimizing a locally weighted Kullback-Leibler distance to $f_{\Z}(\z)$, and we refer to @hjort1996locally and @tjostheim2013local for much more details about this construction, and to Section \@ref(chap:asymptotic) for the relevant asymptotic estimation theory.
@@
 \begin{equation}
-\hrho_{jk}(z_j, z_k) = \arg\max_{\rho_{jk}} n^{-1} \sum_{t=1}^nK_{\hh}(\hZ_t - \z)\log(\hZ_t, \rho_{jk}) - \int K_{\hh}(\y - \z)\psi(\y, \rho_{jk})\,\textrm{d}\y, 
+\hrho_{jk}(z_j, z_k) = \arg\max_{\rho_{jk}} n^{-1} \sum_{t=1}^nK_{\hh}(\hZ_t - \z)\log\psi(\hZ_t, \rho_{jk}) - \int K_{\hh}(\y - \z)\psi(\y, \rho_{jk})\,\textrm{d}\y, 
 (\#eq:locallik2)
@@
 The asymptotic behavior of $\halpha(\z)$ can thus be derived directly from the asymptotic behavior of $\hR(\z)$, which, for the most part, has been established in earlier works, see e.g.
-@tjostheim2013local and @otneim2017locally.
+@tjostheim2013local and Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional].
 Some key results follow below, and we refer to the supplementary material, Sections B--D for proofs of the asymptotic distributional results in this subsection, and to Section A.3 for a discussion of more general situations and for a result on consistency.
@@
 \end{equation}
-where $f_{\Z}(\z) = f_{Z_1,Z_2,Z_3}(z_1,z_2,z_3)$ is the density function of $\Z$, and $\fu(\cdot, \rho_{\hh})$ is the column vector of local score functions $\nabla \log\psi(\cdot, \rho_{\hh})$, and where the gradient is taken with respect to the parameters $\rho_{12}(\z)$, $\rho_{13}(\z)$ and $\rho_{23}(\z)$.
+where $f_{\Z}(\z) = f_{Z_1,Z_2,Z_3}(z_1,z_2,z_3)$ is the density function of $\Z$ (at the point $\z$), and $\fu(\cdot, \rho_{\hh})$ is the column vector of local score functions $\nabla \log\psi(\cdot, \rho_{\hh})$, and where the gradient is taken with respect to the parameters $\rho_{12}(\z)$, $\rho_{13}(\z)$ and $\rho_{23}(\z)$.
 Denote by $\frho = \frho(\z) = \{\rho_{12}(\z), \rho_{13}(\z), \rho_{23}(\z)\}$ the vector of local correlations.
@@
 5.
-The non-negative kernel function $K(\cdot)$ sasisfies $\sup_{\z}K(\z) < \infty$, $\int K(\y)\,\textrm{d}\y < \infty$, $\partial/\partial z_iK(\z) <\infty$, and $\lim_{z_i \rightarrow \infty}|z_iK(\z)| = 0$ for $i = 1,2,3$.
+The non-negative kernel function $K(\cdot)$ satisfies $\sup_{\z}K(\z) < \infty$, $\int K(\y)\,\textrm{d}\y < \infty$, $\partial/\partial z_iK(\z) <\infty$, and $\lim_{z_i \rightarrow \infty}|z_iK(\z)| = 0$ for $i = 1,2,3$.
 
@@
 \begin{equation}
-\sqrt{nb^5}\Jb\Mb^{-1/2}\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
+\sqrt{nb^5}\Mb^{-1/2}\Jb\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
 (\#eq:an3)
@@
 
+It is shown in Section B of the Supplement that the matrices $\Jb$ and $\Mb$ are both of order $b^2$.
 Theorem \@ref(thm:loccor3) and its assumptions above are variations of the assumptions and the results presented by @hjort1996locally, and in particular Theorem 3 by @tjostheim2013local.
@@
 Finally, we present the limiting distribution of the LGPC between $Z_1$ and $Z_2$ given $Z_3, \ldots, Z_p$ when the components in $\Z = (Z_1, \ldots, Z_p)$ are replaced by marginally standard normal *pseudo-observations* as defined in equation \@ref(eq:pseudoobservations).
-The following result ensures that Theorems \@ref(thm:loccor3) and Theorem 3 by @otneim2017conditional, and thus eq.
+The following result ensures that Theorems \@ref(thm:loccor3) and Theorem 2 by @otneim2017conditional, and thus eq.
 \@ref(eq:limitlgpc), still holds.
@@
 We see the results in Figure \@ref(fig:simex2-plots).
-In panel (a), we generate $n=1000$ simulated pairs $(X_1, X_2)$ from this model.
-In panel (b)-(d) we see the estimated LGPC plotted over suitable grids, and where the conditioning variable $X_3 = \rho$ has been fixed at the respective values $-0.9$, $0$ and $0.9$, and we see clearly how the dependence between $X_1$ and $X_2$ dramatically changes.
-It is straightforward to see that $X_1$ and $X_2$ are uncorrelated and that the ordinary partial correlation between $X_1$ and $X_2$ given $X_3$ is equal to zero.
+In panel (a), we display $n=1000$ simulated pairs $(X_1, X_2)$ from this model.
+In panel (b)-(d) we see the estimated LGPC plotted over suitable grids, and where the conditioning variable $X_3 = \rho$ has been fixed at the respective values $-0.9$, $0$ and $0.9$, and we see clearly how the dependence between $X_1$ and $X_2$ dramatically changes in an intuitively reasonable way.
+By a conditioning argument one can see that $X_1$ and $X_2$ are uncorrelated and that the ordinary partial correlation between $X_1$ and $X_2$ given $X_3$ is equal to zero.
 Furthermore, we note that the pairwise simplification defined in Section 3.1.2 would also not be able to measure the *conditional* relationship between $X_1$ and $X_2$ given $X_3$ in this case, as we see clearly from eq.
@@
 
-Departures from many types of conditional independence lead to large values of $T_{n,b}$ that, if larger than a critical value, leads to the rejection of \@ref(eq:ci-test).
+Departures from conditional independence lead to large values of $T_{n,b}$ that, if larger than a critical value, leads to the rejection of \@ref(eq:ci-test).
 Asymptotically, one might expect that approximate $p$-values for our test can be extracted from the limiting distribution of the test statistic $T_{n,b}$, which we derive in the online supplement Section F.
@@
 2. Using the accept-reject algorithm, generate $B$ samples, each of size $n$, from $\widehat f_{Z_1|\Z_3}$ and $\widehat f_{Z_2|\Z_3}$, leading to $B$ replicates 
-$$Z_m^* = \Big\{\{Z_{1t}^*, Z_{2t}^*, \Z_{3t}\}_i\Big\}_m, \qquad t = 1,\ldots,n, \,\, m = 1,\ldots,B,$$
+$$Z_m^* = \Big\{\{Z_{1t}^*, Z_{2t}^*, \Z_{3t}^*\}_i\Big\}_m, \qquad t = 1,\ldots,n, \,\, m = 1,\ldots,B,$$
 of $\Z$ under the null hypothesis.
@@
 2.& $X_{1,t} = 0.5X_{1, t-1} + \epsilon_{1,t}$, $X_{2,t} = 0.5X_{2,t-1} + \epsilon_{2,t}$, $X_{3,t} = X_{1,t-1}$.\\
-3.& $X_{1,t} = \epsilon_{1,t}\sqrt{0.01 + 0.5X_ {t-1}^2}$, $X_{2,t} = 0.5X_{2, t-1} + \epsilon_{2,t}$, $X_{3,t} = X_{1,t-1}$.\\
+3.& $X_{1,t} = \epsilon_{1,t}\sqrt{0.01 + 0.5X_ {1,t-1}^2}$, $X_{2,t} = 0.5X_{2, t-1} + \epsilon_{2,t}$, $X_{3,t} = X_{1,t-1}$.\\
 4.& $X_{1,t} = \epsilon_{1,t}\sqrt{h_{1,t}}$, $X_{2,t} = \epsilon_{2,t}\sqrt{h_{2,t}}$, $X_{3,t} = X_{1,t-1}$, \\ & $h_{1,t}  = 0.01 + 0.9h_{1,t-1} + 0.05X_{1,t-1}^2$, $h_{2,t} = 0.01 + 0.9h_{2,t-1} + 0.05X_{2,t-1}^2$.
@@
 10.
-& $X_{1,t} = \epsilon_{1,t}\sqrt{h_{1,t}}$, $X_{2,t} = \epsilon_{2,t}\sqrt{h_{2,t}}$, $X_{3,t} = X_{1,t-1}$, $h_{1,t} = 0.01 + 0.1h_{1, t-1} + 0.4X_{1,t-1}^2 +$ \\ & $0.5X_{2,t}^2$, $h_{2,t} = 0.01 + 0.9h_{2, t-1} + 0.5Y_t^2$.\\
+& $X_{1,t} = \epsilon_{1,t}\sqrt{h_{1,t}}$, $X_{2,t} = \epsilon_{2,t}\sqrt{h_{2,t}}$, $X_{3,t} = X_{1,t-1}$, $h_{1,t} = 0.01 + 0.1h_{1, t-1} + 0.4X_{1,t-1}^2 +$ \\ & $0.5X_{2,t}^2$, $h_{2,t} = 0.01 + 0.9h_{2, t-1} + 0.5X_t^2$.\\
 \hline
@@
 
-The first set of simulation results are collected from @su2008nonparametric: In Table \ref{tab:n100} LIN is a standard linear Granger causality test, CM and KS are tests by @linton1997conditional that use statistics of the two versions of the nonparametric test developed by @delgado2001significance.
+The first set of simulation results are collected from @su2008nonparametric: In Table \ref{tab:n100} LIN is a standard linear Granger causality test, SCM and SKS are tests by @linton1997conditional that use statistics of the two versions of the nonparametric test developed by @delgado2001significance.
 HEL refers to the Hellinger distance test that @su2008nonparametric present.
@@
 We point out again, however, that testing for conditional independence is merely one, albeit an important one, of several chief objectives for introducing the LGPC.
-Possible refinements of the LGPC testing is mentioned in the beginning of Section \@ref(fauna).
+Possible refinements of the LGPC testing is mentioned in the beginning of Section \@ref(chap:testing).
 
@@
 First we define and develop the LGPC, a signed local measure of conditional dependence that has useful properties and is easy to interpret.
-It ca measure both positive and negative conditional dependence.
+It can measure both positive and negative conditional dependence.
 We then explore the possibility of using the LGPC in estimation of conditional dependence and indicate its potential in testing for conditional independence.
@@
 
-**Acknowledgements.** The authors gratefully acknowledge the editor as well as two anonymus referees, whose comments have greatly improved this paper.
+**Acknowledgements.** The authors gratefully acknowledge the input of an associate editor as well as two anonymuos referees, whose comments have greatly improved this paper.
 

### 2020-09-02 lgpc_JBES.tex
Implementing comments from Dag and one referee

@@
 \makesavenoteenv{longtable}
-\usepackage{graphicx,grffile}
+\usepackage{graphicx}
 \makeatletter
@@
 \definecolor{Gray}{gray}{0.9}
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newenvironment{cslreferences}%
+  {\setlength{\parindent}{0pt}%
+  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
+  {\par}
 
@@
 \newtheorem*{solution}{Solution}
+\let\BeginKnitrBlock\begin \let\EndKnitrBlock\end
 \begin{document}
@@
 \begin{equation}
-\fSigma_{12|3} = \fSigma^{11} - \fSigma^{12}\left(\fSigma^{-1}\right)^{22}\fSigma^{21},
+\fSigma_{12|3} = \fSigma^{11} - \fSigma^{12}\left(\fSigma^{22}\right)^{-1}\fSigma^{21},
 \label{eq:matrixpartial}
@@
 \end{equation}
-The transformation enables us to simplify the local Gaussian representation \eqref{eq:localgaussian} by writing the density \(f_{\Z}\) of \(\Z\) as
+The transformation enables us to simplify the local Gaussian approximation \eqref{eq:localgaussian} by writing the density \(f_{\Z}\) of \(\Z\) at the point \(\fv = \z\) as
 \begin{equation}
@@
 \end{equation}\\
-where, as in Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, \protect\hyperlink{ref-otneim2017conditional}{2018}) in a simplified representation, we have fixed local means and standard deviations \(\mu_j(\z) = 0\) and \(\sigma_j^2(\z) = 1\), \(j = 1,\ldots,p\), and where \(\R(\z) = \{\rho_{jk}(\z)\}\) is the \emph{local correlation matrix}.
+where, as in Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, \protect\hyperlink{ref-otneim2017conditional}{2018}) in a further simplified representation, we have fixed local means and standard deviations \(\mu_j(\z) = 0\) and \(\sigma_j^2(\z) = 1\), \(j = 1,\ldots,p\), and where \(\R(\z) = \{\rho_{jk}(\z)\}\) is the \emph{local correlation matrix}.
 
@@
 
-\[f_{z_1|\z_3}(z_1) = \frac{1}{|2\pi\sigma_{1|3}(z_1)|^{1/2}}\times\exp \left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1)}{\sigma_{1|3}(z_1)}\right)^2\right\},\]
+\[f_{z_1|\z_3}(z_1|\z_3) = \frac{1}{|2\pi\sigma_{1|3}(z_1)|^{1/2}}\times\exp \left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1)}{\sigma_{1|3}(z_1)}\right)^2\right\},\]
 where
@@
 \end{equation}
-in each point \(\z\), where \(K_{\hh}(\x) = |\hh|^{-1/2}K(\hh^{-1/2}\x)\), \(K\) is a non-negative and radially symmetric kernel function that satisfies \(\int K(\x)\,\textrm{d}\x=1\), and \(\hh\) is a diagonal \(3\times3\) matrix of bandwidths that serve as smoothing parameters.
+in each point \(\z\), where \(K_{\hh}(\x) = |\hh|^{-1}K(\hh^{-1}\x)\), \(K\) is a non-negative and radially symmetric kernel function that satisfies \(\int K(\x)\,\textrm{d}\x=1\), and \(\hh\) is a diagonal \(3\times3\) matrix of bandwidths that serve as smoothing parameters.
 The estimate \(\hf_{\Z}(\z) = \psi(\z, \hR(\z))\) aims at minimizing a locally weighted Kullback-Leibler distance to \(f_{\Z}(\z)\), and we refer to Hjort and Jones (\protect\hyperlink{ref-hjort1996locally}{1996}) and Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) for much more details about this construction, and to Section \ref{chap:asymptotic} for the relevant asymptotic estimation theory.
@@
 \begin{equation}
-\hrho_{jk}(z_j, z_k) = \arg\max_{\rho_{jk}} n^{-1} \sum_{t=1}^nK_{\hh}(\hZ_t - \z)\log(\hZ_t, \rho_{jk}) - \int K_{\hh}(\y - \z)\psi(\y, \rho_{jk})\,\textrm{d}\y, 
+\hrho_{jk}(z_j, z_k) = \arg\max_{\rho_{jk}} n^{-1} \sum_{t=1}^nK_{\hh}(\hZ_t - \z)\log\psi(\hZ_t, \rho_{jk}) - \int K_{\hh}(\y - \z)\psi(\y, \rho_{jk})\,\textrm{d}\y, 
 \label{eq:locallik2}
@@
 Equations \eqref{eq:matrixlocalpartial}, \eqref{eq:definition} and their empirical counterparts \eqref{eq:localpartialcov} and \eqref{eq:estimate} demonstrate clearly that the LGPC is nothing more than a deterministic function of the local correlation matrix, in the same way as the ordinary partial correlation is a function of the ordinary correlation matrix.
-The asymptotic behavior of \(\halpha(\z)\) can thus be derived directly from the asymptotic behavior of \(\hR(\z)\), which, for the most part, has been established in earlier works, see e.g.~Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) and Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}).
+The asymptotic behavior of \(\halpha(\z)\) can thus be derived directly from the asymptotic behavior of \(\hR(\z)\), which, for the most part, has been established in earlier works, see e.g.~Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) and Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, \protect\hyperlink{ref-otneim2017conditional}{2018}).
 Some key results follow below, and we refer to the supplementary material, Sections B--D for proofs of the asymptotic distributional results in this subsection, and to Section A.3 for a discussion of more general situations and for a result on consistency.
@@
 \end{equation}
-where \(f_{\Z}(\z) = f_{Z_1,Z_2,Z_3}(z_1,z_2,z_3)\) is the density function of \(\Z\), and \(\fu(\cdot, \rho_{\hh})\) is the column vector of local score functions \(\nabla \log\psi(\cdot, \rho_{\hh})\), and where the gradient is taken with respect to the parameters \(\rho_{12}(\z)\), \(\rho_{13}(\z)\) and \(\rho_{23}(\z)\).
+where \(f_{\Z}(\z) = f_{Z_1,Z_2,Z_3}(z_1,z_2,z_3)\) is the density function of \(\Z\) (at the point \(\z\)), and \(\fu(\cdot, \rho_{\hh})\) is the column vector of local score functions \(\nabla \log\psi(\cdot, \rho_{\hh})\), and where the gradient is taken with respect to the parameters \(\rho_{12}(\z)\), \(\rho_{13}(\z)\) and \(\rho_{23}(\z)\).
 Denote by \(\frho = \frho(\z) = \{\rho_{12}(\z), \rho_{13}(\z), \rho_{23}(\z)\}\) the vector of local correlations.
@@
 
-\begin{theorem}
+\BeginKnitrBlock{theorem}
 \protect\hypertarget{thm:loccor3}{}{\label{thm:loccor3} }Let \(\{\Z_t\}_{t=1}^n\) be observations on the marginally standard normally distributed random vector \(\Z\) having joint density \(f_{\Z}(\z)\).
@@
 \item
-  The non-negative kernel function \(K(\cdot)\) sasisfies \(\sup_{\z}K(\z) < \infty\), \(\int K(\y)\,\textrm{d}\y < \infty\), \(\partial/\partial z_iK(\z) <\infty\), and \(\lim_{z_i \rightarrow \infty}|z_iK(\z)| = 0\) for \(i = 1,2,3\).
+  The non-negative kernel function \(K(\cdot)\) satisfies \(\sup_{\z}K(\z) < \infty\), \(\int K(\y)\,\textrm{d}\y < \infty\), \(\partial/\partial z_iK(\z) <\infty\), and \(\lim_{z_i \rightarrow \infty}|z_iK(\z)| = 0\) for \(i = 1,2,3\).
 \end{enumerate}
@@
 \begin{equation}
-\sqrt{nb^5}\Jb\Mb^{-1/2}\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
+\sqrt{nb^5}\Mb^{-1/2}\Jb\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
 \label{eq:an3}
@@
 \end{align*}
-\end{theorem}
+\EndKnitrBlock{theorem}
 
+It is shown in Section B of the Supplement that the matrices \(\Jb\) and \(\Mb\) are both of order \(b^2\).
 Theorem \ref{thm:loccor3} and its assumptions above are variations of the assumptions and the results presented by Hjort and Jones (\protect\hyperlink{ref-hjort1996locally}{1996}), and in particular Theorem 3 by Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}).
@@
 Finally, we present the limiting distribution of the LGPC between \(Z_1\) and \(Z_2\) given \(Z_3, \ldots, Z_p\) when the components in \(\Z = (Z_1, \ldots, Z_p)\) are replaced by marginally standard normal \emph{pseudo-observations} as defined in equation \eqref{eq:pseudoobservations}.
-The following result ensures that Theorems \ref{thm:loccor3} and Theorem 3 by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}), and thus eq.
+The following result ensures that Theorems \ref{thm:loccor3} and Theorem 2 by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}), and thus eq.
 \eqref{eq:limitlgpc}, still holds.
@@
 
-\begin{theorem}
+\BeginKnitrBlock{theorem}
 \protect\hypertarget{thm:pseudo}{}{\label{thm:pseudo} }Under the conditions in Theorems \ref{thm:loccor3} and Theorem 2 by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}), assume further that the kernel function \(K(\cdot)\) has bounded support.
 Replacing the marginally standard normal vector \(\Z\) with the approximately marginally standard normal vector \(\hZ\), does not change the conclusions in these theorems.
-\end{theorem}
+\EndKnitrBlock{theorem}
 
-\begin{proof}
+\BeginKnitrBlock{proof}
 \iffalse{} {Proof. } \fi{}See Section D of the online supplement.
-\end{proof}
+\EndKnitrBlock{proof}
 
@@
 We see the results in Figure \ref{fig:simex2-plots}.
-In panel (a), we generate \(n=1000\) simulated pairs \((X_1, X_2)\) from this model.
-In panel (b)-(d) we see the estimated LGPC plotted over suitable grids, and where the conditioning variable \(X_3 = \rho\) has been fixed at the respective values \(-0.9\), \(0\) and \(0.9\), and we see clearly how the dependence between \(X_1\) and \(X_2\) dramatically changes.
-It is straightforward to see that \(X_1\) and \(X_2\) are uncorrelated and that the ordinary partial correlation between \(X_1\) and \(X_2\) given \(X_3\) is equal to zero.
+In panel (a), we display \(n=1000\) simulated pairs \((X_1, X_2)\) from this model.
+In panel (b)-(d) we see the estimated LGPC plotted over suitable grids, and where the conditioning variable \(X_3 = \rho\) has been fixed at the respective values \(-0.9\), \(0\) and \(0.9\), and we see clearly how the dependence between \(X_1\) and \(X_2\) dramatically changes in an intuitively reasonable way.
+By a conditioning argument one can see that \(X_1\) and \(X_2\) are uncorrelated and that the ordinary partial correlation between \(X_1\) and \(X_2\) given \(X_3\) is equal to zero.
 Furthermore, we note that the pairwise simplification defined in Section 3.1.2 would also not be able to measure the \emph{conditional} relationship between \(X_1\) and \(X_2\) given \(X_3\) in this case, as we see clearly from eq.
@@
 
-Departures from many types of conditional independence lead to large values of \(T_{n,b}\) that, if larger than a critical value, leads to the rejection of \eqref{eq:ci-test}.
+Departures from conditional independence lead to large values of \(T_{n,b}\) that, if larger than a critical value, leads to the rejection of \eqref{eq:ci-test}.
 Asymptotically, one might expect that approximate \(p\)-values for our test can be extracted from the limiting distribution of the test statistic \(T_{n,b}\), which we derive in the online supplement Section F.
@@
   Using the accept-reject algorithm, generate \(B\) samples, each of size \(n\), from \(\widehat f_{Z_1|\Z_3}\) and \(\widehat f_{Z_2|\Z_3}\), leading to \(B\) replicates
-  \[Z_m^* = \Big\{\{Z_{1t}^*, Z_{2t}^*, \Z_{3t}\}_i\Big\}_m, \qquad t = 1,\ldots,n, \,\, m = 1,\ldots,B,\]
+  \[Z_m^* = \Big\{\{Z_{1t}^*, Z_{2t}^*, \Z_{3t}^*\}_i\Big\}_m, \qquad t = 1,\ldots,n, \,\, m = 1,\ldots,B,\]
   of \(\Z\) under the null hypothesis.
@@
 
-The first set of simulation results are collected from Su and White (\protect\hyperlink{ref-su2008nonparametric}{2008}): In Table \ref{tab:n100} LIN is a standard linear Granger causality test, CM and KS are tests by Linton and Gozalo (\protect\hyperlink{ref-linton1997conditional}{1997}) that use statistics of the two versions of the nonparametric test developed by Delgado and Manteiga (\protect\hyperlink{ref-delgado2001significance}{2001}).
+The first set of simulation results are collected from Su and White (\protect\hyperlink{ref-su2008nonparametric}{2008}): In Table \ref{tab:n100} LIN is a standard linear Granger causality test, SCM and SKS are tests by Linton and Gozalo (\protect\hyperlink{ref-linton1997conditional}{1997}) that use statistics of the two versions of the nonparametric test developed by Delgado and Manteiga (\protect\hyperlink{ref-delgado2001significance}{2001}).
 HEL refers to the Hellinger distance test that Su and White (\protect\hyperlink{ref-su2008nonparametric}{2008}) present.
@@
 We point out again, however, that testing for conditional independence is merely one, albeit an important one, of several chief objectives for introducing the LGPC.
-Possible refinements of the LGPC testing is mentioned in the beginning of Section \ref{fauna}.
+Possible refinements of the LGPC testing is mentioned in the beginning of Section \ref{chap:testing}.
 
@@
 
-\textbf{Acknowledgements.} The authors gratefully acknowledge the editor as well as two anonymus referees, whose comments have greatly improved this paper.
+\textbf{Acknowledgements.} The authors gratefully acknowledge the input of an associate editor as well as two anonymuos referees, whose comments have greatly improved this paper.
 
@@
 \hypertarget{refs}{}
+\begin{cslreferences}
 \leavevmode\hypertarget{ref-baba2004partial}{}%
@@
 Zhang, Kun, Jonas Peters, Dominik Janzing, and Bernhard Schölkopf. 2012. ``Kernel-Based Conditional Independence Test and Application in Causal Discovery.'' \emph{arXiv Preprint arXiv:1202.3775}.
+\end{cslreferences}
 

### 2020-09-02 lgpc_JBES_supp.Rmd
Implementing comments from Dag and one referee

@@
 
-All of these parameters can be described and estimated locally in the exact same way as laid out in the main article, resulting from the locally Gaussian joint density
+All of these parameters can be described and estimated locally in the exact same way as laid out in the main article, resulting from the locally Gaussian joint density $\psi(\z,\fv)$ at the point $\fv = \z$,
 
@@
 \end{equation}
-and the locally Gaussian *conditional* density
+and the locally Gaussian *conditional* quantity
 
@@
 Let $\D = \textrm{diag}\left\{\fSigma_{12|3}\right\}$.
-The local Gaussian partial correlation matrix can be expressed as the lower left hand block matrix of $\D^{-1/2}\fSigma_{12|3}\D^{-1/2}$, and so the LGPC matrix is thus given as the upper right hand block of the matrix
+The local Gaussian partial correlation matrix can be expressed as the lower left hand block matrix of $\D^{-1/2}\fSigma_{12|3}\D^{-1/2}$, and so the LGPC matrix is thus given as the lower left hand block of the matrix
 \begin{equation}
@@
 One can also, based on \eqref{eq:offdiag1} and \eqref{eq:offdiag2}, test for conditional independence between combinations of components given $\Z_3$.
-Equations \eqref{eq:offdiag1}--\eqref{eq:offdiag3} above give explicit expressions for the conditional correlation matrix of $\Z_1$ and $\Z_2$ given $\Z_3$.
+Equations \eqref{eq:offdiag1}--\eqref{eq:offdiag3} above give explicit expressions for the conditional correlation matrix of $\Z_1$ and $\Z_2$ given $\Z_3$, where estimated values give a measure of local conditional dependence between the components.
 
@@
 $$\fSigma^{11} = \begin{pmatrix} 1 & \rho_{12} \\ \rho_{21} & 1 \end{pmatrix}, \fSigma_{13} = \left\{\rho_{1i}^{1,3}\right\}, i=1,\ldots,d_3; \fSigma_{23} = \left\{\rho_{2i}^{2,3}\right\}; i=1,\ldots,d_3,$$
-hence $\begin{pmatrix} \fSigma_{13} \fSigma_{23}\end{pmatrix}^T$ is of dimension $2\times d_3$, whereas $\begin{pmatrix} \fSigma_{31} \fSigma_{32}\end{pmatrix}$ is of dimension $d_3\times 2$ such that
-$$\begin{pmatrix} \fSigma_{13} \\ \fSigma_{23} \end{pmatrix} \fSigma_{33}^{-1} \begin{pmatrix} \fSigma_{31} & \fSigma_32 \end{pmatrix}$$
+hence $\begin{pmatrix} \fSigma_{13} & \fSigma_{23}\end{pmatrix}^T$ is of dimension $2\times d_3$, whereas $\begin{pmatrix} \fSigma_{31} & \fSigma_{32}\end{pmatrix}$ is of dimension $d_3\times 2$ such that
+$$\begin{pmatrix} \fSigma_{13} \\ \fSigma_{23} \end{pmatrix} \fSigma_{33}^{-1} \begin{pmatrix} \fSigma_{31} & \fSigma_{32} \end{pmatrix}$$
 is of dimension $2\times2$.
@@
 \begin{equation}
-\sqrt{nb_n^2}\left(\widehat{\alpha}(z))-\alpha(z)\right)  \stackrel {{\cal L}}{\to} {\cal N}(0,\nabla g(\frho)^{T} {\fSigma} \nabla g (\frho)),
-\end{equation}
-see eq. (22) in the main article, and see Theorem 2 in Otneim and Tjøstheim (2018) for a definition of $\fSigma$. The computation of $\nabla g$ is outlined in Section \ref{app-asvar}.
+\sqrt{nb_n^2}\left(\widehat{\alpha}(z))-\alpha(z)\right)  \stackrel {{\cal L}}{\to} {\cal N}(0,\nabla g(\frho)^{T} {\fLambda} \nabla g (\frho)),
+\end{equation}
+see eq.
+(22) in the main article, and see Theorem 2 in Otneim and Tjøstheim (2018) with $\fSigma = \fLambda$ for a definition of $\fLambda$.
+The computation of $\nabla g$ is outlined in Section \ref{app-asvar}.
 \end{itemize}
@@
 The reader may consult @tjostheim2013local for more details and regularity conditions in this regard.
-The corresponding estimate $\rho_{n, b}(z_i,z_j)$ is obtained by maximizing the log-likelihood in \@ref(eq:loglikelihood).
+The corresponding estimate $\widehat \rho = \rho_{n, b}(z_i,z_j)$ is obtained by maximizing the log-likelihood in \@ref(eq:loglikelihood).
 Note that, writing $\rho$ for $\rho_b$,
@@
 L^{(b)}(Z_{i,t}, Z_{j,t}; \rho) &= K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
-& \qquad\qquad\qquad \int K_b(v-z_i, w- z_j)\rho_b(v,w;\psi_b(z_i, z_j)) \, \di v \di w.
+& \qquad\qquad\qquad - \int K_b(v-z_i, w- z_j)\rho_b(v,w;\psi_b(z_i, z_j)) \, \di v \di w.
 \label{eq:A16}
@@
 Here $\mathcal{F}_t$ is the $\sigma$-algebra generated by $\{Z_{i,s}, Z_{j,s}, s \leq t\}$.
-\item[\textbf{A2}] The parameter value $\rho_{0,b}$ is the unique maximizer of $\E(L_{n,b}(\rho)): \Theta \rightarrow \overline{\mathbb{R}}$, with $\E(L_{n,b}(\rho))$ is defined in \eqref{eq:penalty}.
+\item[\textbf{A2}] The parameter value $\rho_{0,b}$ is the unique maximizer of $\E(L_{n,b}(\rho)): \Theta \rightarrow \overline{\mathbb{R}}$, with $\E(L_{n,b}(\rho))$ defined in \eqref{eq:penalty}.
 \end{itemize}
@@
 
-The proof of Theorem \@ref(thm:A1) is mostly identical to the proof of Theorem 2 of @lacal2018estimating with trivial modifications to take into account the scalar nature of the parameter $\rho_{ij}$ in the present case.
+The proof of Theorem \@ref(thm:A1) is mostly identical to the proof of Theorem 2.1 of @lacal2018estimating with trivial modifications to take into account the scalar nature of the parameter $\rho_{ij}$ in the present case.
 A distributional limit theorem analogous to Theorem 2.2 in @lacal2018estimating can also be rephrased and proved as in that paper, but we refrain from doing this given the distributional results already stated in Section 3 of the main part of the paper.
@@
 
-The second term in $\Jb$ can similarly be written as $II_J = B(\psi(\z, \theta_h(\z)) - f(\z)) + o(b^2)$, where $B = \nabla \fv(\z, \theta_b(\z))$ are second derivatives of the local likelihood function with respect to the local correlations, and $\psi(\z, \theta_b(\z)) - f(\z) = O(b^2)$ according to @hjort1996locally.
+The second term in $\Jb$ can similarly be written as $II_J = B(\psi(\z, \frho_b(\z)) - f(\z)) + o(b^2)$, where $B = \nabla \fv(\z, \frho_b(\z))$ are second derivatives of the local likelihood function with respect to the local correlations, and $\psi(\z, \frho_b(\z)) - f(\z) = O(b^2)$ according to @hjort1996locally.
 Hence, it follows that the leading term of $\Jb^{-1}\Mb(\Jb^{-1})^T$ is $O(b^{-2})$, which again means that the convergence rate in eq.
@@
 Proving joint asymptotic normality of the local Gaussian correlations with marginally normally distributed variables $(Z_1, \ldots, Z_p) = \left(\Phi^{-1}(F_1(X_1)), \ldots, \Phi^{-1}(F_p(X_p))\right)$, where $F_j$ is the cdf of $X_j$ for $i = 1,\ldots,p$, relies on proving asymptotic normality for the variables
-$$Y_n(\z) = \frac{1}{n}\sum_{t=1}^nK_b(\Z_t - \z)\fv(\Z_t, \frho_0),$$
+$$\fY_n(\z) = \frac{1}{n}\sum_{t=1}^nK_b(\Z_t - \z)\fv(\Z_t, \frho_0),$$
 where $\fv(\cdot) = \partial L(\cdot)/\partial\frho$ in the same way as in Section \@ref(app-loccor) in this supplement.
@@
 
-$$\hY_n(\z) = \frac{1}{n}\sum_{t=1}^nK_b(\hZ_t - \z)\fv(\hZ_t,\frho_0),$$
+$$\hfY_n(\z) = \frac{1}{n}\sum_{t=1}^nK_b(\hZ_t - \z)\fv(\hZ_t,\frho_0),$$
 with $\hnZ_t = \Phi^{-1}(F_n(\X_t))$, and $F_n(\X_t)$ is the application of the marginal empirical distribution functions to the corresponding components of the vector $\X_t$.
@@
 In order to ease notation we state the proof in the simplified case only, where the local correlations are estimated using *pairs* $(i,j)$ of variables.
+Then $Y_n$ can be taken to be a scalar.
 It can be extended to the trivariate case by adding the appropriate terms in the derivations below, without changing the final conclusion.
@@
 
-In the following we omit the index $i=1,2$.
-We need a bound on the supremum of $|F_n(X_t) - F(X_t)|$ and $|F_n(x) - F(x)|$ as $n \rightarrow \infty$.
+In the original version of this proof we used an argument based on the law of the iterated logarithm of @rio1995functional, leading to a rate of $n^{-1/2}(2 \log \log n )^{1/2}$, but an anynomous referee provided us with a much simpler argument based on a U-statistic decomposition for a strongly mixing process.
+In the following we omit the index $i=1,2$:
+
+\begin{align*}
+\frac{1}{n}\sum_{i=1}^n \left(F_n(X_{i,t}) - F(X_{i,t})\right) &= 
+\frac{1}{n^2}\sum_{i=1}^n\sum_{j\neq i}^n \left(I(X_{j, t} \leq X_{i,t}) - F(X_{i,t})\right) + \frac{1}{n^2}\sum_{i=1}^n \left(1 - F(X_{i,t}\right) \\
+&= \frac{1}{n} \sum_{i=1}^n \left(0.5 - F(X_{i,t})\right) + O_p\left(n^{-1}\right) \\
+&= O_p\left(n^{-1/2}\right) + O_p\left(n^{-1}\right) \\
+&= O_p\left(n^{-1/2}\right).
+\end{align*}
+
+<!-- We need a bound on the supremum of $|F_n(X_t) - F(X_t)|$ and $|F_n(x) - F(x)|$ as $n \rightarrow \infty$.
 According to @arcones1995law, the best law of the iterated logarithm has been obtained by @rio1995functional.
-One statement of this is as follows:
-
-Let $f(x) \in L_p$, $2<p\leq\infty$, $\sum\alpha_k k^{2/(p-2)} < \infty$, then the sequence $f(X_t)$ satisfies the compact LIL (*law of the iterated logarithms*), i.e.
-$$\left\{\left(n2\log\log n\right)^{-\frac{1}{2}} \sum_{t=1}^n(f(X_t) - E[f(X_t)]) \right\}_{n=1}^{\infty}$$
-is relatively compact and its limit set is $[-\sigma, \sigma]$ where
-$$\sigma^2 = \textrm{Var}\Big(f(X_t)\Big) + 2\sum_{s=1}^{\infty}\textrm{Cov}\Big(f(X_t), f(X_s)\Big).$$
-For a fixed $x$, take $f(X_t) = I(X_t\leq x)$. Then E$(f(X_t)) = F(x)$ and $f(x)\in L_p$ for all $p$, and
-$$\left\{\left(n2\log\log n\right)^{-\frac{1}{2}} \sum_{s=1}^n \left(I(X_s \leq x) - F(x) \right) \right\}$$
-has its limit set in $[-\sigma, \sigma]$. Here 
-$$\textrm{Var}(I(X_t \leq x)) = F(x)(1-F(x)) \leq \frac{1}{4},$$
-and according to Davydov's Lemma (see e.g. @fan2008nonlinear p. 278),
-$$\big| \textrm{Cov}(I(X_t \leq x), I(X_s \leq x))\big| \leq 8 \|I(X_t \leq x)\|_p \|I(X_s \leq x)\|_q \alpha(|s-t)^{1-p^{-1}-q^{-1}}, \,\,\, p,q>1, \,\,\, \frac{1}{p}+\frac{1}{q} < 1,$$
-with $\| I(X_t \leq x)\|_p = \left(\textrm{E}|I(X_t \leq x)|^p\right)^{1/p}$. We may for example take $p=q=4$, leading to $\| I(X_t \leq x) \|_4 = F(x)^{1/4} \leq 1$.
-
-Since we have assumed exponential mixing (much weaker conditions suffice), then Rio's condition $\sum \alpha_k k^{2/(p-2)} < \infty$ is satisfied, and the bounds of $-\sigma$ and $\sigma$ can be made independent of $x$.
-
-Having 
-$$F_n(x) = \frac{1}{n}\sum_{t=1}^n I(X_t \leq x)$$
-means that
-$$n^{\frac{1}{2}}[2\log \log n]^{-\frac{1}{2}} (F_n(x) - F(x))$$
-has its limits in $[-\sigma, \sigma]$ or that the supremum of $|F_n(x) - F(x)|$ is almost surely bounded by $n^{-1/2}(2\log\log n)^{1/2}$.
-
-This immediately implies that
-$$\frac{1}{n}\sum_{t=1}^n\left(F_n(X_t) - F(X_t)\right)$$
-is almost surely (and consequentially in probability) bounded by $n^{-\frac{1}{2}}(2 \log \log n)^{\frac{1}{2}}$ which for all conceivable bandwidths  is faster than the convergence rates of the CLT for $Y_n(\z)$, which is $(nb^2)^{-\frac{1}{2}}$.
+One statement of this is as follows: -->
+
+<!-- Let $f(x) \in L_p$, $2<p\leq\infty$, $\sum\alpha_k k^{2/(p-2)} < \infty$, then the sequence $f(X_t)$ satisfies the compact LIL (*law of the iterated logarithms*), i.e. -->
+<!-- $$\left\{\left(n2\log\log n\right)^{-\frac{1}{2}} \sum_{t=1}^n(f(X_t) - E[f(X_t)]) \right\}_{n=1}^{\infty}$$ -->
+<!-- is relatively compact and its limit set is $[-\sigma, \sigma]$ where -->
+<!-- $$\sigma^2 = \textrm{Var}\Big(f(X_t)\Big) + 2\sum_{s=1}^{\infty}\textrm{Cov}\Big(f(X_t), f(X_s)\Big).$$ -->
+<!-- For a fixed $x$, take $f(X_t) = I(X_t\leq x)$. Then E$(f(X_t)) = F(x)$ and $f(x)\in L_p$ for all $p$, and -->
+<!-- $$\left\{\left(n2\log\log n\right)^{-\frac{1}{2}} \sum_{s=1}^n \left(I(X_s \leq x) - F(x) \right) \right\}$$ -->
+<!-- has its limit set in $[-\sigma, \sigma]$. Here  -->
+<!-- $$\textrm{Var}(I(X_t \leq x)) = F(x)(1-F(x)) \leq \frac{1}{4},$$ -->
+<!-- and according to Davydov's Lemma (see e.g. @fan2008nonlinear p. 278), -->
+<!-- $$\big| \textrm{Cov}(I(X_t \leq x), I(X_s \leq x))\big| \leq 8 \|I(X_t \leq x)\|_p \|I(X_s \leq x)\|_q \alpha(|s-t|)^{1-p^{-1}-q^{-1}}, \,\,\, p,q>1, \,\,\, \frac{1}{p}+\frac{1}{q} < 1,$$ -->
+<!-- with $\| I(X_t \leq x)\|_p = \left(\textrm{E}|I(X_t \leq x)|^p\right)^{1/p}$. We may for example take $p=q=4$, leading to $\| I(X_t \leq x) \|_4 = F(x)^{1/4} \leq 1$. -->
+
+<!-- Since we have assumed exponential mixing (much weaker conditions suffice), then Rio's condition $\sum \alpha_k k^{2/(p-2)} < \infty$ is satisfied, and the bounds of $-\sigma$ and $\sigma$ can be made independent of $x$.
+-->
+
+<!-- Having  -->
+<!-- $$F_n(x) = \frac{1}{n}\sum_{t=1}^n I(X_t \leq x)$$ -->
+<!-- means that -->
+<!-- $$n^{\frac{1}{2}}[2\log \log n]^{-\frac{1}{2}} (F_n(x) - F(x))$$ -->
+<!-- has its limits in $[-\sigma, \sigma]$ or that the supremum of $|F_n(x) - F(x)|$ is almost surely bounded by $n^{-1/2}(2\log\log n)^{1/2}$. -->
+
+<!-- This immediately implies that -->
+<!-- $$\frac{1}{n}\sum_{t=1}^n\left(F_n(X_t) - F(X_t)\right)$$ -->
+<!-- is almost surely (and consequentially in probability) bounded by $n^{-\frac{1}{2}}(2 \log \log n)^{\frac{1}{2}}$ which for all conceivable bandwidths  is faster than the convergence rates of the CLT for $Y_n(\z)$, which is $(nb^2)^{-\frac{1}{2}}$.
+-->
 
@@
 We can use the bootstrap theory by @lacal2018estimating more or less directly, which in turn is based on Gon\c{c}alves and White [-@gonccalves2002bootstrap; -@gonccalves2004maximum], but we spell out some details  below.
-First, we need to specify the analogue of the conditions A1--A5 by @lacal2018estimating, where we adapt their notation to our situation by replacing the parameter $\theta$ with $\rho_{ij}$, and the time series $\{X_t, Y_t\}$ by our multivariate observations $\{\Z_t\}$, possibly a multivariate time series.
+First, we need to specify the analogue of the conditions A1--A5 by @lacal2018estimating, where we adapt their notation to our situation by replacing the parameter $\theta$ with $\rho_{ij}$, and the time series $\{X_t, Y_t\}$ by our multivariate observations $\{\Z_t\}$.
 Since $\rho = \rho_{ij}$ is estimated using the two corresponding data vectors only, we can, for the time being and without loss of generality, assume that the observation series is bivariate: ${\Z_t} = \{Z_{i,t}, Z_{j,t}\}$.
@@
 To state the bootstrap results we need a few more concepts and three additional conditions.
-We block bootstrap the entire vector series $\{\Z_t\}$.
+We block-bootstrap the entire vector series $\{\Z_t\}$.
 This automatically leads to the same block-bootstrap for any pair of components which we have denoted by $\{Z_{i,t}, Z_{j,t}\}$ above.
@@
 This depends on scalar coordinates $(z_1, z_2)$ and a vector coordinate $\z_3$.
-Note that $\alpha_{n,b}(i,j|\Z_3)$ is a continuous function of a vector of local correlations according to one of the alternatives (i)', (ii)--(v) in Section \ref{chap:estimation}.
+Note that $\alpha_{n,b}(i,j|\Z_3)$ is a continuous function of a vector of local correlations according to any of the alternatives (i)', (ii)--(v) in Section \ref{chap:estimation}.
 Then $\alpha_{n,b}(i,j|\Z_3) - \alpha^*_{n,b}(i,j|\Z_3) \buildrel {\mathbb{P}_{\omega}^{*},\mathbb{P}} \over {\longrightarrow} 0$.
@@
 Consider the general case $\Z = (\Z_1, \Z_2, \Z_3)$ where each component is $d_i$-dimensional correspondingly, with $i = 1,2,3$, and $\Z$, naturally, is $d_1+d_2+d_3$-dimensional, and consider also the local partial correlation matrix of $(\Z_1,\Z_2|\Z_3)$.
-In alternative (iii) in Section \@ref(chap:estimation) this matrix is given by equation \@ref(eq:offdiag3) with
+In alternative (iii), where $d_3 = 1$, in Section \@ref(chap:estimation) this matrix is given by equation \@ref(eq:offdiag3) with
 $$
@@
 $$
-L_{ij,n} = \frac{1}{n} \sum_{t=1}^{n} K_b(\Z_{ij,t}-\z_{ij}) \log \psi\left(\Z_{ij}^{k},\rho_{b,ij}(\z_{ij})\right) - \int K_b(\fv_{ij}-\z_{ij}) \psi_{ij}(\fv_{ij},\rho_{b,ij}(\z_{ij})) \di \fv_{ij},
+L_{ij,n} = \frac{1}{n} \sum_{t=1}^{n} K_b(\Z_{ij,t}-\z_{ij}) \log \psi\left(\Z_{ij},\rho_{b,ij}(\z_{ij})\right) - \int K_b(\fv_{ij}-\z_{ij}) \psi_{ij}(\fv_{ij},\rho_{b,ij}(\z_{ij})) \di \fv_{ij},
 $$
@@
 In contradistinction to the 5-parameter case considered by @lacal2018estimating, in our present 1-parameter case $J_{b,ij}$ is not singular as $\bb \to \bm{0}$.
-In fact, it is shown by Otneim and Tjøstheim [-@otneim2017conditional, Theorem 2] that as $\bb \to \bm{0}$, $J_{b,ij} \to u(\z_{ij},\rho_{ij})\psi(\z_{ij},\rho(\z_{ij}))$.
+In fact, it is shown by Otneim and Tjøstheim [-@otneim2017locally, Theorem 2] that as $\bb \to \bm{0}$, $J_{b,ij} \to u(\z_{ij},\rho_{ij})\psi(\z_{ij},\rho(\z_{ij}))$.
 
@@
 \begin{align}
-\rho_{ij,n,b}(\z_{ij}) - \rho_{ij,b}(\z_{ij}) & \sim J_{ij,b}^{-1}(\z_{ij})U_{ij}(\Z_{ij},\rho_{ij,b}(\z_{ij})) \nonumber \\
+\rho_{ij,n,b}(\z_{ij}) - \rho_{ij,b}(\z_{ij}) & \sim J_{ij,b}^{-1}(\z_{ij})U_{ij}(\z_{ij},\rho_{ij,b}(\z_{ij})) \nonumber \\
 & = J_{ij,b}^{-1}(\z_{ij})\Big[\frac{1}{n}\sum_{t=1}^{n} K_b(\Z_{ij,t} - \z_{ij})u_{ij}(\Z_{ij,t},\rho_{ij,b}(\z_{ij})) \nonumber \\
@@
 
-Using the same argument as in @lacal2018estimating separately on the two main terms yields zero means.
+Using the same argument as in @lacal2017local separately on the two main terms yields zero means.
 Moreover, using the definitions of $G_n$ and $A_b(z)$ it can be seen from \@ref(eq:t3) and \@ref(eq:t4) that asymptotically
@@
 This may be somewhat surprising since the convergence rate of $\frho_{n,b}(\z_{ij})-\frho(\z_{ij})$ is $(nb_1b_2)^{-1/2}$, but this is a well-known phenomenon for functionals of type $T_{n,b}$.
-In @lacal2018estimating the convergence rate is changed from $(nb_1^3b_2^3)^{-1/2}$ to $(nb_1^2b_2^2)^{-1/2}$.
+In Lacal and Tjøstheim [-@lacal2018estimating, Remark 3.1] the convergence rate is changed from $(nb_1^3b_2^3)^{-1/2}$ to $(nb_1^2b_2^2)^{-1/2}$.
 These results are consistent with similar results for test functionals for independence in @joe1989estimation.
@@
 \begin{thm}
-Let the assumptions of Theorem \ref{thm:F2} be fulfilled and with a 3-coordinate $\rho$ as defined above. Then,
+Let the assumptions of Theorem \ref{thm:F2} be fulfilled and with a 3-coordinate $\frho$ as defined above. Then,
 as $n \to \infty$,
@@
 \fV_n(z) &= \sqrt{n} (\frho_{n,b}(\z) - \frho_b(\z)) \\
-& \sim \sqrt{n}{\J}_b^{-1}(\z)\left\{\frac{1}{n} \sum_{t=1}^{n}K_b(\Z^t-\z)\fu(\Z^t,\frho_b(\z)) - \int K_b(\fv-\z)\fu(\fv,\frho_b(\z))\psi(\fv,\rho_b(\z)) \di \fv \right\}.
+& \sim \sqrt{n}{\J}_b^{-1}(\z)\left\{\frac{1}{n} \sum_{t=1}^{n}K_b(\Z_t-\z)\fu(\Z_t,\frho_b(\z)) - \int K_b(\fv-\z)\fu(\fv,\frho_b(\z))\psi(\fv,\rho_b(\z)) \di \fv \right\}.
 \end{align*}
@@
 Let $C_H(A)$ be the limit of $C_n(A_b)$ as $n \to \infty$ and $\bb \to \bm{0}$ at the rates specified in the assumptions of Theorem \ref{thm:F2}.
-(This limit exists using the mixing expressions of the proof of Theorem \ref{thm:F3}).
+(This limit exists using the mixing expressions of the proof of Theorem \ref{thm:F2}).
 Then the power of $n^{1/2}(C_{H,n,b}(A_b))^{-1/2}(T_{n,b}-T)$ ($T = 0$ under our $H_0$, but the procedure below can also be carried through with a general ${\falpha}$ under $H_0$) is given by $\mathbb{P}[T_{n,b} \geq z_{\varepsilon}|H[a_n]) \to 1-\Phi(z_{\varepsilon}-\gamma/[C_{H}(A))]^{1/2})$ as $n \to \infty, \bb \to \bm{0}$, where $\Phi(\cdot)$ is the cumulative standard normal distribution function and $z_{\varepsilon}$ is the one-sided critical value of ${\cal N}(0,1)$ at significance level $\varepsilon$.
@@
 $$
-\sqrt{n}[C_{n}(A_b)]^{-\frac{1}{2}}(T_{n,b}^* - T_{n,b}) \stackrel{d_{\mathbb{P}_{\omega}^{*}}}{\to} {\cal N}(0,1).
+\sqrt{n}[C_{n}(A_b)]^{-\frac{1}{2}}(T_{n,b}^* - T_{n,b}) \stackrel{d_{\mathbb{P}_{\omega}^{*}}}{\to} {\cal N}(0,1) \mbox{prob}-\mathbb{P}.
 $$
@@
 Let $(i,j)$ be a fixed pair and $\rho_{ij,n,b}^*(z_i,z_j)$ be the corresponding estimated bootstrap local correlation for the pair $(i,j)$ chosen from the vector $\frho_{n,b}^*$.
-Here the pair of coordinates $(z_i,z_j)$ corresponds to the bivariate $\x$ in Lacal and Tjøstheim [-@lacal2018estimating, proof of Theorem 3.1].
+Here the pair of coordinates $(z_i,z_j)$ corresponds to the bivariate $\x$ in Lacal and Tjøstheim [-@lacal2017local, proof of Theorem 3.1].
 To ease notation and the comparison with Lacal and Tjøstheim [-@lacal2017local; -@lacal2018estimating], we again write $\z_{ij} = (z_i,z_j)$.
@@
 $$
-I_{ij}^*(\z_{ij},\rho_{ij,n}) = \frac{\partial^2}{\partial \rho_{ij,n}^2} L_{ij,n}^{*}(\z_{ij},\rho_{ij,n}) \quad \mbox{and} \quad J_{ij}^*(\z_{ij}) = \E^*\left(-I_{ij}^*(\z_{ij},\rho_{ij,n})\right),
-$$
-where Lacal and Tjøstheim [-@lacal2018estimating, proof of Theorem 4.4] may be consulted for more details in the derivation and a more explicit analog expression for $I_{ij}^*(\z_{ij},\rho_{ij,n})$ for the case considered there.
+I_{ij}^*(\z_{ij},\rho_{ij,n}) = \frac{\partial^2}{\partial \rho_{ij,n}^2} L_{ij,n}^{*}(\z_{ij},\rho_{ij,n}) \quad \mbox{and} \quad J_{ij,n,b}(\z_{ij}) = \E^*\left(-I_{ij}^*(\z_{ij},\rho_{ij,n})\right),
+$$
+where E$^*(\cdot)$ is the mean value under the measure $\mathbb{P}^*_{\omega}$, and where Lacal and Tjøstheim [-@lacal2017local, proof of Theorem 4.4] may be consulted for more details in the derivation and a more explicit analog expression for $I_{ij}^*(\z_{ij},\rho_{ij,n})$ for the case considered there.
 In contradistinction to @lacal2017local, $J_{ij,n,b}$ is not singular as $\bb \to \bm{0}$.
 In fact, it can be shown that, cf.
-Otneim and Tjøstheim [-@otneim2017conditional, Theorem 2], as $\bb \to \bm{0}$, $J_{ij,n,b} \to u(\z_{ij},\rho_{ij,n})\psi(\z_{ij},\rho(\z_{ij,n}))$.
+Otneim and Tjøstheim [-@otneim2017locally, Theorem 2], as $\bb \to \bm{0}$, $J_{ij,n,b}(\z_{ij}) \stackrel{\mathbb{P}}{\to} u(\z_{ij},\rho_{ij,n})\psi(\z_{ij},\rho_{ij,n}(\z_{ij}))$.
 
@@
 0 &= U_{ij}^*(\z_{ij},\rho_{n,b}^*(\z_{ij})) \\
-& \sim U_{ij}^*(\z_{ij},\rho_{n,b}(\z_{ij}))+I_{ij}^*(\z_{ij},\rho_{n,b}(\z_{ij}))(\rho_{n,b}^*(z_{ij})-\rho(z_{ij})) \\
+& \sim U_{ij}^*(\z_{ij},\rho_{n,b}(\z_{ij}))+I_{ij}^*(\z_{ij},\rho_{n,b}(\z_{ij}))(\rho_{n,b}^*(z_{ij})-\rho_{n,b}(z_{ij})) \\
 & \sim U_{ij}^*(\z_{ij},\rho_{n,b}(\z_{ij})) - J_{ij,n,b}(\z_{ij})(\rho_{n,b}^*(\z_{ij})-\rho_{n,b}(\z_{ij}))
@@
 \begin{align}
-\rho_{n,b}^*(\z_{ij}) - \rho_{n,b}(\z_{ij}) &\sim J_{ij,n,b}^{-1}(\z_{ij})u_{ij}(\z_{ij},\rho_{n,b}(\z_{ij})) \nonumber \\
+\rho_{n,b}^*(\z_{ij}) - \rho_{n,b}(\z_{ij}) &\sim J_{ij,n,b}^{-1}(\z_{ij})U^*_{ij}(\z_{ij},\rho_{n,b}(\z_{ij})) \nonumber \\
 & = J_{ij,n,b}^{-1}(\z_{ij})\Big[\frac{1}{n}\sum_{t=1}^{n} K_b(\Z_{ij,t}^{*} - \z_{ij})u_{ij}(\Z_{ij,t}^{*},\rho_{n,b}(\z_{ij})) \nonumber \\
@@
 $$
-\sqrt{n}\left(C_{n}(A_b)\right)^{-\frac{1}{2}}\left((T_{n,b}^* - T_{n,b}\right) \buildrel {d_{\mathbb{P}_{\omega}^*}} \over {\to} {\cal N}(0,1).
+\sqrt{n}\left(C_{n}(A_b)\right)^{-\frac{1}{2}}\left((T_{n,b}^* - T_{n,b}\right) \buildrel {d_{\mathbb{P}_{\omega}^*}} \over {\to} {\cal N}(0,1)\mbox{prob}-\mathbb{P}.
 $$
@@
 
-Further, if $\mathbb{P}^*$ denote the probability conditional on the sample $\{\Z_t\}$, then the level-$\varepsilon$ critical values $\tilde{c}_{\varepsilon}$ are computed as approximate solutions of $\mathbb{P}^{*}(T_n^* > \tilde{c}_{\varepsilon}) = \varepsilon$ and the bootstrap $p$-value is given by $p^* = B^{-1}\sum_{j=1}^{B} 1(T_{nj}^* > T_n)$, where $B$ is the number of bootstrap realization.
+Further, if $\mathbb{P}^*$ denote the probability conditional on the sample $\{\Z_t^*\}$, then the level-$\varepsilon$ critical values $\tilde{c}_{\varepsilon}$ are computed as approximate solutions of $\mathbb{P}^{*}(T_n^* > \tilde{c}_{\varepsilon}) = \varepsilon$ and the bootstrap $p$-value is given by $p^* = B^{-1}\sum_{j=1}^{B} 1(T_{nj}^* > T_n)$, where $B$ is the number of bootstrap realization.
 To integrate this procedure into our context is fairly straightforward.
@@
 For a fixed bandwidth $b$ the population values, following the local Gaussian approach, can be taken as the solution of the integral equation (2) in @tjostheim2013local; see also corresponding equations in e.g.
-Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional].
+Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional] and eq.
+(19) in the main article.
 Note that in the present paper we have largely operated with two alternatives, the general vector case with inference via the pairwise local likelihood, or the the purely scalar case with three coordinates for the local Gaussian correlation.
@@
 The scaling factor $C_n(A_b)$ is of the same type as in the formulation of those results with obvious notational changes.
-Moreover under the present null hypotheses $T = 0$, it may be conjectured that it leads to the asymptotic results that the scaled $T_n^* \buildrel {d} \over {\to} {\cal N}(0,1)$.
+Moreover under the present null hypotheses $T = 0$, it may be conjectured that it leads to the asymptotic results that the scaled $T_n^* \buildrel {d_{P^*_{\omega}}} \over {\to} {\cal N}(0,1)\mbox{prob}-\mathbb{P}$.
 
@@
 
-Let us briefly demonstrate a practical implementation of the LGPC in another situation involving real data. As in the empirical example in the main article, we look at Granger causality,
-
 \begin{figure}
@@
 \end{figure}
+
+Let us briefly demonstrate a practical implementation of the LGPC in another situation involving real data. As in the empirical example in the main article, we look at Granger causality.
 

### 2020-09-02 lgpc_JBES_supp.tex
Implementing comments from Dag and one referee

@@
 \newcommand{\hY}{\widehat{Y}}
+\newcommand{\fY}{\bm{Y}}
+\newcommand{\hfY}{\widehat{\bm{Y}}}
 
@@
 
-All of these parameters can be described and estimated locally in the exact same way as laid out in the main article, resulting from the locally Gaussian joint density
+All of these parameters can be described and estimated locally in the exact same way as laid out in the main article, resulting from the locally Gaussian joint density \(\psi(\z,\fv)\) at the point \(\fv = \z\),
 
@@
 \end{equation}
-and the locally Gaussian \emph{conditional} density
+and the locally Gaussian \emph{conditional} quantity
 
@@
 Let \(\D = \textrm{diag}\left\{\fSigma_{12|3}\right\}\).
-The local Gaussian partial correlation matrix can be expressed as the lower left hand block matrix of \(\D^{-1/2}\fSigma_{12|3}\D^{-1/2}\), and so the LGPC matrix is thus given as the upper right hand block of the matrix
+The local Gaussian partial correlation matrix can be expressed as the lower left hand block matrix of \(\D^{-1/2}\fSigma_{12|3}\D^{-1/2}\), and so the LGPC matrix is thus given as the lower left hand block of the matrix
 \begin{equation}
@@
 The reader may consult Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) for more details and regularity conditions in this regard.
-The corresponding estimate \(\rho_{n, b}(z_i,z_j)\) is obtained by maximizing the log-likelihood in \eqref{eq:loglikelihood}.
+The corresponding estimate \(\widehat \rho = \rho_{n, b}(z_i,z_j)\) is obtained by maximizing the log-likelihood in \eqref{eq:loglikelihood}.
 Note that, writing \(\rho\) for \(\rho_b\),
@@
 
-The proof of Theorem \ref{thm:A1} is mostly identical to the proof of Theorem 2 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) with trivial modifications to take into account the scalar nature of the parameter \(\rho_{ij}\) in the present case.
+The proof of Theorem \ref{thm:A1} is mostly identical to the proof of Theorem 2.1 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) with trivial modifications to take into account the scalar nature of the parameter \(\rho_{ij}\) in the present case.
 A distributional limit theorem analogous to Theorem 2.2 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) can also be rephrased and proved as in that paper, but we refrain from doing this given the distributional results already stated in Section 3 of the main part of the paper.
@@
 
-The second term in \(\Jb\) can similarly be written as \(II_J = B(\psi(\z, \theta_h(\z)) - f(\z)) + o(b^2)\), where \(B = \nabla \fv(\z, \theta_b(\z))\) are second derivatives of the local likelihood function with respect to the local correlations, and \(\psi(\z, \theta_b(\z)) - f(\z) = O(b^2)\) according to Hjort and Jones (\protect\hyperlink{ref-hjort1996locally}{1996}).
+The second term in \(\Jb\) can similarly be written as \(II_J = B(\psi(\z, \frho_b(\z)) - f(\z)) + o(b^2)\), where \(B = \nabla \fv(\z, \frho_b(\z))\) are second derivatives of the local likelihood function with respect to the local correlations, and \(\psi(\z, \frho_b(\z)) - f(\z) = O(b^2)\) according to Hjort and Jones (\protect\hyperlink{ref-hjort1996locally}{1996}).
 Hence, it follows that the leading term of \(\Jb^{-1}\Mb(\Jb^{-1})^T\) is \(O(b^{-2})\), which again means that the convergence rate in eq.
@@
 Proving joint asymptotic normality of the local Gaussian correlations with marginally normally distributed variables \((Z_1, \ldots, Z_p) = \left(\Phi^{-1}(F_1(X_1)), \ldots, \Phi^{-1}(F_p(X_p))\right)\), where \(F_j\) is the cdf of \(X_j\) for \(i = 1,\ldots,p\), relies on proving asymptotic normality for the variables
-\[Y_n(\z) = \frac{1}{n}\sum_{t=1}^nK_b(\Z_t - \z)\fv(\Z_t, \frho_0),\]
+\[\fY_n(\z) = \frac{1}{n}\sum_{t=1}^nK_b(\Z_t - \z)\fv(\Z_t, \frho_0),\]
 where \(\fv(\cdot) = \partial L(\cdot)/\partial\frho\) in the same way as in Section \ref{app-loccor} in this supplement.
@@
 
-\[\hY_n(\z) = \frac{1}{n}\sum_{t=1}^nK_b(\hZ_t - \z)\fv(\hZ_t,\frho_0),\]
+\[\hfY_n(\z) = \frac{1}{n}\sum_{t=1}^nK_b(\hZ_t - \z)\fv(\hZ_t,\frho_0),\]
 with \(\hnZ_t = \Phi^{-1}(F_n(\X_t))\), and \(F_n(\X_t)\) is the application of the marginal empirical distribution functions to the corresponding components of the vector \(\X_t\).
@@
 In order to ease notation we state the proof in the simplified case only, where the local correlations are estimated using \emph{pairs} \((i,j)\) of variables.
+Then \(Y_n\) can be taken to be a scalar.
 It can be extended to the trivariate case by adding the appropriate terms in the derivations below, without changing the final conclusion.
@@
 
-In the following we omit the index \(i=1,2\).
-We need a bound on the supremum of \(|F_n(X_t) - F(X_t)|\) and \(|F_n(x) - F(x)|\) as \(n \rightarrow \infty\).
-According to Arcones (\protect\hyperlink{ref-arcones1995law}{1995}), the best law of the iterated logarithm has been obtained by Rio (\protect\hyperlink{ref-rio1995functional}{1995}).
-One statement of this is as follows:
-
-Let \(f(x) \in L_p\), \(2<p\leq\infty\), \(\sum\alpha_k k^{2/(p-2)} < \infty\), then the sequence \(f(X_t)\) satisfies the compact LIL (\emph{law of the iterated logarithms}), i.e.
-\[\left\{\left(n2\log\log n\right)^{-\frac{1}{2}} \sum_{t=1}^n(f(X_t) - E[f(X_t)]) \right\}_{n=1}^{\infty}\]
-is relatively compact and its limit set is \([-\sigma, \sigma]\) where
-\[\sigma^2 = \textrm{Var}\Big(f(X_t)\Big) + 2\sum_{s=1}^{\infty}\textrm{Cov}\Big(f(X_t), f(X_s)\Big).\]
-For a fixed \(x\), take \(f(X_t) = I(X_t\leq x)\). Then E\((f(X_t)) = F(x)\) and \(f(x)\in L_p\) for all \(p\), and
-\[\left\{\left(n2\log\log n\right)^{-\frac{1}{2}} \sum_{s=1}^n \left(I(X_s \leq x) - F(x) \right) \right\}\]
-has its limit set in \([-\sigma, \sigma]\). Here
-\[\textrm{Var}(I(X_t \leq x)) = F(x)(1-F(x)) \leq \frac{1}{4},\]
-and according to Davydov's Lemma (see e.g.~Fan and Yao (\protect\hyperlink{ref-fan2008nonlinear}{2008}) p.~278),
-\[\big| \textrm{Cov}(I(X_t \leq x), I(X_s \leq x))\big| \leq 8 \|I(X_t \leq x)\|_p \|I(X_s \leq x)\|_q \alpha(|s-t)^{1-p^{-1}-q^{-1}}, \,\,\, p,q>1, \,\,\, \frac{1}{p}+\frac{1}{q} < 1,\]
-with \(\| I(X_t \leq x)\|_p = \left(\textrm{E}|I(X_t \leq x)|^p\right)^{1/p}\). We may for example take \(p=q=4\), leading to \(\| I(X_t \leq x) \|_4 = F(x)^{1/4} \leq 1\).
-
-Since we have assumed exponential mixing (much weaker conditions suffice), then Rio's condition \(\sum \alpha_k k^{2/(p-2)} < \infty\) is satisfied, and the bounds of \(-\sigma\) and \(\sigma\) can be made independent of \(x\).
-
-Having
-\[F_n(x) = \frac{1}{n}\sum_{t=1}^n I(X_t \leq x)\]
-means that
-\[n^{\frac{1}{2}}[2\log \log n]^{-\frac{1}{2}} (F_n(x) - F(x))\]
-has its limits in \([-\sigma, \sigma]\) or that the supremum of \(|F_n(x) - F(x)|\) is almost surely bounded by \(n^{-1/2}(2\log\log n)^{1/2}\).
-
-This immediately implies that
-\[\frac{1}{n}\sum_{t=1}^n\left(F_n(X_t) - F(X_t)\right)\]
-is almost surely (and consequentially in probability) bounded by \(n^{-\frac{1}{2}}(2 \log \log n)^{\frac{1}{2}}\) which for all conceivable bandwidths is faster than the convergence rates of the CLT for \(Y_n(\z)\), which is \((nb^2)^{-\frac{1}{2}}\).
+In the original version of this proof we used an argument based on the law of the iterated logarithm of Rio (\protect\hyperlink{ref-rio1995functional}{1995}), leading to a rate of \(n^{-1/2}(2 \log \log n )^{1/2}\), but an anynomous referee provided us with a much simpler argument based on a U-statistic decomposition for a strongly mixing process.
+In the following we omit the index \(i=1,2\):
+
+\begin{align*}
+\frac{1}{n}\sum_{i=1}^n \left(F_n(X_{i,t}) - F(X_{i,t})\right) &= 
+\frac{1}{n^2}\sum_{i=1}^n\sum_{j\neq i}^n \left(I(X_{j, t} \leq X_{i,t}) - F(X_{i,t})\right) + \frac{1}{n^2}\sum_{i=1}^n \left(1 - F(X_{i,t}\right) \\
+&= \frac{1}{n} \sum_{i=1}^n \left(0.5 - F(X_{i,t})\right) + O_p\left(n^{-1}\right) \\
+&= O_p\left(n^{-1/2}\right) + O_p\left(n^{-1}\right) \\
+&= O_p\left(n^{-1/2}\right).
+\end{align*}
 
@@
 We can use the bootstrap theory by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) more or less directly, which in turn is based on Gon\c{c}alves and White (\protect\hyperlink{ref-gonccalves2002bootstrap}{2002}, \protect\hyperlink{ref-gonccalves2004maximum}{2004}), but we spell out some details below.
-First, we need to specify the analogue of the conditions A1--A5 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}), where we adapt their notation to our situation by replacing the parameter \(\theta\) with \(\rho_{ij}\), and the time series \(\{X_t, Y_t\}\) by our multivariate observations \(\{\Z_t\}\), possibly a multivariate time series.
+First, we need to specify the analogue of the conditions A1--A5 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}), where we adapt their notation to our situation by replacing the parameter \(\theta\) with \(\rho_{ij}\), and the time series \(\{X_t, Y_t\}\) by our multivariate observations \(\{\Z_t\}\).
 Since \(\rho = \rho_{ij}\) is estimated using the two corresponding data vectors only, we can, for the time being and without loss of generality, assume that the observation series is bivariate: \({\Z_t} = \{Z_{i,t}, Z_{j,t}\}\).
@@
 To state the bootstrap results we need a few more concepts and three additional conditions.
-We block bootstrap the entire vector series \(\{\Z_t\}\).
+We block-bootstrap the entire vector series \(\{\Z_t\}\).
 This automatically leads to the same block-bootstrap for any pair of components which we have denoted by \(\{Z_{i,t}, Z_{j,t}\}\) above.
@@
 Consider the general case \(\Z = (\Z_1, \Z_2, \Z_3)\) where each component is \(d_i\)-dimensional correspondingly, with \(i = 1,2,3\), and \(\Z\), naturally, is \(d_1+d_2+d_3\)-dimensional, and consider also the local partial correlation matrix of \((\Z_1,\Z_2|\Z_3)\).
-In alternative (iii) in Section \ref{chap:estimation} this matrix is given by equation \eqref{eq:offdiag3} with
+In alternative (iii), where \(d_3 = 1\), in Section \ref{chap:estimation} this matrix is given by equation \eqref{eq:offdiag3} with
 \[
@@
 \[
-L_{ij,n} = \frac{1}{n} \sum_{t=1}^{n} K_b(\Z_{ij,t}-\z_{ij}) \log \psi\left(\Z_{ij}^{k},\rho_{b,ij}(\z_{ij})\right) - \int K_b(\fv_{ij}-\z_{ij}) \psi_{ij}(\fv_{ij},\rho_{b,ij}(\z_{ij})) \di \fv_{ij},
+L_{ij,n} = \frac{1}{n} \sum_{t=1}^{n} K_b(\Z_{ij,t}-\z_{ij}) \log \psi\left(\Z_{ij},\rho_{b,ij}(\z_{ij})\right) - \int K_b(\fv_{ij}-\z_{ij}) \psi_{ij}(\fv_{ij},\rho_{b,ij}(\z_{ij})) \di \fv_{ij},
 \]
@@
 In contradistinction to the 5-parameter case considered by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}), in our present 1-parameter case \(J_{b,ij}\) is not singular as \(\bb \to \bm{0}\).
-In fact, it is shown by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}, Theorem 2) that as \(\bb \to \bm{0}\), \(J_{b,ij} \to u(\z_{ij},\rho_{ij})\psi(\z_{ij},\rho(\z_{ij}))\).
+In fact, it is shown by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, Theorem 2) that as \(\bb \to \bm{0}\), \(J_{b,ij} \to u(\z_{ij},\rho_{ij})\psi(\z_{ij},\rho(\z_{ij}))\).
 
@@
 
-Using the same argument as in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) separately on the two main terms yields zero means.
+Using the same argument as in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}) separately on the two main terms yields zero means.
 Moreover, using the definitions of \(G_n\) and \(A_b(z)\) it can be seen from \eqref{eq:t3} and \eqref{eq:t4} that asymptotically
@@
 This may be somewhat surprising since the convergence rate of \(\frho_{n,b}(\z_{ij})-\frho(\z_{ij})\) is \((nb_1b_2)^{-1/2}\), but this is a well-known phenomenon for functionals of type \(T_{n,b}\).
-In Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) the convergence rate is changed from \((nb_1^3b_2^3)^{-1/2}\) to \((nb_1^2b_2^2)^{-1/2}\).
+In Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}, Remark 3.1) the convergence rate is changed from \((nb_1^3b_2^3)^{-1/2}\) to \((nb_1^2b_2^2)^{-1/2}\).
 These results are consistent with similar results for test functionals for independence in Joe (\protect\hyperlink{ref-joe1989estimation}{1989}).
@@
 Let \((i,j)\) be a fixed pair and \(\rho_{ij,n,b}^*(z_i,z_j)\) be the corresponding estimated bootstrap local correlation for the pair \((i,j)\) chosen from the vector \(\frho_{n,b}^*\).
-Here the pair of coordinates \((z_i,z_j)\) corresponds to the bivariate \(\x\) in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}, proof of Theorem 3.1).
+Here the pair of coordinates \((z_i,z_j)\) corresponds to the bivariate \(\x\) in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}, proof of Theorem 3.1).
 To ease notation and the comparison with Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}, \protect\hyperlink{ref-lacal2018estimating}{2018}), we again write \(\z_{ij} = (z_i,z_j)\).
@@
 \[
-I_{ij}^*(\z_{ij},\rho_{ij,n}) = \frac{\partial^2}{\partial \rho_{ij,n}^2} L_{ij,n}^{*}(\z_{ij},\rho_{ij,n}) \quad \mbox{and} \quad J_{ij}^*(\z_{ij}) = \E^*\left(-I_{ij}^*(\z_{ij},\rho_{ij,n})\right),
-\]
-where Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}, proof of Theorem 4.4) may be consulted for more details in the derivation and a more explicit analog expression for \(I_{ij}^*(\z_{ij},\rho_{ij,n})\) for the case considered there.
+I_{ij}^*(\z_{ij},\rho_{ij,n}) = \frac{\partial^2}{\partial \rho_{ij,n}^2} L_{ij,n}^{*}(\z_{ij},\rho_{ij,n}) \quad \mbox{and} \quad J_{ij,n,b}(\z_{ij}) = \E^*\left(-I_{ij}^*(\z_{ij},\rho_{ij,n})\right),
+\]
+where E\(^*(\cdot)\) is the mean value under the measure \(\mathbb{P}^*_{\omega}\), and where Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}, proof of Theorem 4.4) may be consulted for more details in the derivation and a more explicit analog expression for \(I_{ij}^*(\z_{ij},\rho_{ij,n})\) for the case considered there.
 In contradistinction to Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}), \(J_{ij,n,b}\) is not singular as \(\bb \to \bm{0}\).
-In fact, it can be shown that, cf.~Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}, Theorem 2), as \(\bb \to \bm{0}\), \(J_{ij,n,b} \to u(\z_{ij},\rho_{ij,n})\psi(\z_{ij},\rho(\z_{ij,n}))\).
+In fact, it can be shown that, cf.~Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, Theorem 2), as \(\bb \to \bm{0}\), \(J_{ij,n,b}(\z_{ij}) \stackrel{\mathbb{P}}{\to} u(\z_{ij},\rho_{ij,n})\psi(\z_{ij},\rho_{ij,n}(\z_{ij}))\).
 
@@
 
-Further, if \(\mathbb{P}^*\) denote the probability conditional on the sample \(\{\Z_t\}\), then the level-\(\varepsilon\) critical values \(\tilde{c}_{\varepsilon}\) are computed as approximate solutions of \(\mathbb{P}^{*}(T_n^* > \tilde{c}_{\varepsilon}) = \varepsilon\) and the bootstrap \(p\)-value is given by \(p^* = B^{-1}\sum_{j=1}^{B} 1(T_{nj}^* > T_n)\), where \(B\) is the number of bootstrap realization.
+Further, if \(\mathbb{P}^*\) denote the probability conditional on the sample \(\{\Z_t^*\}\), then the level-\(\varepsilon\) critical values \(\tilde{c}_{\varepsilon}\) are computed as approximate solutions of \(\mathbb{P}^{*}(T_n^* > \tilde{c}_{\varepsilon}) = \varepsilon\) and the bootstrap \(p\)-value is given by \(p^* = B^{-1}\sum_{j=1}^{B} 1(T_{nj}^* > T_n)\), where \(B\) is the number of bootstrap realization.
 To integrate this procedure into our context is fairly straightforward.
@@
 Once a bootstrap realization \(\{\Z_{1,t}^*,\Z_{2,t}^*,\Z_{3,t}^*\}\) is created, following Su and White (\protect\hyperlink{ref-su2008nonparametric}{2008}), these realizations are now taken as the original observations in the local likelihood estimation procedure.
-For a fixed bandwidth \(b\) the population values, following the local Gaussian approach, can be taken as the solution of the integral equation (2) in Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}); see also corresponding equations in e.g.~Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, \protect\hyperlink{ref-otneim2017conditional}{2018}).
+For a fixed bandwidth \(b\) the population values, following the local Gaussian approach, can be taken as the solution of the integral equation (2) in Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}); see also corresponding equations in e.g.~Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, \protect\hyperlink{ref-otneim2017conditional}{2018}) and eq.
+(19) in the main article.
 Note that in the present paper we have largely operated with two alternatives, the general vector case with inference via the pairwise local likelihood, or the the purely scalar case with three coordinates for the local Gaussian correlation.
@@
 The scaling factor \(C_n(A_b)\) is of the same type as in the formulation of those results with obvious notational changes.
-Moreover under the present null hypotheses \(T = 0\), it may be conjectured that it leads to the asymptotic results that the scaled \(T_n^* \buildrel {d} \over {\to} {\cal N}(0,1)\).
+Moreover under the present null hypotheses \(T = 0\), it may be conjectured that it leads to the asymptotic results that the scaled \(T_n^* \buildrel {d_{P^*_{\omega}}} \over {\to} {\cal N}(0,1)\mbox{prob}-\mathbb{P}\).
 
@@
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-arcones1995law}{}%
-Arcones, Miguel A.
-1995.
-``The Law of the Iterated Logarithm for Empirical Processes Under Absolute Regularity.'' \emph{Journal of Theoretical Probability} 8 (2): 433--51.
-\url{https://doi.org/10.1007/BF02212887}.
-
+\begin{cslreferences}
 \leavevmode\hypertarget{ref-bill:2008}{}%
@@
 Yahoo Finance. 2018. \emph{Accessed June 2018}.
+\end{cslreferences}
 

### 2020-09-09 lgpc_JBES.Rmd
Fixing new round of corrections from Dag

@@
 Helleveien 30, 5045 Bergen, Norway.
-\texttt{hakon.otneim@nhh.no}] and Dag Tjøstheim^[University of Bergen]
+\texttt{hakon.otneim@nhh.no}] and Dag Tjøstheim^[University of Bergen \newline {Acknowledgements:} The authors gratefully acknowledge the input of an associate editor as well as two anonymous referees, whose comments have greatly improved this paper.]
 output: 
@@
 Two examples illustrating the strengths of our approach are given in Section \@ref(chap:example).
-Section \@ref(chap:testing) gives a brief glimpse of the potential in testing.
+Section \@ref(chap:testing) gives a brief account of the potential in testing.
 
@@
 It deals with a number of issues complementing the treatment in the main paper.
-Among them are more examples, extensions of the main concept to more general situations, precise formulations of some theoretical results with proofs, as well as discussions of Pitman efficiency for testing and validity of the use of the bootstrap.
+Among them are more examples, extensions of the main concept to more general situations, precise formulations of some theoretical results with proofs, as well as discussions of local asymptotic power and the Pitman criterion for testing and validity of the use of the bootstrap.
 
@@
 \end{equation}  
-where, as in Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional] in a further simplified representation, we have fixed local means and standard deviations $\mu_j(\z) = 0$ and $\sigma_j^2(\z) = 1$, $j = 1,\ldots,p$, and where $\R(\z) = \{\rho_{jk}(\z)\}$ is the *local correlation matrix*.
+where, as in Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional] in a further simplified approximation, we have fixed local means and standard deviations $\mu_j(\z) = 0$ and $\sigma_j^2(\z) = 1$, $j = 1,\ldots,p$, and where $\R(\z) = \{\rho_{jk}(\z)\}$ is the *local correlation matrix*.
 
 In practice, we do not know $F_{j}$, but we can instead use the empirical distribution function 
-$$F_{j,n}(x) = \frac{1}{n-1}\sum_{i=1}^n 1\left(X_{ji} \leq x\right),$$
+$$\widehat F_{j}(x) = \frac{1}{n-1}\sum_{i=1}^n 1\left(X_{ji} \leq x\right),$$
 where $1(\cdot)$ is the indicator function, $n$ is the number of observations, and $X_{ji}$ is the $i$th observation of $X_j$.
-This results in pseudo standard normal variables $Z_{j,n} = \Phi^{-1}(F_{j,n}(X_j))$.
+This results in pseudo standard normal variables $\widehat Z_{j} = \Phi^{-1}(\widehat F_{j}(X_j))$.
 
-In the following, we will not always distinguish between $Z_j$ and $Z_{j,n}$.
+In the following, we will not always distinguish between $Z_j$ and $\widehat Z_{j}$.
 In fact, we conclude our asymptotic analysis in Section \@ref(chap:asymptotic) by showing that under certain technical conditions, the error made by estimating $\R(\z)$ using the empirically transformed variables $Z_{j,n}$, instead of $Z_j$, is smaller in the limit than the estimation error made when estimating the local correlations themselves.
@@
 We will in this paper refer to $\X$ and its probability density function $f_{\X}$ as being on the \(x\)-scale, and to $\Z$ and its probability density function $f_{\Z}$ as being on the \(z\)-scale.
-For further discussion of the simplified $z$-representation we refer to @otneim2017locally.
+For further discussion of the simplified $z$-approximation we refer to @otneim2017locally.
 
@@
 For a multivariate Gaussian distribution, we have that $\alpha_{\X}(\x) = \alpha_{\Z}(\z) = \alpha$.
-In the remainder of this paper, we write mainly in terms of the $z$-representation using the LGPC $\alpha(\z) = \alpha_{\Z}(\z)$, but when we write the local partial correlation between $X_1$ and $X_2$ given $\X_3 = \x_3$ at the point $(x_1,x_2,\x_3)$, this is simply $\alpha(\z)$ inserted with $z_j = \Phi^{-1}(F_{j}(x_j))$, $j=1,\ldots,p$.
+In the remainder of this paper, we write mainly in terms of the $z$-representation using the LGPC $\alpha(\z) = \alpha_{\Z}(\z)$, but when we write the local partial correlation between $X_1$ and $X_2$ given $\X_3 = \x_3$ at the point $(x_1,x_2,\x_3)$, this is simply $\alpha(\z)$ inserted $z_j = \Phi^{-1}(F_{j}(x_j))$, $j=1,\ldots,p$.
 
@@
 \begin{equation}
-\sqrt{nb^5}\Mb^{-1/2}\Jb\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
+\sqrt{nb^3}\Mb^{-1/2}\Jb\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
 (\#eq:an3)
@@
 \begin{equation}
-\sqrt{nb_n^2}\left(\hfrho(\z) - \frho_0(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \fOmega),
+\sqrt{nb_n^2}\left(\hfrho(\z) - \frho_0(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \fLambda),
 \label{eq:ot2018}
@@
 
-For a proof, we refer to @otneim2017conditional.--> where the asymptotic covariance matrix $\fOmega$ is defined in an analogous way as in Theorem \@ref(thm:loccor3) There are two important differences between these two results, however.
+For a proof, we refer to @otneim2017conditional.--> where the asymptotic covariance matrix $\fLambda$ is defined as $\fSigma$ in Theorem 2 of @otneim2017conditional.
+There are two important differences compared to Theorem \@ref(thm:loccor3), however.
 First of all, we have that the convergence rate in the simplified pairwise case is equal to the usual nonparametric rate of $1/(nb^2)^{1/2}$.
@@
 \end{equation}
-where $m$ is equal to 2 or 5, and $\fLambda$ is either equal to the leading term of $\Jb^{-1}\Mb(\Jb^{-1})^T$, or $\fOmega$ in \@ref(eq:ot2018), depending on whether we use the full trivariate locally Gaussian fit described in Section \@ref(chap:trivariate-full) or the sequentially pairwise simplification described in Section \@ref(chap:bivariate).
+where $m$ is equal to 2 or 5, and $\fLambda$ is either equal to the leading term of $\Jb^{-1}\Mb(\Jb^{-1})^T$, or $\fLambda$ in \@ref(eq:ot2018), depending on whether we use the full trivariate locally Gaussian fit described in Section \@ref(chap:trivariate-full) or the sequentially pairwise simplification described in Section \@ref(chap:bivariate).
 In any case, it is only a matter of basic differentiation to work out an expression for $\nabla g(\cdot)$, and this task has been deferred to Section C of the online supplement in the case where $d_1 = d_2 = 1$.
@@
 We collect daily observations on Oslo Stock Exchange^[Thanks to \emph{Børsprosjektet (The Stock Exchange Project)} at the Norwegian School of Economics for providing the data.] from 1 January 2018 until 31 December 2019.
-In Figure \ref{fig:order1} we see the estimated LGPC along the diagonal $R_{1,t} = R_{2,t-1}$ for three different values $R_{1,t-1} = -0.02, 0$ and $0.02$, along with their block-bootstrapped 95% confidence bands.
+In Figure \ref{fig:order1} we see the estimated LGPC along the diagonal $R_{1,t} = R_{2,t-1}$ for three different values $R_{1,t-1} = -0.02, 0$ and $0.02$, along with their block-bootstrapped 95% confidence intervals.
 The constant $\alpha(\z) = 0$ is indicated as a dotted line.
@@
 Roughly translated to this particular situation this means that if the large-firm portfolio gives a negative return on day $t-1$, then its value on day $t$ depend on the performance small-firm portfolio on day $t-1$ in a nonlinear and non-Gaussian way that is characterized by positive dependence in the first quadrant.
-The evidence displayed in Figure \ref{fig:order1} is quite weak, however, because the confidence bands include zero in most of the sample space.
+The evidence displayed in Figure \ref{fig:order1} is quite weak, however, because the confidence intervals include zero in most of the sample space.
 
@@
 
-We construct a test statistic for testing
-
-$$\textrm{H}_0: X_1 \perp X_2 \,\, | \,\, \X_3$$
-or, equivalently, a test statistic in terms of the marginally Gaussian pseudo observations,
+We construct a test statistic for testing $\textrm{H}_0: X_1 \perp X_2 \,\, | \,\, \X_3,$ or, equivalently, a test statistic in terms of the marginally Gaussian pseudo observations,
 
 \begin{equation}
-\textrm{H}_0: Z_1 \perp Z_2 \,\, | \,\, \Z_3
+\textrm{H}_0: Z_1 \perp Z_2 \,\, | \,\, \Z_3,
 (\#eq:ci-test)
@@
 The rate $n^{-1/2}$ obtained by us must be interpreted with caution, however, because it is based on a pairwise model that is more restrictive.
-We go into more detail regarding asymptotic power and Pitman efficiency in Section G of the online supplement.
+We go into more detail regarding local asymptotic power and the Pitman criterion in Section G of the online supplement.
 
@@
 A next natural step may be structural equations as used in network analysis and Pearl-type causality [@pearl2000causality], and ultimately further exploring the relationship or lack of relationship between that type of causality and Granger causality.
-
 All methods presented in this paper have been implemented in the R programming language [@r], and is available in the package **lg** [@otneim2019lg].
@@
 
-**Acknowledgements.** The authors gratefully acknowledge the input of an associate editor as well as two anonymuos referees, whose comments have greatly improved this paper.
-
 # REFERENCES

### 2020-09-09 lgpc_JBES.tex
Fixing new round of corrections from Dag

@@
 Helleveien 30, 5045 Bergen, Norway.
-\texttt{hakon.otneim@nhh.no}} and Dag Tjøstheim\footnote{University of Bergen}}
+\texttt{hakon.otneim@nhh.no}} and Dag Tjøstheim\footnote{University of Bergen \newline {Acknowledgements:} The authors gratefully acknowledge the input of an associate editor as well as two anonymous referees, whose comments have greatly improved this paper.}}
 \date{}
@@
 Two examples illustrating the strengths of our approach are given in Section \ref{chap:example}.
-Section \ref{chap:testing} gives a brief glimpse of the potential in testing.
+Section \ref{chap:testing} gives a brief account of the potential in testing.
 
@@
 \end{equation}\\
-where, as in Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, \protect\hyperlink{ref-otneim2017conditional}{2018}) in a further simplified representation, we have fixed local means and standard deviations \(\mu_j(\z) = 0\) and \(\sigma_j^2(\z) = 1\), \(j = 1,\ldots,p\), and where \(\R(\z) = \{\rho_{jk}(\z)\}\) is the \emph{local correlation matrix}.
+where, as in Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, \protect\hyperlink{ref-otneim2017conditional}{2018}) in a further simplified approximation, we have fixed local means and standard deviations \(\mu_j(\z) = 0\) and \(\sigma_j^2(\z) = 1\), \(j = 1,\ldots,p\), and where \(\R(\z) = \{\rho_{jk}(\z)\}\) is the \emph{local correlation matrix}.
 
 In practice, we do not know \(F_{j}\), but we can instead use the empirical distribution function
-\[F_{j,n}(x) = \frac{1}{n-1}\sum_{i=1}^n 1\left(X_{ji} \leq x\right),\]
+\[\widehat F_{j}(x) = \frac{1}{n-1}\sum_{i=1}^n 1\left(X_{ji} \leq x\right),\]
 where \(1(\cdot)\) is the indicator function, \(n\) is the number of observations, and \(X_{ji}\) is the \(i\)th observation of \(X_j\).
-This results in pseudo standard normal variables \(Z_{j,n} = \Phi^{-1}(F_{j,n}(X_j))\).
-
-In the following, we will not always distinguish between \(Z_j\) and \(Z_{j,n}\).
+This results in pseudo standard normal variables \(\widehat Z_{j} = \Phi^{-1}(\widehat F_{j}(X_j))\).
+
+In the following, we will not always distinguish between \(Z_j\) and \(\widehat Z_{j}\).
 In fact, we conclude our asymptotic analysis in Section \ref{chap:asymptotic} by showing that under certain technical conditions, the error made by estimating \(\R(\z)\) using the empirically transformed variables \(Z_{j,n}\), instead of \(Z_j\), is smaller in the limit than the estimation error made when estimating the local correlations themselves.
@@
 We will in this paper refer to \(\X\) and its probability density function \(f_{\X}\) as being on the \(x\)-scale, and to \(\Z\) and its probability density function \(f_{\Z}\) as being on the \(z\)-scale.
-For further discussion of the simplified \(z\)-representation we refer to Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}).
+For further discussion of the simplified \(z\)-approximation we refer to Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}).
 
@@
 For a multivariate Gaussian distribution, we have that \(\alpha_{\X}(\x) = \alpha_{\Z}(\z) = \alpha\).
-In the remainder of this paper, we write mainly in terms of the \(z\)-representation using the LGPC \(\alpha(\z) = \alpha_{\Z}(\z)\), but when we write the local partial correlation between \(X_1\) and \(X_2\) given \(\X_3 = \x_3\) at the point \((x_1,x_2,\x_3)\), this is simply \(\alpha(\z)\) inserted with \(z_j = \Phi^{-1}(F_{j}(x_j))\), \(j=1,\ldots,p\).
+In the remainder of this paper, we write mainly in terms of the \(z\)-representation using the LGPC \(\alpha(\z) = \alpha_{\Z}(\z)\), but when we write the local partial correlation between \(X_1\) and \(X_2\) given \(\X_3 = \x_3\) at the point \((x_1,x_2,\x_3)\), this is simply \(\alpha(\z)\) inserted \(z_j = \Phi^{-1}(F_{j}(x_j))\), \(j=1,\ldots,p\).
 
@@
 \begin{equation}
-\sqrt{nb^5}\Mb^{-1/2}\Jb\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
+\sqrt{nb^3}\Mb^{-1/2}\Jb\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
 \label{eq:an3}
@@
 \begin{equation}
-\sqrt{nb_n^2}\left(\hfrho(\z) - \frho_0(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \fOmega),
+\sqrt{nb_n^2}\left(\hfrho(\z) - \frho_0(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \fLambda),
 \label{eq:ot2018}
 \end{equation}
-where the asymptotic covariance matrix \(\fOmega\) is defined in an analogous way as in Theorem \ref{thm:loccor3} There are two important differences between these two results, however.
+where the asymptotic covariance matrix \(\fLambda\) is defined as \(\fSigma\) in Theorem 2 of Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}).
+There are two important differences compared to Theorem \ref{thm:loccor3}, however.
 First of all, we have that the convergence rate in the simplified pairwise case is equal to the usual nonparametric rate of \(1/(nb^2)^{1/2}\).
@@
 \end{equation}
-where \(m\) is equal to 2 or 5, and \(\fLambda\) is either equal to the leading term of \(\Jb^{-1}\Mb(\Jb^{-1})^T\), or \(\fOmega\) in \eqref{eq:ot2018}, depending on whether we use the full trivariate locally Gaussian fit described in Section \ref{chap:trivariate-full} or the sequentially pairwise simplification described in Section \ref{chap:bivariate}.
+where \(m\) is equal to 2 or 5, and \(\fLambda\) is either equal to the leading term of \(\Jb^{-1}\Mb(\Jb^{-1})^T\), or \(\fLambda\) in \eqref{eq:ot2018}, depending on whether we use the full trivariate locally Gaussian fit described in Section \ref{chap:trivariate-full} or the sequentially pairwise simplification described in Section \ref{chap:bivariate}.
 In any case, it is only a matter of basic differentiation to work out an expression for \(\nabla g(\cdot)\), and this task has been deferred to Section C of the online supplement in the case where \(d_1 = d_2 = 1\).
@@
 We collect daily observations on Oslo Stock Exchange\footnote{Thanks to \emph{Børsprosjektet (The Stock Exchange Project)} at the Norwegian School of Economics for providing the data.} from 1 January 2018 until 31 December 2019.
-In Figure \ref{fig:order1} we see the estimated LGPC along the diagonal \(R_{1,t} = R_{2,t-1}\) for three different values \(R_{1,t-1} = -0.02, 0\) and \(0.02\), along with their block-bootstrapped 95\% confidence bands.
+In Figure \ref{fig:order1} we see the estimated LGPC along the diagonal \(R_{1,t} = R_{2,t-1}\) for three different values \(R_{1,t-1} = -0.02, 0\) and \(0.02\), along with their block-bootstrapped 95\% confidence intervals.
 The constant \(\alpha(\z) = 0\) is indicated as a dotted line.
@@
 Roughly translated to this particular situation this means that if the large-firm portfolio gives a negative return on day \(t-1\), then its value on day \(t\) depend on the performance small-firm portfolio on day \(t-1\) in a nonlinear and non-Gaussian way that is characterized by positive dependence in the first quadrant.
-The evidence displayed in Figure \ref{fig:order1} is quite weak, however, because the confidence bands include zero in most of the sample space.
+The evidence displayed in Figure \ref{fig:order1} is quite weak, however, because the confidence intervals include zero in most of the sample space.
 
@@
 The last decade or so has seen the publication of many new tests for conditional independence.
-Su and White have published a series of such tests: Su and White (\protect\hyperlink{ref-su2007consistent}{2007}) is based on detecting differences between estimated characteristic functions (which is also the method used by Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2017})).
+Su and White have published a series of such tests: Su and White (\protect\hyperlink{ref-su2007consistent}{2007}) is based on detecting differences between estimated characteristic functions (which is also the method used by Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2018})).
 Su and White (\protect\hyperlink{ref-su2008nonparametric}{2008}) is based on estimating the Hellinger distance between conditional density estimates, Su and White (\protect\hyperlink{ref-su2012conditional}{2012}) use local polynomial quantile regression to test for conditional independence, and Su and White (\protect\hyperlink{ref-su2014testing}{2014}) present a test based on empirical likelihood.
@@
 
-We construct a test statistic for testing
-
-\[\textrm{H}_0: X_1 \perp X_2 \,\, | \,\, \X_3\]
-or, equivalently, a test statistic in terms of the marginally Gaussian pseudo observations,
-
-\begin{equation}
-\textrm{H}_0: Z_1 \perp Z_2 \,\, | \,\, \Z_3
+We construct a test statistic for testing \(\textrm{H}_0: X_1 \perp X_2 \,\, | \,\, \X_3,\) or, equivalently, a test statistic in terms of the marginally Gaussian pseudo observations,
+
+\begin{equation}
+\textrm{H}_0: Z_1 \perp Z_2 \,\, | \,\, \Z_3,
 \label{eq:ci-test}
@@
 This is measured by the Pitman criterion.
-A discussion of the Pitman criterion for other tests of conditional independence is given on pages 11--12 immediately following Theorem 2 of Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2017}).
+A discussion of the Pitman criterion for other tests of conditional independence is given on pages 11--12 immediately following Theorem 2 of Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2018}).
 The rate \(n^{-1/2}b^{-d_3/4}\) of Wang and Hong compares favorable with other tests.
 The rate \(n^{-1/2}\) obtained by us must be interpreted with caution, however, because it is based on a pairwise model that is more restrictive.
-We go into more detail regarding asymptotic power and Pitman efficiency in Section G of the online supplement.
+We go into more detail regarding local asymptotic power and the Pitman criterion in Section G of the online supplement.
 
@@
 A next natural step may be structural equations as used in network analysis and Pearl-type causality (Pearl \protect\hyperlink{ref-pearl2000causality}{2000}), and ultimately further exploring the relationship or lack of relationship between that type of causality and Granger causality.
-
 All methods presented in this paper have been implemented in the R programming language (R Core Team \protect\hyperlink{ref-r}{2020}), and is available in the package \textbf{lg} (Otneim \protect\hyperlink{ref-otneim2019lg}{2019}).
 In addition, we provide code and data for easy replication of the results presented here in an online appendix.
-
-\textbf{Acknowledgements.} The authors gratefully acknowledge the input of an associate editor as well as two anonymuos referees, whose comments have greatly improved this paper.
 
@@
 \leavevmode\hypertarget{ref-wang2017characteristic}{}%
-Wang, Xia, and Yongmiao Hong. 2017. ``Characteristic Function Based Testing for Conditional Independence: A Nonparametric Regression Approach.'' \emph{Econometric Theory}, 1--35.
+Wang, Xia, and Yongmiao Hong. 2018. ``Characteristic Function Based Testing for Conditional Independence: A Nonparametric Regression Approach.'' \emph{Econometric Theory} 34 (4): 815--49.
 

### 2020-09-09 lgpc_JBES_supp.Rmd
Fixing new round of corrections from Dag

@@
 
-The literature on nonlinear conditional dependence modeling is mostly concerned with the simple case where the vector of conditionally independent variables consists of two scalar random variables, $\Zone = (Z_1, Z_2)$, and this case is therefore the focus of our treatment of the LGPC in the main article.
+The literature on nonlinear conditional dependence modeling is mostly concerned with the simple case where the vector of conditionally dependent variables consists of two scalar random variables, $\Zone = (Z_1, Z_2)$, and this case is therefore the focus of our treatment of the LGPC in the main article.
 This means that we estimate a single local Gaussian partial correlation function $\widehat \alpha(\z)$.
@@
 
-All of these parameters can be described and estimated locally in the exact same way as laid out in the main article, resulting from the locally Gaussian joint density $\psi(\z,\fv)$ at the point $\fv = \z$,
+All of these parameters can be described and estimated locally in the exact same way as laid out in the main article, resulting from the locally approximating Gaussian joint density $\psi(\z,\fv)$ at the point $\fv = \z$,
 
@@
 \item[\textbf{d)}] \textbf{Let} $\bm{d_1, d_2}$ \textbf{and} $\bm{d_3}$ \textbf{be arbitrary}.
-This is the general case, and it seems impossible to obtain explicit formulas for the local Gaussian partial correlations, but the general fact still holds of course: The local partial correlation matrix of $\Z_1$and $\Z_2$ given $\Z_3$ can be identified by the lower left hand block matrix of $\D^{-1/2}\fSigma_{12|3}\D^{-1/2}$, and can be estimated using pairwise local correlations in \eqref{eq:matrixdef}.
+This is the general case, and it seems impossible to obtain simple explicit formulas for the local Gaussian partial correlations, but the general fact still holds of course: The local partial correlation matrix of $\Z_1$and $\Z_2$ given $\Z_3$ can be identified by the lower left hand block matrix of $\D^{-1/2}\fSigma_{12|3}\D^{-1/2}$, and can be estimated using pairwise local correlations in \eqref{eq:matrixdef}.
 \end{itemize}
@@
 \end{equation}
-see eq.
-(22) in the main article, and see Theorem 2 in Otneim and Tjøstheim (2018) with $\fSigma = \fLambda$ for a definition of $\fLambda$.
-The computation of $\nabla g$ is outlined in Section \ref{app-asvar}.
+see eq. (22) in the main article, and see Theorem 2 in Otneim and Tjøstheim (2018) with $\fSigma = \fLambda$ in that theorem. The computation of $\nabla g$ is outlined in Section \ref{app-asvar}.
 \end{itemize}
@@
 \item[\textbf{(v)}] In this scheme, we do not use a pairwise set-up for the local correlation only (the one-parameter case), but for all parameters in a bivariate normal distribution; the correlation $\rho(x_1, x_2)$, the expectations $\mu_1(x_1,x_2), \mu_2(x_1,x_2)$ and the variances $\sigma_1(x_1,x_2),\sigma_2(x_1,x_2)$ (the five-parameter case).
-This setup can also be simplified and fitted to a $p$-variate distribution by estimating $\mu_i = \mu_i(x_1)$ and $\sigma_i = \sigma_i(x_i)$, $i=1,\ldots,p$, marginally.
+This setup can also be simplified and fitted to a $p$-variate distribution by estimating $\mu_i = \mu_i(x_i)$ and $\sigma_i = \sigma_i(x_i)$, $i=1,\ldots,p$, marginally.
 Tjøstheim and Hufthammer (2013) give the asymptotic theory in this case.
@@
 We also state a general consistency result in Theorem \@ref(thm:A1) at the end of this section.
-
-The pairwise log-likelihood in cases (ii) -- (iv) is given by
+Utilizing Theorem 3.2 in the main article we replace the pseudo normals $\hfZ_t$ of equation (12) of the main paper by vectors $\Z_t$ having standard normal marginals.
+
+The pairwise log-likelihood in cases (ii) -- (iv) can then be written as
 
@@
 \end{align}
-where $\Z_{in} = (Z_{i,1}, Z_{i,2}, \ldots, Z_{i,n})$ and $\Z_{jn} = (Z_{j,1}, Z_{j,2}, \ldots, Z_{j,n})$ are the observation vectors for components $i$ and $j$ respectively, $n$ is the number of observations, $z_i, z_j, v, w \in \mathbb{R}$, $\rho_b = \rho_{ij,b}(z_i,z_j)$ is the local Gaussian correlation parameter, $K_b(v-z_i, w-z_j) = (b_1b_2)^{-1}K(b_1^{-1}(v-z_i))K(b_2^{-1}(w-z_j))$ is a product kernel with bandwidth $b=b_n=(b_1,b_2) = (b_{1n}, b_{2n})$, and $K$ is an non-negative kernel function with compact support such that $\int K(v,w) = 1$.
+where $\Z_{in} = (Z_{i,1}, Z_{i,2}, \ldots, Z_{i,n})$ and $\Z_{jn} = (Z_{j,1}, Z_{j,2}, \ldots, Z_{j,n})$ are the $\Z$-vectors for components $i$ and $j$ respectively, $n$ is the number of observations, $z_i, z_j, v, w \in \mathbb{R}$, $\rho_b = \rho_{ij,b}(z_i,z_j)$ is the local Gaussian correlation parameter, $K_b(v-z_i, w-z_j) = (b_1b_2)^{-1}K(b_1^{-1}(v-z_i))K(b_2^{-1}(w-z_j))$ is a product kernel with bandwidth $b=b_n=(b_1,b_2) = (b_{1n}, b_{2n})$, and $K$ is an non-negative kernel function with compact support such that $\int K(v,w) = 1$.
 For a fixed bandwidth $b$, $\rho_{0,b}$ is the true local parameter, that is, the minimizer of the local penalty function
@@
 L^{(b)}(Z_{i,t}, Z_{j,t}; \rho) &= K_b(Z_{i,t} - z_i, Z_{j,t} - z_j)\log\left(\psi(Z_{i,t}, Z_{j,t}; \rho_b(z_i, z_j))\right) \nonumber \\  
-& \qquad\qquad\qquad - \int K_b(v-z_i, w- z_j)\rho_b(v,w;\psi_b(z_i, z_j)) \, \di v \di w.
+& \qquad\qquad\qquad - \int K_b(v-z_i, w- z_j)\psi(v,w;\rho_b(z_i, z_j)) \, \di v \di w.
 \label{eq:A16}
@@
 As $\hh\rightarrow0$, however, we must take extra care when considering the asymptotic covariance matrix.
-@tjostheim2013local, Section 4,  write the Taylor expansions of $\Mb$ and $\Jb$, being functionals of three variables $(x_1, x_2, x_3)$ or $(z_1,z_2,z_3)$, as
+@tjostheim2013local, Section 4,  write the Taylor expansions of $\Mb$ and $\Jb$, now being functionals of three variables $(x_1, x_2, x_3)$ or $(z_1,z_2,z_3)$, as
 $$\Mb = I_{\bm{M}} + II_{\bm{M}} + o(b^2) \,\,\, \textrm{ and } \Jb = I_{\bm{J}} + II_{\bm{J}} + o(b^2),$$
@@
 $$H_{\z,\hh} = K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right)K\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right)v(\Phi^{-1}(\y)),$$
-where  $\y = F(\x)$, $y_i = F(x_i)$, $i=1,2$, and we write $K(x_1,x_2) = K(x_1)K(x_2)$. Writing $k(z) = K'(z)$ and $v_{y_i} = \partial v_i/\partial y_i,$
+where  $\y = F(\x)$, $y_i = F(x_i)$, $i=1,2$, and we write $K(x_1,x_2) = K(x_1)K(x_2)$. Writing $k(z) = K'(z)$ and $v_{i} = \partial v_i/\partial y_i,$
 \begin{align*}
@@
 &= K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right) k\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right) \frac{v(\Phi^{-1}(\y))}{b\phi(\Phi^{-1}(y_2))} \\
-& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad+  K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right) \frac{v_2(\Phi^{-1}(\y))}{\phi(\Phi^{-1}(y_1))}.
+& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad+  K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right) K\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right) \frac{v_2(\Phi^{-1}(\y))}{\phi(\Phi^{-1}(y_2))}.
 \end{align*}
@@
 
-In the original version of this proof we used an argument based on the law of the iterated logarithm of @rio1995functional, leading to a rate of $n^{-1/2}(2 \log \log n )^{1/2}$, but an anynomous referee provided us with a much simpler argument based on a U-statistic decomposition for a strongly mixing process.
-In the following we omit the index $i=1,2$:
-
-\begin{align*}
-\frac{1}{n}\sum_{i=1}^n \left(F_n(X_{i,t}) - F(X_{i,t})\right) &= 
-\frac{1}{n^2}\sum_{i=1}^n\sum_{j\neq i}^n \left(I(X_{j, t} \leq X_{i,t}) - F(X_{i,t})\right) + \frac{1}{n^2}\sum_{i=1}^n \left(1 - F(X_{i,t}\right) \\
-&= \frac{1}{n} \sum_{i=1}^n \left(0.5 - F(X_{i,t})\right) + O_p\left(n^{-1}\right) \\
+In the original version of this proof we used an argument based on the law of the iterated logarithm of @rio1995functional, leading to a rate of $n^{-1/2}(2 \log \log n )^{1/2}$, but an anynomous referee provided us with a much simpler argument based on a U-statistic decomposition for a strongly mixing process:
+
+\begin{align*}
+\frac{1}{n}\sum_{t=1}^n \left(F_n(X_{i,t}) - F(X_{i,t})\right) &= 
+\frac{1}{n^2}\sum_{t=1}^n\sum_{s \neq t}^n \left(I(X_{i,s} \leq X_{i,t}) - F(X_{i,t})\right) + \frac{1}{n^2}\sum_{t=1}^n \left(1 - F(X_{i,t}\right) \\
+&= \frac{1}{n} \sum_{t=1}^n \left(0.5 - F(X_{i,t})\right) + O_p\left(n^{-1}\right) \\
 &= O_p\left(n^{-1/2}\right) + O_p\left(n^{-1}\right) \\
@@
 We have seen that the building blocks of the asymptotic estimation theory of the LGPC between the vectors $\Z_1$ and $\Z_2$ given $\Z_3$ are the pairwise local Gaussian correlations contained in the local correlation matrix $\fSigma$.
-Let us simply use the notation $\rho_{ij} = \rho_{ij}(z_i, z_j)$ for one of these local correlation and $\hrho_{ij}$ for its estimate as outlined in the main article.
+Let us simply use the notation $\rho_{ij} = \rho_{ij}(z_i, z_j)$ for one of these local correlations and $\hrho_{ij}$ for its estimate as outlined in the main article.
 If we can prove that the block bootstrap is valid for such a pairwise local likelihood estimate, then it follows fairly straightforwardly that it is also valid for the appropriate vector of such pairwise estimates used in the asymptotic theory laid out above.
@@
 
-We can use the bootstrap theory by @lacal2018estimating more or less directly, which in turn is based on Gon\c{c}alves and White [-@gonccalves2002bootstrap; -@gonccalves2004maximum], but we spell out some details  below.
+We can use the bootstrap theory by Lacal and Tjøstheim [-@lacal2017local;-@lacal2018estimating] more or less directly, which in turn is based on Gon\c{c}alves and White [-@gonccalves2002bootstrap; -@gonccalves2004maximum], but we spell out some details  below.
 First, we need to specify the analogue of the conditions A1--A5 by @lacal2018estimating, where we adapt their notation to our situation by replacing the parameter $\theta$ with $\rho_{ij}$, and the time series $\{X_t, Y_t\}$ by our multivariate observations $\{\Z_t\}$.
@@
 \begin{cor}
-Let the assumptions of Theorem \ref{thm:E2} hold and let $\fOmega = \fOmega(\frho)$ be the covariance matrix of the $\widehat{\frho}$ vector. Then,
-$$
-\sqrt{nb_1b_2}(\alpha_{n,b}^*(i,j)-\alpha_{n,b}(i,j)) \buildrel {d_{\mathbb{P}_{\omega}^{*}}} \over {\longrightarrow} {\cal N}(0,\nabla g(\rho)^T \fOmega(\frho)) \nabla g(\rho)) \,\, \mbox{prob}-\mathbb{P}.
+Let the assumptions of Theorem \ref{thm:E2} hold and let $\fLambda = \fLambda(\frho)$ be the covariance matrix of the $\widehat{\frho}$ vector. Then,
+$$
+\sqrt{nb_1b_2}(\alpha_{n,b}^*(i,j)-\alpha_{n,b}(i,j)) \buildrel {d_{\mathbb{P}_{\omega}^{*}}} \over {\longrightarrow} {\cal N}(0,\nabla g(\rho)^T \fLambda(\frho)) \nabla g(\rho)) \,\, \mbox{prob}-\mathbb{P}.
 $$ 
@@
 Let $\M_{ij}$ and $\J_{ij}$ be the $M$-matrix and $J$-matrix appearing in the proof of Corollary \ref{cor:E4} for a fixed $\rho_{ij}$.
-Note that in the one-parameter case of $\rho_{ij}$ in Corollary \ref{cor:E4}, to first order the covariance matrix $\fOmega$ is a diagonal matrix having ordered values of $M_{ij}/J_{ij}^2$ along the diagonal.
+Note that in the one-parameter case of $\rho_{ij}$ in Corollary \ref{cor:E4}, to first order the covariance matrix $\fLambda$ is a diagonal matrix having ordered values of $M_{ij}/J_{ij}^2$ along the diagonal.
 This can be seen using the methods of proofs of Theorem 2 of @otneim2017locally and Theorem 2 @otneim2017conditional.
 Note also that the 3-coordinate case of alternative (i) can be stated and proved in the same way, but a different scaling factor is required to avoid a singular covariance matrix as seen in the proof of Theorem 3.1 in the main article.
+The 3-coordinate case is in fact closer to the situation in @lacal2018estimating.
 
@@
 \end{align}
-with $F^{s,t}(\z,\fu)$ being the joint cumulative distribution function of $(\Z_s,\Z_t)$, and where $\{\Z_{t}, t=1,\ldots,n\}$ are the $n$ observations of $\Z$ and $u_{ij}(\Z_{ij}, \rho_{ij,b}(\fv_{ij})) = \frac{\partial}{\partial \rho_{ij,b}}\log\psi(\Z_{ij},\rho_{ij,b}(\fv_{ij}))$.
+with $J_{ij}$ being the bivariate version of $\J_b$ defined in Theorem 3.1, and, with $F^{s,t}(\z,\fu)$ being the joint cumulative distribution function of $(\Z_s,\Z_t)$, and where $\{\Z_{t}, t=1,\ldots,n\}$ are the $n$ observations of $\Z$ and $u_{ij}(\Z_{ij}, \rho_{ij,b}(\fv_{ij})) = \frac{\partial}{\partial \rho_{ij,b}}\log\psi(\Z_{ij},\rho_{ij,b}(\fv_{ij}))$.
 
@@
 Here $K_b$ consists of a triple product of kernels $\prod_{i=1}^{3}\frac{1}{b_i}K(\cdot/b_i)$, $\fu(\z,\frho_b(\z)) = \nabla_{\frho}\log \psi(\z,\frho_b(\z))$ where $\psi$ is the density function of a zero mean trivariate normal distribution having a covariance matrix equal to the correlation matrix and with off diagonal elements equal to the elements in $\frho_b(\z)$.
-This is further developed as in the proof of Theorem \ref{thm:F2}, and the asymptotic normality of $T_{n,b}-T_b$ follows as before from the mixing assumptions and the proof is completed by applying proposition 6.3.9 of @brockwell1991time.
+This is further developed as in the proof of Theorem \ref{thm:F2}, and the asymptotic normality of $T_{n,b}-T_b$ follows as before from the mixing assumptions and the proof is completed by applying Proposition 6.3.9 of @brockwell1991time.
 $\blacksquare$
@@
 
-# ASYMPTOTIC POWER AND PITMAN EFFICIENCY {#chap:pitman}
+# LOCAL ASYMPTOTIC POWER AND THE PITMAN CRITERION {#chap:pitman}
 
@@
 
-We will follow the approach used by @wang2017characteristic, in the proof of their Theorem 2 to derive the results on local asymptotic efficiency.
+We will follow the approach used by @wang2017characteristic, in the proof of their Theorem 2 to derive the results on local asymptotic power.
 We will explicitly use the results in Theorems \@ref(thm:F2) and \@ref(thm:F3) with $h({\falpha}_{n,b}(\z)) = ||{\falpha}_{n,b}(\z)||^2$, so that $T_{n,b} = \int _S||{\falpha}_{n,b}(\z)||^2 \di F_n(\z)$.
@@
 \end{align*}
-Here, using Theorem \@ref(thm:F2), under $H[a_n]$,
+Here, using Theorem \@ref(thm:F2), under $H[a_n]$, and Proposition 6.3.9 of @brockwell1991time,
 $$
@@
 The alternative situation that is covered in Theorem \@ref(thm:F3) can now be stated and proved analogously.
-In this case $b^2\fC_{H,n,b}$ converges towards $\fC^{\prime}(\fA)$, say, where $b$ is the common value of the bandwidths $(b_1,b_2,b_3)$.
+In this case $b^2 C_{H,n,b}$ converges towards $C^{\prime}(\fA)$, say, where $b$ is the common value of the bandwidths $(b_1,b_2,b_3)$.
 
@@
 In fact, it can be shown that, cf.
-Otneim and Tjøstheim [-@otneim2017locally, Theorem 2], as $\bb \to \bm{0}$, $J_{ij,n,b}(\z_{ij}) \stackrel{\mathbb{P}}{\to} u(\z_{ij},\rho_{ij,n})\psi(\z_{ij},\rho_{ij,n}(\z_{ij}))$.
+Otneim and Tjøstheim [-@otneim2017locally, Theorem 2], as $\bb \to \bm{0}$, $J_{ij,n,b}(\z_{ij}) \stackrel{\mathbb{P}}{\to} u(\z_{ij},\rho_{ij})\psi(\z_{ij},\rho_{ij}(\z_{ij}))$.
 

### 2020-09-09 lgpc_JBES_supp.tex
Fixing new round of corrections from Dag

@@
 \newcommand{\hfz}{\widehat{\bm{z}}}
+\newcommand{\hfZ}{\widehat{\bm{Z}}}
 \newcommand{\hZ}{\widehat{\bm{Z}}}
@@
 
-The literature on nonlinear conditional dependence modeling is mostly concerned with the simple case where the vector of conditionally independent variables consists of two scalar random variables, \(\Zone = (Z_1, Z_2)\), and this case is therefore the focus of our treatment of the LGPC in the main article.
+The literature on nonlinear conditional dependence modeling is mostly concerned with the simple case where the vector of conditionally dependent variables consists of two scalar random variables, \(\Zone = (Z_1, Z_2)\), and this case is therefore the focus of our treatment of the LGPC in the main article.
 This means that we estimate a single local Gaussian partial correlation function \(\widehat \alpha(\z)\).
@@
 
-All of these parameters can be described and estimated locally in the exact same way as laid out in the main article, resulting from the locally Gaussian joint density \(\psi(\z,\fv)\) at the point \(\fv = \z\),
+All of these parameters can be described and estimated locally in the exact same way as laid out in the main article, resulting from the locally approximating Gaussian joint density \(\psi(\z,\fv)\) at the point \(\fv = \z\),
 
@@
 We also state a general consistency result in Theorem \ref{thm:A1} at the end of this section.
-
-The pairwise log-likelihood in cases (ii) -- (iv) is given by
+Utilizing Theorem 3.2 in the main article we replace the pseudo normals \(\hfZ_t\) of equation (12) of the main paper by vectors \(\Z_t\) having standard normal marginals.
+
+The pairwise log-likelihood in cases (ii) -- (iv) can then be written as
 
@@
 \end{align}
-where \(\Z_{in} = (Z_{i,1}, Z_{i,2}, \ldots, Z_{i,n})\) and \(\Z_{jn} = (Z_{j,1}, Z_{j,2}, \ldots, Z_{j,n})\) are the observation vectors for components \(i\) and \(j\) respectively, \(n\) is the number of observations, \(z_i, z_j, v, w \in \mathbb{R}\), \(\rho_b = \rho_{ij,b}(z_i,z_j)\) is the local Gaussian correlation parameter, \(K_b(v-z_i, w-z_j) = (b_1b_2)^{-1}K(b_1^{-1}(v-z_i))K(b_2^{-1}(w-z_j))\) is a product kernel with bandwidth \(b=b_n=(b_1,b_2) = (b_{1n}, b_{2n})\), and \(K\) is an non-negative kernel function with compact support such that \(\int K(v,w) = 1\).
+where \(\Z_{in} = (Z_{i,1}, Z_{i,2}, \ldots, Z_{i,n})\) and \(\Z_{jn} = (Z_{j,1}, Z_{j,2}, \ldots, Z_{j,n})\) are the \(\Z\)-vectors for components \(i\) and \(j\) respectively, \(n\) is the number of observations, \(z_i, z_j, v, w \in \mathbb{R}\), \(\rho_b = \rho_{ij,b}(z_i,z_j)\) is the local Gaussian correlation parameter, \(K_b(v-z_i, w-z_j) = (b_1b_2)^{-1}K(b_1^{-1}(v-z_i))K(b_2^{-1}(w-z_j))\) is a product kernel with bandwidth \(b=b_n=(b_1,b_2) = (b_{1n}, b_{2n})\), and \(K\) is an non-negative kernel function with compact support such that \(\int K(v,w) = 1\).
 For a fixed bandwidth \(b\), \(\rho_{0,b}\) is the true local parameter, that is, the minimizer of the local penalty function
@@
 As \(\hh\rightarrow0\), however, we must take extra care when considering the asymptotic covariance matrix.
-Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}), Section 4, write the Taylor expansions of \(\Mb\) and \(\Jb\), being functionals of three variables \((x_1, x_2, x_3)\) or \((z_1,z_2,z_3)\), as
+Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}), Section 4, write the Taylor expansions of \(\Mb\) and \(\Jb\), now being functionals of three variables \((x_1, x_2, x_3)\) or \((z_1,z_2,z_3)\), as
 \[\Mb = I_{\bm{M}} + II_{\bm{M}} + o(b^2) \,\,\, \textrm{ and } \Jb = I_{\bm{J}} + II_{\bm{J}} + o(b^2),\]
@@
 \[H_{\z,\hh} = K\left(\frac{z_1 - \Phi^{-1}(y_1)}{b}\right)K\left(\frac{z_2 - \Phi^{-1}(y_2)}{b}\right)v(\Phi^{-1}(\y)),\]
-where \(\y = F(\x)\), \(y_i = F(x_i)\), \(i=1,2\), and we write \(K(x_1,x_2) = K(x_1)K(x_2)\). Writing \(k(z) = K'(z)\) and \(v_{y_i} = \partial v_i/\partial y_i,\)
+where \(\y = F(\x)\), \(y_i = F(x_i)\), \(i=1,2\), and we write \(K(x_1,x_2) = K(x_1)K(x_2)\). Writing \(k(z) = K'(z)\) and \(v_{i} = \partial v_i/\partial y_i,\)
 \begin{align*}
@@
 
-In the original version of this proof we used an argument based on the law of the iterated logarithm of Rio (\protect\hyperlink{ref-rio1995functional}{1995}), leading to a rate of \(n^{-1/2}(2 \log \log n )^{1/2}\), but an anynomous referee provided us with a much simpler argument based on a U-statistic decomposition for a strongly mixing process.
-In the following we omit the index \(i=1,2\):
-
-\begin{align*}
-\frac{1}{n}\sum_{i=1}^n \left(F_n(X_{i,t}) - F(X_{i,t})\right) &= 
-\frac{1}{n^2}\sum_{i=1}^n\sum_{j\neq i}^n \left(I(X_{j, t} \leq X_{i,t}) - F(X_{i,t})\right) + \frac{1}{n^2}\sum_{i=1}^n \left(1 - F(X_{i,t}\right) \\
-&= \frac{1}{n} \sum_{i=1}^n \left(0.5 - F(X_{i,t})\right) + O_p\left(n^{-1}\right) \\
+In the original version of this proof we used an argument based on the law of the iterated logarithm of Rio (\protect\hyperlink{ref-rio1995functional}{1995}), leading to a rate of \(n^{-1/2}(2 \log \log n )^{1/2}\), but an anynomous referee provided us with a much simpler argument based on a U-statistic decomposition for a strongly mixing process:
+
+\begin{align*}
+\frac{1}{n}\sum_{t=1}^n \left(F_n(X_{i,t}) - F(X_{i,t})\right) &= 
+\frac{1}{n^2}\sum_{t=1}^n\sum_{s \neq t}^n \left(I(X_{i,s} \leq X_{i,t}) - F(X_{i,t})\right) + \frac{1}{n^2}\sum_{t=1}^n \left(1 - F(X_{i,t}\right) \\
+&= \frac{1}{n} \sum_{t=1}^n \left(0.5 - F(X_{i,t})\right) + O_p\left(n^{-1}\right) \\
 &= O_p\left(n^{-1/2}\right) + O_p\left(n^{-1}\right) \\
@@
 We have seen that the building blocks of the asymptotic estimation theory of the LGPC between the vectors \(\Z_1\) and \(\Z_2\) given \(\Z_3\) are the pairwise local Gaussian correlations contained in the local correlation matrix \(\fSigma\).
-Let us simply use the notation \(\rho_{ij} = \rho_{ij}(z_i, z_j)\) for one of these local correlation and \(\hrho_{ij}\) for its estimate as outlined in the main article.
+Let us simply use the notation \(\rho_{ij} = \rho_{ij}(z_i, z_j)\) for one of these local correlations and \(\hrho_{ij}\) for its estimate as outlined in the main article.
 If we can prove that the block bootstrap is valid for such a pairwise local likelihood estimate, then it follows fairly straightforwardly that it is also valid for the appropriate vector of such pairwise estimates used in the asymptotic theory laid out above.
@@
 
-We can use the bootstrap theory by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) more or less directly, which in turn is based on Gon\c{c}alves and White (\protect\hyperlink{ref-gonccalves2002bootstrap}{2002}, \protect\hyperlink{ref-gonccalves2004maximum}{2004}), but we spell out some details below.
+We can use the bootstrap theory by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}, \protect\hyperlink{ref-lacal2018estimating}{2018}) more or less directly, which in turn is based on Gon\c{c}alves and White (\protect\hyperlink{ref-gonccalves2002bootstrap}{2002}, \protect\hyperlink{ref-gonccalves2004maximum}{2004}), but we spell out some details below.
 First, we need to specify the analogue of the conditions A1--A5 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}), where we adapt their notation to our situation by replacing the parameter \(\theta\) with \(\rho_{ij}\), and the time series \(\{X_t, Y_t\}\) by our multivariate observations \(\{\Z_t\}\).
@@
 Let \(\M_{ij}\) and \(\J_{ij}\) be the \(M\)-matrix and \(J\)-matrix appearing in the proof of Corollary \ref{cor:E4} for a fixed \(\rho_{ij}\).
-Note that in the one-parameter case of \(\rho_{ij}\) in Corollary \ref{cor:E4}, to first order the covariance matrix \(\fOmega\) is a diagonal matrix having ordered values of \(M_{ij}/J_{ij}^2\) along the diagonal.
+Note that in the one-parameter case of \(\rho_{ij}\) in Corollary \ref{cor:E4}, to first order the covariance matrix \(\fLambda\) is a diagonal matrix having ordered values of \(M_{ij}/J_{ij}^2\) along the diagonal.
 This can be seen using the methods of proofs of Theorem 2 of Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}) and Theorem 2 Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}).
 Note also that the 3-coordinate case of alternative (i) can be stated and proved in the same way, but a different scaling factor is required to avoid a singular covariance matrix as seen in the proof of Theorem 3.1 in the main article.
+The 3-coordinate case is in fact closer to the situation in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}).
 
@@
 Here \(K_b\) consists of a triple product of kernels \(\prod_{i=1}^{3}\frac{1}{b_i}K(\cdot/b_i)\), \(\fu(\z,\frho_b(\z)) = \nabla_{\frho}\log \psi(\z,\frho_b(\z))\) where \(\psi\) is the density function of a zero mean trivariate normal distribution having a covariance matrix equal to the correlation matrix and with off diagonal elements equal to the elements in \(\frho_b(\z)\).
-This is further developed as in the proof of Theorem \ref{thm:F2}, and the asymptotic normality of \(T_{n,b}-T_b\) follows as before from the mixing assumptions and the proof is completed by applying proposition 6.3.9 of Brockwell and Davis (\protect\hyperlink{ref-brockwell1991time}{2006}).
+This is further developed as in the proof of Theorem \ref{thm:F2}, and the asymptotic normality of \(T_{n,b}-T_b\) follows as before from the mixing assumptions and the proof is completed by applying Proposition 6.3.9 of Brockwell and Davis (\protect\hyperlink{ref-brockwell1991time}{2006}).
 \(\blacksquare\)
@@
 \hypertarget{chap:pitman}{%
-\section{ASYMPTOTIC POWER AND PITMAN EFFICIENCY}\label{chap:pitman}}
+\section{LOCAL ASYMPTOTIC POWER AND THE PITMAN CRITERION}\label{chap:pitman}}
 
@@
 
-We will follow the approach used by Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2017}), in the proof of their Theorem 2 to derive the results on local asymptotic efficiency.
+We will follow the approach used by Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2018}), in the proof of their Theorem 2 to derive the results on local asymptotic power.
 We will explicitly use the results in Theorems \ref{thm:F2} and \ref{thm:F3} with \(h({\falpha}_{n,b}(\z)) = ||{\falpha}_{n,b}(\z)||^2\), so that \(T_{n,b} = \int _S||{\falpha}_{n,b}(\z)||^2 \di F_n(\z)\).
@@
 \end{align*}
-Here, using Theorem \ref{thm:F2}, under \(H[a_n]\),
+Here, using Theorem \ref{thm:F2}, under \(H[a_n]\), and Proposition 6.3.9 of Brockwell and Davis (\protect\hyperlink{ref-brockwell1991time}{2006}),
 \[
@@
 The alternative situation that is covered in Theorem \ref{thm:F3} can now be stated and proved analogously.
-In this case \(b^2\fC_{H,n,b}\) converges towards \(\fC^{\prime}(\fA)\), say, where \(b\) is the common value of the bandwidths \((b_1,b_2,b_3)\).
+In this case \(b^2 C_{H,n,b}\) converges towards \(C^{\prime}(\fA)\), say, where \(b\) is the common value of the bandwidths \((b_1,b_2,b_3)\).
 
@@
 In contradistinction to Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}), \(J_{ij,n,b}\) is not singular as \(\bb \to \bm{0}\).
-In fact, it can be shown that, cf.~Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, Theorem 2), as \(\bb \to \bm{0}\), \(J_{ij,n,b}(\z_{ij}) \stackrel{\mathbb{P}}{\to} u(\z_{ij},\rho_{ij,n})\psi(\z_{ij},\rho_{ij,n}(\z_{ij}))\).
+In fact, it can be shown that, cf.~Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, Theorem 2), as \(\bb \to \bm{0}\), \(J_{ij,n,b}(\z_{ij}) \stackrel{\mathbb{P}}{\to} u(\z_{ij},\rho_{ij})\psi(\z_{ij},\rho_{ij}(\z_{ij}))\).
 
@@
 It was used for this purpose by Paparoditis and Politis (\protect\hyperlink{ref-paparoditis2000local}{2000}) and much more extensively by Su and White (\protect\hyperlink{ref-su2008nonparametric}{2008}).
-It has also been used (without proof) by Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2017}) and by Huang (\protect\hyperlink{ref-huang2010testing}{2010}).
+It has also been used (without proof) by Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2018}) and by Huang (\protect\hyperlink{ref-huang2010testing}{2010}).
 We have used a variation of the local bootstrap in our test of conditional independence with good results in accuracy of level and in power.
@@
 To put the local type bootstrap into our test functional context we need first to state some facts and properties of the local bootstrap as used in the literature.
-Note that our functional is in a sense analogous to that used by Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2017}).
-
-In our notation \(\Z_{3,t}\) is the basic \(d_3\)-dimensional process corresponding to \(X_t\) in Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2017}).
-Under \(H_0\) we assume that \(\Z_{1,t}\) and \(\Z_{2,t}\) of dimensions \(d_1\) and \(d_2\) are conditionally independent given \(\Z_{3,t}\), where \(\Z_{1,t}\) and \(\Z_{2,t}\) correspond to \(Y_t\) and \(Z_t\) in Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2017})
+Note that our functional is in a sense analogous to that used by Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2018}).
+
+In our notation \(\Z_{3,t}\) is the basic \(d_3\)-dimensional process corresponding to \(X_t\) in Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2018}).
+Under \(H_0\) we assume that \(\Z_{1,t}\) and \(\Z_{2,t}\) of dimensions \(d_1\) and \(d_2\) are conditionally independent given \(\Z_{3,t}\), where \(\Z_{1,t}\) and \(\Z_{2,t}\) correspond to \(Y_t\) and \(Z_t\) in Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2018})
 

### 2020-09-09 responses/response_ae.Rmd
First draft of responses to comments

---
title: " "
output: 
  bookdown::pdf_document2:
      dev: tikz
      keep_tex: yes
      number_sections: no
      toc: no
      includes:
          in_header: latex_preamble_supp.tex
classoption: letterpaper
geometry: margin = 1in
bibliography: "../bibliography.bib"
biblio-style: "apalike"
link-citations: yes
indent: yes
fontsize: 12pt
---

\noindent {Here are our responses to the comments from the Associate Editor:}

\vspace{1cm}

**AE’s general comment:**  I have heard from two referees. The views are mixed. Referee 2  is not enthusiastic about the paper and recommends rejection while referee 1 recommends a major revision. I also read the paper carefully and understands both referees concerns

> **Our response:** Thank you for reading our paper so carefully. We are very grateful for your constructive comments which has lead to a much improved paper.

**Referee 2 cover letter:** Definitely the topic is very interesting. However, I do not see a clear contribution of this paper to the literature on estimating dependence and testing independence.

I don't see the added value of this paper when is compared to the nonparametric approaches that use Kernel functions (e.g. Normal Kernel) to locally approximate joint distribution functions.

In addition we have many available tests for conditional independence between a vector and scalar variables (and also between two vectors of random variables) conditional on a third vector random variable. However, the test proposed in the paper is limited to testing conditional independence between two scalars only. In addition, the simulation results are definitely not better than the existing tests.

Also, I believe the asymptotic and bootstrap theories are not complete. I also made other comments.

For the above reasons, I believe that the contributions in this paper is definitely below the level of JBES.

> **Our response:** We have made a response directly to the 6 points in the report of referee 2. We briefly summarize this below, at the same time responding to the cover letter of referee 2: \begin{itemize}
\item[1.] \textbf{The novelty of the paper.} We look as the main contribution of our paper the introduction of the new concept of local partial Gaussian correlation (LGPC). With this concept we have sought to make a nonlinear and local extension of the ordinary (global) partial correlation, whose use in measuring and describing dependence is well known for Gaussian distributions. We claim that it is fundamentally different both in concept and in use from the existing tests of conditional independence. It is hard to obtain a measure of conditional dependence that can be easily interpreted from these test as they would have to be based on distance measures that for example cannot distinguish between negative and positive conditional dependence.
\vspace{.3cm} \newline 
Using the kernel estimates for the purpose of measuring conditional dependence is not easy in the same way as it is not easy to use the joint distribution function for a numeric measure of dependence. See item 1 in our response to referee 2. 
\vspace{.3cm} \newline 
Clearly we did not get our message concerning our main contribution through to this referee. In retrospect we see that this may be our own fault. We did not succeed in adequately presenting the main purpose of our paper. Our main message is not that of proposing just another test of conditional independence,
and we have tried to make this more clear in the present version. Our test did do well in a competition with other test, and it may be possible to have it improved with better focusing and use of more local parameters, but, as perhaps over-stated, the main contribution, we believe, is the introduction and use of the new concept of LGPC in measuring the conditional dependence by considering a nonlinear local version of the partial correlation, a measure that reduces to the ordinary partial correlation for Gaussian variables.
\item [2.] \textbf{Estimation and testing in the full vector case:} 
We have developed a theory for this. More details can be found in the the response to item 3 in the report of referee 2. We have also added a new example where $X_1$ and $X_2$ are not both scalars.
\item [3.] \textbf{Asymptotic theory and bootstrap validity}. 
This has been developed in Sections E, F and H in the Supplement, and details can be found in our response to item 4 in the report of referee 2. Actually, we are happy that referee 2 raised this point. We believe that it has strengthened the paper considerably. We add that we have also examined local power and the Pitman criterion in Section  G of the supplement. See also our response to item 5 of referee 2.
\end{itemize}

**AE further general comment:** Referee 1 enjoys reading the paper and have some comments that can be relatively easily addressed.

Given the fact that both referees think that the paper is interesting and I also feel it is interesting, I lean toward the positive referee and do not object to seeing a revision of the paper. Nevertheless, the authors should keep in mind that they must address the referees' major concerns to their satisfaction in order for the paper to pass the second round of review.

My main comments are as follows, some of which overlap those of the referees.

> **Our response:** We are grateful for the opportunity to revise, and we have responded to all of the comments of the two referees, and we will respond to your comments below.

**Comment 1:** I like the idea to extend the local likelihood framework of Hjort and Jones (1996) to estimate the local correlation and to extend the measure of local Gaussian correlation of Tjøstheim and Hufthammer (2013) to the measure of local Gaussian partial correlation (LGPC).

> **Our response:** Thank you.

**Comment 2:** As both referees mention (see R1's point 1 and R2's point 3), one limitation of the proposed testing procedure is that it only tests the conditional independence between two scalar variables $(X_1,X_2)$ given a random vector $(X_1,\ldots,X_p)$. It is important to generalize the test to allow
both $X_1$ and $X_2$ to be random vectors.

> **Our response:** Thank you for this important input. We now introduce and define the LGPC (matrix) for a general ${\bf X}_1$, ${\bf X}_2$ and ${\bf X}_3$ of dimension $d_1$, $d_2$ and $d_3$ in Section 2.1. The full multivariate framework is analyzed in more detail in Section A: "The multivariate LPGC" in the Supplement. Various special cases are discussed in Section A.2 of that Supplement. There are really no new concepts involved, and large parts of the ensuing estimation and testing theory in the Supplement are carried out in this framework. 
\vspace{.3cm} \newline 
In particular the asymptotic theory for a conditional independence test where both ${\bf X}_1$ and ${\bf X_2}$ (and ${\bf X}_3$) are vectors is given in Section F of the Supplementary material (Theorem F.1 and Theorem F.2). The asymptotic local power is discussed in Section G, and the validity of the bootstrap for the testing statistic are given at this level of generality (Theorems G.1, H.1 and H.2). However, the notation becomes more complex and expressions less explicit, and for these reasons and for reasons of space much of the material in the main article is stated for the case of a scalar $X_1$ and $X_2$. There is a new example of Granger causality in Section 4, though, where the assumption of a scalar $X_1$ and $X_2$ is dropped. The curse of dimensionality can be avoided at the cost of a simplified pairwise model introduced described and used in the main article and in the supplement. Throughout the main paper the reader is referred to appropriate sections of the Supplement for the general vector case.

**Comment 3:** p.5: As Referee 1 points out, the density in (6) is not the real density of ${\Z}$. It is a local Gaussian approximation.

> **Our response:** You are certainly right. This was not well written. We have changed "representation" to "approximation" and also linked this directly to the defining approximative equation (4).

**Comment 4.** p.10: Since you claim that ${\bf b}$ is a diagonal matrix of bandwidths (not squared bandwidths). It seems that you want to define $K_{\bf b} = |{\bf b}|^{-1}k({\bf b}^{-1}{\bf x}$.

> **Our response:** Yes, corrections have been implemented. Thank you

**Comment 5:** pp.12-14:

\begin{itemize}

\item[(a)] p.13: It seems that both ${\bf J}_{\bf b}$ and ${\bf M}_{\bf b}$ are of exact order $O(1)$ (so that the second term in the definition of ${\bf M}_{\bf b}$ is asymptotically negligible). This suggests that the estimator of the local correlations have convergence rate $(nb^5)^{-1/2}$ under certain under smoothing bias conditions. Why is the convergence rate here much slower than the usual $(nb^3)^{-1/2}$-rate for the kernel estimator for a trivariate density? In addition, it is possible to estimate the local conditional correlation or not? Note that for nonparametric tests based on the conditional distribution, typically only the dimension of the conditioning variable matters for the convergence rate.

\item[(b)] p.14, line 2: $1/(nb^3))$ should be $1/\sqrt{nb^3}$, which is the convergence rate of a trivariate density..

\item [(c)] p.14, line 4: $1/(nb^5)$ should be $1/\sqrt{nb^5}$.

\item [(d)] The results are too limited as it is only for the case with $p=3$. A general result for the generic $p$ is needed.

\item [(e)] Theorem 3.1 is a pointwise result. For the inference purpose, one typically needs a uniform result.

\end{itemize}

> **Our response:**
\begin{itemize}
\item[(a)] The reason for the slower convergence can basically be found in the expressions for ${\bf J}_{\bf b}$ and ${\bf M}_{\bf b}$. These expressions contain the matrix ${\bf u}(\cdot) {\bf u}^T(\cdot)$. This matrix has rank 1, and as $b \to 0$ this implies that the matrices ${\bf J}_{\bf b}$ and ${\bf M}_{\bf b}$ are singular and of order $O(b^2)$. It is therefore necessary to do a Taylor expansion to find the correct scaling factor, and this in turn leads to a the stated convergence rate of $(nb^5)^{-1/2}$. (Note that there was a typo in eq. (20). It has now been corrected.) This is explained in the proof of Theorem 3.1 and in more detail in Section 4 of Tj{\o}stheim and Hufthammer (2013)  in  a slightly different situation. Similar results and with  rather different derivations can also be found in Jones (1996, 1998). In the pairwise situation referred to after the statement of Theorem 3.1, the matrices ${\bf J}_{\bf b}$ and ${\bf M}_{\bf b}$ collapse to scalars, the singularity disappears and one is back to the usual rates for the case in nonparametric kernel estimation with the dimension of the conditioning variable determining the convergence rate.
\vspace{.3cm} \newline It is certainly possible to estimate the local conditional correlation (and in the general vector case) using local likelihood for local correlations and then transforming to the LGPC. In the general vector Case this is explained in Sections A.2 and A.3 in the Supplement.
\item[(b)] This has been corrected. Thank you.
\item[(c)] This has been corrected. Thank you.
\item[(d)] Agree and the general vector case has now been implemented. We refer to our response to your comment 2 above.
\item[(e)] We agree that our results are pointwise results, thus resulting in pointwise confidence intervals for example. We have to leave uniform results and confidence bands to future investigations.
\end{itemize}

**Comment 6:** p.16: Is there a data-driven choice of bandwidth? The choice of $c=4$ seems too arbitrary.

> **Our response:** There is a data-driven choice for bandwidth in a density estimation context in Otneim and Tj{\o}stheim (2017). It is based on leave-one-out cross-validation. This is referred to in the beginning of Section 4. It remains to establish such a procedure for the present situation. In the paper we have typically used the asymptotic rates $b=cn^{-1/9}$ for the full trivariate fit and $b=cn^{-1/6}$ for the bivariate simplification. In the testing results (Table 2, $2^*$ and $3^*$ we have used $c=1.0$ and $c=1.4$ to examine sensitivity to bandwidth selection. They gave similar results. Note also that for some of the other competing test several choices of tuning parameters were used. For the plots of the estimated LGPC in Section 4 it is true that we used more smoothing to obtain cleaner plots. Thus $c=4$ for Figure 1, but a smaller bandwidth are used in testing and as mentioned in the description of that example the hypothesis of conditional independence is not rejected for any reasonable significance level.

**Comment 7:** p.24-26: The authors propose a new test for conditional independence based on the LGPC. The test statistics should be formally studied under the null hypothesis of conditional independence and the usual Pitman sequence of local alternatives. For this purpose you need uniform consistent estimate of the local correlations. As referee 2 mentions in his/her fourth comment, you also need to justify the asymptotic validity of the bootstrap.

> **Our response:** The asymptotic properties of the test statistic has been formally studied in Section F of the Supplement. The validity of the bootstrap for the test statistic is established in Section H of the Supplement (for estimation in Section E of the Supplement. The convergence rate of the test statistic is given in Section 5.2 in the main article. It is $n^{-1/2}$ for the pairwise testing case and $(nb^2)^{-1/2}$ for the trivariate testing case. This is briefly compared to rates for other test statistics. In Section G we look at local asymptotic  power and the Pitman criterion. We essentially use the approach of Wang and Hong (2018). (They avoid the uniformity argument presumably by using the weaker condition $D^{\prime}$ in Noether, G.E. (1955).)

**Comment 8:** p.26: For the comparison with existing tests, you need to compare asymptotic power properties too.

> **Our response:** We refer to our response to comment 7, where we have compared convergence rates and indirectly local power

**Comment 9:** The paper contains a lot of typos. Referee 1 pointed out some of them and the authors should correct others too.

> **Our response:** We have corrected the typos kindly pointed out to us by Referee 1. We have read carefully through the revised version, including the Supplement in order to eliminate as many typos as possible.

## References

Jones, M. (1996). The local dependence function. Biometrika, **83**, 899-904.

Jones, M. (1998). Constant local dependence. Journal of Multivariate Analysis, **64**, 148-155.

Noether, G.E. (1955). On a theorem of Pitman. Annals of Mathematical Statistics, **26**, 64-68.

Otneim, H. and Tjøstheim, D. (2017). The locally Gaussian density estimator for multivariate data. Statistics and Computing, **27**, 1595-1616.

Tjæstheim, D. and Hufthammer, K.O. (2013) Local Gaussian correlation: A new measure of dependence. Journal of Econometrics, **172**, 33-48.

Wang, X. and Hong, Y. (2018). Characteristic function based testing for conditional independence: A nonparametric regression approach. Econometric Theory, **34**, 815-849.

### 2020-09-09 responses/response_ae.tex
First draft of responses to comments

--- 2020-09-09 responses/response_ae.Rmd
@@
----
-title: " "
-output: 
-  bookdown::pdf_document2:
-      dev: tikz
-      keep_tex: yes
-      number_sections: no
-      toc: no
-      includes:
-          in_header: latex_preamble_supp.tex
-classoption: letterpaper
-geometry: margin = 1in
-bibliography: "../bibliography.bib"
-biblio-style: "apalike"
-link-citations: yes
-indent: yes
-fontsize: 12pt
----
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+  12pt,
+  letterpaper]{article}
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\hypersetup{
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
+\usepackage[margin = 1in]{geometry}
+\usepackage{longtable,booktabs}
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\usepackage{setspace}
+\usepackage{sectsty}
+\usepackage{graphicx}
+\usepackage{subfig}
+\usepackage{bm}
+\usepackage{amsmath}
+\usepackage{amsfonts}
+\usepackage{amsthm}
+\newtheorem{thm}{Theorem}[section]
+\newtheorem{cor}[thm]{Corollary}
+
+%\doublespacing
+\sectionfont{\centering}
+\renewcommand\thesection{\Alph{section}}
+\numberwithin{equation}{section}
+
+\newcommand{\X}{\bm{X}}
+\newcommand{\Xone}{\bm{X}^{(1)}}
+\newcommand{\Xtwo}{\bm{X}^{(2)}}
+\newcommand{\xtwo}{\bm{x}^{(2)}}
+\newcommand{\x}{\bm{x}}
+\newcommand{\hX}{\widehat{X}}
+\newcommand{\tX}{\tilde{X}}
+
+\newcommand{\Z}{\bm{Z}}
+\newcommand{\z}{\bm{z}}
+\newcommand{\hz}{\widehat{z}}
+\newcommand{\hfz}{\widehat{\bm{z}}}
+\newcommand{\hfZ}{\widehat{\bm{Z}}}
+\newcommand{\hZ}{\widehat{\bm{Z}}}
+\newcommand{\hnZ}{\widehat{Z}}
+\newcommand{\Zone}{\bm{Z}^{(1)}}
+\newcommand{\Ztwo}{\bm{Z}^{(2)}}
+\newcommand{\zone}{\bm{z}^{(1)}}
+\newcommand{\ztwo}{\bm{z}^{(2)}}
+
+\newcommand{\D}{\bm{D}}
+\newcommand{\hD}{\widehat{\bm{D}}}
+
+\newcommand{\y}{\bm{y}}
+\newcommand{\Y}{\bm{Y}}
+\newcommand{\hY}{\widehat{Y}}
+\newcommand{\fY}{\bm{Y}}
+\newcommand{\hfY}{\widehat{\bm{Y}}}
+
+\newcommand{\fv}{\bm{v}}
+\newcommand{\fu}{\bm{u}}
+
+\newcommand{\fC}{\bm{C}}
+\newcommand{\fA}{\bm{A}}
+\newcommand{\fV}{\bm{V}}
+\newcommand{\M}{\bm{M}}
+\newcommand{\J}{\bm{J}}
+
+\newcommand{\R}{\bm{R}}
+\newcommand{\hR}{\widehat{\bm{R}}}
+
+\newcommand{\hF}{\widehat{F}}
+\newcommand{\hf}{\widehat{f}}
+
+\newcommand{\hrho}{\widehat{\rho}}
+\newcommand{\frho}{\bm{\rho}}
+\newcommand{\hfrho}{\widehat{\bm{\rho}}}
+
+\newcommand{\hh}{\bm{b}}
+\newcommand{\bb}{\bm{b}}
+
+\newcommand{\fmu}{\bm{\mu}}
+\newcommand{\hfmu}{\widehat{\bm{\mu}}}
+\newcommand{\hsigma}{\widehat{\sigma}}
+\newcommand{\fSigma}{\bm{\Sigma}}
+\newcommand{\hfSigma}{\widehat{\bm{\Sigma}}}
+\newcommand{\halpha}{\widehat{\alpha}}
+\newcommand{\falpha}{\bm{\alpha}}
+\newcommand{\fOmega}{\bm{\Omega}}
+\newcommand{\fLambda}{\bm{\Lambda}}
+\newcommand{\fepsilon}{\bm{\epsilon}}
+
+\newcommand{\Jb}{\bm{J}_{\hh}}
+\newcommand{\Mb}{\bm{M}_{\hh}}
+
+\newcommand{\Cov}{\textrm{Cov}}
+\newcommand{\E}{\textrm{E}}
+\newcommand{\Var}{\textrm{Var}} 
+\newcommand{\di}{\,\textrm{d}}
+\newcommand{\Corr}{\textrm{Corr}}
+
+\usepackage{color}
+\usepackage{colortbl}
+
+\definecolor{Gray}{gray}{0.9}
+
+\author{}
+\date{\vspace{-2.5em}}
+
+\begin{document}
 
@@
 
-**AE’s general comment:**  I have heard from two referees.
+\textbf{AE's general comment:} I have heard from two referees.
 The views are mixed.
@@
 
-> **Our response:** Thank you for reading our paper so carefully. We are very grateful for your constructive comments which has lead to a much improved paper.
+\begin{quote}
+\textbf{Our response:} Thank you for reading our paper so carefully. We are very grateful for your constructive comments which has lead to a much improved paper.
+\end{quote}
 
-**Referee 2 cover letter:** Definitely the topic is very interesting. However, I do not see a clear contribution of this paper to the literature on estimating dependence and testing independence.
+\textbf{Referee 2 cover letter:} Definitely the topic is very interesting. However, I do not see a clear contribution of this paper to the literature on estimating dependence and testing independence.
 
-I don't see the added value of this paper when is compared to the nonparametric approaches that use Kernel functions (e.g. Normal Kernel) to locally approximate joint distribution functions.
+I don't see the added value of this paper when is compared to the nonparametric approaches that use Kernel functions (e.g.~Normal Kernel) to locally approximate joint distribution functions.
 
@@
 
-> **Our response:** We have made a response directly to the 6 points in the report of referee 2.
-We briefly summarize this below, at the same time responding to the cover letter of referee 2: \begin{itemize}
+\begin{quote}
+\textbf{Our response:} We have made a response directly to the 6 points in the report of referee 2. We briefly summarize this below, at the same time responding to the cover letter of referee 2:
+
+\begin{itemize}
 \item[1.] \textbf{The novelty of the paper.} We look as the main contribution of our paper the introduction of the new concept of local partial Gaussian correlation (LGPC).
@@
 \end{itemize}
+\end{quote}
 
-**AE further general comment:** Referee 1 enjoys reading the paper and have some comments that can be relatively easily addressed.
+\textbf{AE further general comment:} Referee 1 enjoys reading the paper and have some comments that can be relatively easily addressed.
 
@@
 
-> **Our response:** We are grateful for the opportunity to revise, and we have responded to all of the comments of the two referees, and we will respond to your comments below.
+\begin{quote}
+\textbf{Our response:} We are grateful for the opportunity to revise, and we have responded to all of the comments of the two referees, and we will respond to your comments below.
+\end{quote}
 
-**Comment 1:** I like the idea to extend the local likelihood framework of Hjort and Jones (1996) to estimate the local correlation and to extend the measure of local Gaussian correlation of Tjøstheim and Hufthammer (2013) to the measure of local Gaussian partial correlation (LGPC).
+\textbf{Comment 1:} I like the idea to extend the local likelihood framework of Hjort and Jones (1996) to estimate the local correlation and to extend the measure of local Gaussian correlation of Tjøstheim and Hufthammer (2013) to the measure of local Gaussian partial correlation (LGPC).
 
-> **Our response:** Thank you.
+\begin{quote}
+\textbf{Our response:} Thank you.
+\end{quote}
 
-**Comment 2:** As both referees mention (see R1's point 1 and R2's point 3), one limitation of the proposed testing procedure is that it only tests the conditional independence between two scalar variables $(X_1,X_2)$ given a random vector $(X_1,\ldots,X_p)$.
+\textbf{Comment 2:} As both referees mention (see R1's point 1 and R2's point 3), one limitation of the proposed testing procedure is that it only tests the conditional independence between two scalar variables \((X_1,X_2)\) given a random vector \((X_1,\ldots,X_p)\).
 It is important to generalize the test to allow
-both $X_1$ and $X_2$ to be random vectors.
+both \(X_1\) and \(X_2\) to be random vectors.
 
-> **Our response:** Thank you for this important input.
-We now introduce and define the LGPC (matrix) for a general ${\bf X}_1$, ${\bf X}_2$ and ${\bf X}_3$ of dimension $d_1$, $d_2$ and $d_3$ in Section 2.1.
-The full multivariate framework is analyzed in more detail in Section A: "The multivariate LPGC" in the Supplement.
+\begin{quote}
+\textbf{Our response:} Thank you for this important input.
+We now introduce and define the LGPC (matrix) for a general \({\bf X}_1\), \({\bf X}_2\) and \({\bf X}_3\) of dimension \(d_1\), \(d_2\) and \(d_3\) in Section 2.1.
+The full multivariate framework is analyzed in more detail in Section A: ``The multivariate LPGC'' in the Supplement.
 Various special cases are discussed in Section A.2 of that Supplement.
@@
 \vspace{.3cm} \newline 
-In particular the asymptotic theory for a conditional independence test where both ${\bf X}_1$ and ${\bf X_2}$ (and ${\bf X}_3$) are vectors is given in Section F of the Supplementary material (Theorem F.1 and Theorem F.2).
+In particular the asymptotic theory for a conditional independence test where both \({\bf X}_1\) and \({\bf X_2}\) (and \({\bf X}_3\)) are vectors is given in Section F of the Supplementary material (Theorem F.1 and Theorem F.2).
 The asymptotic local power is discussed in Section G, and the validity of the bootstrap for the testing statistic are given at this level of generality (Theorems G.1, H.1 and H.2).
-However, the notation becomes more complex and expressions less explicit, and for these reasons and for reasons of space much of the material in the main article is stated for the case of a scalar $X_1$ and $X_2$.
-There is a new example of Granger causality in Section 4, though, where the assumption of a scalar $X_1$ and $X_2$ is dropped.
+However, the notation becomes more complex and expressions less explicit, and for these reasons and for reasons of space much of the material in the main article is stated for the case of a scalar \(X_1\) and \(X_2\).
+There is a new example of Granger causality in Section 4, though, where the assumption of a scalar \(X_1\) and \(X_2\) is dropped.
 The curse of dimensionality can be avoided at the cost of a simplified pairwise model introduced described and used in the main article and in the supplement.
 Throughout the main paper the reader is referred to appropriate sections of the Supplement for the general vector case.
+\end{quote}
 
-**Comment 3:** p.5: As Referee 1 points out, the density in (6) is not the real density of ${\Z}$. It is a local Gaussian approximation.
+\textbf{Comment 3:} p.5: As Referee 1 points out, the density in (6) is not the real density of \({\Z}\). It is a local Gaussian approximation.
 
-> **Our response:** You are certainly right. This was not well written. We have changed "representation" to "approximation" and also linked this directly to the defining approximative equation (4).
+\begin{quote}
+\textbf{Our response:} You are certainly right.
+This was not well written.
+We have changed ``representation'' to ``approximation'' and also linked this directly to the defining approximative equation (4).
+\end{quote}
 
-**Comment 4.** p.10: Since you claim that ${\bf b}$ is a diagonal matrix of bandwidths (not squared bandwidths). It seems that you want to define $K_{\bf b} = |{\bf b}|^{-1}k({\bf b}^{-1}{\bf x}$.
+\textbf{Comment 4.} p.10: Since you claim that \({\bf b}\) is a diagonal matrix of bandwidths (not squared bandwidths).
+It seems that you want to define \(K_{\bf b} = |{\bf b}|^{-1}k({\bf b}^{-1}{\bf x}\).
 
-> **Our response:** Yes, corrections have been implemented. Thank you
+\begin{quote}
+\textbf{Our response:} Yes, corrections have been implemented. Thank you
+\end{quote}
 
-**Comment 5:** pp.12-14:
+\textbf{Comment 5:} pp.12-14:
 
@@
 
-> **Our response:**
+\begin{quote}
+\textbf{Our response:}
+
 \begin{itemize}
@@
 \end{itemize}
+\end{quote}
 
-**Comment 6:** p.16: Is there a data-driven choice of bandwidth? The choice of $c=4$ seems too arbitrary.
+\textbf{Comment 6:} p.16: Is there a data-driven choice of bandwidth? The choice of \(c=4\) seems too arbitrary.
 
-> **Our response:** There is a data-driven choice for bandwidth in a density estimation context in Otneim and Tj{\o}stheim (2017).
+\begin{quote}
+\textbf{Our response:} There is a data-driven choice for bandwidth in a density estimation context in Otneim and Tj\{\o\}stheim (2017).
 It is based on leave-one-out cross-validation.
@@
 It remains to establish such a procedure for the present situation.
-In the paper we have typically used the asymptotic rates $b=cn^{-1/9}$ for the full trivariate fit and $b=cn^{-1/6}$ for the bivariate simplification.
-In the testing results (Table 2, $2^*$ and $3^*$ we have used $c=1.0$ and $c=1.4$ to examine sensitivity to bandwidth selection.
+In the paper we have typically used the asymptotic rates \(b=cn^{-1/9}\) for the full trivariate fit and \(b=cn^{-1/6}\) for the bivariate simplification.
+In the testing results (Table 2, \(2^*\) and \(3^*\) we have used \(c=1.0\) and \(c=1.4\) to examine sensitivity to bandwidth selection.
 They gave similar results.
@@
 For the plots of the estimated LGPC in Section 4 it is true that we used more smoothing to obtain cleaner plots.
-Thus $c=4$ for Figure 1, but a smaller bandwidth are used in testing and as mentioned in the description of that example the hypothesis of conditional independence is not rejected for any reasonable significance level.
+Thus \(c=4\) for Figure 1, but a smaller bandwidth are used in testing and as mentioned in the description of that example the hypothesis of conditional independence is not rejected for any reasonable significance level.
+\end{quote}
 
-**Comment 7:** p.24-26: The authors propose a new test for conditional independence based on the LGPC.
+\textbf{Comment 7:} p.24-26: The authors propose a new test for conditional independence based on the LGPC.
 The test statistics should be formally studied under the null hypothesis of conditional independence and the usual Pitman sequence of local alternatives.
@@
 
-> **Our response:** The asymptotic properties of the test statistic has been formally studied in Section F of the Supplement.
+\begin{quote}
+\textbf{Our response:} The asymptotic properties of the test statistic has been formally studied in Section F of the Supplement.
 The validity of the bootstrap for the test statistic is established in Section H of the Supplement (for estimation in Section E of the Supplement.
 The convergence rate of the test statistic is given in Section 5.2 in the main article.
-It is $n^{-1/2}$ for the pairwise testing case and $(nb^2)^{-1/2}$ for the trivariate testing case.
+It is \(n^{-1/2}\) for the pairwise testing case and \((nb^2)^{-1/2}\) for the trivariate testing case.
 This is briefly compared to rates for other test statistics.
@@
 We essentially use the approach of Wang and Hong (2018).
-(They avoid the uniformity argument presumably by using the weaker condition $D^{\prime}$ in Noether, G.E.
+(They avoid the uniformity argument presumably by using the weaker condition \(D^{\prime}\) in Noether, G.E.
 (1955).)
+\end{quote}
 
-**Comment 8:** p.26: For the comparison with existing tests, you need to compare asymptotic power properties too.
+\textbf{Comment 8:} p.26: For the comparison with existing tests, you need to compare asymptotic power properties too.
 
-> **Our response:** We refer to our response to comment 7, where we have compared convergence rates and indirectly local power
+\begin{quote}
+\textbf{Our response:} We refer to our response to comment 7, where we have compared convergence rates and indirectly local power
+\end{quote}
 
-**Comment 9:** The paper contains a lot of typos. Referee 1 pointed out some of them and the authors should correct others too.
+\textbf{Comment 9:} The paper contains a lot of typos. Referee 1 pointed out some of them and the authors should correct others too.
 
-> **Our response:** We have corrected the typos kindly pointed out to us by Referee 1.
+\begin{quote}
+\textbf{Our response:} We have corrected the typos kindly pointed out to us by Referee 1.
 We have read carefully through the revised version, including the Supplement in order to eliminate as many typos as possible.
+\end{quote}
 
-## References
+\hypertarget{references}{%
+\subsection{References}\label{references}}
 
-Jones, M. (1996). The local dependence function. Biometrika, **83**, 899-904.
+Jones, M. (1996). The local dependence function. Biometrika, \textbf{83}, 899-904.
 
-Jones, M. (1998). Constant local dependence. Journal of Multivariate Analysis, **64**, 148-155.
+Jones, M. (1998). Constant local dependence. Journal of Multivariate Analysis, \textbf{64}, 148-155.
 
-Noether, G.E. (1955). On a theorem of Pitman. Annals of Mathematical Statistics, **26**, 64-68.
+Noether, G.E. (1955). On a theorem of Pitman. Annals of Mathematical Statistics, \textbf{26}, 64-68.
 
-Otneim, H. and Tjøstheim, D. (2017). The locally Gaussian density estimator for multivariate data. Statistics and Computing, **27**, 1595-1616.
+Otneim, H. and Tjøstheim, D. (2017). The locally Gaussian density estimator for multivariate data. Statistics and Computing, \textbf{27}, 1595-1616.
 
-Tjæstheim, D. and Hufthammer, K.O. (2013) Local Gaussian correlation: A new measure of dependence. Journal of Econometrics, **172**, 33-48.
+Tjæstheim, D. and Hufthammer, K.O. (2013) Local Gaussian correlation: A new measure of dependence. Journal of Econometrics, \textbf{172}, 33-48.
 
-Wang, X. and Hong, Y. (2018). Characteristic function based testing for conditional independence: A nonparametric regression approach. Econometric Theory, **34**, 815-849.
+Wang, X. and Hong, Y. (2018). Characteristic function based testing for conditional independence: A nonparametric regression approach. Econometric Theory, \textbf{34}, 815-849.
+
+\end{document}

### 2020-09-09 responses/response_editor.Rmd
First draft of responses to comments

---
title: " "
output: 
  bookdown::pdf_document2:
      dev: tikz
      keep_tex: yes
      number_sections: yes
      toc: no
      includes:
          in_header: latex_preamble_supp.tex
classoption: letterpaper
geometry: margin = 1in
bibliography: "bibliography.bib"
biblio-style: "apalike"
link-citations: yes
indent: yes
fontsize: 12pt
---

\noindent \textbf{Letter to the editor.}

\vspace{1cm}

\noindent Dear Professor Fan.

\vspace{1cm}

We refer to your letter of 12 February 2020 where we are invited to revise our manuscript JBES-P-2019-0538, "The locally Gaussian partial correlation".

We grateful for the opportunity to revise. Enclosed please find our revised version, including supplementary material to be posted online, as well as individual responses to the Associate and two referees.

We apologize for the somewhat long time taken to complete the revision. This is due to the implementations of a rather large number of changes prompted by the detailed and constructive comments of reviewers. We thank them for their positive attitude to our paper and for reading the paper so carefully.

We believe that we have taken all of the queries of the referees and the Associate Editor into account, and in the accompanying files we have made point by point responses. As a result of the input received we think that the paper is much improved. Here are the major changes and improvements:

The local Gaussian partial correlation (LPGC) has been introduced and analyzed in a general vector case looking at the local joint conditional distribution of two random vectors $\X_1$ and $\X_2$ of dimension $d_1$ and $d_2$, respectively, given a third vector $\X_3$ of dimension $d_3$. An example concerning Granger causality testing where $\X_1$ and $\X_2$ are not scalars has been added. We have added asymptotic theory for the test statistic, theory on the validity of the bootstrap for the test statistic and for the estimation of LGPC. We have also derived local power and compared with other test statistics.

For reasons of space most of these changes have been included in a Supplement intended to be available online. This has made it possible for us to keep the length of the main article at 35 pages, as advised by the Editor.

In addition there are numerous other changes directly related to the comments of the referees and the AE. In particular,  we have tried to make it more clear that our paper is not just concerned with presenting just another test of conditional independence. But rather it is concerned with the introduction of a new framework for measuring conditional dependence, and that this framework reduces to the ordinary partial correlation coefficient in the Gaussian case.

We hope that you find the present version acceptable for JBES.

\vspace{.5cm}

Sincerely,

\vspace{.5cm}

Håkon Otneim and Dag Tjøstheim

### 2020-09-09 responses/response_editor.tex
First draft of responses to comments

--- 2020-09-09 responses/response_ae.tex
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\setcounter{secnumdepth}{5}
 \usepackage{setspace}
@@
 
-\noindent {Here are our responses to the comments from the Associate Editor:}
+\noindent \textbf{Letter to the editor.}
 
@@
 
-\textbf{AE's general comment:} I have heard from two referees.
-The views are mixed.
-Referee 2 is not enthusiastic about the paper and recommends rejection while referee 1 recommends a major revision.
-I also read the paper carefully and understands both referees concerns
+\noindent Dear Professor Fan.
 
-\begin{quote}
-\textbf{Our response:} Thank you for reading our paper so carefully. We are very grateful for your constructive comments which has lead to a much improved paper.
-\end{quote}
+\vspace{1cm}
 
-\textbf{Referee 2 cover letter:} Definitely the topic is very interesting. However, I do not see a clear contribution of this paper to the literature on estimating dependence and testing independence.
+We refer to your letter of 12 February 2020 where we are invited to revise our manuscript JBES-P-2019-0538, ``The locally Gaussian partial correlation''.
 
-I don't see the added value of this paper when is compared to the nonparametric approaches that use Kernel functions (e.g.~Normal Kernel) to locally approximate joint distribution functions.
+We grateful for the opportunity to revise.
+Enclosed please find our revised version, including supplementary material to be posted online, as well as individual responses to the Associate and two referees.
 
-In addition we have many available tests for conditional independence between a vector and scalar variables (and also between two vectors of random variables) conditional on a third vector random variable.
-However, the test proposed in the paper is limited to testing conditional independence between two scalars only.
-In addition, the simulation results are definitely not better than the existing tests.
+We apologize for the somewhat long time taken to complete the revision.
+This is due to the implementations of a rather large number of changes prompted by the detailed and constructive comments of reviewers.
+We thank them for their positive attitude to our paper and for reading the paper so carefully.
 
-Also, I believe the asymptotic and bootstrap theories are not complete. I also made other comments.
+We believe that we have taken all of the queries of the referees and the Associate Editor into account, and in the accompanying files we have made point by point responses.
+As a result of the input received we think that the paper is much improved.
+Here are the major changes and improvements:
 
-For the above reasons, I believe that the contributions in this paper is definitely below the level of JBES.
+The local Gaussian partial correlation (LPGC) has been introduced and analyzed in a general vector case looking at the local joint conditional distribution of two random vectors \(\X_1\) and \(\X_2\) of dimension \(d_1\) and \(d_2\), respectively, given a third vector \(\X_3\) of dimension \(d_3\).
+An example concerning Granger causality testing where \(\X_1\) and \(\X_2\) are not scalars has been added.
+We have added asymptotic theory for the test statistic, theory on the validity of the bootstrap for the test statistic and for the estimation of LGPC.
+We have also derived local power and compared with other test statistics.
 
-\begin{quote}
-\textbf{Our response:} We have made a response directly to the 6 points in the report of referee 2. We briefly summarize this below, at the same time responding to the cover letter of referee 2:
+For reasons of space most of these changes have been included in a Supplement intended to be available online.
+This has made it possible for us to keep the length of the main article at 35 pages, as advised by the Editor.
 
-\begin{itemize}
-\item[1.] \textbf{The novelty of the paper.} We look as the main contribution of our paper the introduction of the new concept of local partial Gaussian correlation (LGPC).
-With this concept we have sought to make a nonlinear and local extension of the ordinary (global) partial correlation, whose use in measuring and describing dependence is well known for Gaussian distributions.
-We claim that it is fundamentally different both in concept and in use from the existing tests of conditional independence.
-It is hard to obtain a measure of conditional dependence that can be easily interpreted from these test as they would have to be based on distance measures that for example cannot distinguish between negative and positive conditional dependence.
-\vspace{.3cm} \newline 
-Using the kernel estimates for the purpose of measuring conditional dependence is not easy in the same way as it is not easy to use the joint distribution function for a numeric measure of dependence.
-See item 1 in our response to referee 2.
-\vspace{.3cm} \newline 
-Clearly we did not get our message concerning our main contribution through to this referee.
-In retrospect we see that this may be our own fault.
-We did not succeed in adequately presenting the main purpose of our paper.
-Our main message is not that of proposing just another test of conditional independence,
-and we have tried to make this more clear in the present version.
-Our test did do well in a competition with other test, and it may be possible to have it improved with better focusing and use of more local parameters, but, as perhaps over-stated, the main contribution, we believe, is the introduction and use of the new concept of LGPC in measuring the conditional dependence by considering a nonlinear local version of the partial correlation, a measure that reduces to the ordinary partial correlation for Gaussian variables.
-\item [2.] \textbf{Estimation and testing in the full vector case:} 
-We have developed a theory for this. More details can be found in the the response to item 3 in the report of referee 2. We have also added a new example where $X_1$ and $X_2$ are not both scalars.
-\item [3.] \textbf{Asymptotic theory and bootstrap validity}. 
-This has been developed in Sections E, F and H in the Supplement, and details can be found in our response to item 4 in the report of referee 2.
-Actually, we are happy that referee 2 raised this point.
-We believe that it has strengthened the paper considerably.
-We add that we have also examined local power and the Pitman criterion in Section  G of the supplement.
-See also our response to item 5 of referee 2.
-\end{itemize}
-\end{quote}
+In addition there are numerous other changes directly related to the comments of the referees and the AE.
+In particular, we have tried to make it more clear that our paper is not just concerned with presenting just another test of conditional independence.
+But rather it is concerned with the introduction of a new framework for measuring conditional dependence, and that this framework reduces to the ordinary partial correlation coefficient in the Gaussian case.
 
-\textbf{AE further general comment:} Referee 1 enjoys reading the paper and have some comments that can be relatively easily addressed.
+We hope that you find the present version acceptable for JBES.
 
-Given the fact that both referees think that the paper is interesting and I also feel it is interesting, I lean toward the positive referee and do not object to seeing a revision of the paper.
-Nevertheless, the authors should keep in mind that they must address the referees' major concerns to their satisfaction in order for the paper to pass the second round of review.
+\vspace{.5cm}
 
-My main comments are as follows, some of which overlap those of the referees.
+Sincerely,
 
-\begin{quote}
-\textbf{Our response:} We are grateful for the opportunity to revise, and we have responded to all of the comments of the two referees, and we will respond to your comments below.
-\end{quote}
+\vspace{.5cm}
 
-\textbf{Comment 1:} I like the idea to extend the local likelihood framework of Hjort and Jones (1996) to estimate the local correlation and to extend the measure of local Gaussian correlation of Tjøstheim and Hufthammer (2013) to the measure of local Gaussian partial correlation (LGPC).
-
-\begin{quote}
-\textbf{Our response:} Thank you.
-\end{quote}
-
-\textbf{Comment 2:} As both referees mention (see R1's point 1 and R2's point 3), one limitation of the proposed testing procedure is that it only tests the conditional independence between two scalar variables \((X_1,X_2)\) given a random vector \((X_1,\ldots,X_p)\).
-It is important to generalize the test to allow
-both \(X_1\) and \(X_2\) to be random vectors.
-
-\begin{quote}
-\textbf{Our response:} Thank you for this important input.
-We now introduce and define the LGPC (matrix) for a general \({\bf X}_1\), \({\bf X}_2\) and \({\bf X}_3\) of dimension \(d_1\), \(d_2\) and \(d_3\) in Section 2.1.
-The full multivariate framework is analyzed in more detail in Section A: ``The multivariate LPGC'' in the Supplement.
-Various special cases are discussed in Section A.2 of that Supplement.
-There are really no new concepts involved, and large parts of the ensuing estimation and testing theory in the Supplement are carried out in this framework.
-\vspace{.3cm} \newline 
-In particular the asymptotic theory for a conditional independence test where both \({\bf X}_1\) and \({\bf X_2}\) (and \({\bf X}_3\)) are vectors is given in Section F of the Supplementary material (Theorem F.1 and Theorem F.2).
-The asymptotic local power is discussed in Section G, and the validity of the bootstrap for the testing statistic are given at this level of generality (Theorems G.1, H.1 and H.2).
-However, the notation becomes more complex and expressions less explicit, and for these reasons and for reasons of space much of the material in the main article is stated for the case of a scalar \(X_1\) and \(X_2\).
-There is a new example of Granger causality in Section 4, though, where the assumption of a scalar \(X_1\) and \(X_2\) is dropped.
-The curse of dimensionality can be avoided at the cost of a simplified pairwise model introduced described and used in the main article and in the supplement.
-Throughout the main paper the reader is referred to appropriate sections of the Supplement for the general vector case.
-\end{quote}
-
-\textbf{Comment 3:} p.5: As Referee 1 points out, the density in (6) is not the real density of \({\Z}\). It is a local Gaussian approximation.
-
-\begin{quote}
-\textbf{Our response:} You are certainly right.
-This was not well written.
-We have changed ``representation'' to ``approximation'' and also linked this directly to the defining approximative equation (4).
-\end{quote}
-
-\textbf{Comment 4.} p.10: Since you claim that \({\bf b}\) is a diagonal matrix of bandwidths (not squared bandwidths).
-It seems that you want to define \(K_{\bf b} = |{\bf b}|^{-1}k({\bf b}^{-1}{\bf x}\).
-
-\begin{quote}
-\textbf{Our response:} Yes, corrections have been implemented. Thank you
-\end{quote}
-
-\textbf{Comment 5:} pp.12-14:
-
-\begin{itemize}
-
-\item[(a)] p.13: It seems that both ${\bf J}_{\bf b}$ and ${\bf M}_{\bf b}$ are of exact order $O(1)$ (so that the second term in the definition of ${\bf M}_{\bf b}$ is asymptotically negligible).
-This suggests that the estimator of the local correlations have convergence rate $(nb^5)^{-1/2}$ under certain under smoothing bias conditions.
-Why is the convergence rate here much slower than the usual $(nb^3)^{-1/2}$-rate for the kernel estimator for a trivariate density?
-In addition, it is possible to estimate the local conditional correlation or not?
-Note that for nonparametric tests based on the conditional distribution, typically only the dimension of the conditioning variable matters for the convergence rate.
-
-\item[(b)] p.14, line 2: $1/(nb^3))$ should be $1/\sqrt{nb^3}$, which is the convergence rate of a trivariate density..
-
-\item [(c)] p.14, line 4: $1/(nb^5)$ should be $1/\sqrt{nb^5}$.
-
-\item [(d)] The results are too limited as it is only for the case with $p=3$. A general result for the generic $p$ is needed.
-
-\item [(e)] Theorem 3.1 is a pointwise result. For the inference purpose, one typically needs a uniform result.
-
-\end{itemize}
-
-\begin{quote}
-\textbf{Our response:}
-
-\begin{itemize}
-\item[(a)] The reason for the slower convergence can basically be found in the expressions for ${\bf J}_{\bf b}$ and ${\bf M}_{\bf b}$.
-These expressions contain the matrix ${\bf u}(\cdot) {\bf u}^T(\cdot)$.
-This matrix has rank 1, and as $b \to 0$ this implies that the matrices ${\bf J}_{\bf b}$ and ${\bf M}_{\bf b}$ are singular and of order $O(b^2)$.
-It is therefore necessary to do a Taylor expansion to find the correct scaling factor, and this in turn leads to a the stated convergence rate of $(nb^5)^{-1/2}$.
-(Note that there was a typo in eq.
-(20).
-It has now been corrected.) This is explained in the proof of Theorem 3.1 and in more detail in Section 4 of Tj{\o}stheim and Hufthammer (2013)  in  a slightly different situation.
-Similar results and with  rather different derivations can also be found in Jones (1996, 1998).
-In the pairwise situation referred to after the statement of Theorem 3.1, the matrices ${\bf J}_{\bf b}$ and ${\bf M}_{\bf b}$ collapse to scalars, the singularity disappears and one is back to the usual rates for the case in nonparametric kernel estimation with the dimension of the conditioning variable determining the convergence rate.
-\vspace{.3cm} \newline It is certainly possible to estimate the local conditional correlation (and in the general vector case) using local likelihood for local correlations and then transforming to the LGPC.
-In the general vector Case this is explained in Sections A.2 and A.3 in the Supplement.
-\item[(b)] This has been corrected. Thank you.
-\item[(c)] This has been corrected. Thank you.
-\item[(d)] Agree and the general vector case has now been implemented. We refer to our response to your comment 2 above.
-\item[(e)] We agree that our results are pointwise results, thus resulting in pointwise confidence intervals for example.
-We have to leave uniform results and confidence bands to future investigations.
-\end{itemize}
-\end{quote}
-
-\textbf{Comment 6:} p.16: Is there a data-driven choice of bandwidth? The choice of \(c=4\) seems too arbitrary.
-
-\begin{quote}
-\textbf{Our response:} There is a data-driven choice for bandwidth in a density estimation context in Otneim and Tj\{\o\}stheim (2017).
-It is based on leave-one-out cross-validation.
-This is referred to in the beginning of Section 4.
-It remains to establish such a procedure for the present situation.
-In the paper we have typically used the asymptotic rates \(b=cn^{-1/9}\) for the full trivariate fit and \(b=cn^{-1/6}\) for the bivariate simplification.
-In the testing results (Table 2, \(2^*\) and \(3^*\) we have used \(c=1.0\) and \(c=1.4\) to examine sensitivity to bandwidth selection.
-They gave similar results.
-Note also that for some of the other competing test several choices of tuning parameters were used.
-For the plots of the estimated LGPC in Section 4 it is true that we used more smoothing to obtain cleaner plots.
-Thus \(c=4\) for Figure 1, but a smaller bandwidth are used in testing and as mentioned in the description of that example the hypothesis of conditional independence is not rejected for any reasonable significance level.
-\end{quote}
-
-\textbf{Comment 7:} p.24-26: The authors propose a new test for conditional independence based on the LGPC.
-The test statistics should be formally studied under the null hypothesis of conditional independence and the usual Pitman sequence of local alternatives.
-For this purpose you need uniform consistent estimate of the local correlations.
-As referee 2 mentions in his/her fourth comment, you also need to justify the asymptotic validity of the bootstrap.
-
-\begin{quote}
-\textbf{Our response:} The asymptotic properties of the test statistic has been formally studied in Section F of the Supplement.
-The validity of the bootstrap for the test statistic is established in Section H of the Supplement (for estimation in Section E of the Supplement.
-The convergence rate of the test statistic is given in Section 5.2 in the main article.
-It is \(n^{-1/2}\) for the pairwise testing case and \((nb^2)^{-1/2}\) for the trivariate testing case.
-This is briefly compared to rates for other test statistics.
-In Section G we look at local asymptotic power and the Pitman criterion.
-We essentially use the approach of Wang and Hong (2018).
-(They avoid the uniformity argument presumably by using the weaker condition \(D^{\prime}\) in Noether, G.E.
-(1955).)
-\end{quote}
-
-\textbf{Comment 8:} p.26: For the comparison with existing tests, you need to compare asymptotic power properties too.
-
-\begin{quote}
-\textbf{Our response:} We refer to our response to comment 7, where we have compared convergence rates and indirectly local power
-\end{quote}
-
-\textbf{Comment 9:} The paper contains a lot of typos. Referee 1 pointed out some of them and the authors should correct others too.
-
-\begin{quote}
-\textbf{Our response:} We have corrected the typos kindly pointed out to us by Referee 1.
-We have read carefully through the revised version, including the Supplement in order to eliminate as many typos as possible.
-\end{quote}
-
-\hypertarget{references}{%
-\subsection{References}\label{references}}
-
-Jones, M. (1996). The local dependence function. Biometrika, \textbf{83}, 899-904.
-
-Jones, M. (1998). Constant local dependence. Journal of Multivariate Analysis, \textbf{64}, 148-155.
-
-Noether, G.E. (1955). On a theorem of Pitman. Annals of Mathematical Statistics, \textbf{26}, 64-68.
-
-Otneim, H. and Tjøstheim, D. (2017). The locally Gaussian density estimator for multivariate data. Statistics and Computing, \textbf{27}, 1595-1616.
-
-Tjæstheim, D. and Hufthammer, K.O. (2013) Local Gaussian correlation: A new measure of dependence. Journal of Econometrics, \textbf{172}, 33-48.
-
-Wang, X. and Hong, Y. (2018). Characteristic function based testing for conditional independence: A nonparametric regression approach. Econometric Theory, \textbf{34}, 815-849.
+Håkon Otneim and Dag Tjøstheim
 

### 2020-09-09 responses/response_reviewers.Rmd
First draft of responses to comments

---
title: " "
output: 
  bookdown::pdf_document2:
      dev: tikz
      keep_tex: yes
      number_sections: no
      toc: no
      includes:
          in_header: latex_preamble_supp.tex
classoption: letterpaper
geometry: margin = 1in
bibliography: "../bibliography.bib"
biblio-style: "apalike"
link-citations: yes
indent: yes
fontsize: 12pt
---

\noindent {Here are our responses to the comments from the referees:}

## Responses to referee 1

**Referee’s general comment:** The authors propose a new conditional dependence measure (called the local Gaussian partial coefficient, LGPC). They then construct an alternative test for conditional independence and test Granger causality in time series. I enjoy reading the paper and only have a few specific comments/suggestions that matter for the exposition of the paper. And I recommend a revision. In the following, I list the expositional issues that I would like to point out and discuss with the authors.

> **Our response:**  We are extremely grateful for your thoughtful comments and for your careful reading of our manuscript. Your comments have lead to very substantial improvements in the paper. Below is our response to the detailed comments 1 -- 12.

**Comment 1:** The whole paper discussed the LGPC coefficient as the local version of the partial correlation coefficient between $X_1$ and $X_2$ given ${\bf X}^{(2)}$, with ${\bf X}^{(1)} = (X_1,X_2)$ and ${\bf X}^{(2)} = (X_3,\ldots,X_p)$. While it is easy to understand, it seems that the discussed dependence measure is only limited to bivariate case conditional on the third vector. Many of the existing conditional independence tests mentioned in Section 4 are, however, designed to work for general dimensions; for example, when testing ${\bf X} \perp {\bf Y}|{\bf Z}$ with $({\bf X},{\bf Y}, {\bf Z}) \in \mathbb{R}^{d_x+d_y+d_z}$. Of course, ``curse of dimensionality'' would always be a serious issue depending on the specific testing methodology adopted; among them the empirical process based tests would be less affected by the dimension of ${\bf Z}$ in theory. Perhaps, more importantly, the aforementioned conditional independence tests do not place restrictions on the dimension of ${\bf X}$ and ${\bf Y}$, and theoretically speaking, they should work for very general settings in which the components $X_1$ and $X_2$ in ${\bf X}^{(1)}$ can be multivariate. How can the current LGPC measure adapt to the multivariate $X_1$ and $X_2$ case? Could you provide some insights or remarks towards this direction?

> **Our response:** Thank you for this important input. We now introduce and define the LGPC (matrix) for a general ${\bf X}_1$, ${\bf X}_2$ and ${\bf X}_3$ of dimension $d_1$, $d_2$ and $d_3$ in Section 2.1. The full multivariate framework is analyzed in more detail in section A:''The multivariate LPGC'' in the supplement. Various special cases are discussed in Section A.2 of that supplement. There are really no new concepts involved, and large parts of the ensuing estimation and testing theory in the Supplement are carried out in this framework. In particular, the asymptotic theory for a conditional independence test where bath ${\bf X}_1$ and ${\bf X}_2$ (and ${\bf X}_3$ are vectors, is given in Section F of the Supplement (Theorem F.1 and Theorem F.2). The asymptotic power is discussed in Section G, and the validity of the bootstrap for the testing statistic are given at this level of generality (Theorems G.1, H.1 and H:2). However, the notation becomes more complex and expressions less explicit, and for these reasons and for reasons of space much of the material in the main article is stated for the case of a scalar $X_1$ and $X_2$. There is a new example of Granger causality in Section 4, though, where the assumption of a scalar $X_1$ and $X_2$ is dropped. The curse of dimensionality can be avoided at the cost of a simplified pairwise model introduced, described and used in the main article and in the Supplement. Throughout the main paper we refer to the appropriate sections of the Supplement for the general vector case.

**Comment 2:** In Equation (6), isn't it an approximation of the density $f_Z$ of ${\bf Z}$ in the neighbourhood of ${\bf z}$? The way it writes now seems to indicate that the expression is exactly the density $f_{\bf Z}$ of ${\bf Z}$, but $f_{\bf Z}({\bf z})$ can certainly differ from $\psi({\bf z}, R({\bf z}))$ in the general case.

> **Our response:** You are certainly right. This was not well written. We have changed "representation" to "approximation" and also linked this directly to the defining approximative equation (4).

**Comment 3:** In the definition of $F_{j,n}(x)$ of page 5, $X_i$ should be $X_{ji}$. Also the empirical analogue for each $Z_j$ can only be written as $Z_{ij,n} = \Phi^{-1}(F_{j,n}(X_{ji}))$ for $j=1,\ldots,p$, and $i=1,\ldots,n$.

> **Our response:** Thank you. This has been corrected on page 6 of the new version.

**Comment 4:** Above Equation (10), for the conditional density, use notation $f_{Z_1|{\bf Z}_3}(z_1|{\bf z}_3)$ to be consistent with subsequent notations.

> **Our response:** Has been done. Thank you.

**Comment 5:** In Equation (15), change $\hat{\bf z}$ to ${\bf z}$. And isn't the definition of kernel function $K_b({\bf x} = |{\bf b}|^{-1}K({\bf b}^{-1}{\bf x})$?

> **Our response:** Yes, both corrections have been implemented.

**Comment 6:** In Equation (17), change $\hat{\bf z}$ to ${\bf z}$, $(\hat{z}_j,\hat{z}_k)$ to $(z_j,z_k)$, and there is a $\psi$ missing in log.

> **Our response:** We have corrected these mistakes. Thanks.

**Comment 7:** In Equation (19), I suspect the expression should be ${\bf M}_b^{-1/2}{\bf J}_b$ instead of ${\bf J}_{b}M_{b}^{-1/2}$. Pls double check. Although the exact expression of the asymptotic covariance matrix ${\bf J}_b^{-1}{\bf M}_b{\bf J}_b^{-1}$ is not available, could you state the orders of ${\bf M}_b$ and ${\bf J}_b$? It is hard to understand the convergence rate without knowing the respective orders of ${\bf M}_b$ and ${\bf J}_b$?

> **Our response:** Thanks a lot for spotting this error in the sequencing of those two matrices. (I am afraid that this mistake has gone undetected in some earlier publications). The orders of ${\bf M}_b$ and ${\bf J}_b$ are $O(b^2)$. This has now been stated with a reference to the Supplement for their derivation. We have also corrected a typo in eq. (20).

**Comment 8:** In page 18, line 10 from below, starting from "It is straightforward to see that". The sentence is very confusing as it holds only when $\rho = 0$ in the context of Figure 3.

> **Our response:** We believe that it holds for every $\rho$ via a conditioning argument. We have rephrased the sentence.

**Comment 9:** In page 25, line 13 from below, starting from "Departures from many types of conditional independence". This sentence is also confusing as conditional independence is a restriction to the underlying populations under the null hypothesis. Are there many other different types of conditional independence?

> **Our response:** We are sorry for this absolutely silly statement. It has been corrected.

**Comment 10:** In Table 1, correct the typos in DGPs 3( change $X_{t-1}^2$ and 10 (Change $Y_t^2$).

> **Our response:** Thank you: Done.

**Comment 11:** In Table 2, what are the tests SCM and SKS?. I couldn't find these definitions anywhere.

> **Our response:** Sorry, has now been included.

**Comment 12:* In the proof of Theorem 3.2, to show that
$$
\frac{1}{n}\sum_{i=1}^{n}(F_n(X_i)-F(x_i)) \to_p 0,
$$
a conventional $U$-statistic argument (e.g. Hoeffding decomposition) for dependent data (e.g. strong mixing) is perhaps much easier and preferable, by noting that
\begin{align*}
\frac{1}{n}\sum_{i=1}^{n}(F_n(X_i)-F(X_i)) &= \frac{1}{n^2}\sum_{i=1}^{n}\sum_{j \neq i}^{n}\{I(X_j \leq X_i)-F(X_i)\}+\frac{1}{n^2}\sum_{i=1}^{n}\{1-F(x_i)\} \\
& = \frac{1}{n}\sum\{0.5 - F(X_i)\} + O_p(n^{-1}) \\
&= O_p(n^{-1/2})+O_p(n^{-1}) = O_p(n^{-1/2}).
\end{align*}
This also implies that the order of $\frac{1}{n}\sum_{i=1}^{n}(F_n(X_i)-F(X_i))$ is sharper than the obtained order $n^{-1/2}(2 \log \log n)^{1/2}$.

> **Our response:** Thanks very much indeed. This is much easier, more elegant and preferable. We have taken the liberty to include it with acknowledgement to "an anonymous referee".

## Responses to referee 2

**Referee's general comment:** The paper is interesting and well written. However, I have the following comments:

> **Our response:** Thank you! We are grateful for your constructive comments. They have lead to great improvements of the paper.

**Comment 1:** What is the difference between your approach and existing nonparametric approaches that use kernel functions (e.g. Normal Kernel) to locally approximate joint distribution functions?

> **Our response:** These differences have been looked at in two papers. For the estimation of the joint distributions we refer to Otneim and Tj{\o}stheim (2017), and for the conditional distribution to Otneim and Tj{\o}stheim (2018). The authors of these two papers show that the local Gaussian approach is much less vulnerable to the curse of dimensionality using a pairwise local Gaussian approach. However, both the kernel and local Gaussian approach to estimating conditional distributions are ill-fitted to actually *measuring* the conditional dependence numerically, for the same reason that the joint distribution is not a well fitted tool to actually measuring the dependence between two or more stochastic variables numerically.

**Comment 2:** Using estimated $Z$ (obtained from the empirical distribution of $X$ instead of the true $Z$ introduces some estimation effect/error. Might be this error is not negligible for finite samples. Did you examine the impact of this estimation error on the estimation of dependence on testing independence when the sample size is small?

> **Our response:** Asymptotically it is negligible as demonstrated by Theorem 3.2 and its proof. We can examine finite sample effects through our simulation experiments where the truth is known. This is the case for the estimation of the level (5\% is correct) for the DGPs created under the conditional independence of the null hypothesis. The results from these experiments are shown in Table 2 in the main article and in Tables $2^* - 4^*$ in the Supplement, Section K. These experiments were carried out for sample sizes of 100, 200 and 400, and even for a sample size as small as 100, results close to the correct ones were obtained.

**Comment 3:** One limitation of your testing procedure is you can only test independence between two scalars (X1 and X2). This is a big issue for testing Granger non-causality, because $X1$ and $X2$ has to be a vector if many lags are needed for Granger causality analysis.

> **Our response:** Thank you for this important comment to which we agree. As a consequence we have rewritten the paper extensively. We now introduce and define the LGPC (matrix) for a general ${\bf X}_1$, ${\bf X}_2$ and ${\bf X}_3$ of dimension $d_1$, $d_2$ and $d_3$ in Section 2.1. The full multivariate framework is analyzed in more detail in Section A: ``The multivariate LPGC'' of the Supplement. Various special cases are discussed in Section A.2 of that supplement. There are really no new concepts involved, and large parts of the ensuing estimation and testing theory in the Supplement are carried out in this framework. In particular, the asymptotic theory for a conditional independence test where both ${\bf X}_1$ and ${\bf X}_2$ (and ${\bf X}_3$) are vectors, is given in Section F of the Supplement (Theorem F.1 and Theorem F.2). The local asymptotic power is discussed in Section G, and the validity of the bootstrap for the testing statistic are given at this level of generality (Theorems G.1, H.1 and H.2). However, the notation becomes more complex and expressions less explicit, and for these reasons and for reasons of space much of the material in the main article is stated for a scalar $X_1$ and $X_2$. There is a new example of Granger causality in Section 4, though, where the assumption of a scalar $X_1$ and $X_2$ is dropped. The curse of dimensionality can be avoided at the cost of a simplified pairwise model introduced, described and used in the main article and in the Supplement. Throughout the paper the reader is referred to the appropriate sections of the Supplement for the general vector case.

**Comment 4:** The asymptotic and bootstrap theories for the test statistic of conditional independence are missing. I appreciate the authors refer to some papers for this, but it will be valuable for readers if these theories can be adapted in the context of the current paper and the results can be included in the paper.

> **Our response:** We  have followed up your suggestion which we think has resulted in a great improvement of the paper. For reasons of space we have had to put these developments in the Supplement, mostly for a general ${\bf X} = \{{\bf X}_1,{\bf X}_2, {\bf X}_3\}$, but with appropriate references in the main article to where they can be found. The validity of the bootstrap for estimation can be found in Section E of that Supplement, the asymptotic theory for the test statistic in Section F, the bootstrap validity for the test statistic in Section G.

**Comment 5:** A theoretical discussion on how your tests compares to existing test (in terms of the rate of convergence ...) will be appreciated.

> **Our response:** Obviously this is another important point. As a consequence we have included a discussion of this in Section 5.2 in the main article and in Section G of the Supplement on local asymptotic power and the Pitman criterion.

**Comment 6:** In the simulation part, I think you should consider other distributions other than the normal distribution. Your test is optimal when the joint distribution is normally distributed, Thus, I believe it is an unfair comparison with Su and White tests and other tests that are fully non-parametric.

> **Our response:** It is somewhat hard not to disagree with this comment. The optimal test in the Gaussian case is the  LIN, the ordinary linear Granger test. It does well (best, see Table 2 of main article) on DGP 5 which is linear Gaussian, but it does miserably on the other DGPs, which have a non-Gaussian distribution. Our test does slightly worse than the Granger test on DGP 5, but it does well on the other DGPs which has a non-Gaussian joint distribution. We think that the competition with Sun and White is fair, when we have chosen *their* examples to compete against.

## References

Otneim, H. and Tjøstheim, D: (2017). The locally Gaussian density estimator for multivariate data. Statistics and Computing, **27**, 1595-1616.

Otneim, H. and Tjøstheim, D. (2018). Conditional density estimation using local Gaussian correlation. Statistics and Computing, **28**, 303-321.

### 2020-09-09 responses/response_reviewers.tex
First draft of responses to comments

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
  letterpaper]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin = 1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{setspace}
\usepackage{sectsty}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\newtheorem{thm}{Theorem}[section]
\newtheorem{cor}[thm]{Corollary}

%\doublespacing
\sectionfont{\centering}
\renewcommand\thesection{\Alph{section}}
\numberwithin{equation}{section}

\newcommand{\X}{\bm{X}}
\newcommand{\Xone}{\bm{X}^{(1)}}
\newcommand{\Xtwo}{\bm{X}^{(2)}}
\newcommand{\xtwo}{\bm{x}^{(2)}}
\newcommand{\x}{\bm{x}}
\newcommand{\hX}{\widehat{X}}
\newcommand{\tX}{\tilde{X}}

\newcommand{\Z}{\bm{Z}}
\newcommand{\z}{\bm{z}}
\newcommand{\hz}{\widehat{z}}
\newcommand{\hfz}{\widehat{\bm{z}}}
\newcommand{\hfZ}{\widehat{\bm{Z}}}
\newcommand{\hZ}{\widehat{\bm{Z}}}
\newcommand{\hnZ}{\widehat{Z}}
\newcommand{\Zone}{\bm{Z}^{(1)}}
\newcommand{\Ztwo}{\bm{Z}^{(2)}}
\newcommand{\zone}{\bm{z}^{(1)}}
\newcommand{\ztwo}{\bm{z}^{(2)}}

\newcommand{\D}{\bm{D}}
\newcommand{\hD}{\widehat{\bm{D}}}

\newcommand{\y}{\bm{y}}
\newcommand{\Y}{\bm{Y}}
\newcommand{\hY}{\widehat{Y}}
\newcommand{\fY}{\bm{Y}}
\newcommand{\hfY}{\widehat{\bm{Y}}}

\newcommand{\fv}{\bm{v}}
\newcommand{\fu}{\bm{u}}

\newcommand{\fC}{\bm{C}}
\newcommand{\fA}{\bm{A}}
\newcommand{\fV}{\bm{V}}
\newcommand{\M}{\bm{M}}
\newcommand{\J}{\bm{J}}

\newcommand{\R}{\bm{R}}
\newcommand{\hR}{\widehat{\bm{R}}}

\newcommand{\hF}{\widehat{F}}
\newcommand{\hf}{\widehat{f}}

\newcommand{\hrho}{\widehat{\rho}}
\newcommand{\frho}{\bm{\rho}}
\newcommand{\hfrho}{\widehat{\bm{\rho}}}

\newcommand{\hh}{\bm{b}}
\newcommand{\bb}{\bm{b}}

\newcommand{\fmu}{\bm{\mu}}
\newcommand{\hfmu}{\widehat{\bm{\mu}}}
\newcommand{\hsigma}{\widehat{\sigma}}
\newcommand{\fSigma}{\bm{\Sigma}}
\newcommand{\hfSigma}{\widehat{\bm{\Sigma}}}
\newcommand{\halpha}{\widehat{\alpha}}
\newcommand{\falpha}{\bm{\alpha}}
\newcommand{\fOmega}{\bm{\Omega}}
\newcommand{\fLambda}{\bm{\Lambda}}
\newcommand{\fepsilon}{\bm{\epsilon}}

\newcommand{\Jb}{\bm{J}_{\hh}}
\newcommand{\Mb}{\bm{M}_{\hh}}

\newcommand{\Cov}{\textrm{Cov}}
\newcommand{\E}{\textrm{E}}
\newcommand{\Var}{\textrm{Var}} 
\newcommand{\di}{\,\textrm{d}}
\newcommand{\Corr}{\textrm{Corr}}

\usepackage{color}
\usepackage{colortbl}

\definecolor{Gray}{gray}{0.9}

\author{}
\date{\vspace{-2.5em}}

\begin{document}

\noindent {Here are our responses to the comments from the referees:}

\hypertarget{responses-to-referee-1}{%
\subsection{Responses to referee 1}\label{responses-to-referee-1}}

\textbf{Referee's general comment:} The authors propose a new conditional dependence measure (called the local Gaussian partial coefficient, LGPC). They then construct an alternative test for conditional independence and test Granger causality in time series. I enjoy reading the paper and only have a few specific comments/suggestions that matter for the exposition of the paper. And I recommend a revision. In the following, I list the expositional issues that I would like to point out and discuss with the authors.

\begin{quote}
\textbf{Our response:} We are extremely grateful for your thoughtful comments and for your careful reading of our manuscript. Your comments have lead to very substantial improvements in the paper. Below is our response to the detailed comments 1 -- 12.
\end{quote}

\textbf{Comment 1:} The whole paper discussed the LGPC coefficient as the local version of the partial correlation coefficient between \(X_1\) and \(X_2\) given \({\bf X}^{(2)}\), with \({\bf X}^{(1)} = (X_1,X_2)\) and \({\bf X}^{(2)} = (X_3,\ldots,X_p)\). While it is easy to understand, it seems that the discussed dependence measure is only limited to bivariate case conditional on the third vector. Many of the existing conditional independence tests mentioned in Section 4 are, however, designed to work for general dimensions; for example, when testing \({\bf X} \perp {\bf Y}|{\bf Z}\) with \(({\bf X},{\bf Y}, {\bf Z}) \in \mathbb{R}^{d_x+d_y+d_z}\). Of course, ``curse of dimensionality'\,' would always be a serious issue depending on the specific testing methodology adopted; among them the empirical process based tests would be less affected by the dimension of \({\bf Z}\) in theory. Perhaps, more importantly, the aforementioned conditional independence tests do not place restrictions on the dimension of \({\bf X}\) and \({\bf Y}\), and theoretically speaking, they should work for very general settings in which the components \(X_1\) and \(X_2\) in \({\bf X}^{(1)}\) can be multivariate. How can the current LGPC measure adapt to the multivariate \(X_1\) and \(X_2\) case? Could you provide some insights or remarks towards this direction?

\begin{quote}
\textbf{Our response:} Thank you for this important input. We now introduce and define the LGPC (matrix) for a general \({\bf X}_1\), \({\bf X}_2\) and \({\bf X}_3\) of dimension \(d_1\), \(d_2\) and \(d_3\) in Section 2.1. The full multivariate framework is analyzed in more detail in section A:`'The multivariate LPGC'' in the supplement. Various special cases are discussed in Section A.2 of that supplement. There are really no new concepts involved, and large parts of the ensuing estimation and testing theory in the Supplement are carried out in this framework. In particular, the asymptotic theory for a conditional independence test where bath \({\bf X}_1\) and \({\bf X}_2\) (and \({\bf X}_3\) are vectors, is given in Section F of the Supplement (Theorem F.1 and Theorem F.2). The asymptotic power is discussed in Section G, and the validity of the bootstrap for the testing statistic are given at this level of generality (Theorems G.1, H.1 and H:2). However, the notation becomes more complex and expressions less explicit, and for these reasons and for reasons of space much of the material in the main article is stated for the case of a scalar \(X_1\) and \(X_2\). There is a new example of Granger causality in Section 4, though, where the assumption of a scalar \(X_1\) and \(X_2\) is dropped. The curse of dimensionality can be avoided at the cost of a simplified pairwise model introduced, described and used in the main article and in the Supplement. Throughout the main paper we refer to the appropriate sections of the Supplement for the general vector case.
\end{quote}

\textbf{Comment 2:} In Equation (6), isn't it an approximation of the density \(f_Z\) of \({\bf Z}\) in the neighbourhood of \({\bf z}\)? The way it writes now seems to indicate that the expression is exactly the density \(f_{\bf Z}\) of \({\bf Z}\), but \(f_{\bf Z}({\bf z})\) can certainly differ from \(\psi({\bf z}, R({\bf z}))\) in the general case.

\begin{quote}
\textbf{Our response:} You are certainly right. This was not well written. We have changed ``representation'' to ``approximation'' and also linked this directly to the defining approximative equation (4).
\end{quote}

\textbf{Comment 3:} In the definition of \(F_{j,n}(x)\) of page 5, \(X_i\) should be \(X_{ji}\). Also the empirical analogue for each \(Z_j\) can only be written as \(Z_{ij,n} = \Phi^{-1}(F_{j,n}(X_{ji}))\) for \(j=1,\ldots,p\), and \(i=1,\ldots,n\).

\begin{quote}
\textbf{Our response:} Thank you. This has been corrected on page 6 of the new version.
\end{quote}

\textbf{Comment 4:} Above Equation (10), for the conditional density, use notation \(f_{Z_1|{\bf Z}_3}(z_1|{\bf z}_3)\) to be consistent with subsequent notations.

\begin{quote}
\textbf{Our response:} Has been done. Thank you.
\end{quote}

\textbf{Comment 5:} In Equation (15), change \(\hat{\bf z}\) to \({\bf z}\). And isn't the definition of kernel function \(K_b({\bf x} = |{\bf b}|^{-1}K({\bf b}^{-1}{\bf x})\)?

\begin{quote}
\textbf{Our response:} Yes, both corrections have been implemented.
\end{quote}

\textbf{Comment 6:} In Equation (17), change \(\hat{\bf z}\) to \({\bf z}\), \((\hat{z}_j,\hat{z}_k)\) to \((z_j,z_k)\), and there is a \(\psi\) missing in log.

\begin{quote}
\textbf{Our response:} We have corrected these mistakes. Thanks.
\end{quote}

\textbf{Comment 7:} In Equation (19), I suspect the expression should be \({\bf M}_b^{-1/2}{\bf J}_b\) instead of \({\bf J}_{b}M_{b}^{-1/2}\). Pls double check. Although the exact expression of the asymptotic covariance matrix \({\bf J}_b^{-1}{\bf M}_b{\bf J}_b^{-1}\) is not available, could you state the orders of \({\bf M}_b\) and \({\bf J}_b\)? It is hard to understand the convergence rate without knowing the respective orders of \({\bf M}_b\) and \({\bf J}_b\)?

\begin{quote}
\textbf{Our response:} Thanks a lot for spotting this error in the sequencing of those two matrices. (I am afraid that this mistake has gone undetected in some earlier publications). The orders of \({\bf M}_b\) and \({\bf J}_b\) are \(O(b^2)\). This has now been stated with a reference to the Supplement for their derivation. We have also corrected a typo in eq. (20).
\end{quote}

\textbf{Comment 8:} In page 18, line 10 from below, starting from ``It is straightforward to see that''. The sentence is very confusing as it holds only when \(\rho = 0\) in the context of Figure 3.

\begin{quote}
\textbf{Our response:} We believe that it holds for every \(\rho\) via a conditioning argument. We have rephrased the sentence.
\end{quote}

\textbf{Comment 9:} In page 25, line 13 from below, starting from ``Departures from many types of conditional independence''. This sentence is also confusing as conditional independence is a restriction to the underlying populations under the null hypothesis. Are there many other different types of conditional independence?

\begin{quote}
\textbf{Our response:} We are sorry for this absolutely silly statement. It has been corrected.
\end{quote}

\textbf{Comment 10:} In Table 1, correct the typos in DGPs 3( change \(X_{t-1}^2\) and 10 (Change \(Y_t^2\)).

\begin{quote}
\textbf{Our response:} Thank you: Done.
\end{quote}

\textbf{Comment 11:} In Table 2, what are the tests SCM and SKS?. I couldn't find these definitions anywhere.

\begin{quote}
\textbf{Our response:} Sorry, has now been included.
\end{quote}

**Comment 12:* In the proof of Theorem 3.2, to show that
\[
\frac{1}{n}\sum_{i=1}^{n}(F_n(X_i)-F(x_i)) \to_p 0,
\]
a conventional \(U\)-statistic argument (e.g.~Hoeffding decomposition) for dependent data (e.g.~strong mixing) is perhaps much easier and preferable, by noting that
\begin{align*}
\frac{1}{n}\sum_{i=1}^{n}(F_n(X_i)-F(X_i)) &= \frac{1}{n^2}\sum_{i=1}^{n}\sum_{j \neq i}^{n}\{I(X_j \leq X_i)-F(X_i)\}+\frac{1}{n^2}\sum_{i=1}^{n}\{1-F(x_i)\} \\
& = \frac{1}{n}\sum\{0.5 - F(X_i)\} + O_p(n^{-1}) \\
&= O_p(n^{-1/2})+O_p(n^{-1}) = O_p(n^{-1/2}).
\end{align*}
This also implies that the order of \(\frac{1}{n}\sum_{i=1}^{n}(F_n(X_i)-F(X_i))\) is sharper than the obtained order \(n^{-1/2}(2 \log \log n)^{1/2}\).

\begin{quote}
\textbf{Our response:} Thanks very much indeed. This is much easier, more elegant and preferable. We have taken the liberty to include it with acknowledgement to ``an anonymous referee''.
\end{quote}

\hypertarget{responses-to-referee-2}{%
\subsection{Responses to referee 2}\label{responses-to-referee-2}}

\textbf{Referee's general comment:} The paper is interesting and well written. However, I have the following comments:

\begin{quote}
\textbf{Our response:} Thank you! We are grateful for your constructive comments. They have lead to great improvements of the paper.
\end{quote}

\textbf{Comment 1:} What is the difference between your approach and existing nonparametric approaches that use kernel functions (e.g.~Normal Kernel) to locally approximate joint distribution functions?

\begin{quote}
\textbf{Our response:} These differences have been looked at in two papers. For the estimation of the joint distributions we refer to Otneim and Tj\{\o\}stheim (2017), and for the conditional distribution to Otneim and Tj\{\o\}stheim (2018). The authors of these two papers show that the local Gaussian approach is much less vulnerable to the curse of dimensionality using a pairwise local Gaussian approach. However, both the kernel and local Gaussian approach to estimating conditional distributions are ill-fitted to actually \emph{measuring} the conditional dependence numerically, for the same reason that the joint distribution is not a well fitted tool to actually measuring the dependence between two or more stochastic variables numerically.
\end{quote}

\textbf{Comment 2:} Using estimated \(Z\) (obtained from the empirical distribution of \(X\) instead of the true \(Z\) introduces some estimation effect/error. Might be this error is not negligible for finite samples. Did you examine the impact of this estimation error on the estimation of dependence on testing independence when the sample size is small?

\begin{quote}
\textbf{Our response:} Asymptotically it is negligible as demonstrated by Theorem 3.2 and its proof. We can examine finite sample effects through our simulation experiments where the truth is known. This is the case for the estimation of the level (5\% is correct) for the DGPs created under the conditional independence of the null hypothesis. The results from these experiments are shown in Table 2 in the main article and in Tables \(2^* - 4^*\) in the Supplement, Section K. These experiments were carried out for sample sizes of 100, 200 and 400, and even for a sample size as small as 100, results close to the correct ones were obtained.
\end{quote}

\textbf{Comment 3:} One limitation of your testing procedure is you can only test independence between two scalars (X1 and X2). This is a big issue for testing Granger non-causality, because \(X1\) and \(X2\) has to be a vector if many lags are needed for Granger causality analysis.

\begin{quote}
\textbf{Our response:} Thank you for this important comment to which we agree. As a consequence we have rewritten the paper extensively. We now introduce and define the LGPC (matrix) for a general \({\bf X}_1\), \({\bf X}_2\) and \({\bf X}_3\) of dimension \(d_1\), \(d_2\) and \(d_3\) in Section 2.1. The full multivariate framework is analyzed in more detail in Section A: ``The multivariate LPGC'\,' of the Supplement. Various special cases are discussed in Section A.2 of that supplement. There are really no new concepts involved, and large parts of the ensuing estimation and testing theory in the Supplement are carried out in this framework. In particular, the asymptotic theory for a conditional independence test where both \({\bf X}_1\) and \({\bf X}_2\) (and \({\bf X}_3\)) are vectors, is given in Section F of the Supplement (Theorem F.1 and Theorem F.2). The local asymptotic power is discussed in Section G, and the validity of the bootstrap for the testing statistic are given at this level of generality (Theorems G.1, H.1 and H.2). However, the notation becomes more complex and expressions less explicit, and for these reasons and for reasons of space much of the material in the main article is stated for a scalar \(X_1\) and \(X_2\). There is a new example of Granger causality in Section 4, though, where the assumption of a scalar \(X_1\) and \(X_2\) is dropped. The curse of dimensionality can be avoided at the cost of a simplified pairwise model introduced, described and used in the main article and in the Supplement. Throughout the paper the reader is referred to the appropriate sections of the Supplement for the general vector case.
\end{quote}

\textbf{Comment 4:} The asymptotic and bootstrap theories for the test statistic of conditional independence are missing. I appreciate the authors refer to some papers for this, but it will be valuable for readers if these theories can be adapted in the context of the current paper and the results can be included in the paper.

\begin{quote}
\textbf{Our response:} We have followed up your suggestion which we think has resulted in a great improvement of the paper. For reasons of space we have had to put these developments in the Supplement, mostly for a general \({\bf X} = \{{\bf X}_1,{\bf X}_2, {\bf X}_3\}\), but with appropriate references in the main article to where they can be found. The validity of the bootstrap for estimation can be found in Section E of that Supplement, the asymptotic theory for the test statistic in Section F, the bootstrap validity for the test statistic in Section G.
\end{quote}

\textbf{Comment 5:} A theoretical discussion on how your tests compares to existing test (in terms of the rate of convergence \ldots) will be appreciated.

\begin{quote}
\textbf{Our response:} Obviously this is another important point. As a consequence we have included a discussion of this in Section 5.2 in the main article and in Section G of the Supplement on local asymptotic power and the Pitman criterion.
\end{quote}

\textbf{Comment 6:} In the simulation part, I think you should consider other distributions other than the normal distribution. Your test is optimal when the joint distribution is normally distributed, Thus, I believe it is an unfair comparison with Su and White tests and other tests that are fully non-parametric.

\begin{quote}
\textbf{Our response:} It is somewhat hard not to disagree with this comment. The optimal test in the Gaussian case is the LIN, the ordinary linear Granger test. It does well (best, see Table 2 of main article) on DGP 5 which is linear Gaussian, but it does miserably on the other DGPs, which have a non-Gaussian distribution. Our test does slightly worse than the Granger test on DGP 5, but it does well on the other DGPs which has a non-Gaussian joint distribution. We think that the competition with Sun and White is fair, when we have chosen \emph{their} examples to compete against.
\end{quote}

\hypertarget{references}{%
\subsection{References}\label{references}}

Otneim, H. and Tjøstheim, D: (2017). The locally Gaussian density estimator for multivariate data. Statistics and Computing, \textbf{27}, 1595-1616.

Otneim, H. and Tjøstheim, D. (2018). Conditional density estimation using local Gaussian correlation. Statistics and Computing, \textbf{28}, 303-321.

\end{document}

### 2020-09-10 lgpc_JBES.Rmd
Re-compiled all figures with random seed for replication. R-code ready.

@@
 \begin{figure}
-    \subfloat[$n = $ 100\label{fig:gaussian-example-plots1}]{\includegraphics[width=0.31\linewidth]{figures/gaussian-example-plots-1} }
-    \subfloat[$n = $ 500\label{fig:gaussian-example-plots2}]{\includegraphics[width=0.31\linewidth]{figures/gaussian-example-plots-2} }
-    \subfloat[$n = $ 2000\label{fig:gaussian-example-plots3}]{\includegraphics[width=0.31\linewidth]{figures/gaussian-example-plots-3} }
+    \subfloat[$n = $ 100\label{fig:gaussian-example-plots1}]{\includegraphics[width=0.31\linewidth]{rcode_and_data/figures/gaussian-example-plots-1} }
+    \subfloat[$n = $ 500\label{fig:gaussian-example-plots2}]{\includegraphics[width=0.31\linewidth]{rcode_and_data/figures/gaussian-example-plots-2} }
+    \subfloat[$n = $ 2000\label{fig:gaussian-example-plots3}]{\includegraphics[width=0.31\linewidth]{rcode_and_data/figures/gaussian-example-plots-3} }
     \caption{Estimated local Gaussian partial correlation between $X_1$ and $X_2$ given $X_3 = 0$, where $(X_1, X_2, X_3)$ is a jointly standard normally distributed vector.
@@
 \begin{figure}[p]
-    \subfloat[Observed pairs $(X_1, X_2)$ from the conditionally \newline Gaussian model.\label{fig:simex2-plots1}]{\includegraphics[width=0.49\linewidth]{figures/simex2-plots-1} }
-    \subfloat[The estimated LGPC between $X_1$ and $X_2$ given $\rho = -0.9$.\label{fig:simex2-plots2}]{\includegraphics[width=0.49\linewidth]{figures/simex2-plots-2} }\newline
-    \subfloat[The estimated LGPC between $X_1$ and $X_2$ \newline  given $\rho = 0$\label{fig:simex2-plots3}]{\includegraphics[width=0.49\linewidth]{figures/simex2-plots-3} }
-    \subfloat[The estimated LGPC between $X_1$ and $X_2$ \newline given $\rho = 0.9$\label{fig:simex2-plots4}]{\includegraphics[width=0.49\linewidth]{figures/simex2-plots-4} }
+    \subfloat[Observed pairs $(X_1, X_2)$ from the conditionally \newline Gaussian model.\label{fig:simex2-plots1}]{\includegraphics[width=0.49\linewidth]{rcode_and_data/figures/simex2-plots-1} }
+    \subfloat[The estimated LGPC between $X_1$ and $X_2$ given $\rho = -0.9$.\label{fig:simex2-plots2}]{\includegraphics[width=0.49\linewidth]{rcode_and_data/figures/simex2-plots-2} }\newline
+    \subfloat[The estimated LGPC between $X_1$ and $X_2$ \newline  given $\rho = 0$\label{fig:simex2-plots3}]{\includegraphics[width=0.49\linewidth]{rcode_and_data/figures/simex2-plots-3} }
+    \subfloat[The estimated LGPC between $X_1$ and $X_2$ \newline given $\rho = 0.9$\label{fig:simex2-plots4}]{\includegraphics[width=0.49\linewidth]{rcode_and_data/figures/simex2-plots-4} }
     \caption{The conditionally Gaussian model}\label{fig:simex2-plots}
@@
 \centering
-\includegraphics{figures/granger-order1}
+\includegraphics{rcode_and_data/figures/granger-order1}
 \caption{The estimated LGPC between $R_{1,t}$ and $R_{2,t-1}$ given $R_{1,t-1}$ along the diagonal of the sample space, for three values of the conditioning variable}
@@
 \centering
-\includegraphics{figures/granger-order2}
+\includegraphics{rcode_and_data/figures/granger-order2}
 \caption{The estimated local Gaussian partial correlations for the three variable pairs in $(R_{1,t}, R_{2,t-1}, R_{2, t-2})$ given $(R_{1, t-1}, R_{1, t-2})$ estimated along the diagonal in the sample space.

### 2020-09-10 lgpc_JBES_supp.Rmd
Re-compiled all figures with random seed for replication. R-code ready.

@@
 
-\begin{figure}[t]
-    \subfloat[Estimated local Gaussian partial correlation \newline between $X_1$ and $X_2$ given $X_3=0$ in the \newline structural model $X_2 = X_1^2 + X_3.$ The color scale \newline goes from $-1$ (dark blue) to $+1$ (dark red).\label{fig:structural-plot1}]{\includegraphics[width=0.49\linewidth]{figures/structural-plot-1} }
-    \subfloat[The estimated LGPC along the curve $X_2 = X_1^2$ with 95\% confidence band.\label{fig:structural-plot2}]{\includegraphics[width=0.49\linewidth]{figures/structural-plot-2} }
-    \caption{The structural model. The fully trivariate model of Section 3.1.1 is used.}\label{fig:structural-plot}
-\end{figure}
-
 \begin{equation}
@@
 The opposite phenomenon is clearly visible in the right hand half of the plot.
-In Figure \@ref(fig:structural-plot2) we have plotted the estimated LGPC along the curve $X_2 = X_1^2$, along with an asymptotic 95% confidence band.
+In Figure \@ref(fig:structural-plot2) we have plotted the estimated LGPC along the curve $X_2 = X_1^2$.
 The full trivariate fit presented in Section 3.1.1 gives a very similar picture.
@@
 
+\begin{figure}[t]
+    \subfloat[Estimated local Gaussian partial correlation \newline between $X_1$ and $X_2$ given $X_3=0$ in the \newline structural model $X_2 = X_1^2 + X_3.$ The color scale \newline goes from $-1$ (dark blue) to $+1$ (dark red).\label{fig:structural-plot1}]{\includegraphics[width=0.49\linewidth]{rcode_and_data/figures/structural-plot-1} }
+    \subfloat[The estimated LGPC along the curve $X_2 = X_1^2$.\label{fig:structural-plot2}]{\includegraphics[width=0.49\linewidth]{rcode_and_data/figures/structural-plot-2} }
+    \caption{The structural model.}\label{fig:structural-plot}
+\end{figure}
+
 \begin{figure}
-\subfloat[Local partial dependence map of $V_t^*$ vs. $R_{t-1}$ \newline on the plane defined by $V_{t-1}^* = 0$\label{fig:realplots1}]{\includegraphics[width=0.49\linewidth]{figures/realplots-1} }
+\subfloat[Local partial dependence map of $V_t^*$ vs.
+$R_{t-1}$ \newline on the plane defined by $V_{t-1}^* = 0$\label{fig:realplots1}]{\includegraphics[width=0.49\linewidth]{rcode_and_data/figures/realplots-1} }
 \subfloat[Local partial dependence map of $R_t$ vs.
-$V_{t-1}^*$ \newline on the plane defined by $R_{t-1} = 0$\label{fig:realplots2}]{\includegraphics[width=0.49\linewidth]{figures/realplots-2} }\newline
-\subfloat[Local partial dependence map of $R_t$ vs. $V_{t-1}^*$ \newline on the  plane defined by $R_{t-1} = -0.05$\label{fig:realplots3}]{\includegraphics[width=0.49\linewidth]{figures/realplots-3} }
-\subfloat[Local partial dependence map of $R_t$ vs. $V_{t-1}^*$ \newline on the  plane defined by $R_{t-1} = 0.05$\label{fig:realplots4}]{\includegraphics[width=0.49\linewidth]{figures/realplots-4} }
+$V_{t-1}^*$ \newline on the plane defined by $R_{t-1} = 0$\label{fig:realplots2}]{\includegraphics[width=0.49\linewidth]{rcode_and_data/figures/realplots-2} }\newline
+\subfloat[Local partial dependence map of $R_t$ vs.
+$V_{t-1}^*$ \newline on the  plane defined by $R_{t-1} = -0.05$\label{fig:realplots3}]{\includegraphics[width=0.49\linewidth]{rcode_and_data/figures/realplots-3} }
+\subfloat[Local partial dependence map of $R_t$ vs.
+$V_{t-1}^*$ \newline on the  plane defined by $R_{t-1} = 0.05$\label{fig:realplots4}]{\includegraphics[width=0.49\linewidth]{rcode_and_data/figures/realplots-4} }
 \caption{Local partial dependence maps for the S\&P 500 data}\label{fig:realplots}

### 2020-09-10 lgpc_JBES_supp.tex
Re-compiled all figures with random seed for replication. R-code ready.

@@
 The opposite phenomenon is clearly visible in the right hand half of the plot.
-In Figure \ref{fig:structural-plot2} we have plotted the estimated LGPC along the curve \(X_2 = X_1^2\), along with an asymptotic 95\% confidence band.
+In Figure \ref{fig:structural-plot2} we have plotted the estimated LGPC along the curve \(X_2 = X_1^2\).
 The full trivariate fit presented in Section 3.1.1 gives a very similar picture.

### 2020-09-10 rcode.R

@@ -1,684 +1,0 @@
-# R-code for reproducing the simulations and empirical results in ?The locally
-# Gaussian partial correlation?
-
-# Packages needed
-library(lg)          # <-- The main package, see documentation for details.
-library(mvtnorm)
-library(dplyr)
-library(ggplot2)
-library(readr)
-library(lubridate)
-library(tidyr)
-library(readxl)
-library(tseries)
-
-# Load the necessary packages and set random seed for full reproducibility
-SEED <- 1
-set.seed(SEED)
-
-# EXAMPLE: THE MULTIVARIATE GAUSSIAN DISTRIBUTION
-# -----------------------------------------------
-
-# We generate samples from three-variate Gaussian distributions with different
-# sample sizes. We then calculate the partial correlations between X1 and X2
-# given X3 and save the results.
-
-# Estimation parameters
-n_gaussian                           <- c(100, 500, 2000)       # The sample sizes
-level                                <- .95                     # Conf. level for asymptotic conf. int.
-grid_size                            <- 7                       # No. of grid points in each direction
-bw_select                            <- "plugin"                # Bandwidth selection method
-plugin_constant                      <- 4                       # Proportionality constant for plugin bandwidth selector
-est_method                           <- "1par"                  # Estimation method
-transform_to_marginal_normality      <- TRUE                    # Transform to marginal normality
-
-# Create the grid
-grid <- expand.grid(X1 = seq(-3.5, 3.5, length.out = grid_size),
-                    X2 = seq(-3.5, 3.5, length.out = grid_size))
-
-# This will be a list of plots that we print out later
-gaussian_example <- list()
-
-# One set of plots for each sample size
-for(i in seq_along(n_gaussian)) {
-    
-    # Generate data
-    x <- rmvnorm(n_gaussian[i], sigma = diag(3))
-    
-    # Estimate the local correlation
-    lg_object <- lg_main(x, 
-                         bw_method = bw_select,
-                         plugin_constant_joint = plugin_constant,
-                         est_method = "1par", 
-                         transform_to_marginal_normality = TRUE)
-    
-    partial_correlations <- partial_cor(lg_object, grid = grid, condition = 0)
-    
-    cplot <- corplot(partial_correlations,
[…]

### 2020-09-10 rcode_and_data/rcode.R

@@ -1,0 +1,454 @@
+# R-code for reproducing the simulations and empirical results in ?The locally
+# Gaussian partial correlation? For prefect reproduction, run this script in its
+# entirety, and notice that the random seed is sett just below the package
+# loadings. Output from the sessionInfo() of the system used to create these
+# figures is listed in the end of this document.
+
+# Packages needed
+library(lg)          # <-- The main package, see documentation for details.
+library(mvtnorm)
+library(dplyr)
+library(ggplot2)
+library(readr)
+library(lubridate)
+library(tidyr)
+library(readxl)
+library(tseries)
+
+# Load the necessary packages and set random seed for full reproducibility
+SEED <- 1
+set.seed(SEED)
+
+# EXAMPLE: THE MULTIVARIATE GAUSSIAN DISTRIBUTION
+# -----------------------------------------------
+
+# We generate samples from three-variate Gaussian distributions with different
+# sample sizes. We then calculate the partial correlations between X1 and X2
+# given X3 and save the results.
+
+# Estimation parameters
+n_gaussian                           <- c(100, 500, 2000)       # The sample sizes
+level                                <- .95                     # Conf. level for asymptotic conf. int.
+grid_size                            <- 7                       # No. of grid points in each direction
+bw_select                            <- "plugin"                # Bandwidth selection method
+plugin_constant                      <- 4                       # Proportionality constant for plugin bandwidth selector
+est_method                           <- "1par"                  # Estimation method
+transform_to_marginal_normality      <- TRUE                    # Transform to marginal normality
+
+# Create the grid
+grid <- expand.grid(X1 = seq(-3.5, 3.5, length.out = grid_size),
+                    X2 = seq(-3.5, 3.5, length.out = grid_size))
+
+# This will be a list of plots that we print out later
+gaussian_example <- list()
+
+# One set of plots for each sample size
+for(i in seq_along(n_gaussian)) {
+    
+    # Generate data
+    x <- rmvnorm(n_gaussian[i], sigma = diag(3))
+    
+    # Estimate the local correlation
+    lg_object <- lg_main(x, 
+                         bw_method = bw_select,
+                         plugin_constant_joint = plugin_constant,
[…]

### 2020-09-16 lgpc_JBES.Rmd
New set of comments from Dag

@@
 It deals with a number of issues complementing the treatment in the main paper.
-Among them are more examples, extensions of the main concept to more general situations, precise formulations of some theoretical results with proofs, as well as discussions of local asymptotic power and the Pitman criterion for testing and validity of the use of the bootstrap.
+Among them are more examples, extensions of the main concept to more general situations, precise formulations of some theoretical results with proofs, as well as discussions of local asymptotic power and the Pitman criterion for testing, and validity of the use of the bootstrap.
 
@@
 In the following, we will not always distinguish between $Z_j$ and $\widehat Z_{j}$.
-In fact, we conclude our asymptotic analysis in Section \@ref(chap:asymptotic) by showing that under certain technical conditions, the error made by estimating $\R(\z)$ using the empirically transformed variables $Z_{j,n}$, instead of $Z_j$, is smaller in the limit than the estimation error made when estimating the local correlations themselves.
+In fact, we conclude our asymptotic analysis in Section \@ref(chap:asymptotic) by showing that under certain technical conditions, the error made by estimating $\R(\z)$ using the empirically transformed variables $\widehat Z_{j}$, instead of $Z_j$, is smaller in the limit than the estimation error made when estimating the local correlations themselves.
 
@@
 To see that conditional independence between $X_1$ and $X_2$ given $\X_3$ is equivalent to $\alpha(\z) \equiv 0$, or equivalently $\alpha(\x) \equiv 0$, we need to follow a few simple steps that would also work in the global Gaussian case.
-Working on the standard normal $z$-scale and assuming that $(z_1, z_2, \z_3)$ has density function $f_{\Z}(\z)$ as given by \@ref(eq:lgdeapprox), it follows from @otneim2017conditional that we can calculate conditional distributions in the same way as in the global Gaussian case.
-For example, the conditional density of $Z_1|\Z_3$ at the point $z_1$ is given by
+Working on the standard normal $z$-scale and assuming that $(Z_1, Z_2, \Z_3)$ has density function $f_{\Z}(\z)$ as in \@ref(eq:lgdeapprox), it follows from @otneim2017conditional that we can calculate conditional distributions in the same way as in the global Gaussian case.
+For example, the conditional density of $Z_1|\Z_3$ at the point $(z_1, \z_3)$ is given by
 
@@
 For the visual display of conditional dependence maps, however, we tend to prefer a fair amount of smoothing, and in Figure \@ref(fig:gaussian-example-plots) we have used $c=4$.
-@otneim2017locally also suggest a algorithm for bandwidth selection in the density estimation context, maximizing the leave-one-out cross validated log-likelihood of the sample;
+@otneim2017locally also suggest an algorithm for bandwidth selection in the density estimation context, maximizing the leave-one-out cross validated log-likelihood of the sample;
 $$\bb_{CV} = \textrm{arg max}_{\bb} \sum_{t = 1}^n \widehat{f}_{\bb}^{(-t)}(\Z_t),$$
@@
 We restrict ourselves to the case where $X_1$ and $X_2$ are scalars, since this is the situation considered in the papers we compare with, but we reiterate (end of Section \@ref(chap:definition)) that there are no conceptual difficulties in extending to the vector case as is seen in the online supplement.
-The theory underlying LGPC testing is essentially based on @lacal2018estimating, and is outlined in Sections F--H of the supplement.
+The theory underlying LGPC testing is essentially based on @lacal2018estimating, and is outlined for the full vector case in Sections F--H of the supplement.
 
@@
 @cheng2012conditional provide simulations for their tests based on the maximal conditional correlation (MCC).
-In the online supplement to this article we provide a richer version of Table \@ref(tab:n100) below, namely Table 2*, that includes more results for the HEL- and MCC-tests at different levels of smoothing.
+In the online supplement to this article we provide a richer version of Table \@ref(tab:n100), namely Table 2*, that includes more results for the HEL- and MCC-tests at different levels of smoothing.
 
@@
 For 3 scalar variables $X_1, X_2, X_3$ a full trivariate analysis depending on all 3 coordinates can be undertaken as in Section \@ref(chap:trivariate-full).
-If any of the components $X_1, X_2$ or $X_3$ are vectors, a pairwise approximation and simplification has been described in Section \@ref(chap:bivariate).
+If any of the components $X_1, X_2$ or $X_3$ are vectors, a pairwise approximation and simplification has been described in Section \@ref(chap:bivariate) and in the supplement.
 This can be likened to additive approximation in nonparametric regression.

### 2020-09-16 lgpc_JBES.tex
New set of comments from Dag

@@
 In the following, we will not always distinguish between \(Z_j\) and \(\widehat Z_{j}\).
-In fact, we conclude our asymptotic analysis in Section \ref{chap:asymptotic} by showing that under certain technical conditions, the error made by estimating \(\R(\z)\) using the empirically transformed variables \(Z_{j,n}\), instead of \(Z_j\), is smaller in the limit than the estimation error made when estimating the local correlations themselves.
+In fact, we conclude our asymptotic analysis in Section \ref{chap:asymptotic} by showing that under certain technical conditions, the error made by estimating \(\R(\z)\) using the empirically transformed variables \(\widehat Z_{j}\), instead of \(Z_j\), is smaller in the limit than the estimation error made when estimating the local correlations themselves.
 
@@
 To see that conditional independence between \(X_1\) and \(X_2\) given \(\X_3\) is equivalent to \(\alpha(\z) \equiv 0\), or equivalently \(\alpha(\x) \equiv 0\), we need to follow a few simple steps that would also work in the global Gaussian case.
-Working on the standard normal \(z\)-scale and assuming that \((z_1, z_2, \z_3)\) has density function \(f_{\Z}(\z)\) as given by \eqref{eq:lgdeapprox}, it follows from Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}) that we can calculate conditional distributions in the same way as in the global Gaussian case.
-For example, the conditional density of \(Z_1|\Z_3\) at the point \(z_1\) is given by
+Working on the standard normal \(z\)-scale and assuming that \((Z_1, Z_2, \Z_3)\) has density function \(f_{\Z}(\z)\) as in \eqref{eq:lgdeapprox}, it follows from Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}) that we can calculate conditional distributions in the same way as in the global Gaussian case.
+For example, the conditional density of \(Z_1|\Z_3\) at the point \((z_1, \z_3)\) is given by
 
@@
 In applications we may circumvent this problem by using the bootstrap.
-The validity of the bootstrap can be proved using a direct adaptations of the techniques of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}).
+The validity of the bootstrap can be proved using a direct adaptations of the techniques of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}).
 It is spelled out in some detail for the pairwise case in Theorems E.1, E.2 and corollaries E.3 and E.4 in Section E of the online supplement.
-The same arguments can be used in the present 3-parameter case (in fact, Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) treat the more general five-parameter case).
+The same arguments can be used in the present 3-parameter case (in fact, Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) treat the more general five-parameter case).
 
@@
 For the visual display of conditional dependence maps, however, we tend to prefer a fair amount of smoothing, and in Figure \ref{fig:gaussian-example-plots} we have used \(c=4\).
-Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}) also suggest a algorithm for bandwidth selection in the density estimation context, maximizing the leave-one-out cross validated log-likelihood of the sample;
+Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}) also suggest an algorithm for bandwidth selection in the density estimation context, maximizing the leave-one-out cross validated log-likelihood of the sample;
 \[\bb_{CV} = \textrm{arg max}_{\bb} \sum_{t = 1}^n \widehat{f}_{\bb}^{(-t)}(\Z_t),\]
@@
 We restrict ourselves to the case where \(X_1\) and \(X_2\) are scalars, since this is the situation considered in the papers we compare with, but we reiterate (end of Section \ref{chap:definition}) that there are no conceptual difficulties in extending to the vector case as is seen in the online supplement.
-The theory underlying LGPC testing is essentially based on Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}), and is outlined in Sections F--H of the supplement.
+The theory underlying LGPC testing is essentially based on Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}), and is outlined for the full vector case in Sections F--H of the supplement.
 
@@
 We have therefore approximated the distribution of \(T_{n,b}\) using the bootstrap.
-The validity of the bootstrap in such contexts has been examined by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) and we adapt their arguments to our situation in the online supplement, Section H.
+The validity of the bootstrap in such contexts has been examined by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) and we adapt their arguments to our situation in the online supplement, Section H.
 
@@
 Cheng and Huang (\protect\hyperlink{ref-cheng2012conditional}{2012}) provide simulations for their tests based on the maximal conditional correlation (MCC).
-In the online supplement to this article we provide a richer version of Table \ref{tab:n100} below, namely Table 2*, that includes more results for the HEL- and MCC-tests at different levels of smoothing.
+In the online supplement to this article we provide a richer version of Table \ref{tab:n100}, namely Table 2*, that includes more results for the HEL- and MCC-tests at different levels of smoothing.
 
@@
 For 3 scalar variables \(X_1, X_2, X_3\) a full trivariate analysis depending on all 3 coordinates can be undertaken as in Section \ref{chap:trivariate-full}.
-If any of the components \(X_1, X_2\) or \(X_3\) are vectors, a pairwise approximation and simplification has been described in Section \ref{chap:bivariate}.
+If any of the components \(X_1, X_2\) or \(X_3\) are vectors, a pairwise approximation and simplification has been described in Section \ref{chap:bivariate} and in the supplement.
 This can be likened to additive approximation in nonparametric regression.
@@
 \leavevmode\hypertarget{ref-lacal2018estimating}{}%
----------. 2018. ``Estimating and Testing Nonlinear Local Dependence Between Two Time Series.'' \emph{Journal of Business \& Economic Statistics}, 1--13.
+---------. 2019. ``Estimating and Testing Nonlinear Local Dependence Between Two Time Series.'' \emph{Journal of Business \& Economic Statistics} 37 (4): 648--60.
 

### 2020-09-16 lgpc_JBES_supp.Rmd
New set of comments from Dag

@@
 We proceed now to establish the notion of conditional independence between $\Z_1$ and $\Z_2$ using the LGPC in the multivariate case.
-The conditional densities of $\Z_1|\Z_3$ and $\Z_2|\Z_3$ in terms of the local Gaussian parameters are given by
+The conditional densities of $\Z_1|\Z_3$ and $\Z_2|\Z_3$ in terms of the local Gaussian parameters at the point $(\z_1, \z_2, \z_3)$ are given by
 
@@
 \end{equation}
-Under conditional independence between $\Z_1$ and $\Z_2$ given $\Z_3$, the joint conditional density of $(\Z_1, \Z_2|\Z_3)$ is given by
+Under conditional independence between $\Z_1$ and $\Z_2$ given $\Z_3$, the joint conditional density of $(\Z_1, \Z_2|\Z_3)$ at the point $(\z_1, \z_2, \z_3)$ is given by
 
@@
 Here, since all variables are normalized with variance one, we have that
-$$\fSigma = \begin{pmatrix} 1 & \rho_{12} & \rho_{13} \\ \rho_{12} & 1 & \rho_{23} \\ \rho_{31} & \rho_{32} & 1 \end{pmatrix},$$
+$$\fSigma = \begin{pmatrix} 1 & \rho_{12} & \rho_{13} \\ \rho_{21} & 1 & \rho_{23} \\ \rho_{31} & \rho_{32} & 1 \end{pmatrix},$$
 and 
@@
 where $K$ is a product kernel, and, in our case, $b_w$ is the vector defined by $b_w^T = \begin{pmatrix} 1 & b_1w_1 & b_2w_2 & b_3w_3 \end{pmatrix}$, and $A$ is the $3\times 4$ matrix $A = \begin{pmatrix} v & v_{z_1} & v_{z_2} & v_{z_3}\end{pmatrix}$ with $\fv^T =  \begin{pmatrix} v_1 & v_2 & v_3 \end{pmatrix}$, $v_i = \partial L_n/\partial \rho_i$, where the index $i=1,2,3$ represents the three local correlations, and $v_{z_i} = \partial \fv/\partial z_i$.
-In the next step, they compute the matrix
-
-\noindent $\int K^2(w_1, w_2, w_3)b_wb_w^T\,\textrm{d}w_1\,\textrm{d}w_2\,\textrm{d}w_3$, which, by omitting all constant factors and exploiting that $\int w^k K^2(w)\,\textrm{d}w = 0$ for $k = 1,3$, in our case becomes the diagonal matrix
+In the next step, they Taylor expand $f$ and compute the matrix $\int K^2(w_1, w_2, w_3)b_wb_w^T\,\textrm{d}w_1\,\textrm{d}w_2\,\textrm{d}w_3$, which, by omitting all constant factors and exploiting that $\int w^k K^2(w)\,\textrm{d}w = 0$ for $k = 1,3$, in our case becomes the diagonal matrix
 $$H = \begin{pmatrix} 1 & 0 & 0 & 0 \\ 0 & b_1^2w_1^2 & 0 & 0 \\ 0 & 0 & b_2^2w_2^2 & 0 \\ 0 & 0 & 0 & b_3^2w_3^2 \end{pmatrix}.$$
@@
 \frac{1}{n}\sum_{t=1}^n \left(F_n(X_{i,t}) - F(X_{i,t})\right) &= 
-\frac{1}{n^2}\sum_{t=1}^n\sum_{s \neq t}^n \left(I(X_{i,s} \leq X_{i,t}) - F(X_{i,t})\right) + \frac{1}{n^2}\sum_{t=1}^n \left(1 - F(X_{i,t}\right) \\
+\frac{1}{n^2}\sum_{t=1}^n\sum_{s \neq t}^n \left(I(X_{i,s} \leq X_{i,t}) - F(X_{i,t})\right) + \frac{1}{n^2}\sum_{t=1}^n \left(1 - F(X_{i,t})\right) \\
 &= \frac{1}{n} \sum_{t=1}^n \left(0.5 - F(X_{i,t})\right) + O_p\left(n^{-1}\right) \\
@@
 Let assumptions A1, A2 and A4 hold, and  let $\{L^{(b)}((Z_{i,t},Z_{j,t}),\rho)\}$, $\rho \in \Theta$ be Lipschitz continuous and 4-dominated on $\Theta$ uniformly in $t,n$.
-Moreover, assume that, as $b=b_n \to 0, \sigma_b^2 \stackrel{\textrm{def}}{=} \Var(L^{(b)}((Z_{i,t},Z_{j,t}),\rho) \to \sigma^2 < \infty$ and $\mu_b \stackrel{\textrm{def}}{=} \E(L^{(b)}(Z_{i,t},Z_{j,t}),\rho)) \to \mu < \infty$, for all $t=1,\ldots,n$ and $\rho \in \Theta$.
+Moreover, assume that, as $b=b_n \to 0, \sigma_b^2 \stackrel{\textrm{def}}{=} \Var(L^{(b)}((Z_{i,t},Z_{j,t}),\rho)) \to \sigma^2 < \infty$ and $\mu_b \stackrel{\textrm{def}}{=} \E(L^{(b)}(Z_{i,t},Z_{j,t}),\rho)) \to \mu < \infty$, for all $t=1,\ldots,n$ and $\rho \in \Theta$.
 Then,
@@
 A more complicated version is given by @lacal2018estimating concerning the five parameter case.
-The proofs are essentially the same and are spelled out in some detail in the supplementary material of that paper as Theorems D1 and D2.
+The proofs are essentially the same and are spelled out in some detail in the supplementary material of that paper as the proofs of Theorems D1 and D2.
 
@@
 
-Let $\M_{ij}$ and $\J_{ij}$ be the $M$-matrix and $J$-matrix appearing in the proof of Corollary \ref{cor:E4} for a fixed $\rho_{ij}$.
+Let $\M_{ij}$ and $\J_{ij}$ be as in the proof of Corollary \ref{cor:E4} for a fixed $\rho_{ij}$.
 Note that in the one-parameter case of $\rho_{ij}$ in Corollary \ref{cor:E4}, to first order the covariance matrix $\fLambda$ is a diagonal matrix having ordered values of $M_{ij}/J_{ij}^2$ along the diagonal.
@@
 One possible choice of the function $h$ is $h(\frho(\z)) = \sum_{i,j} \alpha_{i,j}^2(z_{1,i},z_{2,j}|\ztwo)$, which is a function of $\frho$ since the $\alpha$-s are functions of $\frho$.
-The corresponding function in @lacal2018estimating is $h(\theta(x))$, where $\theta(x)$ is at most 5-dimensional and $\x$ is bivariate.
+The corresponding function in @lacal2018estimating is $h(\theta(\x), \x)$, where $\theta(\x)$ is at most 5-dimensional and $\x$ is bivariate.
 As in that paper we proceed in two stages.
@@
 It is more difficult to obtain the limiting distribution.
-We will follow the technique in the proof of Theorem 3.1 in @lacal2018estimating.
+We will follow the technique in the proof of Theorem 3.2 in @lacal2018estimating.
 See also @joe1989estimation.
@@
 \begin{align*}
-A_b(\z)  & = \int_{S} \left[\sum_{i,j}\left\{\frac{\partial}{\partial \rho_{ij}} h(\frho_b,(\fv)\right\}J_{ij}^{-1}(\rho_{ij,b}(\fv_{ij}))K_b(\z_{ij}- \fv_{ij})u_{ij}(\z_{ij},\rho_{ij,b}(\fv_{ij}))\right] \di F(\fv) \\
+A_b(\z)  & = \int_{S} \left[\sum_{i,j}\left\{\frac{\partial}{\partial \rho_{ij}} h(\frho_b,(\fv))\right\}J_{ij}^{-1}(\rho_{ij,b}(\fv_{ij}))K_b(\z_{ij}- \fv_{ij})u_{ij}(\z_{ij},\rho_{ij,b}(\fv_{ij}))\right] \di F(\fv) \\
 & \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad + h(\frho_b(\z))1_S(\z)
@@
 & = \frac{1}{n}\sum_{t=1}^{n}K_b(\Z_{ij,t}-\z_{ij})u_{ij}(\Z_{ij,t},\rho_{ij}) - 
-\int K_b(\fv_{ij}-\z_{ij})u(\fv_{ij},\rho_{ij})\psi(\fv_{ij},\rho_{ij}) \di \fv_{ij}
+\int K_b(\fv_{ij}-\z_{ij})u_{ij}(\fv_{ij},\rho_{ij})\psi(\fv_{ij},\rho_{ij}) \di \fv_{ij}
 \end{align*}
@@
 $$
-where $G_n(z) = \sqrt{n}(F_n(z)-F(z))$ and
-\begin{align*}
-\fV_n(z) &= \sqrt{n} (\frho_{n,b}(\z) - \frho_b(\z)) \\
+where $G_n(\z) = \sqrt{n}(F_n(\z)-F(\z))$ and
+\begin{align*}
+\fV_n(\z) &= \sqrt{n} (\frho_{n,b}(\z) - \frho_b(\z)) \\
 & \sim \sqrt{n}{\J}_b^{-1}(\z)\left\{\frac{1}{n} \sum_{t=1}^{n}K_b(\Z_t-\z)\fu(\Z_t,\frho_b(\z)) - \int K_b(\fv-\z)\fu(\fv,\frho_b(\z))\psi(\fv,\rho_b(\z)) \di \fv \right\}.
@@
 \end{align*}
-Here, using Theorem \@ref(thm:F2), under $H[a_n]$, and Proposition 6.3.9 of @brockwell1991time,
+Analogously to @wang2017characteristic, arguing as in the proof of Theorem \@ref(thm:F2) and using Proposition 6.3.9 of @brockwell1991time, as $b_n, a_n \to 0$,
 $$
@@
 \begin{align*}
-A_b(z) = \int_{S} \Big[\sum_{i,j}\frac{\partial}{\partial \rho_{ij,n,b}} h(\rho_{n,b}(\fv)J_{ij}^{-1}(\rho_{ij,n,b}(\fv_{ij}))K_b(\z_{ij}-\fv_{ij})&u_{ij}(\z_{ij},\rho_{ij,n,b}(\fv_{ij})\Big]  \di F(\fv) \\
-& + h(\rho_{ij,n,b}(\z)1_S(\z),
+A_b(z) = \int_{S} \Big[\sum_{i,j}\frac{\partial}{\partial \rho_{ij,n,b}} h(\rho_{n,b}(\fv))J_{ij}^{-1}(\rho_{ij,n,b}(\fv_{ij}))K_b(\z_{ij}-\fv_{ij})&u_{ij}(\z_{ij},\rho_{ij,n,b}(\fv_{ij})\Big]  \di F(\fv) \\
+& + h(\rho_{ij,n,b}(\z))1_S(\z),
 \end{align*}
@@
 where E$^*(\cdot)$ is the mean value under the measure $\mathbb{P}^*_{\omega}$, and where Lacal and Tjøstheim [-@lacal2017local, proof of Theorem 4.4] may be consulted for more details in the derivation and a more explicit analog expression for $I_{ij}^*(\z_{ij},\rho_{ij,n})$ for the case considered there.
-In contradistinction to @lacal2017local, $J_{ij,n,b}$ is not singular as $\bb \to \bm{0}$.
+In contradistinction to @lacal2017local, $J_{ij,n,b}$ is not singular as $n\to \infty, \bb \to 0$.
 In fact, it can be shown that, cf.
-Otneim and Tjøstheim [-@otneim2017locally, Theorem 2], as $\bb \to \bm{0}$, $J_{ij,n,b}(\z_{ij}) \stackrel{\mathbb{P}}{\to} u(\z_{ij},\rho_{ij})\psi(\z_{ij},\rho_{ij}(\z_{ij}))$.
+Otneim and Tjøstheim [-@otneim2017locally, Theorem 2], as $n \to \infty$ and $\bb \to \bm{0}$, $J_{ij,n,b}(\z_{ij}) \stackrel{\mathbb{P}}{\to} u(\z_{ij},\rho_{ij})\psi(\z_{ij},\rho_{ij}(\z_{ij}))$.
 
@@
 To be able to do that one must device a test statistic that is constructed with a probability structure as under the null hypothesis of conditional independence.
-It is difficult to obtain such a test based on asymptotic arguments for at least two reasons: As mentioned before the asymptotic distribution of test functionals such as $T_{n,b}$ is known to be inaccurate except for extremely large sample sizes.
+It is difficult to obtain such a test based on asymptotic arguments: As mentioned before the asymptotic distribution of test functionals such as $T_{n,b}$ is known to be inaccurate except for extremely large sample sizes.
 It is also difficult to build the probability null structure into the scaling factor of $T_{n,b}$ since this would amount to building into that structure certain relationships between pairwise local correlations.

### 2020-09-16 lgpc_JBES_supp.tex
New set of comments from Dag

@@
 We proceed now to establish the notion of conditional independence between \(\Z_1\) and \(\Z_2\) using the LGPC in the multivariate case.
-The conditional densities of \(\Z_1|\Z_3\) and \(\Z_2|\Z_3\) in terms of the local Gaussian parameters are given by
+The conditional densities of \(\Z_1|\Z_3\) and \(\Z_2|\Z_3\) in terms of the local Gaussian parameters at the point \((\z_1, \z_2, \z_3)\) are given by
 
@@
 \end{equation}
-Under conditional independence between \(\Z_1\) and \(\Z_2\) given \(\Z_3\), the joint conditional density of \((\Z_1, \Z_2|\Z_3)\) is given by
+Under conditional independence between \(\Z_1\) and \(\Z_2\) given \(\Z_3\), the joint conditional density of \((\Z_1, \Z_2|\Z_3)\) at the point \((\z_1, \z_2, \z_3)\) is given by
 
@@
 
-Observe, finally, that a more general pairwise scheme can be devised, as treated in detail by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}), Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) and Jordanger and Tjøstheim (\protect\hyperlink{ref-jordanger2017nonlinear}{2020}):
+Observe, finally, that a more general pairwise scheme can be devised, as treated in detail by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}), Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) and Jordanger and Tjøstheim (\protect\hyperlink{ref-jordanger2017nonlinear}{2020}):
 
@@
 
-We must next proceed by introducing some new concepts before stating the equivalents of assumptions A1 -- A2 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}).
+We must next proceed by introducing some new concepts before stating the equivalents of assumptions A1 -- A2 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}).
 These assumptions are similar but distinct from those used in Theorem 3.1 in the main article and Theorem 2 in Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}).
@@
 
-The conditions A1-- A2 are quite mild and discussed by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}).
+The conditions A1-- A2 are quite mild and discussed by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}).
 Note that they have been formulated for a scalar parameter \(\rho = \rho_{ij}\).
@@
 
-The proof of Theorem \ref{thm:A1} is mostly identical to the proof of Theorem 2.1 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) with trivial modifications to take into account the scalar nature of the parameter \(\rho_{ij}\) in the present case.
-A distributional limit theorem analogous to Theorem 2.2 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) can also be rephrased and proved as in that paper, but we refrain from doing this given the distributional results already stated in Section 3 of the main part of the paper.
+The proof of Theorem \ref{thm:A1} is mostly identical to the proof of Theorem 2.1 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) with trivial modifications to take into account the scalar nature of the parameter \(\rho_{ij}\) in the present case.
+A distributional limit theorem analogous to Theorem 2.2 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) can also be rephrased and proved as in that paper, but we refrain from doing this given the distributional results already stated in Section 3 of the main part of the paper.
 
@@
 where \(K\) is a product kernel, and, in our case, \(b_w\) is the vector defined by \(b_w^T = \begin{pmatrix} 1 & b_1w_1 & b_2w_2 & b_3w_3 \end{pmatrix}\), and \(A\) is the \(3\times 4\) matrix \(A = \begin{pmatrix} v & v_{z_1} & v_{z_2} & v_{z_3}\end{pmatrix}\) with \(\fv^T = \begin{pmatrix} v_1 & v_2 & v_3 \end{pmatrix}\), \(v_i = \partial L_n/\partial \rho_i\), where the index \(i=1,2,3\) represents the three local correlations, and \(v_{z_i} = \partial \fv/\partial z_i\).
-In the next step, they compute the matrix
-
-\noindent \(\int K^2(w_1, w_2, w_3)b_wb_w^T\,\textrm{d}w_1\,\textrm{d}w_2\,\textrm{d}w_3\), which, by omitting all constant factors and exploiting that \(\int w^k K^2(w)\,\textrm{d}w = 0\) for \(k = 1,3\), in our case becomes the diagonal matrix
+In the next step, they Taylor expand \(f\) and compute the matrix \(\int K^2(w_1, w_2, w_3)b_wb_w^T\,\textrm{d}w_1\,\textrm{d}w_2\,\textrm{d}w_3\), which, by omitting all constant factors and exploiting that \(\int w^k K^2(w)\,\textrm{d}w = 0\) for \(k = 1,3\), in our case becomes the diagonal matrix
 \[H = \begin{pmatrix} 1 & 0 & 0 & 0 \\ 0 & b_1^2w_1^2 & 0 & 0 \\ 0 & 0 & b_2^2w_2^2 & 0 \\ 0 & 0 & 0 & b_3^2w_3^2 \end{pmatrix}.\]
@@
 
-We can use the bootstrap theory by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}, \protect\hyperlink{ref-lacal2018estimating}{2018}) more or less directly, which in turn is based on Gon\c{c}alves and White (\protect\hyperlink{ref-gonccalves2002bootstrap}{2002}, \protect\hyperlink{ref-gonccalves2004maximum}{2004}), but we spell out some details below.
-First, we need to specify the analogue of the conditions A1--A5 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}), where we adapt their notation to our situation by replacing the parameter \(\theta\) with \(\rho_{ij}\), and the time series \(\{X_t, Y_t\}\) by our multivariate observations \(\{\Z_t\}\).
+We can use the bootstrap theory by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}, \protect\hyperlink{ref-lacal2018estimating}{2019}) more or less directly, which in turn is based on Gon\c{c}alves and White (\protect\hyperlink{ref-gonccalves2002bootstrap}{2002}, \protect\hyperlink{ref-gonccalves2004maximum}{2004}), but we spell out some details below.
+First, we need to specify the analogue of the conditions A1--A5 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}), where we adapt their notation to our situation by replacing the parameter \(\theta\) with \(\rho_{ij}\), and the time series \(\{X_t, Y_t\}\) by our multivariate observations \(\{\Z_t\}\).
 Since \(\rho = \rho_{ij}\) is estimated using the two corresponding data vectors only, we can, for the time being and without loss of generality, assume that the observation series is bivariate: \({\Z_t} = \{Z_{i,t}, Z_{j,t}\}\).
@@
 This automatically leads to the same block-bootstrap for any pair of components which we have denoted by \(\{Z_{i,t}, Z_{j,t}\}\) above.
-Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) use both the block bootstrap as well as the stationary bootstrap, and the results below hold irrespective of this particular choice.
-We will therefore not distinguish between them here, but rather refer the reader to Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) for further details on this distinction.
+Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) use both the block bootstrap as well as the stationary bootstrap, and the results below hold irrespective of this particular choice.
+We will therefore not distinguish between them here, but rather refer the reader to Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) for further details on this distinction.
 Bootstrapped quantities will be denoted with a star (\(*\)) in the following.
@@
 Assumption A4 sets the stage for the block bootstrapping, whereas A5 is needed in proving the validity of the bootstrap central limit theorem.
-The assumption A4 is not restrictive, and again we refer to Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) for more discussion.
+The assumption A4 is not restrictive, and again we refer to Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) for more discussion.
 
@@
 Theorems \ref{thm:E1} and \ref{thm:E2} are formulated in their simplest case here, which concerns only one local parameter.
-A more complicated version is given by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) concerning the five parameter case.
-The proofs are essentially the same and are spelled out in some detail in the supplementary material of that paper as Theorems D1 and D2.
+A more complicated version is given by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) concerning the five parameter case.
+The proofs are essentially the same and are spelled out in some detail in the supplementary material of that paper as the proofs of Theorems D1 and D2.
 
@@
 
-\emph{Proof:} The proof of Theorem \ref{thm:E2} is given as the proof of Theorem D2 in the online appendix accompanying Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}).
+\emph{Proof:} The proof of Theorem \ref{thm:E2} is given as the proof of Theorem D2 in the online appendix accompanying Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}).
 This proof can be adapted to prove Corollary \ref{cor:E4}.
 The starting point consists in replacing \(L_{n,b}^*\) and \(L_{n,b}\) by \(L^* = \sum_{i,j} a_{ij}L_{ij,n,b}^*\) and \(L = \sum_{i,j}a_{ij}L_{ij,n,b}\), respectively, where the sum runs over all possible pairs \((i,j)\) corresponding to \(\rho_{ij}^{1,2}\), \(\rho_{ij}^{1,3}\), \(\rho_{ij}^{2,3}\) that are included in the \(\frho\)-vector.
-The derivations in the proof of Theorem D2 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) deal with the more complicated five-parameter case.
+The derivations in the proof of Theorem D2 by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) deal with the more complicated five-parameter case.
 In the present case, however, the vector \(\frho\) consists of correlations \(\rho_{ij}\) that have been estimated using the one-parameter pairwise setup, where, cf.~Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}), the quantities \(J_{ij}\) and \(M_{ij}\) are scalars.
 Here \(L_{ij,n.b}\) is the log likelihood for the pair \((i,j)\) treated in the proof of Theorem D2, and \(a_{ij}\) is a corresponding sequence of real numbers.
-Because of the linearity of the expressions for \(L^*\) and \(L\) it is easily checked by going step by step that the proof of Theorem D2 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) can be carried through for \(L^*\) and \(L\).
+Because of the linearity of the expressions for \(L^*\) and \(L\) it is easily checked by going step by step that the proof of Theorem D2 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) can be carried through for \(L^*\) and \(L\).
 The only difference is that the quantities \(J_{n,b}\) and \(M_{n,b}\) now have to be replaced by \(J_{n.b}(a) = \E(-\sum_{i,j}a_{ij}\nabla^2 L_{ij,n,b}(\rho_{0,b}))\) and \(\Var(\sqrt{nb_1b_2}\sum_{i,j} a_{ij}\nabla L_{ij,n,b})\).
@@
 
-Let \(\M_{ij}\) and \(\J_{ij}\) be the \(M\)-matrix and \(J\)-matrix appearing in the proof of Corollary \ref{cor:E4} for a fixed \(\rho_{ij}\).
+Let \(\M_{ij}\) and \(\J_{ij}\) be as in the proof of Corollary \ref{cor:E4} for a fixed \(\rho_{ij}\).
 Note that in the one-parameter case of \(\rho_{ij}\) in Corollary \ref{cor:E4}, to first order the covariance matrix \(\fLambda\) is a diagonal matrix having ordered values of \(M_{ij}/J_{ij}^2\) along the diagonal.
@@
 Note also that the 3-coordinate case of alternative (i) can be stated and proved in the same way, but a different scaling factor is required to avoid a singular covariance matrix as seen in the proof of Theorem 3.1 in the main article.
-The 3-coordinate case is in fact closer to the situation in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}).
+The 3-coordinate case is in fact closer to the situation in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}).
 
@@
 A more convenient test statistic is \(\sum_{i,j} \alpha_{ij}^2\).
-To follow the notation of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) we write, letting \(\z=(\z_1,\z_2,\z_3)\),
+To follow the notation of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) we write, letting \(\z=(\z_1,\z_2,\z_3)\),
 \[
@@
 One possible choice of the function \(h\) is \(h(\frho(\z)) = \sum_{i,j} \alpha_{i,j}^2(z_{1,i},z_{2,j}|\ztwo)\), which is a function of \(\frho\) since the \(\alpha\)-s are functions of \(\frho\).
-The corresponding function in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) is \(h(\theta(x))\), where \(\theta(x)\) is at most 5-dimensional and \(\x\) is bivariate.
+The corresponding function in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) is \(h(\theta(\x), \x)\), where \(\theta(\x)\) is at most 5-dimensional and \(\x\) is bivariate.
 As in that paper we proceed in two stages.
@@
 It is more difficult to obtain the limiting distribution.
-We will follow the technique in the proof of Theorem 3.1 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}).
+We will follow the technique in the proof of Theorem 3.2 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}).
 See also Joe (\protect\hyperlink{ref-joe1989estimation}{1989}).
@@
 This will be exploited in Section \ref{chap:pitman}.
-See also Theorem 3.2 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) and its relation to Corollary 3.1 in that paper.
+See also Theorem 3.2 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) and its relation to Corollary 3.1 in that paper.
 
@@
 \]
-where Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}, \protect\hyperlink{ref-lacal2018estimating}{2018}) may be consulted for more details in the derivation and a more explicit expression for \(I_{ij}(\z_{ij},\rho_{ij})\) for the case considered there.
-In contradistinction to the 5-parameter case considered by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}), in our present 1-parameter case \(J_{b,ij}\) is not singular as \(\bb \to \bm{0}\).
+where Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}, \protect\hyperlink{ref-lacal2018estimating}{2019}) may be consulted for more details in the derivation and a more explicit expression for \(I_{ij}(\z_{ij},\rho_{ij})\) for the case considered there.
+In contradistinction to the 5-parameter case considered by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}), in our present 1-parameter case \(J_{b,ij}\) is not singular as \(\bb \to \bm{0}\).
 In fact, it is shown by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, Theorem 2) that as \(\bb \to \bm{0}\), \(J_{b,ij} \to u(\z_{ij},\rho_{ij})\psi(\z_{ij},\rho(\z_{ij}))\).
@@
 The vector \(\fV_n(\z) = \sqrt{n}[\frho_{n,b}(\z)-\frho_b(\z)]\) has the dimension of \(\frho(\z)\) and consists of components \(\{\rho_{ij,n,b}(\z_{ij})-\rho_{ij,b}(\z_{ij})\}\) where \(i\) and \(j\) in the most general case run over all possible pairs formed by the components of \(\frho(\z)\).
-To be able to use the mixing central limit theorem we follow Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) and express this more explicitly in terms of the variables \(\Z_{t}\), \(t=1,\ldots,n\).
-By Taylor expansion as in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}),
+To be able to use the mixing central limit theorem we follow Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) and express this more explicitly in terms of the variables \(\Z_{t}\), \(t=1,\ldots,n\).
+By Taylor expansion as in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}),
 \begin{align*}
@@
 The sum in the second term taken together with the last term above converges due to the mixing condition.
-Details of this argument are identical to those of the last part of the proof of Theorem 3.1 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}), as is the case of \(\bb_n \rightarrow \bm{0}\), completing the proof by referring to Proposition 6.3.9 of Brockwell and Davis (\protect\hyperlink{ref-brockwell1991time}{2006}).
+Details of this argument are identical to those of the last part of the proof of Theorem 3.1 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}), as is the case of \(\bb_n \rightarrow \bm{0}\), completing the proof by referring to Proposition 6.3.9 of Brockwell and Davis (\protect\hyperlink{ref-brockwell1991time}{2006}).
 \(\blacksquare\)
@@
 This may be somewhat surprising since the convergence rate of \(\frho_{n,b}(\z_{ij})-\frho(\z_{ij})\) is \((nb_1b_2)^{-1/2}\), but this is a well-known phenomenon for functionals of type \(T_{n,b}\).
-In Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}, Remark 3.1) the convergence rate is changed from \((nb_1^3b_2^3)^{-1/2}\) to \((nb_1^2b_2^2)^{-1/2}\).
+In Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}, Remark 3.1) the convergence rate is changed from \((nb_1^3b_2^3)^{-1/2}\) to \((nb_1^2b_2^2)^{-1/2}\).
 These results are consistent with similar results for test functionals for independence in Joe (\protect\hyperlink{ref-joe1989estimation}{1989}).
 
-The variance in \eqref{eq:t4} can be evaluated in more detail, as has been done by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}), in particular using the mixing inequality and Hall and Heyde (\protect\hyperlink{ref-hall1980martingale}{1980}, Corollary A2).
+The variance in \eqref{eq:t4} can be evaluated in more detail, as has been done by Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}), in particular using the mixing inequality and Hall and Heyde (\protect\hyperlink{ref-hall1980martingale}{1980}, Corollary A2).
 But as shown by Teräsvirta, Tjøstheim, and Granger (\protect\hyperlink{ref-terasvirta2010modelling}{2010}) and in numerous other papers, the asymptotic variance in the distribution of such functionals as \(T_{n,b}\) cannot be expected to be accurate unless one has extremely large sample sizes.
@@
 \]
-where \(G_n(z) = \sqrt{n}(F_n(z)-F(z))\) and
-\begin{align*}
-\fV_n(z) &= \sqrt{n} (\frho_{n,b}(\z) - \frho_b(\z)) \\
+where \(G_n(\z) = \sqrt{n}(F_n(\z)-F(\z))\) and
+\begin{align*}
+\fV_n(\z) &= \sqrt{n} (\frho_{n,b}(\z) - \frho_b(\z)) \\
 & \sim \sqrt{n}{\J}_b^{-1}(\z)\left\{\frac{1}{n} \sum_{t=1}^{n}K_b(\Z_t-\z)\fu(\Z_t,\frho_b(\z)) - \int K_b(\fv-\z)\fu(\fv,\frho_b(\z))\psi(\fv,\rho_b(\z)) \di \fv \right\}.
@@
 It converges towards \(C_{H}(\fA_b)\) as depicted in the mixing expression of the proof of Theorem \ref{thm:F2}.
-See also Theorem 3.2 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) which is formulated without using independence.
+See also Theorem 3.2 of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) which is formulated without using independence.
 
@@
 \end{align*}
-Here, using Theorem \ref{thm:F2}, under \(H[a_n]\), and Proposition 6.3.9 of Brockwell and Davis (\protect\hyperlink{ref-brockwell1991time}{2006}),
+Analogously to Wang and Hong (\protect\hyperlink{ref-wang2017characteristic}{2018}), arguing as in the proof of Theorem \ref{thm:F2} and using Proposition 6.3.9 of Brockwell and Davis (\protect\hyperlink{ref-brockwell1991time}{2006}), as \(b_n, a_n \to 0\),
 \[
@@
 Here the pair of coordinates \((z_i,z_j)\) corresponds to the bivariate \(\x\) in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}, proof of Theorem 3.1).
-To ease notation and the comparison with Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}, \protect\hyperlink{ref-lacal2018estimating}{2018}), we again write \(\z_{ij} = (z_i,z_j)\).
+To ease notation and the comparison with Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}, \protect\hyperlink{ref-lacal2018estimating}{2019}), we again write \(\z_{ij} = (z_i,z_j)\).
 We will also omit some indices or parts of indices when it is clear what is meant.
@@
 where E\(^*(\cdot)\) is the mean value under the measure \(\mathbb{P}^*_{\omega}\), and where Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}, proof of Theorem 4.4) may be consulted for more details in the derivation and a more explicit analog expression for \(I_{ij}^*(\z_{ij},\rho_{ij,n})\) for the case considered there.
-In contradistinction to Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}), \(J_{ij,n,b}\) is not singular as \(\bb \to \bm{0}\).
-In fact, it can be shown that, cf.~Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, Theorem 2), as \(\bb \to \bm{0}\), \(J_{ij,n,b}(\z_{ij}) \stackrel{\mathbb{P}}{\to} u(\z_{ij},\rho_{ij})\psi(\z_{ij},\rho_{ij}(\z_{ij}))\).
+In contradistinction to Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}), \(J_{ij,n,b}\) is not singular as \(n\to \infty, \bb \to 0\).
+In fact, it can be shown that, cf.~Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, Theorem 2), as \(n \to \infty\) and \(\bb \to \bm{0}\), \(J_{ij,n,b}(\z_{ij}) \stackrel{\mathbb{P}}{\to} u(\z_{ij},\rho_{ij})\psi(\z_{ij},\rho_{ij}(\z_{ij}))\).
 
@@
 \end{equation}
-and we have thus reached an expression where we can apply the mixing CLT as in Gonçalves and White (\protect\hyperlink{ref-gonccalves2002bootstrap}{2002}) and an application of their result, quoted as Theorems C.1 and E.1 in the supplementary materials of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}) and Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) respectively, to the expressions in \eqref{eq:t22} and \eqref{eq:t33}.
+and we have thus reached an expression where we can apply the mixing CLT as in Gonçalves and White (\protect\hyperlink{ref-gonccalves2002bootstrap}{2002}) and an application of their result, quoted as Theorems C.1 and E.1 in the supplementary materials of Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}) and Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}) respectively, to the expressions in \eqref{eq:t22} and \eqref{eq:t33}.
 This yields asymptotic normality of \(\sqrt{n}(T_{n,b}^*-T_{b,n})\).
@@
 To be able to do that one must device a test statistic that is constructed with a probability structure as under the null hypothesis of conditional independence.
-It is difficult to obtain such a test based on asymptotic arguments for at least two reasons: As mentioned before the asymptotic distribution of test functionals such as \(T_{n,b}\) is known to be inaccurate except for extremely large sample sizes.
+It is difficult to obtain such a test based on asymptotic arguments: As mentioned before the asymptotic distribution of test functionals such as \(T_{n,b}\) is known to be inaccurate except for extremely large sample sizes.
 It is also difficult to build the probability null structure into the scaling factor of \(T_{n,b}\) since this would amount to building into that structure certain relationships between pairwise local correlations.
@@
 In the \(\{\Z_{1,t}^*,\Z_{2,t}^*,\Z_{3,t}^*\}\)-framework we obtain an estimate \(\hfrho^*\) of the local Gaussian correlation vector conditional on the bootstrap realizations.
-An asymptotic theory of this estimate can now be established using the method of proof in Theorem 2 in Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}), and subsequently an asymptotic theory for the test statistic \(T_n^*\) using the method of proof in Theorem 4.4 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}) and Theorem D.4 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}).
+An asymptotic theory of this estimate can now be established using the method of proof in Theorem 2 in Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}), and subsequently an asymptotic theory for the test statistic \(T_n^*\) using the method of proof in Theorem 4.4 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2017local}{2017}) and Theorem D.4 in Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2019}).
 See also the formulation of such proofs in the present paper, theorems \ref{thm:H1} and \ref{thm:H2}.
@@
 \leavevmode\hypertarget{ref-gonccalves2002bootstrap}{}%
-Gonçalves, Si@articlelacal2017local, title=Local Gaussian autocorrelation and tests for serial independence, author=Lacal, Virginia and TjØstheim, Dag, journal=Journal of Time Series Analysis, volume=38, number=1, pages=51--71, year=2017, publisher=Wiley Online Library lvia, and Halbert White.
-2002.
-``The Bootstrap of the Mean for Dependent Heterogeneous Arrays.'' \emph{Econometric Theory} 18 (6): 1367--84.
+Gonçalves, Silvia, and Halbert White. 2002. ``The Bootstrap of the Mean for Dependent Heterogeneous Arrays.'' \emph{Econometric Theory} 18 (6): 1367--84.
 
 \leavevmode\hypertarget{ref-gonccalves2004maximum}{}%
-Gonçalves, Silvia, and Halbert White. 2004. ``Maximum Likelihood and the Bootstrap for Nonlinear Dynamic Models.'' \emph{Journal of Econometrics} 119 (1): 199--219.
+---------. 2004. ``Maximum Likelihood and the Bootstrap for Nonlinear Dynamic Models.'' \emph{Journal of Econometrics} 119 (1): 199--219.
 

### 2020-09-16 lgpc_JBES.Rmd
A few minor details in the main paper by Håkon

@@
 @tjostheim2013local and Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional].
-Some key results follow below, and we refer to the supplementary material, Sections B--D for proofs of the asymptotic distributional results in this subsection, and to Section A.3 for a discussion of more general situations and for a result on consistency.
+Some key results follow below, and we refer to the supplementary material, Sections B--D, for proofs of the asymptotic distributional results in this subsection, and to Section A.3 for a discussion of more general situations and for a result on consistency.
 
@@
 \begin{equation}
-\sqrt{nb_n^2}\left(\hfrho(\z) - \frho_0(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \fLambda),
+\sqrt{nb^2}\left(\hfrho(\z) - \frho_0(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \fLambda),
 \label{eq:ot2018}
@@
 \begin{equation}
-\sqrt{nb_n^m}\left(\halpha(\z) - \alpha(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}\left(0, \nabla g(\frho)^T\fLambda\nabla g(\frho)\right),
+\sqrt{nb^m}\left(\halpha(\z) - \alpha(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}\left(0, \nabla g(\frho)^T\fLambda\nabla g(\frho)\right),
 (\#eq:limitlgpc)

### 2020-09-16 lgpc_JBES.tex
A few minor details in the main paper by Håkon

@@
 The asymptotic behavior of \(\halpha(\z)\) can thus be derived directly from the asymptotic behavior of \(\hR(\z)\), which, for the most part, has been established in earlier works, see e.g.~Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) and Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, \protect\hyperlink{ref-otneim2017conditional}{2018}).
-Some key results follow below, and we refer to the supplementary material, Sections B--D for proofs of the asymptotic distributional results in this subsection, and to Section A.3 for a discussion of more general situations and for a result on consistency.
+Some key results follow below, and we refer to the supplementary material, Sections B--D, for proofs of the asymptotic distributional results in this subsection, and to Section A.3 for a discussion of more general situations and for a result on consistency.
 
@@
 \begin{equation}
-\sqrt{nb_n^m}\left(\halpha(\z) - \alpha(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}\left(0, \nabla g(\frho)^T\fLambda\nabla g(\frho)\right),
+\sqrt{nb^m}\left(\halpha(\z) - \alpha(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}\left(0, \nabla g(\frho)^T\fLambda\nabla g(\frho)\right),
 \label{eq:limitlgpc}

### 2020-09-16 responses/response_ae.Rmd
Corrections and spell check of responses

@@
 
-> **Our response:** Thank you for reading our paper so carefully. We are very grateful for your constructive comments which has lead to a much improved paper.
+> **Our response:** Thank you for reading our paper so carefully. We are very grateful for your constructive comments which have lead to a much improved paper.
 
@@
 With this concept we have sought to make a nonlinear and local extension of the ordinary (global) partial correlation, whose use in measuring and describing dependence is well known for Gaussian distributions.
-We claim that it is fundamentally different both in concept and in use from the existing tests of conditional independence.
-It is hard to obtain a measure of conditional dependence that can be easily interpreted from these test as they would have to be based on distance measures that for example cannot distinguish between negative and positive conditional dependence.
+We claim that it is fundamentally different, both in concept and in use, from the existing tests of conditional independence.
+It is hard to obtain a measure of conditional dependence that can be easily interpreted from these tests as it would have to be based on distance measures that for example cannot distinguish between negative and positive conditional dependence.
 \vspace{.3cm} \newline 
@@
 and we have tried to make this more clear in the present version.
-Our test did do well in a competition with other test, and it may be possible to have it improved with better focusing and use of more local parameters, but, as perhaps over-stated, the main contribution, we believe, is the introduction and use of the new concept of LGPC in measuring the conditional dependence by considering a nonlinear local version of the partial correlation, a measure that reduces to the ordinary partial correlation for Gaussian variables.
+Our test did do well in a competition with other test, and it may be possible to have it improved with better focusing and use of more local parameters, but, at the risk of overstating this, the main contribution, we believe, is the introduction and use of the new concept of LGPC in measuring the conditional dependence by considering a nonlinear local version of the partial correlation, a measure that reduces to the ordinary partial correlation for Gaussian variables.
 \item [2.] \textbf{Estimation and testing in the full vector case:} 
@@
 \item [3.] \textbf{Asymptotic theory and bootstrap validity}. 
-This has been developed in Sections E, F and H in the Supplement, and details can be found in our response to item 4 in the report of referee 2.
+This has been developed in Sections E, F and H in the supplement, and details can be found in our response to item 4 in the report of referee 2.
 Actually, we are happy that referee 2 raised this point.
@@
 
-**AE further general comment:** Referee 1 enjoys reading the paper and have some comments that can be relatively easily addressed.
+**AE's further general comment:** Referee 1 enjoys reading the paper and have some comments that can be relatively easily addressed.
 
@@
 
-**Comment 2:** As both referees mention (see R1's point 1 and R2's point 3), one limitation of the proposed testing procedure is that it only tests the conditional independence between two scalar variables $(X_1,X_2)$ given a random vector $(X_1,\ldots,X_p)$.
+**Comment 2:** As both referees mention (see R1's point 1 and R2's point 3), one limitation of the proposed testing procedure is that it only tests the conditional independence between two scalar variables $(X_1,X_2)$ given a random vector $(X_3,\ldots,X_p)$.
 It is important to generalize the test to allow
@@
 We now introduce and define the LGPC (matrix) for a general ${\bf X}_1$, ${\bf X}_2$ and ${\bf X}_3$ of dimension $d_1$, $d_2$ and $d_3$ in Section 2.1.
-The full multivariate framework is analyzed in more detail in Section A: "The multivariate LPGC" in the Supplement.
-Various special cases are discussed in Section A.2 of that Supplement.
-There are really no new concepts involved, and large parts of the ensuing estimation and testing theory in the Supplement are carried out in this framework.
+The full multivariate framework is analyzed in more detail in Section A: "The multivariate LPGC" in the supplement.
+Various special cases are discussed in Section A.2 of that supplement.
+There are really no new concepts involved, and large parts of the ensuing estimation and testing theory in the supplement are carried out in this framework.
 \vspace{.3cm} \newline 
-In particular the asymptotic theory for a conditional independence test where both ${\bf X}_1$ and ${\bf X_2}$ (and ${\bf X}_3$) are vectors is given in Section F of the Supplementary material (Theorem F.1 and Theorem F.2).
-The asymptotic local power is discussed in Section G, and the validity of the bootstrap for the testing statistic are given at this level of generality (Theorems G.1, H.1 and H.2).
+In particular the asymptotic theory for a conditional independence test where both ${\bf X}_1$ and ${\bf X_2}$ (and ${\bf X}_3$) are vectors is given in Section F of the supplementary material (Theorem F.1 and Theorem F.2).
+The asymptotic local power is discussed in Section G, and the validity of the bootstrap for the testing statistic are given at this level of generality (Theorems H.1 and H.2).
 However, the notation becomes more complex and expressions less explicit, and for these reasons and for reasons of space much of the material in the main article is stated for the case of a scalar $X_1$ and $X_2$.
 There is a new example of Granger causality in Section 4, though, where the assumption of a scalar $X_1$ and $X_2$ is dropped.
-The curse of dimensionality can be avoided at the cost of a simplified pairwise model introduced described and used in the main article and in the supplement.
-Throughout the main paper the reader is referred to appropriate sections of the Supplement for the general vector case.
+The curse of dimensionality can be avoided at the cost of a simplified pairwise model introduced, described and used in the main article and in the supplement.
+Throughout the main paper the reader is referred to appropriate sections of the supplement for the general vector case.
 
@@
 
-**Comment 4.** p.10: Since you claim that ${\bf b}$ is a diagonal matrix of bandwidths (not squared bandwidths). It seems that you want to define $K_{\bf b} = |{\bf b}|^{-1}k({\bf b}^{-1}{\bf x}$.
+**Comment 4.** p.10: Since you claim that ${\bf b}$ is a diagonal matrix of bandwidths (not squared bandwidths). It seems that you want to define $K_{\bf b} = |{\bf b}|^{-1}K({\bf b}^{-1}{\bf x})$.
 
-> **Our response:** Yes, corrections have been implemented. Thank you
+> **Our response:** Yes, corrections have been implemented. Thank you.
 
@@
 (20).
-It has now been corrected.) This is explained in the proof of Theorem 3.1 and in more detail in Section 4 of Tj{\o}stheim and Hufthammer (2013)  in  a slightly different situation.
+It has now been corrected.) This is explained in the proof of Theorem 3.1 and in more detail in Section 4 of Tjøstheim and Hufthammer (2013)  in  a slightly different situation.
 Similar results and with  rather different derivations can also be found in Jones (1996, 1998).
@@
 \vspace{.3cm} \newline It is certainly possible to estimate the local conditional correlation (and in the general vector case) using local likelihood for local correlations and then transforming to the LGPC.
-In the general vector Case this is explained in Sections A.2 and A.3 in the Supplement.
+In the general vector case this is explained in Sections A.2 and A.3 in the supplement.
 \item[(b)] This has been corrected. Thank you.
@@
 
-> **Our response:** There is a data-driven choice for bandwidth in a density estimation context in Otneim and Tj{\o}stheim (2017).
+> **Our response:** There is a data-driven choice for bandwidth in a density estimation context in Otneim and Tjøstheim (2017).
 It is based on leave-one-out cross-validation.
@@
 It remains to establish such a procedure for the present situation.
-In the paper we have typically used the asymptotic rates $b=cn^{-1/9}$ for the full trivariate fit and $b=cn^{-1/6}$ for the bivariate simplification.
-In the testing results (Table 2, $2^*$ and $3^*$ we have used $c=1.0$ and $c=1.4$ to examine sensitivity to bandwidth selection.
+In the paper we have typically used the asymptotic rates $b=cn^{-1/9}$ (with $c$ empirically determined to 1.75 in the density estimation context) for the full trivariate fit and $b=cn^{-1/6}$ for the bivariate simplification.
+In the testing results (Table 2, $2^*$ and $3^*$) we have used $c=1.0$ and $c=1.4$ to examine sensitivity to bandwidth selection.
 They gave similar results.
-Note also that for some of the other competing test several choices of tuning parameters were used.
+Note also that for some of the other competing tests several choices of tuning parameters were used.
 For the plots of the estimated LGPC in Section 4 it is true that we used more smoothing to obtain cleaner plots.
-Thus $c=4$ for Figure 1, but a smaller bandwidth are used in testing and as mentioned in the description of that example the hypothesis of conditional independence is not rejected for any reasonable significance level.
+Thus $c=4$ for Figure 1, but a smaller bandwidth is used in testing and as mentioned in the description of that example the hypothesis of conditional independence is not rejected for any reasonable significance level.
 
@@
 
-> **Our response:** The asymptotic properties of the test statistic has been formally studied in Section F of the Supplement.
-The validity of the bootstrap for the test statistic is established in Section H of the Supplement (for estimation in Section E of the Supplement.
+> **Our response:** The asymptotic properties of the test statistic has been formally studied in Section F of the supplement.
+The validity of the bootstrap for the test statistic is established in Section H of the supplement (for estimation in Section E of the supplement.
 The convergence rate of the test statistic is given in Section 5.2 in the main article.
@@
 > **Our response:** We have corrected the typos kindly pointed out to us by Referee 1.
-We have read carefully through the revised version, including the Supplement in order to eliminate as many typos as possible.
+We have read carefully through the revised version, including the supplement in order to eliminate as many typos as possible.
 
 ## References
+
+${ }$
 
@@
 
-Tjæstheim, D. and Hufthammer, K.O. (2013) Local Gaussian correlation: A new measure of dependence. Journal of Econometrics, **172**, 33-48.
+Tjøstheim, D. and Hufthammer, K.O. (2013) Local Gaussian correlation: A new measure of dependence. Journal of Econometrics, **172**, 33-48.
 

### 2020-09-16 responses/response_ae.tex
Corrections and spell check of responses

@@
 \begin{quote}
-\textbf{Our response:} Thank you for reading our paper so carefully. We are very grateful for your constructive comments which has lead to a much improved paper.
+\textbf{Our response:} Thank you for reading our paper so carefully. We are very grateful for your constructive comments which have lead to a much improved paper.
 \end{quote}
@@
 
-\textbf{AE further general comment:} Referee 1 enjoys reading the paper and have some comments that can be relatively easily addressed.
+\textbf{AE's further general comment:} Referee 1 enjoys reading the paper and have some comments that can be relatively easily addressed.
 
@@
 
-\textbf{Comment 2:} As both referees mention (see R1's point 1 and R2's point 3), one limitation of the proposed testing procedure is that it only tests the conditional independence between two scalar variables \((X_1,X_2)\) given a random vector \((X_1,\ldots,X_p)\).
+\textbf{Comment 2:} As both referees mention (see R1's point 1 and R2's point 3), one limitation of the proposed testing procedure is that it only tests the conditional independence between two scalar variables \((X_1,X_2)\) given a random vector \((X_3,\ldots,X_p)\).
 It is important to generalize the test to allow
@@
 We now introduce and define the LGPC (matrix) for a general \({\bf X}_1\), \({\bf X}_2\) and \({\bf X}_3\) of dimension \(d_1\), \(d_2\) and \(d_3\) in Section 2.1.
-The full multivariate framework is analyzed in more detail in Section A: ``The multivariate LPGC'' in the Supplement.
-Various special cases are discussed in Section A.2 of that Supplement.
-There are really no new concepts involved, and large parts of the ensuing estimation and testing theory in the Supplement are carried out in this framework.
+The full multivariate framework is analyzed in more detail in Section A: ``The multivariate LPGC'' in the supplement.
+Various special cases are discussed in Section A.2 of that supplement.
+There are really no new concepts involved, and large parts of the ensuing estimation and testing theory in the supplement are carried out in this framework.
 \vspace{.3cm} \newline 
-In particular the asymptotic theory for a conditional independence test where both \({\bf X}_1\) and \({\bf X_2}\) (and \({\bf X}_3\)) are vectors is given in Section F of the Supplementary material (Theorem F.1 and Theorem F.2).
-The asymptotic local power is discussed in Section G, and the validity of the bootstrap for the testing statistic are given at this level of generality (Theorems G.1, H.1 and H.2).
+In particular the asymptotic theory for a conditional independence test where both \({\bf X}_1\) and \({\bf X_2}\) (and \({\bf X}_3\)) are vectors is given in Section F of the supplementary material (Theorem F.1 and Theorem F.2).
+The asymptotic local power is discussed in Section G, and the validity of the bootstrap for the testing statistic are given at this level of generality (Theorems H.1 and H.2).
 However, the notation becomes more complex and expressions less explicit, and for these reasons and for reasons of space much of the material in the main article is stated for the case of a scalar \(X_1\) and \(X_2\).
 There is a new example of Granger causality in Section 4, though, where the assumption of a scalar \(X_1\) and \(X_2\) is dropped.
-The curse of dimensionality can be avoided at the cost of a simplified pairwise model introduced described and used in the main article and in the supplement.
-Throughout the main paper the reader is referred to appropriate sections of the Supplement for the general vector case.
+The curse of dimensionality can be avoided at the cost of a simplified pairwise model introduced, described and used in the main article and in the supplement.
+Throughout the main paper the reader is referred to appropriate sections of the supplement for the general vector case.
 \end{quote}
@@
 \textbf{Comment 4.} p.10: Since you claim that \({\bf b}\) is a diagonal matrix of bandwidths (not squared bandwidths).
-It seems that you want to define \(K_{\bf b} = |{\bf b}|^{-1}k({\bf b}^{-1}{\bf x}\).
+It seems that you want to define \(K_{\bf b} = |{\bf b}|^{-1}K({\bf b}^{-1}{\bf x})\).
 
 \begin{quote}
-\textbf{Our response:} Yes, corrections have been implemented. Thank you
+\textbf{Our response:} Yes, corrections have been implemented. Thank you.
 \end{quote}
@@
 \begin{quote}
-\textbf{Our response:} There is a data-driven choice for bandwidth in a density estimation context in Otneim and Tj\{\o\}stheim (2017).
+\textbf{Our response:} There is a data-driven choice for bandwidth in a density estimation context in Otneim and Tjøstheim (2017).
 It is based on leave-one-out cross-validation.
@@
 It remains to establish such a procedure for the present situation.
-In the paper we have typically used the asymptotic rates \(b=cn^{-1/9}\) for the full trivariate fit and \(b=cn^{-1/6}\) for the bivariate simplification.
-In the testing results (Table 2, \(2^*\) and \(3^*\) we have used \(c=1.0\) and \(c=1.4\) to examine sensitivity to bandwidth selection.
+In the paper we have typically used the asymptotic rates \(b=cn^{-1/9}\) (with \(c\) empirically determined to 1.75 in the density estimation context) for the full trivariate fit and \(b=cn^{-1/6}\) for the bivariate simplification.
+In the testing results (Table 2, \(2^*\) and \(3^*\)) we have used \(c=1.0\) and \(c=1.4\) to examine sensitivity to bandwidth selection.
 They gave similar results.
-Note also that for some of the other competing test several choices of tuning parameters were used.
+Note also that for some of the other competing tests several choices of tuning parameters were used.
 For the plots of the estimated LGPC in Section 4 it is true that we used more smoothing to obtain cleaner plots.
-Thus \(c=4\) for Figure 1, but a smaller bandwidth are used in testing and as mentioned in the description of that example the hypothesis of conditional independence is not rejected for any reasonable significance level.
+Thus \(c=4\) for Figure 1, but a smaller bandwidth is used in testing and as mentioned in the description of that example the hypothesis of conditional independence is not rejected for any reasonable significance level.
 \end{quote}
@@
 \begin{quote}
-\textbf{Our response:} The asymptotic properties of the test statistic has been formally studied in Section F of the Supplement.
-The validity of the bootstrap for the test statistic is established in Section H of the Supplement (for estimation in Section E of the Supplement.
+\textbf{Our response:} The asymptotic properties of the test statistic has been formally studied in Section F of the supplement.
+The validity of the bootstrap for the test statistic is established in Section H of the supplement (for estimation in Section E of the supplement.
 The convergence rate of the test statistic is given in Section 5.2 in the main article.
@@
 \textbf{Our response:} We have corrected the typos kindly pointed out to us by Referee 1.
-We have read carefully through the revised version, including the Supplement in order to eliminate as many typos as possible.
+We have read carefully through the revised version, including the supplement in order to eliminate as many typos as possible.
 \end{quote}
@@
 \subsection{References}\label{references}}
+
+\({ }\)
 
@@
 
-Tjæstheim, D. and Hufthammer, K.O. (2013) Local Gaussian correlation: A new measure of dependence. Journal of Econometrics, \textbf{172}, 33-48.
+Tjøstheim, D. and Hufthammer, K.O. (2013) Local Gaussian correlation: A new measure of dependence. Journal of Econometrics, \textbf{172}, 33-48.
 

### 2020-09-16 responses/response_editor.Rmd
Corrections and spell check of responses

@@
 geometry: margin = 1in
-bibliography: "bibliography.bib"
 biblio-style: "apalike"
@@
 
-\noindent \textbf{Letter to the editor.}
+\noindent \textbf{Letter to the Editor.}
+
+\vspace{1cm}
+
+\noindent Professor Jianqing Fan
+
+\noindent Editor JBES
 
@@
 We grateful for the opportunity to revise.
-Enclosed please find our revised version, including supplementary material to be posted online, as well as individual responses to the Associate and two referees.
+Enclosed please find our revised version, including supplementary material to be posted online, as well as individual responses to the Associate Editor and two referees.
 
@@
 An example concerning Granger causality testing where $\X_1$ and $\X_2$ are not scalars has been added.
-We have added asymptotic theory for the test statistic, theory on the validity of the bootstrap for the test statistic and for the estimation of LGPC.
+We have added asymptotic theory for the test statistic, theory on the validity of the bootstrap for the test statistic, and for the estimation of LGPC.
 We have also derived local power and compared with other test statistics.
 
-For reasons of space most of these changes have been included in a Supplement intended to be available online.
+For reasons of space most of these changes have been included in a supplement intended to be available online.
 This has made it possible for us to keep the length of the main article at 35 pages, as advised by the Editor.
 
-In addition there are numerous other changes directly related to the comments of the referees and the AE.
+In addition there are numerous other changes directly related to the comments of the referees and the Associate Editor.
 In particular,  we have tried to make it more clear that our paper is not just concerned with presenting just another test of conditional independence.
-But rather it is concerned with the introduction of a new framework for measuring conditional dependence, and that this framework reduces to the ordinary partial correlation coefficient in the Gaussian case.
+But rather it is concerned with the introduction of a new framework for measuring conditional dependence, where this framework reduces to the ordinary partial correlation coefficient in the Gaussian case.
 

### 2020-09-16 responses/response_editor.tex
Corrections and spell check of responses

@@
 An example concerning Granger causality testing where \(\X_1\) and \(\X_2\) are not scalars has been added.
-We have added asymptotic theory for the test statistic, theory on the validity of the bootstrap for the test statistic and for the estimation of LGPC.
+We have added asymptotic theory for the test statistic, theory on the validity of the bootstrap for the test statistic, and for the estimation of LGPC.
 We have also derived local power and compared with other test statistics.
 
-For reasons of space most of these changes have been included in a Supplement intended to be available online.
+For reasons of space most of these changes have been included in a supplement intended to be available online.
 This has made it possible for us to keep the length of the main article at 35 pages, as advised by the Editor.
 
-In addition there are numerous other changes directly related to the comments of the referees and the AE.
+In addition there are numerous other changes directly related to the comments of the referees and the Associate Editor.
 In particular, we have tried to make it more clear that our paper is not just concerned with presenting just another test of conditional independence.
-But rather it is concerned with the introduction of a new framework for measuring conditional dependence, and that this framework reduces to the ordinary partial correlation coefficient in the Gaussian case.
+But rather it is concerned with the introduction of a new framework for measuring conditional dependence, where this framework reduces to the ordinary partial correlation coefficient in the Gaussian case.
 

### 2020-09-16 responses/response_reviewers.Rmd
Corrections and spell check of responses

@@
 We now introduce and define the LGPC (matrix) for a general ${\bf X}_1$, ${\bf X}_2$ and ${\bf X}_3$ of dimension $d_1$, $d_2$ and $d_3$ in Section 2.1.
-The full multivariate framework is analyzed in more detail in section A:''The multivariate LPGC'' in the supplement.
+The full multivariate framework is analyzed in more detail in section A: ''The multivariate LPGC'' in the supplement.
 Various special cases are discussed in Section A.2 of that supplement.
-There are really no new concepts involved, and large parts of the ensuing estimation and testing theory in the Supplement are carried out in this framework.
-In particular, the asymptotic theory for a conditional independence test where bath ${\bf X}_1$ and ${\bf X}_2$ (and ${\bf X}_3$ are vectors, is given in Section F of the Supplement (Theorem F.1 and Theorem F.2).
-The asymptotic power is discussed in Section G, and the validity of the bootstrap for the testing statistic are given at this level of generality (Theorems G.1, H.1 and H:2).
+There are really no new concepts involved, and large parts of the ensuing estimation and testing theory in the supplement are carried out in this framework.
+In particular, the asymptotic theory for a conditional independence test where both ${\bf X}_1$ and ${\bf X}_2$ (and ${\bf X}_3$ are vectors, is given in Section F of the supplement (Theorem F.1 and Theorem F.2).
+The asymptotic power is discussed in Section G, and the validity of the bootstrap for the testing statistic are given at this level of generality (H.1 and H.2).
 However, the notation becomes more complex and expressions less explicit, and for these reasons and for reasons of space much of the material in the main article is stated for the case of a scalar $X_1$ and $X_2$.
 There is a new example of Granger causality in Section 4, though, where the assumption of a scalar $X_1$ and $X_2$ is dropped.
-The curse of dimensionality can be avoided at the cost of a simplified pairwise model introduced, described and used in the main article and in the Supplement.
-Throughout the main paper we refer to the appropriate sections of the Supplement for the general vector case.
+The curse of dimensionality can be avoided at the cost of a simplified pairwise model introduced, described and used in the main article and in the supplement.
+Throughout the main paper we refer to the appropriate sections of the supplement for the general vector case.
 
-**Comment 2:** In Equation (6), isn't it an approximation of the density $f_Z$ of ${\bf Z}$ in the neighbourhood of ${\bf z}$?
+**Comment 2:** In Equation (6), isn't it an approximation of the density $f_Z$ of ${\bf Z}$ in the neighborhood of ${\bf z}$?
 The way it writes now seems to indicate that the expression is exactly the density $f_{\bf Z}$ of ${\bf Z}$, but $f_{\bf Z}({\bf z})$ can certainly differ from $\psi({\bf z}, R({\bf z}))$ in the general case.
@@
 
-**Comment 5:** In Equation (15), change $\hat{\bf z}$ to ${\bf z}$. And isn't the definition of kernel function $K_b({\bf x} = |{\bf b}|^{-1}K({\bf b}^{-1}{\bf x})$?
+**Comment 5:** In Equation (15), change $\hat{\bf z}$ to ${\bf z}$. And isn't the definition of kernel function $K_b({\bf x}) = |{\bf b}|^{-1}K({\bf b}^{-1}{\bf x})$?
 
@@
 > **Our response:** Thanks a lot for spotting this error in the sequencing of those two matrices.
-(I am afraid that this mistake has gone undetected in some earlier publications).
+(We are afraid that this mistake has gone undetected in some earlier publications).
 The orders of ${\bf M}_b$ and ${\bf J}_b$ are $O(b^2)$.
-This has now been stated with a reference to the Supplement for their derivation.
+This has now been stated with a reference to the supplement for their derivation.
 We have also corrected a typo in eq.
@@
 
-**Comment 12:* In the proof of Theorem 3.2, to show that
+**Comment 12:** In the proof of Theorem 3.2, to show that
 $$
@@
 > **Our response:** These differences have been looked at in two papers.
-For the estimation of the joint distributions we refer to Otneim and Tj{\o}stheim (2017), and for the conditional distribution to Otneim and Tj{\o}stheim (2018).
+For the estimation of the joint distributions we refer to Otneim and Tjøstheim (2017), and for the conditional distribution to Otneim and Tjøstheim (2018).
 The authors of these two papers show that the local Gaussian approach is much less vulnerable to the curse of dimensionality using a pairwise local Gaussian approach.
-However, both the kernel and local Gaussian approach to estimating conditional distributions are ill-fitted to actually *measuring* the conditional dependence numerically, for the same reason that the joint distribution is not a well fitted tool to actually measuring the dependence between two or more stochastic variables numerically.
+However, both the kernel and local Gaussian approach to estimating conditional distributions are ill-fitted to actually *measuring* the conditional dependence numerically, for the same reason that the joint distribution is not a well fitted tool to actually measuring numerically the dependence between two or more stochastic variables.
 
@@
 > **Our response:** Asymptotically it is negligible as demonstrated by Theorem 3.2 and its proof.
-We can examine finite sample effects through our simulation experiments where the truth is known.
+We can examine finite sample effects through our simulation experiments when the truth is known.
 This is the case for the estimation of the level (5\% is correct) for the DGPs created under the conditional independence of the null hypothesis.
-The results from these experiments are shown in Table 2 in the main article and in Tables $2^* - 4^*$ in the Supplement, Section K.
+The results from these experiments are shown in Table 2 in the main article and in Tables $2^* - 4^*$ in the supplement, Section K.
 These experiments were carried out for sample sizes of 100, 200 and 400, and even for a sample size as small as 100, results close to the correct ones were obtained.
@@
 We now introduce and define the LGPC (matrix) for a general ${\bf X}_1$, ${\bf X}_2$ and ${\bf X}_3$ of dimension $d_1$, $d_2$ and $d_3$ in Section 2.1.
-The full multivariate framework is analyzed in more detail in Section A: ``The multivariate LPGC'' of the Supplement.
+The full multivariate framework is analyzed in more detail in Section A: ``The multivariate LPGC'' of the supplement.
 Various special cases are discussed in Section A.2 of that supplement.
-There are really no new concepts involved, and large parts of the ensuing estimation and testing theory in the Supplement are carried out in this framework.
-In particular, the asymptotic theory for a conditional independence test where both ${\bf X}_1$ and ${\bf X}_2$ (and ${\bf X}_3$) are vectors, is given in Section F of the Supplement (Theorem F.1 and Theorem F.2).
-The local asymptotic power is discussed in Section G, and the validity of the bootstrap for the testing statistic are given at this level of generality (Theorems G.1, H.1 and H.2).
+There are really no new concepts involved, and large parts of the ensuing estimation and testing theory in the supplement are carried out in this framework.
+In particular, the asymptotic theory for a conditional independence test where both ${\bf X}_1$ and ${\bf X}_2$ (and ${\bf X}_3$) are vectors, is given in Section F of the supplement (Theorem F.1 and Theorem F.2).
+The local asymptotic power is discussed in Section G, and the validity of the bootstrap for the testing statistic are given at this level of generality (Theorems H.1 and H.2).
 However, the notation becomes more complex and expressions less explicit, and for these reasons and for reasons of space much of the material in the main article is stated for a scalar $X_1$ and $X_2$.
 There is a new example of Granger causality in Section 4, though, where the assumption of a scalar $X_1$ and $X_2$ is dropped.
-The curse of dimensionality can be avoided at the cost of a simplified pairwise model introduced, described and used in the main article and in the Supplement.
-Throughout the paper the reader is referred to the appropriate sections of the Supplement for the general vector case.
+The curse of dimensionality can be avoided at the cost of a simplified pairwise model introduced, described and used in the main article and in the supplement.
+Throughout the paper the reader is referred to the appropriate sections of the supplement for the general vector case.
 
@@
 > **Our response:** We  have followed up your suggestion which we think has resulted in a great improvement of the paper.
-For reasons of space we have had to put these developments in the Supplement, mostly for a general ${\bf X} = \{{\bf X}_1,{\bf X}_2, {\bf X}_3\}$, but with appropriate references in the main article to where they can be found.
-The validity of the bootstrap for estimation can be found in Section E of that Supplement, the asymptotic theory for the test statistic in Section F, the bootstrap validity for the test statistic in Section G.
+For reasons of space we have had to put these developments in the supplement, for a general ${\bf X} = \{{\bf X}_1,{\bf X}_2, {\bf X}_3\}$, but with appropriate references in the main article to where they can be found.
+The validity of the bootstrap for estimation can be found in Section E of that supplement, the asymptotic theory for the test statistic in Section F, the bootstrap validity for the test statistic in Section H.
 
@@
 > **Our response:** Obviously this is another important point.
-As a consequence we have included a discussion of this in Section 5.2 in the main article and in Section G of the Supplement on local asymptotic power and the Pitman criterion.
+As a consequence we have included a discussion of this in Section 5.2 in the main article and in Section G of the supplement on local asymptotic power and the Pitman criterion.
 
@@
 It does well (best, see Table 2 of main article) on DGP 5 which is linear Gaussian, but it does miserably on the other DGPs, which have a non-Gaussian distribution.
-Our test does slightly worse than the Granger test on DGP 5, but it does well on the other DGPs which has a non-Gaussian joint distribution.
-We think that the competition with Sun and White is fair, when we have chosen *their* examples to compete against.
+Our test does slightly worse than the Granger test on DGP 5, but it does well on the other DGPs which have a non-Gaussian joint distribution.
+We think that the competition with Su and White is fair, when we have chosen *their* examples to compete against.
 
 ## References
+
+${ }$
 

### 2020-09-16 responses/response_reviewers.tex
Corrections and spell check of responses

@@
 We now introduce and define the LGPC (matrix) for a general \({\bf X}_1\), \({\bf X}_2\) and \({\bf X}_3\) of dimension \(d_1\), \(d_2\) and \(d_3\) in Section 2.1.
-The full multivariate framework is analyzed in more detail in section A:`'The multivariate LPGC'' in the supplement.
+The full multivariate framework is analyzed in more detail in section A: `'The multivariate LPGC'' in the supplement.
 Various special cases are discussed in Section A.2 of that supplement.
-There are really no new concepts involved, and large parts of the ensuing estimation and testing theory in the Supplement are carried out in this framework.
-In particular, the asymptotic theory for a conditional independence test where bath \({\bf X}_1\) and \({\bf X}_2\) (and \({\bf X}_3\) are vectors, is given in Section F of the Supplement (Theorem F.1 and Theorem F.2).
-The asymptotic power is discussed in Section G, and the validity of the bootstrap for the testing statistic are given at this level of generality (Theorems G.1, H.1 and H:2).
+There are really no new concepts involved, and large parts of the ensuing estimation and testing theory in the supplement are carried out in this framework.
+In particular, the asymptotic theory for a conditional independence test where both \({\bf X}_1\) and \({\bf X}_2\) (and \({\bf X}_3\) are vectors, is given in Section F of the supplement (Theorem F.1 and Theorem F.2).
+The asymptotic power is discussed in Section G, and the validity of the bootstrap for the testing statistic are given at this level of generality (H.1 and H.2).
 However, the notation becomes more complex and expressions less explicit, and for these reasons and for reasons of space much of the material in the main article is stated for the case of a scalar \(X_1\) and \(X_2\).
 There is a new example of Granger causality in Section 4, though, where the assumption of a scalar \(X_1\) and \(X_2\) is dropped.
-The curse of dimensionality can be avoided at the cost of a simplified pairwise model introduced, described and used in the main article and in the Supplement.
-Throughout the main paper we refer to the appropriate sections of the Supplement for the general vector case.
+The curse of dimensionality can be avoided at the cost of a simplified pairwise model introduced, described and used in the main article and in the supplement.
+Throughout the main paper we refer to the appropriate sections of the supplement for the general vector case.
 \end{quote}
 
-\textbf{Comment 2:} In Equation (6), isn't it an approximation of the density \(f_Z\) of \({\bf Z}\) in the neighbourhood of \({\bf z}\)?
+\textbf{Comment 2:} In Equation (6), isn't it an approximation of the density \(f_Z\) of \({\bf Z}\) in the neighborhood of \({\bf z}\)?
 The way it writes now seems to indicate that the expression is exactly the density \(f_{\bf Z}\) of \({\bf Z}\), but \(f_{\bf Z}({\bf z})\) can certainly differ from \(\psi({\bf z}, R({\bf z}))\) in the general case.
@@
 
-\textbf{Comment 5:} In Equation (15), change \(\hat{\bf z}\) to \({\bf z}\). And isn't the definition of kernel function \(K_b({\bf x} = |{\bf b}|^{-1}K({\bf b}^{-1}{\bf x})\)?
+\textbf{Comment 5:} In Equation (15), change \(\hat{\bf z}\) to \({\bf z}\). And isn't the definition of kernel function \(K_b({\bf x}) = |{\bf b}|^{-1}K({\bf b}^{-1}{\bf x})\)?
 
@@
 \textbf{Our response:} Thanks a lot for spotting this error in the sequencing of those two matrices.
-(I am afraid that this mistake has gone undetected in some earlier publications).
+(We are afraid that this mistake has gone undetected in some earlier publications).
 The orders of \({\bf M}_b\) and \({\bf J}_b\) are \(O(b^2)\).
-This has now been stated with a reference to the Supplement for their derivation.
+This has now been stated with a reference to the supplement for their derivation.
 We have also corrected a typo in eq.
@@
 
-**Comment 12:* In the proof of Theorem 3.2, to show that
+\textbf{Comment 12:} In the proof of Theorem 3.2, to show that
 \[
@@
 \textbf{Our response:} These differences have been looked at in two papers.
-For the estimation of the joint distributions we refer to Otneim and Tj\{\o\}stheim (2017), and for the conditional distribution to Otneim and Tj\{\o\}stheim (2018).
+For the estimation of the joint distributions we refer to Otneim and Tjøstheim (2017), and for the conditional distribution to Otneim and Tjøstheim (2018).
 The authors of these two papers show that the local Gaussian approach is much less vulnerable to the curse of dimensionality using a pairwise local Gaussian approach.
-However, both the kernel and local Gaussian approach to estimating conditional distributions are ill-fitted to actually \emph{measuring} the conditional dependence numerically, for the same reason that the joint distribution is not a well fitted tool to actually measuring the dependence between two or more stochastic variables numerically.
+However, both the kernel and local Gaussian approach to estimating conditional distributions are ill-fitted to actually \emph{measuring} the conditional dependence numerically, for the same reason that the joint distribution is not a well fitted tool to actually measuring numerically the dependence between two or more stochastic variables.
 \end{quote}
@@
 \textbf{Our response:} Asymptotically it is negligible as demonstrated by Theorem 3.2 and its proof.
-We can examine finite sample effects through our simulation experiments where the truth is known.
+We can examine finite sample effects through our simulation experiments when the truth is known.
 This is the case for the estimation of the level (5\% is correct) for the DGPs created under the conditional independence of the null hypothesis.
-The results from these experiments are shown in Table 2 in the main article and in Tables \(2^* - 4^*\) in the Supplement, Section K.
+The results from these experiments are shown in Table 2 in the main article and in Tables \(2^* - 4^*\) in the supplement, Section K.
 These experiments were carried out for sample sizes of 100, 200 and 400, and even for a sample size as small as 100, results close to the correct ones were obtained.
@@
 We now introduce and define the LGPC (matrix) for a general \({\bf X}_1\), \({\bf X}_2\) and \({\bf X}_3\) of dimension \(d_1\), \(d_2\) and \(d_3\) in Section 2.1.
-The full multivariate framework is analyzed in more detail in Section A: ``The multivariate LPGC'\,' of the Supplement.
+The full multivariate framework is analyzed in more detail in Section A: ``The multivariate LPGC'\,' of the supplement.
 Various special cases are discussed in Section A.2 of that supplement.
-There are really no new concepts involved, and large parts of the ensuing estimation and testing theory in the Supplement are carried out in this framework.
-In particular, the asymptotic theory for a conditional independence test where both \({\bf X}_1\) and \({\bf X}_2\) (and \({\bf X}_3\)) are vectors, is given in Section F of the Supplement (Theorem F.1 and Theorem F.2).
-The local asymptotic power is discussed in Section G, and the validity of the bootstrap for the testing statistic are given at this level of generality (Theorems G.1, H.1 and H.2).
+There are really no new concepts involved, and large parts of the ensuing estimation and testing theory in the supplement are carried out in this framework.
+In particular, the asymptotic theory for a conditional independence test where both \({\bf X}_1\) and \({\bf X}_2\) (and \({\bf X}_3\)) are vectors, is given in Section F of the supplement (Theorem F.1 and Theorem F.2).
+The local asymptotic power is discussed in Section G, and the validity of the bootstrap for the testing statistic are given at this level of generality (Theorems H.1 and H.2).
 However, the notation becomes more complex and expressions less explicit, and for these reasons and for reasons of space much of the material in the main article is stated for a scalar \(X_1\) and \(X_2\).
 There is a new example of Granger causality in Section 4, though, where the assumption of a scalar \(X_1\) and \(X_2\) is dropped.
-The curse of dimensionality can be avoided at the cost of a simplified pairwise model introduced, described and used in the main article and in the Supplement.
-Throughout the paper the reader is referred to the appropriate sections of the Supplement for the general vector case.
+The curse of dimensionality can be avoided at the cost of a simplified pairwise model introduced, described and used in the main article and in the supplement.
+Throughout the paper the reader is referred to the appropriate sections of the supplement for the general vector case.
 \end{quote}
@@
 \textbf{Our response:} We have followed up your suggestion which we think has resulted in a great improvement of the paper.
-For reasons of space we have had to put these developments in the Supplement, mostly for a general \({\bf X} = \{{\bf X}_1,{\bf X}_2, {\bf X}_3\}\), but with appropriate references in the main article to where they can be found.
-The validity of the bootstrap for estimation can be found in Section E of that Supplement, the asymptotic theory for the test statistic in Section F, the bootstrap validity for the test statistic in Section G.
+For reasons of space we have had to put these developments in the supplement, for a general \({\bf X} = \{{\bf X}_1,{\bf X}_2, {\bf X}_3\}\), but with appropriate references in the main article to where they can be found.
+The validity of the bootstrap for estimation can be found in Section E of that supplement, the asymptotic theory for the test statistic in Section F, the bootstrap validity for the test statistic in Section H.
 \end{quote}
@@
 \textbf{Our response:} Obviously this is another important point.
-As a consequence we have included a discussion of this in Section 5.2 in the main article and in Section G of the Supplement on local asymptotic power and the Pitman criterion.
+As a consequence we have included a discussion of this in Section 5.2 in the main article and in Section G of the supplement on local asymptotic power and the Pitman criterion.
 \end{quote}
@@
 It does well (best, see Table 2 of main article) on DGP 5 which is linear Gaussian, but it does miserably on the other DGPs, which have a non-Gaussian distribution.
-Our test does slightly worse than the Granger test on DGP 5, but it does well on the other DGPs which has a non-Gaussian joint distribution.
-We think that the competition with Sun and White is fair, when we have chosen \emph{their} examples to compete against.
+Our test does slightly worse than the Granger test on DGP 5, but it does well on the other DGPs which have a non-Gaussian joint distribution.
+We think that the competition with Su and White is fair, when we have chosen \emph{their} examples to compete against.
 \end{quote}

### 2020-09-18 lgpc_JBES_blinded.Rmd
Created blinded versions and zipped extra material

--- 2020-09-16 lgpc_JBES.Rmd
@@
 title: "The locally Gaussian partial correlation"
-author: Håkon Otneim^[Corresponding author.
-Department of Business and Management Science, NHH Norwegian School of Economics.
-Helleveien 30, 5045 Bergen, Norway.
-\texttt{hakon.otneim@nhh.no}] and Dag Tjøstheim^[University of Bergen \newline {Acknowledgements:} The authors gratefully acknowledge the input of an associate editor as well as two anonymous referees, whose comments have greatly improved this paper.]
+author: \footnote{{Acknowledgements:} The authors gratefully acknowledge the input of an associate editor as well as two anonymous referees, whose comments have greatly improved this paper.}
 output: 

### 2020-09-18 lgpc_JBES_blinded.tex
Created blinded versions and zipped extra material

--- 2020-09-16 lgpc_JBES.tex
@@
   pdftitle={The locally Gaussian partial correlation},
-  pdfauthor={Håkon Otneim and Dag Tjøstheim},
   hidelinks,
@@
 \title{The locally Gaussian partial correlation}
-\author{Håkon Otneim\footnote{Corresponding author.
-Department of Business and Management Science, NHH Norwegian School of Economics.
-Helleveien 30, 5045 Bergen, Norway.
-\texttt{hakon.otneim@nhh.no}} and Dag Tjøstheim\footnote{University of Bergen \newline {Acknowledgements:} The authors gratefully acknowledge the input of an associate editor as well as two anonymous referees, whose comments have greatly improved this paper.}}
+\author{\footnote{{Acknowledgements:} The authors gratefully acknowledge the input of an associate editor as well as two anonymous referees, whose comments have greatly improved this paper.}}
 \date{}

### 2020-09-18 lgpc_JBES_supp_blinded.Rmd
Created blinded versions and zipped extra material

--- 2020-09-16 lgpc_JBES_supp.Rmd
@@
 title: "Online supplement accompanying 'The Locally Gaussian Partial Correlation'"
-author: Håkon Otneim^[Corresponding author.
-Department of Business and Management Science, NHH Norwegian School of Economics.
-Helleveien 30, 5045 Bergen, Norway.
-\texttt{hakon.otneim@nhh.no}] and Dag Tjøstheim^[University of Bergen]
+author: 
 output: 

### 2020-09-18 lgpc_JBES_supp_blinded.tex
Created blinded versions and zipped extra material

--- 2020-09-16 lgpc_JBES_supp.tex
@@
   pdftitle={Online supplement accompanying `The Locally Gaussian Partial Correlation'},
-  pdfauthor={Håkon Otneim and Dag Tjøstheim},
   hidelinks,
@@
 \title{Online supplement accompanying `The Locally Gaussian Partial Correlation'}
-\author{Håkon Otneim\footnote{Corresponding author.
-Department of Business and Management Science, NHH Norwegian School of Economics.
-Helleveien 30, 5045 Bergen, Norway.
-\texttt{hakon.otneim@nhh.no}} and Dag Tjøstheim\footnote{University of Bergen}}
-\date{}
+\author{}
+\date{\vspace{-2.5em}}
 

### 2021-01-22 lgpc_JBES.Rmd
Paper akseptert, rettet opp i noen små detaljer iht kommentar fra referee

@@
 In all of the following we assume that the components of $\X$ are continuous with joint density function $f_{\X}$.
-Given a point $\x$, we approximate $f_{\X}$ in a neighborhood of $\x$ by a multivariate Gaussian density
+Given a point $\x$, we approximate $f_{\X}$ in a neighborhood of $\x$ by a multivariate Gaussian density $\psi({\x},{\fv})$, which at the point ${\fv} = {\x}$ is
 
 \begin{equation} 
-\psi(\x, \fv) = \frac{1}{(2\pi)^{p/2}|\fSigma(\x)|^{1/2}} \exp \left\{-\frac{1}{2}(\fv - \fmu(\x))^T\fSigma^{-1}(\x)(\fv - \fmu(\x))\right\},
+\psi(\x) = \frac{1}{(2\pi)^{p/2}|\fSigma(\x)|^{1/2}} \exp \left\{-\frac{1}{2}(\x - \fmu(\x))^T\fSigma^{-1}(\x)(\x - \fmu(\x))\right\},
 (\#eq:localgaussian)
@@
 In practice, we do not know $F_{j}$, but we can instead use the empirical distribution function 
-$$\widehat F_{j}(x) = \frac{1}{n-1}\sum_{i=1}^n 1\left(X_{ji} \leq x\right),$$
-where $1(\cdot)$ is the indicator function, $n$ is the number of observations, and $X_{ji}$ is the $i$th observation of $X_j$.
-This results in pseudo standard normal variables $\widehat Z_{j} = \Phi^{-1}(\widehat F_{j}(X_j))$.
+$$\widehat F_{j}(x) = \frac{1}{n-1}\sum_{i=1}^n 1\left(X_{j,t} \leq x\right),$$
+where $1(\cdot)$ is the indicator function, $n$ is the number of observations, and $X_{j,t}$ is the $t$th observation of $X_{j}$.
+This results in pseudo standard normal variables $\widehat Z_{j,t} = \Phi^{-1}(\widehat F_{j}(X_{j,t}))$.
 
-In the following, we will not always distinguish between $Z_j$ and $\widehat Z_{j}$.
-In fact, we conclude our asymptotic analysis in Section \@ref(chap:asymptotic) by showing that under certain technical conditions, the error made by estimating $\R(\z)$ using the empirically transformed variables $\widehat Z_{j}$, instead of $Z_j$, is smaller in the limit than the estimation error made when estimating the local correlations themselves.
+In the following, we will not always distinguish between $Z_{j,t}$ and $\widehat Z_{j,t}$.
+In fact, we conclude our asymptotic analysis in Section \@ref(chap:asymptotic) by showing that under certain technical conditions, the error made by estimating $\R(\z)$ using the empirically transformed variables $\widehat Z_{j,t}$, instead of $Z_{j,t}$, is smaller in the limit than the estimation error made when estimating the local correlations themselves.
 
@@
 
-$$f_{z_1|\z_3}(z_1|\z_3) = \frac{1}{|2\pi\sigma_{1|3}(z_1)|^{1/2}}\times\exp \left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1)}{\sigma_{1|3}(z_1)}\right)^2\right\},$$
+$$f_{Z_1|\Z_3}(z_1|\z_3) = \frac{1}{|2\pi\sigma_{1|3}(z_1, \z_3)|^{1/2}}\times\exp \left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1, \z_3)}{\sigma_{1|3}(z_1, \z_3)}\right)^2\right\},$$
 where
@@
 \begin{equation}
-\mu_{1|3}(z_1) = \R_{12}(z_1, \z_3)\R_{22}(\z_3)^{-1}\z_3 \,\,\, \textrm{ and } \,\,\, \sigma_{1|3}(z_1) = \fSigma_{Z_1|\Z_3},
+\mu_{1|3}(z_1, \z_3) = \R_{12}(z_1, \z_3)\R_{22}(\z_3)^{-1}\z_3 \,\,\, \textrm{ and } \,\,\, \sigma_{1|3}(z_1, \z_3) = \fSigma_{Z_1|\Z_3}(z_1,\z_3),
 (\#eq:localconditionalpar)
@@
 The conditional density $f_{Z_2|\Z_3}(z_2|\z_3)$ is defined in the same way.
-If $Z_1$ and $Z_2$ are conditionally independent given $\Z_3$, then
+If $Z_1$ and $Z_2$ are conditionally independent given $\Z_3$, then, by definition,
 
@@
 f_{Z_1,Z_2|\Z_3}(z_1, z_2|\z_3) &= f_{Z_1|\Z_3}(z_1|\z_3)f_{Z_2|\Z_3}(z_2|\z_3) \\
-&= \frac{1}{2\pi\sqrt{1-\sigma_{1|3}(z_1)}\sqrt{1-\sigma_{2|3}(z_2)}} \\ 
-&\qquad \qquad \times\exp\left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1)}{\sigma_{1|3}(z_1)}\right)^2\right\}\exp\left\{-\frac{1}{2}\left(\frac{z_2 - \mu_{2|3}(z_2)}{\sigma_{2|3}(z_2)}\right)^2\right\},
+&= \frac{1}{2\pi\sqrt{1-\sigma_{1|3}(z_1, \z_3)}\sqrt{1-\sigma_{2|3}(z_2, \z_3)}} \\ 
+&\qquad \qquad \times\exp\left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1, \z_3)}{\sigma_{1|3}(z_1, \z_3)}\right)^2\right\}\exp\left\{-\frac{1}{2}\left(\frac{z_2 - \mu_{2|3}(z_2, \z_3)}{\sigma_{2|3}(z_2, \z_3)}\right)^2\right\},
 \end{align*}
@@
 \begin{equation}
-\hfSigma_{12|3}(\z) = \hR^{11}(\z) -  \hR^{12}(\z)\left(\hR^{22}(\z)\right)^{-1}\hR^{21}(\z),
+\hfSigma_{12|3}(\z) = \hR^{11}(\zone) -  \hR^{12}(\z)\left(\hR^{22}(\ztwo)\right)^{-1}\hR^{21}(\z),
 (\#eq:localpartialcov)
@@
 We transform the observations to the $z$-scale as in \@ref(eq:pseudoobservations), but now estimate the individual components in $\R(\z)$ one by one, taking only into consideration the *pair* of variables in question:
-\begin{equation}
-\hrho_{jk}(z_j, z_k) = \arg\max_{\rho_{jk}} n^{-1} \sum_{t=1}^nK_{\hh}(\hZ_t - \z)\log\psi(\hZ_t, \rho_{jk}) - \int K_{\hh}(\y - \z)\psi(\y, \rho_{jk})\,\textrm{d}\y, 
+\begin{align}
+\hrho_{jk}(z_j, z_k) &= \arg\max_{\rho_{jk}} \Bigg[\frac{1}{n} \sum_{t=1}^nK_b(\widehat{Z}_{j,t} - z_j, \widehat{Z}_{k,t}- z_k) \log(\psi(\widehat{Z}_{j,t}, \widehat{Z}_{k,t}; \rho_{jk}(z_j, z_k))) \nonumber \\
+& \qquad\qquad\qquad\qquad\qquad - \int K_b(v - z_j, w- z_k)\psi(v,w; \rho_{jk}(z_j,z_k))\di v \di w,
 (\#eq:locallik2)
-\end{equation}
+\end{align}
 where all running variables and samples are bivariate subsets corresponding to the indices $(j,k)$, $\psi(\cdot, \rho)$ is the bivariate version of \@ref(eq:lgdeapprox), and $\hh$ now is a $2\times2$ diagonal matrix of bandwidths.
@@
 We estimate the local correlation matrix $\R(\z)$ by maximizing the local likelihood function in \@ref(eq:locallik3), which we denote by $L_n(\R(\z), \z)$, with the only exception that we, for now, assume the true transformation between the \(x\)- and the \(z\)-scale to be known.
-For a fixed matrix of bandwidths $\hh$, denote by $\R_{\hh}(z)$ the local correlation that satisfies, as $n \rightarrow \infty$,
+For a fixed matrix of bandwidths $\hh$, denote by $\frho_{\hh}(\z) = \{\rho_{12,\hh}(\z), \rho_{13,\hh}(\z), \rho_{23,\hh}(\z)\}$ the vector of local correlations that satisfiy, as $n \rightarrow \infty$,
 
 \begin{equation}
-\nabla L_n(\R_{\hh}, \z) \rightarrow \int K_{\hh}(\y - \z)\fu(\y, \R_{\hh}(\z))\{f_{\Z}(\y) - \psi(\y, \R_{\hh})\}\,\textrm{d}\y = 0,
+\nabla L_n(\frho_{\hh}(\z), \z) \rightarrow \int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\{f_{\Z}(\y) - \psi(\y, \frho_{\hh})\}\,\textrm{d}\y = 0,
 (\#eq:popfixedh3)
 \end{equation}
-where $f_{\Z}(\z) = f_{Z_1,Z_2,Z_3}(z_1,z_2,z_3)$ is the density function of $\Z$ (at the point $\z$), and $\fu(\cdot, \rho_{\hh})$ is the column vector of local score functions $\nabla \log\psi(\cdot, \rho_{\hh})$, and where the gradient is taken with respect to the parameters $\rho_{12}(\z)$, $\rho_{13}(\z)$ and $\rho_{23}(\z)$.
-Denote by $\frho = \frho(\z) = \{\rho_{12}(\z), \rho_{13}(\z), \rho_{23}(\z)\}$ the vector of local correlations.
-The joint limiting distribution of $\hfrho$ (for convenience only stated on the \(z\)-scale below) is given by the following result as $n \rightarrow \infty$ and $\hh \rightarrow \bm{0}$:
+where $f_{\Z}(\z) = f_{Z_1,Z_2,Z_3}(z_1,z_2,z_3)$ is the density function of $\Z$ (at the point $\z$), and $\fu(\cdot, \frho_{\hh})$ is the column vector of local score functions $\nabla \log\psi(\cdot, \frho_{\hh})$, and where the gradient is taken with respect to the parameters $\rho_{12}(\z)$, $\rho_{13}(\z)$ and $\rho_{23}(\z)$.
+The joint limiting distribution of $\hfrho(\z)$ (for convenience only stated on the \(z\)-scale below) is given by the following result as $n \rightarrow \infty$ and $\hh \rightarrow \bm{0}$:
 
@@
 \begin{equation}
-\sqrt{nb^3}\Mb^{-1/2}\Jb\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
+\sqrt{nb^3}\Mb^{-1/2}\Jb\left(\hfrho_n(\z) - \frho_0(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
 (\#eq:an3)
@@
 The constant $\alpha(\z) = 0$ is indicated as a dotted line.
+A desirable property of the LGPC is that one is able to distinguish between positive and negative conditional dependence.
+In the context of Granger causality, we view this as a nice empirical device in contrast to the conventional Granger causality testing procedure that fails to say anything about the potential asymmetric pattern in the underlying problem.
 
@@
 \end{figure}
-
 One could perhaps make the argument that $R_{1,t}$ and $R_{2,t-1}$ seem to be somewhat positively dependent in the right half of the graph for *small* values of the conditioning variable $R_{1,t-1}$.
@@
 Furthermore, due to our transformation of the data \@ref(eq:trans) to marginal standard normality, we do not necessarily have to specify weight functions in our test functional to lessen the impact of outliers - unless of course we wish to test for conditional independence in a specific portion of the sample space.
-
 On the other hand, we have to pay a price when imposing structure on the dependence.
@@
 \end{equation}
-where $h(\cdot)$ is a real-valued function that is typically even and non-negative for most standard applications, and $S \subseteq \mathbb{R}^p$ is an integration area that can be altered in order to focus the test to specific portions of the sample space.
+where $h(\cdot)$ is a real-valued function that is typically even and non-negative for most standard applications (we use $h(x) = x^2$ in all our examples), and $S \subseteq \mathbb{R}^p$ is an integration area that can be altered in order to focus the test to specific portions of the sample space.
 Standard laws of large numbers ensure that, under regularity conditions, see Theorem F.1 of the supplement, $T_{n,b}$ converges in probability towards its population value $$T = \int_S h\left(\alpha(\z)\right) \di F(\z).$$

### 2021-01-22 lgpc_JBES.tex
Paper akseptert, rettet opp i noen små detaljer iht kommentar fra referee

@@
 \newtheorem*{solution}{Solution}
-\let\BeginKnitrBlock\begin \let\EndKnitrBlock\end
 \begin{document}
@@
 In all of the following we assume that the components of \(\X\) are continuous with joint density function \(f_{\X}\).
-Given a point \(\x\), we approximate \(f_{\X}\) in a neighborhood of \(\x\) by a multivariate Gaussian density
+Given a point \(\x\), we approximate \(f_{\X}\) in a neighborhood of \(\x\) by a multivariate Gaussian density \(\psi({\x},{\fv})\), which at the point \({\fv} = {\x}\) is
 
 \begin{equation} 
-\psi(\x, \fv) = \frac{1}{(2\pi)^{p/2}|\fSigma(\x)|^{1/2}} \exp \left\{-\frac{1}{2}(\fv - \fmu(\x))^T\fSigma^{-1}(\x)(\fv - \fmu(\x))\right\},
+\psi(\x) = \frac{1}{(2\pi)^{p/2}|\fSigma(\x)|^{1/2}} \exp \left\{-\frac{1}{2}(\x - \fmu(\x))^T\fSigma^{-1}(\x)(\x - \fmu(\x))\right\},
 \label{eq:localgaussian}
@@
 In practice, we do not know \(F_{j}\), but we can instead use the empirical distribution function
-\[\widehat F_{j}(x) = \frac{1}{n-1}\sum_{i=1}^n 1\left(X_{ji} \leq x\right),\]
-where \(1(\cdot)\) is the indicator function, \(n\) is the number of observations, and \(X_{ji}\) is the \(i\)th observation of \(X_j\).
-This results in pseudo standard normal variables \(\widehat Z_{j} = \Phi^{-1}(\widehat F_{j}(X_j))\).
+\[\widehat F_{j}(x) = \frac{1}{n-1}\sum_{i=1}^n 1\left(X_{j,t} \leq x\right),\]
+where \(1(\cdot)\) is the indicator function, \(n\) is the number of observations, and \(X_{j,t}\) is the \(t\)th observation of \(X_{j}\).
+This results in pseudo standard normal variables \(\widehat Z_{j,t} = \Phi^{-1}(\widehat F_{j}(X_{j,t}))\).
 
-In the following, we will not always distinguish between \(Z_j\) and \(\widehat Z_{j}\).
-In fact, we conclude our asymptotic analysis in Section \ref{chap:asymptotic} by showing that under certain technical conditions, the error made by estimating \(\R(\z)\) using the empirically transformed variables \(\widehat Z_{j}\), instead of \(Z_j\), is smaller in the limit than the estimation error made when estimating the local correlations themselves.
+In the following, we will not always distinguish between \(Z_{j,t}\) and \(\widehat Z_{j,t}\).
+In fact, we conclude our asymptotic analysis in Section \ref{chap:asymptotic} by showing that under certain technical conditions, the error made by estimating \(\R(\z)\) using the empirically transformed variables \(\widehat Z_{j,t}\), instead of \(Z_{j,t}\), is smaller in the limit than the estimation error made when estimating the local correlations themselves.
 
@@
 
-\[f_{z_1|\z_3}(z_1|\z_3) = \frac{1}{|2\pi\sigma_{1|3}(z_1)|^{1/2}}\times\exp \left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1)}{\sigma_{1|3}(z_1)}\right)^2\right\},\]
+\[f_{Z_1|\Z_3}(z_1|\z_3) = \frac{1}{|2\pi\sigma_{1|3}(z_1, \z_3)|^{1/2}}\times\exp \left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1, \z_3)}{\sigma_{1|3}(z_1, \z_3)}\right)^2\right\},\]
 where
@@
 \begin{equation}
-\mu_{1|3}(z_1) = \R_{12}(z_1, \z_3)\R_{22}(\z_3)^{-1}\z_3 \,\,\, \textrm{ and } \,\,\, \sigma_{1|3}(z_1) = \fSigma_{Z_1|\Z_3},
+\mu_{1|3}(z_1, \z_3) = \R_{12}(z_1, \z_3)\R_{22}(\z_3)^{-1}\z_3 \,\,\, \textrm{ and } \,\,\, \sigma_{1|3}(z_1, \z_3) = \fSigma_{Z_1|\Z_3}(z_1,\z_3),
 \label{eq:localconditionalpar}
@@
 The conditional density \(f_{Z_2|\Z_3}(z_2|\z_3)\) is defined in the same way.
-If \(Z_1\) and \(Z_2\) are conditionally independent given \(\Z_3\), then
+If \(Z_1\) and \(Z_2\) are conditionally independent given \(\Z_3\), then, by definition,
 
@@
 \begin{equation}
-\hfSigma_{12|3}(\z) = \hR^{11}(\z) -  \hR^{12}(\z)\left(\hR^{22}(\z)\right)^{-1}\hR^{21}(\z),
+\hfSigma_{12|3}(\z) = \hR^{11}(\zone) -  \hR^{12}(\z)\left(\hR^{22}(\ztwo)\right)^{-1}\hR^{21}(\z),
 \label{eq:localpartialcov}
@@
 We transform the observations to the \(z\)-scale as in \eqref{eq:pseudoobservations}, but now estimate the individual components in \(\R(\z)\) one by one, taking only into consideration the \emph{pair} of variables in question:
-\begin{equation}
-\hrho_{jk}(z_j, z_k) = \arg\max_{\rho_{jk}} n^{-1} \sum_{t=1}^nK_{\hh}(\hZ_t - \z)\log\psi(\hZ_t, \rho_{jk}) - \int K_{\hh}(\y - \z)\psi(\y, \rho_{jk})\,\textrm{d}\y, 
+\begin{align}
+\hrho_{jk}(z_j, z_k) &= \arg\max_{\rho_{jk}} \Bigg[\frac{1}{n} \sum_{t=1}^nK_b(\widehat{Z}_{j,t} - z_j, \widehat{Z}_{k,t}- z_k) \log(\psi(\widehat{Z}_{j,t}, \widehat{Z}_{k,t}; \rho_{jk}(z_j, z_k))) \nonumber \\
+& \qquad\qquad\qquad\qquad\qquad - \int K_b(v - z_j, w- z_k)\psi(v,w; \rho_{jk}(z_j,z_k))\di v \di w,
 \label{eq:locallik2}
-\end{equation}
+\end{align}
 where all running variables and samples are bivariate subsets corresponding to the indices \((j,k)\), \(\psi(\cdot, \rho)\) is the bivariate version of \eqref{eq:lgdeapprox}, and \(\hh\) now is a \(2\times2\) diagonal matrix of bandwidths.
@@
 We estimate the local correlation matrix \(\R(\z)\) by maximizing the local likelihood function in \eqref{eq:locallik3}, which we denote by \(L_n(\R(\z), \z)\), with the only exception that we, for now, assume the true transformation between the \(x\)- and the \(z\)-scale to be known.
-For a fixed matrix of bandwidths \(\hh\), denote by \(\R_{\hh}(z)\) the local correlation that satisfies, as \(n \rightarrow \infty\),
+For a fixed matrix of bandwidths \(\hh\), denote by \(\frho_{\hh}(\z) = \{\rho_{12,\hh}(\z), \rho_{13,\hh}(\z), \rho_{23,\hh}(\z)\}\) the vector of local correlations that satisfiy, as \(n \rightarrow \infty\),
 
 \begin{equation}
-\nabla L_n(\R_{\hh}, \z) \rightarrow \int K_{\hh}(\y - \z)\fu(\y, \R_{\hh}(\z))\{f_{\Z}(\y) - \psi(\y, \R_{\hh})\}\,\textrm{d}\y = 0,
+\nabla L_n(\frho_{\hh}(\z), \z) \rightarrow \int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\{f_{\Z}(\y) - \psi(\y, \frho_{\hh})\}\,\textrm{d}\y = 0,
 \label{eq:popfixedh3}
 \end{equation}
-where \(f_{\Z}(\z) = f_{Z_1,Z_2,Z_3}(z_1,z_2,z_3)\) is the density function of \(\Z\) (at the point \(\z\)), and \(\fu(\cdot, \rho_{\hh})\) is the column vector of local score functions \(\nabla \log\psi(\cdot, \rho_{\hh})\), and where the gradient is taken with respect to the parameters \(\rho_{12}(\z)\), \(\rho_{13}(\z)\) and \(\rho_{23}(\z)\).
-Denote by \(\frho = \frho(\z) = \{\rho_{12}(\z), \rho_{13}(\z), \rho_{23}(\z)\}\) the vector of local correlations.
-The joint limiting distribution of \(\hfrho\) (for convenience only stated on the \(z\)-scale below) is given by the following result as \(n \rightarrow \infty\) and \(\hh \rightarrow \bm{0}\):
+where \(f_{\Z}(\z) = f_{Z_1,Z_2,Z_3}(z_1,z_2,z_3)\) is the density function of \(\Z\) (at the point \(\z\)), and \(\fu(\cdot, \frho_{\hh})\) is the column vector of local score functions \(\nabla \log\psi(\cdot, \frho_{\hh})\), and where the gradient is taken with respect to the parameters \(\rho_{12}(\z)\), \(\rho_{13}(\z)\) and \(\rho_{23}(\z)\).
+The joint limiting distribution of \(\hfrho(\z)\) (for convenience only stated on the \(z\)-scale below) is given by the following result as \(n \rightarrow \infty\) and \(\hh \rightarrow \bm{0}\):
 
-\BeginKnitrBlock{theorem}
+\begin{theorem}
 \protect\hypertarget{thm:loccor3}{}{\label{thm:loccor3} }Let \(\{\Z_t\}_{t=1}^n\) be observations on the marginally standard normally distributed random vector \(\Z\) having joint density \(f_{\Z}(\z)\).
@@
 \begin{equation}
-\sqrt{nb^3}\Mb^{-1/2}\Jb\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
+\sqrt{nb^3}\Mb^{-1/2}\Jb\left(\hfrho_n(\z) - \frho_0(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
 \label{eq:an3}
@@
 \end{align*}
-\EndKnitrBlock{theorem}
+\end{theorem}
 
@@
 
-\BeginKnitrBlock{theorem}
+\begin{theorem}
 \protect\hypertarget{thm:pseudo}{}{\label{thm:pseudo} }Under the conditions in Theorems \ref{thm:loccor3} and Theorem 2 by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}), assume further that the kernel function \(K(\cdot)\) has bounded support.
 Replacing the marginally standard normal vector \(\Z\) with the approximately marginally standard normal vector \(\hZ\), does not change the conclusions in these theorems.
-\EndKnitrBlock{theorem}
+\end{theorem}
 
-\BeginKnitrBlock{proof}
+\begin{proof}
 \iffalse{} {Proof. } \fi{}See Section D of the online supplement.
-\EndKnitrBlock{proof}
+\end{proof}
 
@@
 The constant \(\alpha(\z) = 0\) is indicated as a dotted line.
+A desirable property of the LGPC is that one is able to distinguish between positive and negative conditional dependence.
+In the context of Granger causality, we view this as a nice empirical device in contrast to the conventional Granger causality testing procedure that fails to say anything about the potential asymmetric pattern in the underlying problem.
 
@@
 Furthermore, due to our transformation of the data \eqref{eq:trans} to marginal standard normality, we do not necessarily have to specify weight functions in our test functional to lessen the impact of outliers - unless of course we wish to test for conditional independence in a specific portion of the sample space.
-
 On the other hand, we have to pay a price when imposing structure on the dependence.
@@
 \end{equation}
-where \(h(\cdot)\) is a real-valued function that is typically even and non-negative for most standard applications, and \(S \subseteq \mathbb{R}^p\) is an integration area that can be altered in order to focus the test to specific portions of the sample space.
+where \(h(\cdot)\) is a real-valued function that is typically even and non-negative for most standard applications (we use \(h(x) = x^2\) in all our examples), and \(S \subseteq \mathbb{R}^p\) is an integration area that can be altered in order to focus the test to specific portions of the sample space.
 Standard laws of large numbers ensure that, under regularity conditions, see Theorem F.1 of the supplement, \(T_{n,b}\) converges in probability towards its population value \[T = \int_S h\left(\alpha(\z)\right) \di F(\z).\]

### 2021-01-22 lgpc_JBES_supp.Rmd
Paper akseptert, rettet opp i noen små detaljer iht kommentar fra referee

@@
 
-# THE MULTIVARIATE LGPC
+This supplement accompanies the article entitled *The Locally Gaussian Partial Correlation*, and contains several technical derivations and discussions.
+In Section \@ref(multi) we write out the details of the multivariate version of the LGPC.
+Section \@ref(app-loccor) contains details regarding the proof of Theorem 3.1 in the main article.
+In Section \@ref(app-asvar) we include some calulations needed to obtain exolicit expressions of the asymptotic variance of the LGPC.
+Section \@ref(proofpseudo) contains a proof of Theorem 3.2, and in Section \@ref(validity-bootstrap) we present conditions for the validity of the bootstrap when estimating the LGPC.
+Sections \@ref(asym-test) and \@ref(chap:pitman) present asymptotic theory for the test statistic in the test for conditional independence, while Sections \@ref(boot-test) and \@ref(local-boot) contain various details about bootstrapping the test statistic.
+In Section \@ref(examples) we include two additional examples -- one simulated and one using real data.
+Finally, in Section \@ref(power), we present an extension of the power study found in the main article.
+
+# THE MULTIVARIATE LGPC {#multi}
 
@@
 where $\Z_{in} = (Z_{i,1}, Z_{i,2}, \ldots, Z_{i,n})$ and $\Z_{jn} = (Z_{j,1}, Z_{j,2}, \ldots, Z_{j,n})$ are the $\Z$-vectors for components $i$ and $j$ respectively, $n$ is the number of observations, $z_i, z_j, v, w \in \mathbb{R}$, $\rho_b = \rho_{ij,b}(z_i,z_j)$ is the local Gaussian correlation parameter, $K_b(v-z_i, w-z_j) = (b_1b_2)^{-1}K(b_1^{-1}(v-z_i))K(b_2^{-1}(w-z_j))$ is a product kernel with bandwidth $b=b_n=(b_1,b_2) = (b_{1n}, b_{2n})$, and $K$ is an non-negative kernel function with compact support such that $\int K(v,w) = 1$.
-For a fixed bandwidth $b$, $\rho_{0,b}$ is the true local parameter, that is, the minimizer of the local penalty function
-
-\begin{equation}
-\int K_b(v-z_i, w-z_j)\big[\psi(v,w; \rho_b(z_i, z_j)) - \log(\psi(v,w; \rho_b(z_i, z_j)))f(v,w)\big] \di v \di w,
+For a fixed bandwidth $b$, $\rho_{0,b}$ is the true local parameter, that is, the maximizer of the local penalty function
+
+\begin{equation}
+\int K_b(v-z_i, w-z_j)\big[\log(\psi(v,w; \rho_b(z_i, z_j)))f(v,w) - \psi(v,w; \rho_b(z_i, z_j))\big] \di v \di w,
 (\#eq:penalty)
@@
 
-# ASYMPTOTIC THEORY FOR THE TEST STATISTIC
+# ASYMPTOTIC THEORY FOR THE TEST STATISTIC {#asym-test}
 
@@
 We will put the emphasis on a) and only comment briefly on b).
+Following Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) we neglect the asymptotic bias in the asymptotic expressions in Theorems F.2 and F.3.
+We are then implicitly assuming a suitable bandwidth condition as in Assumption 3 in Theorem 3.1.
+Deriving explicit expressions for the bias under general circumstances remanins to be worked out.
 
@@
 
-# BOOTSTRAPPING THE TEST STATISTIC
+# BOOTSTRAPPING THE TEST STATISTIC {#boot-test}
 
@@
 
-# THE LOCAL BOOTSTRAP
+# THE LOCAL BOOTSTRAP {#local-boot}
 
@@
 
-# TWO ADDITIONAL EXAMPLES OF ESTIMATION
+# TWO ADDITIONAL EXAMPLES OF ESTIMATION {#examples}
 
@@
 
-# POWER STUDIES
+# POWER STUDIES {#power}
 

### 2021-01-22 lgpc_JBES_supp.tex
Paper akseptert, rettet opp i noen små detaljer iht kommentar fra referee

@@
 
-\hypertarget{the-multivariate-lgpc}{%
-\section{THE MULTIVARIATE LGPC}\label{the-multivariate-lgpc}}
+This supplement accompanies the article entitled \emph{The Locally Gaussian Partial Correlation}, and contains several technical derivations and discussions.
+In Section \ref{multi} we write out the details of the multivariate version of the LGPC.
+Section \ref{app-loccor} contains details regarding the proof of Theorem 3.1 in the main article.
+In Section \ref{app-asvar} we include some calulations needed to obtain exolicit expressions of the asymptotic variance of the LGPC.
+Section \ref{proofpseudo} contains a proof of Theorem 3.2, and in Section \ref{validity-bootstrap} we present conditions for the validity of the bootstrap when estimating the LGPC.
+Sections \ref{asym-test} and \ref{chap:pitman} present asymptotic theory for the test statistic in the test for conditional independence, while Sections \ref{boot-test} and \ref{local-boot} contain various details about bootstrapping the test statistic.
+In Section \ref{examples} we include two additional examples -- one simulated and one using real data.
+Finally, in Section \ref{power}, we present an extension of the power study found in the main article.
+
+\hypertarget{multi}{%
+\section{THE MULTIVARIATE LGPC}\label{multi}}
 
@@
 where \(\Z_{in} = (Z_{i,1}, Z_{i,2}, \ldots, Z_{i,n})\) and \(\Z_{jn} = (Z_{j,1}, Z_{j,2}, \ldots, Z_{j,n})\) are the \(\Z\)-vectors for components \(i\) and \(j\) respectively, \(n\) is the number of observations, \(z_i, z_j, v, w \in \mathbb{R}\), \(\rho_b = \rho_{ij,b}(z_i,z_j)\) is the local Gaussian correlation parameter, \(K_b(v-z_i, w-z_j) = (b_1b_2)^{-1}K(b_1^{-1}(v-z_i))K(b_2^{-1}(w-z_j))\) is a product kernel with bandwidth \(b=b_n=(b_1,b_2) = (b_{1n}, b_{2n})\), and \(K\) is an non-negative kernel function with compact support such that \(\int K(v,w) = 1\).
-For a fixed bandwidth \(b\), \(\rho_{0,b}\) is the true local parameter, that is, the minimizer of the local penalty function
-
-\begin{equation}
-\int K_b(v-z_i, w-z_j)\big[\psi(v,w; \rho_b(z_i, z_j)) - \log(\psi(v,w; \rho_b(z_i, z_j)))f(v,w)\big] \di v \di w,
+For a fixed bandwidth \(b\), \(\rho_{0,b}\) is the true local parameter, that is, the maximizer of the local penalty function
+
+\begin{equation}
+\int K_b(v-z_i, w-z_j)\big[\log(\psi(v,w; \rho_b(z_i, z_j)))f(v,w) - \psi(v,w; \rho_b(z_i, z_j))\big] \di v \di w,
 \label{eq:penalty}
@@
 
-\hypertarget{asymptotic-theory-for-the-test-statistic}{%
-\section{ASYMPTOTIC THEORY FOR THE TEST STATISTIC}\label{asymptotic-theory-for-the-test-statistic}}
+\hypertarget{asym-test}{%
+\section{ASYMPTOTIC THEORY FOR THE TEST STATISTIC}\label{asym-test}}
 
@@
 
-\hypertarget{bootstrapping-the-test-statistic}{%
-\section{BOOTSTRAPPING THE TEST STATISTIC}\label{bootstrapping-the-test-statistic}}
+\hypertarget{boot-test}{%
+\section{BOOTSTRAPPING THE TEST STATISTIC}\label{boot-test}}
 
@@
 
-\hypertarget{the-local-bootstrap}{%
-\section{THE LOCAL BOOTSTRAP}\label{the-local-bootstrap}}
+\hypertarget{local-boot}{%
+\section{THE LOCAL BOOTSTRAP}\label{local-boot}}
 
@@
 
-\hypertarget{two-additional-examples-of-estimation}{%
-\section{TWO ADDITIONAL EXAMPLES OF ESTIMATION}\label{two-additional-examples-of-estimation}}
+\hypertarget{examples}{%
+\section{TWO ADDITIONAL EXAMPLES OF ESTIMATION}\label{examples}}
 
@@
 
-\hypertarget{power-studies}{%
-\section{POWER STUDIES}\label{power-studies}}
+\hypertarget{power}{%
+\section{POWER STUDIES}\label{power}}
 

### 2021-01-22 responses2/responses.Rmd
Paper akseptert, rettet opp i noen små detaljer iht kommentar fra referee

---
title: ""
output: pdf_document
---

\begin{center}
\Large \textbf{Responses to comments made my the Associate Editor and the reviewer to the manuscript JBES-P-2019-0538.R1: "\textit{The locally Gaussian partial correlation}"}
\end{center}

# Responses to comments made by the Associate Editor

The associate editor writes the following:

*I have only one editorial comment: The authors should include a short abstract for and in the online supplement that explains each section briefly.*

**Our response:** We have added a paragraph in the beginning of the supplement detailing the contents of all the sections in the document.

# Responses to comments made by Referee 2.

We are very grateful to the reviewer for reading our revised manuscript so carefully once again. Her are our point-by-point responses to the comments:

\begin{itemize}

\item[1.] \textit{It seems to me that it is better to replace $v$ in Equation (4) by $x$ instead. Otherwise, when $f_x$ itself is a Gaussian density, it is hard to simply reduce $\psi(x,v)$ to $f_x(x)$. This is also in line with the expression in Equation (6).}

\textbf{Our response:} Thank you, has been done. We have also added a small comment just before eq. (4) in order to make this notation clear.

\item[2.] \textit{The authors need to change $\widehat{Z}_j$ in page 6 to be consistent with the one in Equation (12) in page 11, because the estimated version can only be defined as $Phi^{-1}\left(\widehat F_j(X_{jt})\right)$ for each component $j = 1,\ldots,p$. Also please make the notation consistent, using either $X_{j,t}$ or $X_{jt}$ . There are other notation issues across the main text and the supplemental material, with some listed below.}

\textbf{Our response:} Thank you for noticing this detail. Has been fixed. We have also searched the document for other occurrences of this \textit{particular} error, but could not find any.

\item[3.] \textit{Make notations like $f_{z_1|z_3}$ and $f_{Z_1|Z_3}$ consistent in page 9. In Equation (10), the definition of $\mu_{1|3}$ depends on both $z_1$ and $z_3$ , so it is better defined as $\mu_{1|3}(z_1,z_3)$. Similarly, $\sigma_{1|3}(z_1,z_3)$ should be used and $\Sigma_{Z_1|Z_3}$ should be replaced by $\Sigma_{Z_1|Z_3}(z_1,z_3)$ in Equation (10). In addition, I do not see the product of $f_{z_1|z_3}$ and $f_{z_2|z_3}$ (with their expressions given by Equation (10) and alike) leads directly to the expression below Equation (10). Pls double-check.}

\textbf{Our response:} Thank you, we have updated the notation according to your comment. The expression below eq. (10) follows directly from the definition of conditional independence in terms of density functions. Two variables are conditionally independent given a third variable if and only if their corresponding \textit{joint} conditional density given the third variable is equal to the product of their corresponding \textit{marginal} conditional densities given the third variable. We have made this clear by adding "\textit{by definition}" just before the expression.

\item[4.] \textit{Make Equation (13) consistent with Equation (7).}

\textbf{Our response:} Has been done.

\item[5.] \textit{Is it beneficial to exploit the symmetric structure in Equation (17) in Section 3.1.2, say considering only the cases of $j<k$?}

\textbf{Our response:} That is certainly possible. However, we prefer to keep eq. (17) in its current form.

\item[6.] \textit{I’m afraid that in page 13 replacing Equation (18) by Equation (A.13) (defined in page 42) helps understand better the optimization problem for the pairwise case than described in Equation (18).}

\textbf{Our response:} We agree, and we have written eq. (18) in the same way as eq. (A.13) in the Appendix.

\item[7.] \textit{I would suggest the authors to unify the notations in Equation (19) and below, because $u(\cdot,\rho_b)$ and $u(\cdot,R_b)$ (or $\psi(\cdot,\rho_b)$ and $\psi(\cdot,R_b)$) are used as equivalent expressions. Although I understand that they could serve the purpose, $u(\cdot,\rho_b)$ and $\psi(\cdot,R_b)$ should be reserved for the case of general dimensions.}

\textbf{Our response:} Thank you for pointing this out. We have changed the notation in eq. 19 and the following paragraph so that it is in line with Theorem 3.1.

\item[8.] \textit{Make Equation (20) consistent with Equations (21) and (22); i.e, deciding to add the argument $z$ or not.}

\textbf{Our response:} We have added the argument $z$ to eq. (20).

\item[9.] \textit{The authors stress one desirable property of LGPC being able to distinguish between positive and negative conditional dependence. I cannot find too much discussion in Section 4 about this property, which could be viewed as a nice empirical device in contrast to the conventional Granger causality testing procedure that fails to say anything about the potentially asymmetric pattern of the underlying problem.}

\textbf{Our response:} This is a good point indeed. We have added a comment about this in the paragraph following eq. (25).

\item[10.] \textit{In Section 5.3, what is the exact form of function $h(x)$ as given in Equation (28), say the standard $h(x) = x^2$?}

\textbf{Our response:} We point out just after eq. (28) that we use $h(x) = x^2$ in all our examples.

\item[11.] \textit{The online supplement contains a detailed analysis and is very useful. I appreciate your efforts very much. Equation (A.14) is defined for the minimization problem, but Equation (A.13) is define for the maximization problem. Pls change Equation (A.14) accordingly as in the subsequent theoretical analysis including the assumptions it is about maximization problem.}

\textbf{Our response:} Thank you for reading this long appendix. We have rephrased eq. (A.14) to be a maximization problem in accordance with (A.13).

\item[12.] \textit{In the second parts of both Theorem F.2 and Theorem F.3, they do not involve any bias terms that typically diverge to infinity as $n \to \infty$ and $b \to 0$; i.e., $T_b$ can be simply replaced by $T$ under the condition of $n \to \infty$ and $b \to 0$. This may seem to be a bit surprising to me given the popular $L_2$ form of the proposed test statistic in Equation (28). Is it because of the particular bandwidth condition imposed in Assumption 3 in Theorem 3.1? Is it somewhat undersmoothing to take care of the potential bias terms?}

\textbf{Our response:} You are correct. This is a consequence of such a bandwidth condition, and we have added a comment pointing this out just before Theorem F.2.
\end{itemize}

### 2021-01-26 responses2/responses.Rmd
Few corrections to the responses, added empty transfer documents

@@
 
-\textbf{Our response:} Thank you, we have updated the notation according to your comment.
-The expression below eq.
-(10) follows directly from the definition of conditional independence in terms of density functions.
-Two variables are conditionally independent given a third variable if and only if their corresponding \textit{joint} conditional density given the third variable is equal to the product of their corresponding \textit{marginal} conditional densities given the third variable.
-We have made this clear by adding "\textit{by definition}" just before the expression.
+\textbf{Our response:} Thank you, we have updated the notation according to your comment. The expression below eq. (10) contained an error under each of the square root signs which has been corrected.
 

### 2021-01-27 accepted_submit/source/lgpc_JBES.tex
Submitted final version to JBES

--- 2021-01-22 lgpc_JBES.tex
@@
 f_{Z_1,Z_2|\Z_3}(z_1, z_2|\z_3) &= f_{Z_1|\Z_3}(z_1|\z_3)f_{Z_2|\Z_3}(z_2|\z_3) \\
-&= \frac{1}{2\pi\sqrt{1-\sigma_{1|3}(z_1, \z_3)}\sqrt{1-\sigma_{2|3}(z_2, \z_3)}} \\ 
+&= \frac{1}{2\pi\sqrt{\sigma_{1|3}(z_1, \z_3)\sigma_{2|3}(z_2, \z_3)}} \\ 
 &\qquad \qquad \times\exp\left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1, \z_3)}{\sigma_{1|3}(z_1, \z_3)}\right)^2\right\}\exp\left\{-\frac{1}{2}\left(\frac{z_2 - \mu_{2|3}(z_2, \z_3)}{\sigma_{2|3}(z_2, \z_3)}\right)^2\right\},
@@
 \begin{figure}
-    \subfloat[$n = $ 100\label{fig:gaussian-example-plots1}]{\includegraphics[width=0.31\linewidth]{rcode_and_data/figures/gaussian-example-plots-1} }
-    \subfloat[$n = $ 500\label{fig:gaussian-example-plots2}]{\includegraphics[width=0.31\linewidth]{rcode_and_data/figures/gaussian-example-plots-2} }
-    \subfloat[$n = $ 2000\label{fig:gaussian-example-plots3}]{\includegraphics[width=0.31\linewidth]{rcode_and_data/figures/gaussian-example-plots-3} }
+    \subfloat[$n = $ 100\label{fig:gaussian-example-plots1}]{\includegraphics[width=0.31\linewidth]{gaussian-example-plots-1} }
+    \subfloat[$n = $ 500\label{fig:gaussian-example-plots2}]{\includegraphics[width=0.31\linewidth]{gaussian-example-plots-2} }
+    \subfloat[$n = $ 2000\label{fig:gaussian-example-plots3}]{\includegraphics[width=0.31\linewidth]{gaussian-example-plots-3} }
     \caption{Estimated local Gaussian partial correlation between $X_1$ and $X_2$ given $X_3 = 0$, where $(X_1, X_2, X_3)$ is a jointly standard normally distributed vector.
@@
 \begin{figure}[p]
-    \subfloat[Observed pairs $(X_1, X_2)$ from the conditionally \newline Gaussian model.\label{fig:simex2-plots1}]{\includegraphics[width=0.49\linewidth]{rcode_and_data/figures/simex2-plots-1} }
-    \subfloat[The estimated LGPC between $X_1$ and $X_2$ given $\rho = -0.9$.\label{fig:simex2-plots2}]{\includegraphics[width=0.49\linewidth]{rcode_and_data/figures/simex2-plots-2} }\newline
-    \subfloat[The estimated LGPC between $X_1$ and $X_2$ \newline  given $\rho = 0$\label{fig:simex2-plots3}]{\includegraphics[width=0.49\linewidth]{rcode_and_data/figures/simex2-plots-3} }
-    \subfloat[The estimated LGPC between $X_1$ and $X_2$ \newline given $\rho = 0.9$\label{fig:simex2-plots4}]{\includegraphics[width=0.49\linewidth]{rcode_and_data/figures/simex2-plots-4} }
+    \subfloat[Observed pairs $(X_1, X_2)$ from the conditionally \newline Gaussian model.\label{fig:simex2-plots1}]{\includegraphics[width=0.49\linewidth]{simex2-plots-1} }
+    \subfloat[The estimated LGPC between $X_1$ and $X_2$ given $\rho = -0.9$.\label{fig:simex2-plots2}]{\includegraphics[width=0.49\linewidth]{simex2-plots-2} }\newline
+    \subfloat[The estimated LGPC between $X_1$ and $X_2$ \newline  given $\rho = 0$\label{fig:simex2-plots3}]{\includegraphics[width=0.49\linewidth]{simex2-plots-3} }
+    \subfloat[The estimated LGPC between $X_1$ and $X_2$ \newline given $\rho = 0.9$\label{fig:simex2-plots4}]{\includegraphics[width=0.49\linewidth]{simex2-plots-4} }
     \caption{The conditionally Gaussian model}\label{fig:simex2-plots}
@@
 \centering
-\includegraphics{rcode_and_data/figures/granger-order1}
+\includegraphics{granger-order1}
 \caption{The estimated LGPC between $R_{1,t}$ and $R_{2,t-1}$ given $R_{1,t-1}$ along the diagonal of the sample space, for three values of the conditioning variable}
@@
 \centering
-\includegraphics{rcode_and_data/figures/granger-order2}
+\includegraphics{granger-order2}
 \caption{The estimated local Gaussian partial correlations for the three variable pairs in $(R_{1,t}, R_{2,t-1}, R_{2, t-2})$ given $(R_{1, t-1}, R_{1, t-2})$ estimated along the diagonal in the sample space.
@@
 
-\input{tables/power_table_dim3_n100_short}
+\input{power_table_dim3_n100_short}
 

### 2021-01-27 lgpc_JBES_blinded.Rmd
Submitted final version to JBES

@@
 title: "The locally Gaussian partial correlation"
-author: \footnote{{Acknowledgements:} The authors gratefully acknowledge the input of an associate editor as well as two anonymous referees, whose comments have greatly improved this paper.}
+author: 
 output: 
@@
 f_{Z_1,Z_2|\Z_3}(z_1, z_2|\z_3) &= f_{Z_1|\Z_3}(z_1|\z_3)f_{Z_2|\Z_3}(z_2|\z_3) \\
-&= \frac{1}{2\pi\sqrt{1-\sigma_{1|3}(z_1)}\sqrt{1-\sigma_{2|3}(z_2)}} \\ 
-&\qquad \qquad \times\exp\left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1)}{\sigma_{1|3}(z_1)}\right)^2\right\}\exp\left\{-\frac{1}{2}\left(\frac{z_2 - \mu_{2|3}(z_2)}{\sigma_{2|3}(z_2)}\right)^2\right\},
+&= \frac{1}{2\pi\sqrt{\sigma_{1|3}(z_1, \z_3)\sigma_{2|3}(z_2, \z_3)}} \\ 
+&\qquad \qquad \times\exp\left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1, \z_3)}{\sigma_{1|3}(z_1, \z_3)}\right)^2\right\}\exp\left\{-\frac{1}{2}\left(\frac{z_2 - \mu_{2|3}(z_2, \z_3)}{\sigma_{2|3}(z_2, \z_3)}\right)^2\right\},
 \end{align*}

### 2021-01-27 lgpc_JBES_blinded.tex
Submitted final version to JBES

@@
 \title{The locally Gaussian partial correlation}
-\author{\footnote{{Acknowledgements:} The authors gratefully acknowledge the input of an associate editor as well as two anonymous referees, whose comments have greatly improved this paper.}}
-\date{}
+\author{}
+\date{\vspace{-2.5em}}
 

### 2021-01-27 lgpc_JBES_supp_blinded.Rmd
Submitted final version to JBES

@@
 We will put the emphasis on a) and only comment briefly on b).
+Following Lacal and Tjøstheim (\protect\hyperlink{ref-lacal2018estimating}{2018}) we neglect the asymptotic bias in the asymptotic expressions in Theorems F.2 and F.3.
+We are then implicitly assuming a suitable bandwidth condition as in Assumption 3 in Theorem 3.1.
+Deriving explicit expressions for the bias under general circumstances remains to be worked out.
 

### 2021-01-27 accepted_submit/rcode_and_data/rcode.R

@@ -1,0 +1,454 @@
+# R-code for reproducing the simulations and empirical results in "The locally
+# Gaussian partial correlation". For prefect reproduction, run this script in its
+# entirety, and notice that the random seed is set just below the package
+# loadings. Output from the sessionInfo() of the system used to create these
+# figures is listed in the end of this document.
+
+# Packages needed
+library(lg)          # <-- The main package, see documentation for details.
+library(mvtnorm)
+library(dplyr)
+library(ggplot2)
+library(readr)
+library(lubridate)
+library(tidyr)
+library(readxl)
+library(tseries)
+
+# Load the necessary packages and set random seed for full reproducibility
+SEED <- 1
+set.seed(SEED)
+
+# EXAMPLE: THE MULTIVARIATE GAUSSIAN DISTRIBUTION
+# -----------------------------------------------
+
+# We generate samples from three-variate Gaussian distributions with different
+# sample sizes. We then calculate the partial correlations between X1 and X2
+# given X3 and save the results.
+
+# Estimation parameters
+n_gaussian                           <- c(100, 500, 2000)       # The sample sizes
+level                                <- .95                     # Conf. level for asymptotic conf. int.
+grid_size                            <- 7                       # No. of grid points in each direction
+bw_select                            <- "plugin"                # Bandwidth selection method
+plugin_constant                      <- 4                       # Proportionality constant for plugin bandwidth selector
+est_method                           <- "1par"                  # Estimation method
+transform_to_marginal_normality      <- TRUE                    # Transform to marginal normality
+
+# Create the grid
+grid <- expand.grid(X1 = seq(-3.5, 3.5, length.out = grid_size),
+                    X2 = seq(-3.5, 3.5, length.out = grid_size))
+
+# This will be a list of plots that we print out later
+gaussian_example <- list()
+
+# One set of plots for each sample size
+for(i in seq_along(n_gaussian)) {
+    
+    # Generate data
+    x <- rmvnorm(n_gaussian[i], sigma = diag(3))
+    
+    # Estimate the local correlation
+    lg_object <- lg_main(x, 
+                         bw_method = bw_select,
+                         plugin_constant_joint = plugin_constant,
[…]

### 2021-01-28 accepted_submit/rcode_and_data/rcode.R
Cleaned up, zipped the submitted materiel

@@ -1,454 +1,0 @@
-# R-code for reproducing the simulations and empirical results in "The locally
-# Gaussian partial correlation". For prefect reproduction, run this script in its
-# entirety, and notice that the random seed is set just below the package
-# loadings. Output from the sessionInfo() of the system used to create these
-# figures is listed in the end of this document.
-
-# Packages needed
-library(lg)          # <-- The main package, see documentation for details.
-library(mvtnorm)
-library(dplyr)
-library(ggplot2)
-library(readr)
-library(lubridate)
-library(tidyr)
-library(readxl)
-library(tseries)
-
-# Load the necessary packages and set random seed for full reproducibility
-SEED <- 1
-set.seed(SEED)
-
-# EXAMPLE: THE MULTIVARIATE GAUSSIAN DISTRIBUTION
-# -----------------------------------------------
-
-# We generate samples from three-variate Gaussian distributions with different
-# sample sizes. We then calculate the partial correlations between X1 and X2
-# given X3 and save the results.
-
-# Estimation parameters
-n_gaussian                           <- c(100, 500, 2000)       # The sample sizes
-level                                <- .95                     # Conf. level for asymptotic conf. int.
-grid_size                            <- 7                       # No. of grid points in each direction
-bw_select                            <- "plugin"                # Bandwidth selection method
-plugin_constant                      <- 4                       # Proportionality constant for plugin bandwidth selector
-est_method                           <- "1par"                  # Estimation method
-transform_to_marginal_normality      <- TRUE                    # Transform to marginal normality
-
-# Create the grid
-grid <- expand.grid(X1 = seq(-3.5, 3.5, length.out = grid_size),
-                    X2 = seq(-3.5, 3.5, length.out = grid_size))
-
-# This will be a list of plots that we print out later
-gaussian_example <- list()
-
-# One set of plots for each sample size
-for(i in seq_along(n_gaussian)) {
-    
-    # Generate data
-    x <- rmvnorm(n_gaussian[i], sigma = diag(3))
-    
-    # Estimate the local correlation
-    lg_object <- lg_main(x, 
-                         bw_method = bw_select,
-                         plugin_constant_joint = plugin_constant,
[…]
